### 2017-02-05 nbe/icfp17.tex
[ ICFP 17 ] Typing rules completed.

%% For double-blind review submission
\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format should change 'acmlarge' to
%% 'sigplan,10pt'.

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\makeatletter\if@ACM@journal\makeatother
%% Journal information (used by PACMPL format)
%% Supplied to authors by publisher for camera-ready submission
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmYear{2017}
\acmMonth{1}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\else\makeatother
%% Conference information (used by SIGPLAN proceedings format)
%% Supplied to authors by publisher for camera-ready submission
\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
\acmYear{2017}
\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\fi

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission
\setcopyright{none}             %% For review submission
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

\input{macros}

\begin{document}

%% Title information
\title[NbE for Sized Types]{Normalization by Evaluation for Sized Dependent Types}
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{Andreas Abel}
\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \department{Department of Computer Science and Engineering}
  \institution{Gothenburg University}
  \streetaddress{Rännvägen 6b}
  \city{Göteborg}
%  \state{State1}
  \postcode{41296}
  \country{Sweden}
}
\email{andreas.abel@gu.se}          %% \email is recommended

%% Author with two affiliations and emails.
\author{First2 Last2}
\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position2a}
  \department{Department2a}             %% \department is recommended
  \institution{Institution2a}           %% \institution is required
  \streetaddress{Street2a Address2a}
  \city{City2a}
  \state{State2a}
  \postcode{Post-Code2a}
  \country{Country2a}
}
\email{first2.last2@inst2a.com}         %% \email is recommended
\affiliation{
  \position{Position2b}
  \department{Department2b}             %% \department is recommended
  \institution{Institution2b}           %% \institution is required
  \streetaddress{Street3b Address2b}
  \city{City2b}
  \state{State2b}
  \postcode{Post-Code2b}
  \country{Country2b}
}
\email{first2.last2@inst2b.org}         %% \email is recommended

%% Paper note
%% The \thanks command may be used to create a "paper note" ---
%% similar to a title note or an author note, but not explicitly
%% associated with a particular element.  It will appear immediately
%% above the permission/copyright statement.
\thanks{with paper note}                %% \thanks is optional
                                        %% can be repeated if necesary
                                        %% contents suppressed with 'anonymous'

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Text of abstract \ldots.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords is optional

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}
\label{sec:intro}

\newpage

\section{Syntax}

Syntax (de Bruijn terms).
\[
\begin{array}{lllrl@{\qquad}l}
\Sort & \ni & s
    & ::= & \Set[\ell] ~(\ell \in \NN)& \mbox{sort (universe)} \\
% \Ann & \ni & \star & ::= & \erased \mid \nonstrict \mid \noterased & \mbox{annotation (irrelevant, shape-irrelevant, relevant)}\\
\Ann & \ni & \star & ::= & \erased \mid \noterased & \mbox{annotation (irrelevant, relevant)}\\
\Exp & \ni & t,u,T,U
    & ::= & w \mid t\,e & \mbox{expressions} \\
\Whnf & \ni & w, W
    & ::= & n \mid s \mid \Size \mid \epiT U T % \mid \piSizeT T
    % \mid \forallT T
    \mid \lambda t% & \mbox{functions} \\ &&
    \mid \Nat a \mid \tzero \mid \suc a t & \mbox{weak head normal forms} \\ % \mbox{natural numbers} \\
\Ne & \ni & n
    & ::= & \ind i \mid n \, e & \mbox{neutral expressions} \\
\Elim & \ni & e
    & ::= & t \mid a \mid \ann a \mid \case[\ell] T {t_1} t_2 \mid \fix[\ell] T t & \mbox{eliminations} \\
\SizeExp & \ni & a,b,c & ::= & \infty \mid \ind i + o ~(o \in \NN) & \mbox{size expressions} \\
\Cxt & \ni & \Gamma, \Delta & ::= & \cempty \mid \erext[\noterased] \Gamma T \mid \erext \Gamma \Size \\
\end{array}
\]
We omit the ``$\noterased$''-markers from $\Gamma$ and $\Pi$ by default.  We write $\forallT T$ for $\erpiT \Size T$.
For concrete function types, we may use a (named dependent) function type notation as syntactic sugar for the corresponding de Bruijn representation.  For instance, $\funT i \Size \Nat i \to \Set[\ell]$ is sugar for $\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.  Similar, we use named lambda abstraction as sugar for de Bruijn.  Named abstract takes care of proper lifting of de Bruijn indices.

Resurrection\cite{pfenning:lics01} \fbox{$\resurrect\Gamma$} removes
the ``$\erased$''-markers from all bindings in $\Gamma$ (\ie, replaces
them by ``$\noterased$''-markers).  Note that, trivially, resurrection is idempotent: $\resurrecttwice\Gamma = \resurrect{\Gamma}$.

We write $\sigma$ and $\tau$ for substitutions, which are lists of terms, and $t \sigma$ for the application of substitution $\sigma$ to term $t$.  We write $\slift^n_m$ for the substitution $(v_{n+m-1},\dots,v_{n+1},v_n)$ which applies to a term with $m$ free indices and increases all of them by $n$.  We write $\slift_m$ for the lifting $\slift^1_m$ and $\sid_m$ for the identity substitution $ \slift^0_m$. The substitution $[u]_m = (\sid_m, u)$ replaces free index $\ind 0$ by term $u$ and decrements all other free indices by $1$. We drop subscript $m$ when clear from the context.

Size increment \fbox{$a + k$} for $k \in \NN$ is defined by $\infty + k = \infty$ and $(\ind i + o) + k = \ind i + (o + k)$.  Size comparison \fbox{$a \leq b$} holds if either $b = \infty$ or $a = \ind i + o$ and $b = \ind i + o'$ with $o \leq o'$.

Judgements.
\[
\begin{array}{l@{\qquad}l}
  \der \Gamma & \mbox{context $\Gamma$ is well-formed} \\
  \Gamma(i) = \eann T & \mbox{in context $\Gamma$, de Bruijn index $i$ has type $T$ and annotation $\evar$} \\
  \Gamma \der a : \Size & \mbox{in context $\Gamma$, size expression $a$ is well-formed} \\
  \Gamma \der t : T & \mbox{in context $\Gamma$, term $t$ has type $T$} \\
  \Gamma \der t = t' : T & \mbox{in context $\Gamma$, terms $t$ and $t'$ are equal of type $T$} \\
%  \Gamma \mid U \der e : T & \mbox{in context $\Gamma$, elimination $e$ goes from $U$ to $T$} \\
%  \Gamma \mid U \der e = e' : T & \mbox{in context $\Gamma$, eliminations $e$ and $e'$ are equal} \\
  \Gamma \der T \leq T' & \mbox{in context $\Gamma$, type $T$ is a subtype of $T'$} \\
  \suty \Gamma \sigma \tau \Delta & \mbox{$\sigma$/$\tau$ is a valid term/type-level substitution for $\Delta$} \\
  \sueq \Gamma \sigma {\sigma'} \tau \Delta & \mbox{$\sigma$/$\sigma'$/$\tau$ are a equal term/term/type-level substitutions for $\Delta$} \\
\end{array}
\]
Derived judgements.
\[
\begin{array}{lll@{~}l@{\qquad}l}
  \Gamma \der T & \defas & \Gamma \der T : s & \mforsome s
    & \mbox{in context $\Gamma$, type $T$ is well-formed} \\
  \Gamma \der T = T' & \defas & \Gamma \der T = T' : s & \mforsome s
    & \mbox{in context $\Gamma$, types $T$ and $T'$ are equal} \\
  \Gamma \der a \leq b : \Size & \defas & \multicolumn 2 {l@{\qquad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
    & \mbox{sizes $a$ and $b$ are well-formed and comparable} \\
%    & \mbox{in context $\Gamma$, well-formed size $a$ is less or equal than well-formed size $b$} \\
\end{array}
\]
Inference rules.

\fbox{$\der \Gamma$} (implies $\der \resurrect \Gamma$).
\[
  \ru{}{\der \cempty}
\qquad
  \ru{\der \Gamma \qquad
      \resurrect \Gamma \der T
    }{\der \cext \Gamma T}
\qquad
  \ru{\der \Gamma
    }{\der \cext \Gamma \Size}
\qquad
  \ru{\der \Gamma
    }{\der \erext \Gamma \Size}
\]

\fbox{$\Gamma(i) = \eann T$} (implies $\Gamma \der T$).
\[
  \ru{}{(\eext \Gamma T)(0) = \eann T \slift}
\qquad
  \ru{\Gamma(i) = \eann T
    }{(\cext \Gamma \_)(i+1) = \eann T \slift}
\]

\fbox{$\Gamma \der a : \Size$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der a : \Size$).
\[
  \ru{\der \Gamma}{\Gamma \der \infty : \Size}
\qquad
  \ru{\der \Gamma \qquad \Gamma(i) = \erann[\noterased]\Size
    }{\Gamma \der \ind i + o : \Size
    }
\]

\fbox{$\Gamma \der t : T$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der T$ [and $\resurrect \Gamma \der t : T$]. Note: no rule for $\Gamma \der \Size : s$.)
\begin{gather*}
  \ru{\der \Gamma
    }{\Gamma \der \Set[\ell] : \Set[\ell']
    }{\ell {<} \ell'}
\qquad
  \ru{\Gamma \der U : s \qquad
      \cext \Gamma U \der T : s
    }{\Gamma \der \piT U T : s}
\qquad
  \ru{\cext \Gamma \Size \der T : s
    }{\Gamma \der \epiT \Size T : s}
\qquad
  \ru{%\der \Gamma \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \Nat a : s}
\\[2ex]
  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
    }{\Gamma \der \ind i : T
    }{T \not= \Size}
\qquad
  \ru{\eext \Gamma U \der t : T
    }{\Gamma \der \lambda t : \epiT U T}
\qquad
  \ru{\Gamma \der t : \piT U T \qquad
      \Gamma \der u : U
    }{\Gamma \der t\,u : T[u]}
% \qquad
%   \ru{\Gamma \der t : T \qquad
%       \Gamma \mid T \der e : T'
%     }{\Gamma \der t\,e : T'}
\qquad
  \ru{\Gamma \der t : \piT \Size T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a : T[a]}
\\[2ex] % \qquad
  \ru{\resurrect\Gamma \der b : \Size
    }{\Gamma \der \tzero : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der a : \Size \qquad
      \Gamma \der t : \Nat b
    }{\Gamma \der \suc a t : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der t : \forallT T \qquad
      \Gamma \der a : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der t \ann a : T[b]}
\\[2ex]
  \ru{\Gamma \der u : \Nat b \qquad
      \Gamma \der T : \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x
     }{\Gamma \der u\,\fix[\ell] T t : T \,b \, u}
\\[2ex]
  \ru{\Gamma \der u : \Nat (b + 1) \qquad
      \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
      \Gamma \der t_z : T \, \tzero \qquad
      \Gamma \der a : \Size \qquad
      \Gamma \der t_s : (n : \Nat b) \to T \, (\suc a n)
    }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
\end{gather*}
Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
% \fbox{$\Gamma \mid U \der e : T$}
% \begin{gather*}
%   \ru{\cext \Gamma U \der T \qquad \Gamma \der u : U
%     }{\Gamma \mid \piT U T \der u : T[u]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \piT \Size T \der a : T[a]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \forallT T \der \ann a : T[a]}
% \\
%   \ru{
%     }{\Gamma \mid \Nat a \der \fix T t : T \ann a n}
% \end{gather*}

\fbox{$\Gamma \der t = t' : T$} (implies $\resurrect \Gamma \der T$ and $\Gamma \der t,t' : T$ [and $\resurrect\Gamma \der t = t' : T$].)

Computation rules.
\begin{gather*}
  \ru{\cext \Gamma U \der t : T \qquad
      \Gamma \der u : U
    }{\Gamma \der (\lambda t)\,u = t[u] : T[u]}
\qquad
  \ru{\cext \Gamma \Size \der t : T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der (\lambda t)\,a = t[a] : T[a]}
\qquad
  \ru{\erext \Gamma \Size \der t : T \qquad
      \Gamma \der a : \Size \qquad
      \resurrect \Gamma \der b : \Size
    }{\Gamma \der (\lambda t)\,\ann a = t[a] : T[b]}
\\[2ex]
  \rux{\Gamma \der u : \Nat b \qquad
      \Gamma \der a : \Size \qquad
      \Gamma \der T : \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x
     }{\Gamma \der u\,\fix[\ell] T t = t\ann a (\lambda x. x\,\fix[\ell] T t)\, u : T \,b \, u
     }{u \in \{\tzero, \suc a t'\}}
\\[2ex]
  \ru{\resurrect\Gamma \der b : \Size \qquad
      \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
      \Gamma \der t_z : T\,\tzero \qquad
      \Gamma \der a : \Size \qquad
      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc a n)
    }{\Gamma \der \tzero\,\case[\ell] T {t_z} {t_s} = t_z : T\,\tzero}
\\[2ex]
  \ru{\Gamma \der t : \Nat b \qquad
      \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
      \Gamma \der t_z : T\,\tzero \qquad
      \Gamma \der a : \Size \qquad
      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc a n)
    }{\Gamma \der \suc a t\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,\suc a t}
\end{gather*}
Extensionality rules.
\begin{gather*}
  \ru{\Gamma \der t : \piT U T
    }{\Gamma \der t = \lambda x. t\,x : \piT U T}
\qquad
  \ru{\Gamma \der t : \forallT T \qquad \erext \Gamma \Size \der a : \Size
    }{\Gamma \der t = \lambda i. t \ann a : \forallT T}
\end{gather*}
Congruence rules.
\begin{gather*}
  \ru{\der \Gamma
    }{\Gamma \der \Set[\ell] = \Set[\ell]: \Set[\ell']
    }{\ell {<} \ell'}
\qquad
  \ru{\Gamma \der U = U' : s \qquad
      \cext \Gamma U \der T = T' : s
    }{\Gamma \der \piT U T = \piT{U'}T': s}
\qquad
  \ru{\cext \Gamma \Size \der T = T' : s
    }{\Gamma \der \epiT \Size T = \epiT \Size T' : s}
\qquad
  \ru{%\der \Gamma \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \Nat a = \Nat a : s}
\\[2ex]
  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
    }{\Gamma \der \ind i = \ind i : T
    }{T \not= \Size}
\qquad
  \ru{\eext \Gamma U \der t = t' : T
    }{\Gamma \der \lambda t = \lambda t' : \epiT U T}
\qquad
  \ru{\Gamma \der t = t' : \piT U T \qquad
      \Gamma \der u = u' : U
    }{\Gamma \der t\,u = t'\,u' : T[u]}
% \qquad
%   \ru{\Gamma \der t : T \qquad
%       \Gamma \mid T \der e : T'
%     }{\Gamma \der t\,e : T'}
\qquad
  \ru{\Gamma \der t = t' : \piT \Size T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a = t'\,a : T[a]}
\\[2ex] % \qquad
  \ru{\resurrect\Gamma \der b : \Size
    }{\Gamma \der \tzero = \tzero : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der a, a' : \Size \qquad
      \Gamma \der t = t' : \Nat b
    }{\Gamma \der \suc a t = \suc {a'} t' : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der t = t' : \forallT T \qquad
      \Gamma \der a, a' : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der t \ann a = t' \ann{a'} : T[b]}
\\[2ex]
  \ru{\Gamma \der u = u' : \Nat a \qquad
      \Gamma \der T = T' : \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t = t' : \allT i \Size (\funT x {\Nat i} T \ann i x) \to \funT x {\Nat (i + 1)} T \ann{i+1} x
     }{\Gamma \der u\,\fix[\ell] T t = u'\,\fix[\ell] {T'} t' : T \ann a u}
\\[2ex]
  \ru{\Gamma \der u = u' : \Nat (a + 1) \qquad
      \Gamma \der T = T' : Nat (a + 1) \to \Set[\ell] \qquad
      \Gamma \der t_z = t_z' : T \, \tzero \qquad
      \Gamma \der t_s = t_s' : (n : \Nat a) \to T \, (\suc a n)
    }{\Gamma \der u\,\case[\ell] T {t_z} t_s = u'\,\case[\ell] {T'} {t_z'} t_s' : T \, u}
\end{gather*}
Equivalence rules.  (Reflexivity is admissible.)
\begin{gather*}
  \ru{\Gamma \der t = t' : T
    }{\Gamma \der t' = t : T}
\qquad
  \ru{\Gamma \der t_1 = t_2 : T \qquad
      \Gamma \der t_2 = t_3 : T
    }{\Gamma \der t_1 = t_3 : T}
\end{gather*}

\fbox{$\Gamma \der T \leq T'$} (implies $\resurrect \Gamma \der T,T'$)
\begin{gather*}
  \ru{\der \Gamma \qquad \ell {\leq} \ell'
    }{\Gamma \der \Set[\ell] \leq \Set[\ell']}
\qquad
  \ru{\Gamma \der a \leq b : \Size
    }{\Gamma \der \Nat a \leq \Nat b}
\qquad
  \ru{\Gamma \der T = T'
    }{\Gamma \der T \leq T'}
\\[2ex]
  \ru{\Gamma \der U' \leq U \qquad
      \cext \Gamma {U'} \der T \leq T'
    }{\Gamma \der \piT U T \leq \piT {U'} {T'}}
\qquad
  \ru{\Gamma \der T_1 \leq T_2 \qquad
      \Gamma \der T_2 \leq T_3
    }{\Gamma \der T_1 \leq T_3}
\qquad
\end{gather*}

Substitution typing. % and equality.
\fbox{$\Gamma \der \tau : \Delta$}
(implies $\der \Gamma$ and $\der \Delta$
 [and $\resurrect\Gamma \der \tau : \Delta$]
 and $\resurrect\Gamma \der \tau : \resurrect \Delta$).

\begin{gather*}
  \ru{\der \Gamma \qquad
    }{\Gamma \der \sempty : \cempty}
\qquad
  \ru{\Gamma \der \tau : \Delta \qquad
      \resurrect \Delta \der T \qquad
      \Gamma \der t : T \tau
    }{\Gamma \der \sext \tau t : \cext \Delta T}
\qquad
  \ru{\Gamma \der \tau : \Delta \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \sext \tau a : \eext \Delta \Size}
\end{gather*}

% Substitution typing. % and equality.
% \fbox{$\Gamma \der \sigma : \Delta$}
% (implies $\der \Gamma$ and $\der \Delta$
%  and $\resurrect\Gamma \der \sigma : \Delta$
%  and $\resurrect\Gamma \der \sigma : \resurrect \Delta$).
% % and
% % \fbox{$\Gamma \der \sigma = \sigma' : \Delta$}.

% \begin{gather*}
%   \ru{\der \Gamma \qquad
%     }{\Gamma \der \sempty : \cempty}
% \qquad
%   \ru{\Gamma \der \sigma : \Delta \qquad
%       \resurrect \Delta \der T \qquad
%       \Gamma \der t : T \sigma
%     }{\Gamma \der \sext \sigma t : \cext \Delta T}
% \qquad
%   \ru{\Gamma \der \sigma : \Delta \qquad
%       \Gamma \der a : \Size
%     }{\Gamma \der \sext \sigma a : \eext \Delta \Size}
% \end{gather*}

% \begin{gather*}
%   \ru{\der \Gamma \qquad
%     }{\Gamma \der \sempty = \sempty : \cempty}
% \qquad
%   \ru{\Gamma \der \sigma = \sigma': \Delta \qquad
%       \resurrect \Delta \der T \qquad
%       \Gamma \der t = t' : T \sigma
%     }{\Gamma \der \sext \sigma t = \sext {\sigma'} t': \cext \Delta T}
% \end{gather*}

% Substitution typing and equality.
% \fbox{$\suty \Gamma \sigma \tau \Delta$} and

Substitution equality
\fbox{$\sueq \Gamma \sigma {\sigma'} \tau \Delta$}
(implies $\der \Gamma$ and $\der \Delta$ and $\Gamma \der \tau : \Delta$ and $\sueq \Gamma {\sigma'} {\sigma} \tau \Delta$).

% \begin{gather*}
%   \ru{\der \Gamma \qquad
%     }{\suty \Gamma \sempty \sempty \cempty}
% \qquad
%   \ru{\suty \Gamma \sigma \tau \Delta \qquad
%       \resurrect \Delta \der T \qquad
%       \Gamma \der u = t : T \tau
%     }{\suty \Gamma {\sext \sigma u} {\sext \tau t} \cext \Delta T}
% \\[2ex]
%   \ru{\suty \Gamma \sigma \tau \Delta \qquad
%       \Gamma \der a = b : \Size
%     }{\suty \Gamma {\sext \sigma a} {\sext \tau b} \cext \Delta \Size}
% \qquad
%   \ru{\suty \Gamma \sigma \tau \Delta \qquad
%       \Gamma \der a, b : \Size
%     }{\suty \Gamma {\sext \sigma a} {\sext \tau b} \erext \Delta \Size}
% \end{gather*}

\begin{gather*}
  \ru{\der \Gamma \qquad
    }{\sueq \Gamma \sempty \sempty \sempty \cempty}
\qquad
  \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
      \resurrect \Delta \der T \qquad
      \Gamma \der u = u' = t : T \tau
    }{\sueq \Gamma {\sext \sigma u } {\sext {\sigma'} {u'}} {\sext \tau t} {\cext \Delta T}}
\\[2ex]
  \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
      \Gamma \der a = a' = b : T \tau
    }{\sueq \Gamma {\sext \sigma a } {\sext {\sigma'} {a'}} {\sext \tau b} {\cext \Delta \Size}}
\qquad
  \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
      \Gamma \der a, a', b : T \tau
    }{\sueq \Gamma {\sext \sigma a } {\sext {\sigma'} {a'}} {\sext \tau b} {\erext \Delta \Size}}
\end{gather*}

\begin{lemma}[Reflexivity]
  If\/ $\Gamma \der t : T$ then $\Gamma \der t = t : T$.
\end{lemma}

In the following, let $J$ %, $L$, and $R$
match a part of a judgement.

\begin{lemma}[Context well-formedness]
\label{lem:cxtwf}\bla
  \begin{enumerate}
  \item If $\der \cext \Gamma \Delta$ then $\der \Gamma$
  \item If\/ $\Gamma \der J$ then $\der \Gamma$.
  \end{enumerate}
\end{lemma}

\begin{lemma}[Resurrection]
\label{lem:res} \bla
\begin{enumerate}
\item If\/ $\der \Gamma$ then $\der \resurrect\Gamma$ and $\resurrect \Gamma \der \sid : \Gamma$.
  % Consequently, if\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
  % Further, if $\Gamma \der J : \resurrect\Delta$ then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
\item If\/ $\Gamma \der J : \resurrect \Delta$ and $\der \Delta$ then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der \sigma : \Delta$ then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
\end{enumerate}
\end{lemma}

\begin{lemma}[Substitution]
\label{lem:sub}\bla
  If\/ $\Gamma \der \sigma : \Delta$ and $\Delta \der J$ then $\Gamma \der J\sigma$.
\end{lemma}

\begin{lemma}[Specific substitutions]
\label{lem:specsub}\bla
\begin{enumerate}
\item If $\der \cext \Gamma \Delta$ then $\cext \Gamma \Delta \der \slift_{|\Gamma|}^{|\Delta|} : \Gamma$.
      If $\der \cext \Gamma T$ then $\cext \Gamma T \der \slift : \Gamma$.
\item If $\der \Gamma$ then $\Gamma \der \sid : \Gamma$.
\item If $\Gamma \der u : U$ then $\Gamma \der \sg u : \cext \Gamma U$.
\end{enumerate}
\end{lemma}

\begin{lemma}[Substitution equality]
\label{lem:subeq}\bla
\begin{enumerate}
\item Conversion: If\/ $\sueq \Gamma \sigma {\sigma'} {\tau_1} \Delta$ and $\sueq {\resurrect\Gamma} {\tau_1} {\tau_2} {\tau} {\resurrect\Delta}$ then $\sueq \Gamma \sigma {\sigma'} {\tau_2} \Delta$.
\item Reflexivity: If\/ $\Gamma \der \sigma : \Delta$ then $\sueq \Gamma \sigma \sigma \sigma \Delta$.
\item Symmetry:  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then $\sueq \Gamma {\sigma'} {\sigma} {\tau} \Delta$.
\item Transitivity:  If\/ $\sueq \Gamma {\sigma_1} {\sigma_2} {\tau} \Delta$ and $\sueq \Gamma {\sigma_2} {\sigma_3} {\tau} \Delta$ then $\sueq \Gamma {\sigma_1} {\sigma-3} {\tau} \Delta$.
\item Functionality: Let $\sueq \Gamma \sigma {\sigma'} \tau \Delta$.
  \begin{enumerate}
  \item
  If $\Delta \der t : T$ then $\Gamma \der t\sigma = t\sigma' : T\tau$.
  \item
  If $\Delta \der t = t' : T$ then $\Gamma \der t\sigma = t'\sigma' : T\tau$.
  \item
  Corollary: If $\Delta \der T \leq T'$ then $\Gamma \der T\sigma \leq T\sigma'$.
  \end{enumerate}
\end{enumerate}
\end{lemma}

\section{Semantics}
\label{sec:sem}

Judgements.
\[
\begin{array}{l@{\qquad}l}
  t \evalsto w & \mbox{term $t$ evaluates to weak head normal form $w$} \\
  w @ e \evalsto w' & \mbox{eliminating $w$ with $e$ evaluates to $w'$} \\
%  w @ \vec e \evalsto w' & \mbox{applying $w$ to eliminations $\vec e$ evaluates to $w'$} \\
\end{array}
\]

\fbox{$t \evalsto w$}
\begin{gather*}
  \ru{}{w \evalsto w}
\qquad
  \ru{t \evalsto w \qquad w @ e \evalsto w'}{t\,e \evalsto w'}
\end{gather*}

% \fbox{$w @ \vec e \evalsto w'$}
% \begin{gather*}
%   \ru{}{w @ () \evalsto w}
% \qquad
%   \ru{w @ e \evalsto w' \qquad
%       w' @ \vec e \evalsto w''
%     }{w @ (e, \vec e) \evalsto w''}
% \end{gather*}

\fbox{$w @ e \evalsto w'$}
\begin{gather*}
  \ru{t[u] \evalsto w
    }{(\lambda t) @ u \evalsto w}
\qquad
  \ru{t[a] \evalsto w
    }{(\lambda t) @ a \evalsto w}
\qquad
  \ru{t[a] \evalsto w
    }{(\lambda t) @ \ann a \evalsto w}
\\[2ex]
  \ru{t_z \evalsto w
    }{\tzero @ \casett \evalsto w}
\qquad
  \ru{t_s\, t \evalsto w
    }{(\suc a t) @ \casett \evalsto w}
\qquad
  \rux{t\,(\lambda x.\, x@ \fixtt)\,w \evalsto w'
     }{w @ \fixtt \evalsto w'
     }{w \in \{ \tzero, \suc a u \}}
\end{gather*}

\newpage

\section{Conclusions}
\label{sec:concl}

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Bibliography
\bibliography{medium}

%% Appendix
\appendix
\section{Appendix}

Text of appendix \ldots

\end{document}

### 2017-02-05 nbe/icfp17.tex
[ ICFP 17 ] Introduced category Data (c).

@@
     \mid \lambda t% & \mbox{functions} \\ &&
-    \mid \Nat a \mid \tzero \mid \suc a t & \mbox{weak head normal forms} \\ % \mbox{natural numbers} \\
+    \mid \Nat a \mid c & \mbox{weak head normal forms} \\ % \mbox{natural numbers} \\
+\Data & \ni & c
+   & ::= & \tzero \mid \suc a t & \mbox{constructed data} \\
 \Ne & \ni & n
@@
     & ::= & t \mid a \mid \ann a \mid \case[\ell] T {t_1} t_2 \mid \fix[\ell] T t & \mbox{eliminations} \\
-\SizeExp & \ni & a,b,c & ::= & \infty \mid \ind i + o ~(o \in \NN) & \mbox{size expressions} \\
+\SizeExp & \ni & a,b & ::= & \infty \mid \ind i + o ~(o \in \NN) & \mbox{size expressions} \\
 \Cxt & \ni & \Gamma, \Delta & ::= & \cempty \mid \erext[\noterased] \Gamma T \mid \erext \Gamma \Size \\
@@
 For instance, $\funT i \Size \Nat i \to \Set[\ell]$ is sugar for $\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.
-Similar, we use named lambda abstraction as sugar for de Bruijn.
+We abbreviate this type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for $\allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$.
+Similar as for $\Pi$, we use named lambda abstraction as sugar for de Bruijn abstraction.
 Named abstract takes care of proper lifting of de Bruijn indices.
@@
 \\[2ex]
-  \ru{\Gamma \der u : \Nat b \qquad
-      \Gamma \der T : \funT i \Size \Nat i \to \Set[\ell] \\
-      \Gamma \der t : \allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x
-     }{\Gamma \der u\,\fix[\ell] T t : T \,b \, u}
-\\[2ex]
   \ru{\Gamma \der u : \Nat (b + 1) \qquad
@@
     }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
+\\[2ex]
+  \ru{\Gamma \der u : \Nat b \qquad
+      \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
+      \Gamma \der t : \FixT\, T
+     }{\Gamma \der u\,\fix[\ell] T t : T \,b \, u}
 \end{gather*}
@@
 \\[2ex]
-  \rux{\Gamma \der u : \Nat b \qquad
-      \Gamma \der a : \Size \qquad
-      \Gamma \der T : \funT i \Size \Nat i \to \Set[\ell] \\
-      \Gamma \der t : \allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x
-     }{\Gamma \der u\,\fix[\ell] T t = t\ann a (\lambda x. x\,\fix[\ell] T t)\, u : T \,b \, u
-     }{u \in \{\tzero, \suc a t'\}}
-\\[2ex]
   \ru{\resurrect\Gamma \der b : \Size \qquad
@@
       \Gamma \der t_s : (n : \Nat b) \to T\,(\suc a n)
-    }{\Gamma \der \suc a t\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,\suc a t}
+    }{\Gamma \der (\suc a t)\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,(\suc a t)}
+\\[2ex]
+  \ru{\Gamma \der c : \Nat b \qquad
+      \Gamma \der a : \Size \qquad
+      \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
+      \Gamma \der t : \FixT\,T %% \allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x
+     }{\Gamma \der c\,\fix[\ell] T t = t\ann a (\lambda x. x\,\fix[\ell] T t)\, c : T \,b \, c
+     }%{u \in \{\tzero, \suc a t'\}}
 \end{gather*}
@@
 \\[2ex]
-  \ru{\Gamma \der u = u' : \Nat a \qquad
-      \Gamma \der T = T' : \funT i \Size \Nat i \to \Set[\ell] \\
-      \Gamma \der t = t' : \allT i \Size (\funT x {\Nat i} T \ann i x) \to \funT x {\Nat (i + 1)} T \ann{i+1} x
-     }{\Gamma \der u\,\fix[\ell] T t = u'\,\fix[\ell] {T'} t' : T \ann a u}
-\\[2ex]
   \ru{\Gamma \der u = u' : \Nat (a + 1) \qquad
-      \Gamma \der T = T' : Nat (a + 1) \to \Set[\ell] \qquad
+      \Gamma \der T = T' : \Nat (a + 1) \to \Set[\ell] \qquad
       \Gamma \der t_z = t_z' : T \, \tzero \qquad
@@
     }{\Gamma \der u\,\case[\ell] T {t_z} t_s = u'\,\case[\ell] {T'} {t_z'} t_s' : T \, u}
+\\[2ex]
+  \ru{\Gamma \der u = u' : \Nat a \qquad
+      \Gamma \der T = T' : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
+      \Gamma \der t = t' : \FixT\,T \qquad %% \allT i \Size (\funT x {\Nat i} T \,i\, x) \to \funT x {\Nat (i + 1)} T \,{i+1}\, x
+     }{\Gamma \der u\,\fix[\ell] T t = u'\,\fix[\ell] {T'} t' : T \, a \, u}
 \end{gather*}
@@
 \qquad
-  \rux{t\,(\lambda x.\, x@ \fixtt)\,w \evalsto w'
-     }{w @ \fixtt \evalsto w'
-     }{w \in \{ \tzero, \suc a u \}}
+  \ru{t\,\ann\infty\,(\lambda x.\, x@ \fixtt)\,c \evalsto w
+     }{c @ \fixtt \evalsto w
+     }%{w \in \{ \tzero, \suc a u \}}
 \end{gather*}

### 2017-02-06 nbe/icfp17.tex
[ ICFP 17 ] Before removing term derivation witness from equality.

@@
 %\documentclass[acmlarge]{acmart}\settopmatter{}
+
+\usepackage[cal=boondoxo]{mathalfa}
 
@@
     & ::= & t \mid a \mid \ann a \mid \case[\ell] T {t_1} t_2 \mid \fix[\ell] T t & \mbox{eliminations} \\
-\SizeExp & \ni & a,b & ::= & \infty \mid \ind i + o ~(o \in \NN) & \mbox{size expressions} \\
+\SizeExp & \ni & a,b & ::= & \infty \mid k \mid \ind i + k ~(k \in \NN) & \mbox{size expressions} \\
 \Cxt & \ni & \Gamma, \Delta & ::= & \cempty \mid \erext[\noterased] \Gamma T \mid \erext \Gamma \Size \\
@@
 Similar as for $\Pi$, we use named lambda abstraction as sugar for de Bruijn abstraction.
-Named abstract takes care of proper lifting of de Bruijn indices.
+Named abstract takes care of proper lifting of de Bruijn indices, for instance, $\lambda x.\, t x = \lambda.
+(t \slift)\,\ind 0$ if $t$ is outside the scope of $x$.
 
@@
 We write $\sigma$ and $\tau$ for substitutions, which are lists of terms, and $t \sigma$ for the application of substitution $\sigma$ to term $t$.
-We write $\slift^n_m$ for the substitution $(v_{n+m-1},\dots,v_{n+1},v_n)$ which applies to a term with $m$ free indices and increases all of them by $n$.
+We write $\slift^k_m$ for the substitution $(v_{k+m-1},\dots,v_{k+1},v_k)$ which applies to a term with $m$ free indices and increases all of them by $k$.
 We write $\slift_m$ for the lifting $\slift^1_m$ and $\sid_m$ for the identity substitution $ \slift^0_m$.
@@
 
-Size increment \fbox{$a + k$} for $k \in \NN$ is defined by $\infty + k = \infty$ and $(\ind i + o) + k = \ind i + (o + k)$.
-Size comparison \fbox{$a \leq b$} holds if either $b = \infty$ or $a = \ind i + o$ and $b = \ind i + o'$ with $o \leq o'$.
+Size increment \fbox{$a + k$} for $k \in \NN$ extends addition by $\infty + k = \infty$ and $(\ind i + k') + k = \ind i + (k' + k)$.
+Size comparison \fbox{$a \leq b$} holds if either $b = \infty$ or  $k \leq k'$ where either $a = k$ and $b = k'$ or $a \in \{ k, \ind i + k \}$ and $b = \ind i + k'$.
 
@@
 
-\fbox{$\Gamma(i) = \eann T$} (implies $\Gamma \der T$).
+\fbox{$\Gamma(i) = \eann T$} (implies $\resurrect\Gamma \der T$).
 \[
@@
       \Gamma \der a : \Size \qquad
-      \Gamma \der t_s : (n : \Nat b) \to T \, (\suc a n)
+      \Gamma \der t_s : (t : \Nat b) \to T \, (\suc a t)
     }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
@@
       \Gamma \der a : \Size \qquad
-      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc a n)
+      \Gamma \der t_s : (t : \Nat b) \to T\,(\suc a t)
     }{\Gamma \der \tzero\,\case[\ell] T {t_z} {t_s} = t_z : T\,\tzero}
@@
 
+\begin{lemma}[Inversion of typing]
+\label{lem:invty} \bla
+\begin{enumerate}
+\item TODO
+\end{enumerate}
+\end{lemma}
+
 \section{Semantics}
 \label{sec:sem}
+
+\subsection{Operational semantics}
 
@@
 
+We write \fbox{$t \approx t'$} iff $t$ and $t'$ are syntactically equal if we ignore sizes $\ann a$ in irrelevant positions.
+For example, $\suc a \tzero \approx \suc 0 \tzero$, but $\Nat a \not\approx \Nat 0$.
+
+\subsection{Logical predicates}
+
+We write $\forall \rho : \Gamma \leq \Delta.~ P(\Gamma,\rho)$ short for
+$\forall \Gamma, \rho.~ \Gamma \der \rho : \Delta \implies P(\Gamma,\rho)$.
+We apply implicit binding (like $\Gamma$ here) in other, analogous cases as well.
+
+We write \fbox{$\Gamma \der t :=: t' : T$} for the conjunction of $\Gamma \der t,t' : T$ and $\Gamma \der t = t' : T$.  We write \fbox{$\Gamma \der t \evalsto w : T$} if additionally $t \evalsto w$.
+
+We define a family of PERs \fbox{$\sneq \Delta \_ \_ T$} that characterize the neutrals of type $T$.
+\[
+  \sneq \Delta t {t'} T \miff \Delta \der t \evalsto n : T \mand \Delta \der t' \evalsto n' : T
+\]
+
+Assuming $\Delta \der b : \Size$, we define inductively a family \fbox{$\snateq \Delta \_ \_ b$} of relations that characterize the semantic natural numbers.
+\begin{gather*}
+  \ru{\sneq \Delta t {t'} {\Nat b}
+    }{\snateq \Delta t {t'} b
+    }
+\\[2ex]
+  \rux{\Delta \der t \evalsto \tzero : \Nat b \qquad
+      \Delta \der t' \evalsto \tzero : \Nat b
+    }{\snateq \Delta t {t'} b
+    }{b > 0}
+\\[2ex]
+  \rux{\Delta \der t \evalsto \suc a u : \Nat b \qquad
+      \Delta \der t' \evalsto \suc{a'}{u'} : \Nat b \qquad
+      \snateq \Delta u {u'} {b'}
+    }{\snateq \Delta t {t'} b
+    }{b = b' + 1}
+\end{gather*}
+Note that in the last rule $b$ and $b'$ could be both $\infty$.
+We write \fbox{$\snat \Delta t b$} iff $\snateq \Delta t t b$.
+
+By recursion on level $\ell \in \NN$ we define inductively what it means to be a semantic type, \fbox{$\sty \Delta T \ell$}, and simultaneously
+\begin{enumerate}
+\item what it means for semantic types to be equal, \fbox{$\steqd \Delta T {T'} \ell {\T}$},
+\item what it means for a term to have a semantic type, \fbox{$\tcal :: \sttmd \Delta t T {\T}$}, and
+\item what it means for two such terms to be equal, \fbox{$\sttmeqd \Delta t {t'} T {\T} {\tcal}$},
+\end{enumerate}
+all three by recursion on a derivation $\T :: \sty \Delta T \ell$.
+For each $\ell$, this is an instance of an inductive-recursive definition \cite{dybjer:jsl00}.
+
+We first list the rules for \fbox{$\sty \Delta T \ell$}:
+\begin{gather*}
+  \rux{\Delta \der T \evalsto \Set[\ell'] : \Set[\ell]
+     }{\sty \Delta T \ell
+     }{\ell' {<} \ell}
+\qquad
+  \ru{\Delta \der T \evalsto N : \Set[\ell]
+    }{\sty \Delta T \ell}
+\qquad
+  \ru{\Delta \der T \evalsto \Nat a : \Set[\ell]
+    }{\sty \Delta T \ell}
+\\[2ex]
+  \rul{\Delta \der T' \evalsto \epiT \Size T : \Set[\ell] \qquad
+       \eext \Delta \Size \der T : \Set[\ell] \qquad
+      \forall \rho : \Gamma \leq \Delta,~\Gamma \der a : \Size.~
+        \sty \Gamma {T(\rho,a)} \ell \\
+    }{\sty \Delta {T'} \ell}
+\\[2ex]
+  \rul{\Delta \der T' \evalsto \piT U T : \Set[\ell] \qquad \cext \Delta U \der T : \Set[\ell] \\ \relax
+      \U :: \forall \rho : \Gamma \leq \Delta.~ \sty{\Gamma}{U\rho} {\ell} \\ \relax
+      \T :: \forall \rho : \Gamma \leq \Delta,~\sttmd \Gamma u {U\rho} {\U(\rho)}.~
+        \sty \Gamma {T(\rho,u)} \ell \\
+      \forall \rho : \Gamma \leq \Delta,~ \uu :: \sttmd \Gamma u {U\rho} {\U(\rho)}.~
+        \sttmeqd \Gamma u {u'} {U\rho} {\U(\rho)} {\uu} \implies
+        \steqd \Gamma {T(\rho,u)} {T(\rho,u')} \ell {\T(\rho,\uu)}
+    }{\sty \Delta {T'} \ell}
+\end{gather*}
+Then, we give the cases for \fbox{$\steqd \Delta T {T'} \ell {\T}$}.
+\begin{caselist}
+
+\nextcase Universes:
+\[
+  \rux{\Delta \der T \evalsto \Set[\ell'] : \Set[\ell]
+     }{\sty \Delta T \ell
+     }{\ell' {<} \ell}
+\]
+Then $\steqd \Delta T {T'} \ell {\T}$ iff $\Delta \der T' \evalsto \Set[\ell'] : \Set[\ell]$.
+
+\nextcase Neutrals:
+\[
+  \ru{\Delta \der T \evalsto N : \Set[\ell]
+    }{\sty \Delta T \ell}
+\]
+Then $\steqd \Delta T {T'} \ell {\T}$ iff $\Delta \der T' \evalsto N' : \Set[\ell]$ and
+$\Delta \der N = N' : \Set[\ell]$.
+
+\nextcase Natural number type:
+\[
+  \ru{\Delta \der T \evalsto \Nat a : \Set[\ell]
+    }{\sty \Delta T \ell}
+\]
+Then $\steqd \Delta T {T'} \ell {\T}$ iff $\Delta \der T' \evalsto \Nat a : \Set[\ell]$.
+
+\nextcase Size quantification:
+\[
+  \rul{\Delta \der T \evalsto \epiT \Size U : \Set[\ell] \qquad
+       \eext \Delta \Size \der U : \Set[\ell] \qquad
+      \U :: \forall \rho : \Gamma \leq \Delta,~\Gamma \der a : \Size.~
+        \sty \Gamma {U(\rho,a)} \ell \\
+    }{\sty \Delta {T} \ell}
+\]
+Then $\steqd \Delta T {T'} \ell {\T}$ iff $\Delta \der T' \evalsto \epiT \Size {U'} : \Set[\ell]$
+and $\eext \Delta \Size \der U' : \Set[\ell]$ and forall $\rho : \Gamma \leq \Delta$ and $\Gamma \der a : \Size$ we have $\steqd \Gamma {U(\rho,a)} {U'(\rho,a)} \ell {\U(\rho,a)}$.
+
+\nextcase Function spaces:
+\[
+  \rul{\Delta \der T_1' \evalsto \piT {U_1} {T_1} : \Set[\ell] \qquad \cext \Delta U_1 \der T_1 : \Set[\ell] \\ \relax
+      \U :: \forall \rho : \Gamma \leq \Delta.~ \sty{\Gamma}{U_1\rho} {\ell} \\ \relax
+      \T :: \forall \rho : \Gamma \leq \Delta,~\sttmd \Gamma u {U_1\rho} {\U(\rho)}.~
+        \sty \Gamma {T_1(\rho,u)} \ell \\
+      \forall \rho : \Gamma \leq \Delta,~ \uu :: \sttmd \Gamma u {U_1\rho} {\U(\rho)}.~
+        \sttmeqd \Gamma u {u'} {U_1\rho} {\U(\rho)} {\uu} \implies
+        \steqd \Gamma {T_1(\rho,u)} {T_1(\rho,u')} \ell {\T(\rho,\uu)}
+    }{\sty \Delta {T_1'} \ell}
+\]
+Then $\steqd \Delta {T_1'} {T_2'} \ell {\T}$ iff $\Delta \der T_2' \evalsto \piT{U_2}{T_2} : \Set[\ell]$ and
+$\cext \Delta{U_2} \der T_2 : \Set[\ell]$ and
+forall $\rho : \Gamma \leq \Delta$ we have $\steqd \Gamma {U_1\rho} {U_2\rho} \ell {\U(\rho)}$
+and forall $\uu :: \sttmd \Gamma u {U_1\rho} {\U(\rho)}$ finally
+$\steqd \Gamma {T_1(\rho,u)} {T_2(\rho,u)} \ell {\T(\rho,\uu)}$.
+
+\end{caselist}
+
+Next, we define semantic membership \fbox{$\sttmd \Delta t T {\T}$}.
+\begin{caselist}
+
+\nextcase
+\[
+  \rux{\Delta \der T \evalsto \Set[\ell'] : \Set[\ell]
+     }{\sty \Delta T \ell
+     }{\ell' {<} \ell}
+\]
+Then $\sttmd \Delta t T {\T}$ iff $\sty \Delta t {\ell'}$.
+
+\nextcase
+\[
+  \ru{\Delta \der T \evalsto N : \Set[\ell]
+    }{\sty \Delta T \ell}
+\]
+Then $\sttmd \Delta t T {\T}$ iff $\Delta \der t \evalsto n : N$.
+
+\nextcase
+\[
+  \ru{\Delta \der T \evalsto \Nat a : \Set[\ell]
+    }{\sty \Delta T \ell}
+\]
+Then $\sttmd \Delta t T {\T}$ iff $\snat \Delta t a$.
+
+\nextcase %  Let $\T' :: \sty \Delta {T'} \ell$ be derived by
+\[
+  \T' ::
+  \rul{\Delta \der T' \evalsto \piT \Size T : \Set[\ell] \qquad
+       \cext \Delta \Size \der T : \Set[\ell] \qquad
+      \T :: \forall \rho : \Gamma \leq \Delta,~\Gamma \der a : \Size.~
+        \sty \Gamma {T(\rho,a)} \ell \\
+    }{\sty \Delta {T'} \ell}
+\]
+Then $\sttmd \Delta t {T'} {\T'}$ iff $\Delta \der t : T'$ and
+forall $\rho : \Gamma \leq \Delta$ and $\acal :: \Gamma \der a : \Size$
+we have $\sttmd \Gamma {t\,a} {T(\rho,a)} {\T(\rho,\acal)}$.
+
+\nextcase % Let $\T' :: \sty \Delta {T'} \ell$ be derived by
+\[
+  \T' ::
+  \rul{\Delta \der T' \evalsto \erpiT \Size T : \Set[\ell] \qquad
+       \erext \Delta \Size \der T : \Set[\ell] \qquad
+      \T :: \forall \rho : \Gamma \leq \Delta,~\Gamma \der b : \Size.~
+        \sty \Gamma {T(\rho,b)} \ell \\
+    }{\sty \Delta {T'} \ell}
+\]
+Then $\sttmd \Delta t {T'} {\T'}$ iff $\Delta \der t : T'$ and
+forall $\rho : \Gamma \leq \Delta$ and $\bb :: \Gamma \der b : \Size$ and  $\resurrect\Gamma \der a : \Size$,
+\begin{enumerate}
+\item $\tcal :: \sttmd \Gamma {t \ann a} {T(\rho,b)} {\T(\rho,\bb)}$, and
+\item $\sttmeqd \Gamma {t \ann a} {t \ann {a'}} {T(\rho,b)} {\T(\rho,\bb)} {\tcal}$ for all $\resurrect\Gamma \der a' : \Size$.
+\end{enumerate}
+
+\nextcase
+\[
+  \T' ::
+  \rul{\Delta \der T' \evalsto \piT U T : \Set[\ell] \qquad \cext \Delta U \der T : \Set[\ell] \\ \relax
+      \U :: \forall \rho : \Gamma \leq \Delta.~ \sty{\Gamma}{U\rho} {\ell} \\ \relax
+      \T :: \forall \rho : \Gamma \leq \Delta,~\sttmd \Gamma u {U\rho} {\U(\rho)}.~
+        \sty \Gamma {T(\rho,u)} \ell \\
+      \forall \rho : \Gamma \leq \Delta,~ \uu :: \sttmd \Gamma u {U\rho} {\U(\rho)}.~
+        \sttmeqd \Gamma u {u'} {U\rho} {\U(\rho)} {\uu} \implies
+        \steqd \Gamma {T(\rho,u)} {T(\rho,u')} \ell {\T(\rho,\uu)}
+    }{\sty \Delta {T'} \ell}
+\]
+Then $\sttmd \Delta t {T'} {\T'}$ iff $\Delta \der t : T'$ and
+forall $\rho : \Gamma \leq \Delta$ and $\uu :: \sttmd \Gamma u {U\rho} {\U(\rho)}$,
+\begin{enumerate}
+\item $\tcal :: \sttmd \Gamma {t\rho\,u} {T(\rho,u)} {\T(\rho,\uu)}$, and
+\item $\sttmeqd \Gamma {t\rho\,u} {t\rho\,u'} {T(\rho,u)} {\T(\rho,\uu)} {\tcal}$ for all
+  $\sttmeqd \Gamma u {u'} {U\rho} {\U(\rho)} {\uu}$.
+\end{enumerate}
+
+\end{caselist}
+
+Finally, we define \fbox{$\sttmeqd \Delta t {t'} T {\T} {\tcal}$}
+where $\tcal :: \sttmd \Delta t T {\T}$.
+\begin{caselist}
+
+\nextcase
+\[
+  \T ::
+  \rux{\Delta \der T \evalsto \Set[\ell'] : \Set[\ell]
+     }{\sty \Delta T \ell
+     }{\ell' {<} \ell}
+\]
+Then $\sttmeqd \Delta t {t'} T {\T} {\tcal}$ iff
+% $\T' : \sty \Delta {t} \ell$ and
+$\steqd \Delta t {t'} {\ell'} {\tcal}$.
+
+\nextcase
+\[
+  \ru{\Delta \der T \evalsto N : \Set[\ell]
+    }{\sty \Delta T \ell}
+\]
+Then $\sttmeqd \Delta t {t'}T {\T}$ iff $\Delta \der t \evalsto n : N$.
+
+\nextcase
+\[
+  \ru{\Delta \der T \evalsto \Nat a : \Set[\ell]
+    }{\sty \Delta T \ell}
+\]
+Then $\sttmd \Delta t T {\T}$ iff $\snat \Delta t a$.
+
+\nextcase %  Let $\T' :: \sty \Delta {T'} \ell$ be derived by
+\[
+  \T' ::
+  \rul{\Delta \der T' \evalsto \piT \Size T : \Set[\ell] \qquad
+       \cext \Delta \Size \der T : \Set[\ell] \qquad
+      \T :: \forall \rho : \Gamma \leq \Delta,~\Gamma \der a : \Size.~
+        \sty \Gamma {T(\rho,a)} \ell \\
+    }{\sty \Delta {T'} \ell}
+\]
+Then $\sttmd \Delta t {T'} {\T'}$ iff $\Delta \der t : T'$ and
+forall $\rho : \Gamma \leq \Delta$ and $\acal :: \Gamma \der a : \Size$
+we have $\sttmd \Gamma {t\,a} {T(\rho,a)} {\T(\rho,\acal)}$.
+
+\nextcase % Let $\T' :: \sty \Delta {T'} \ell$ be derived by
+\[
+  \T' ::
+  \rul{\Delta \der T' \evalsto \erpiT \Size T : \Set[\ell] \qquad
+       \erext \Delta \Size \der T : \Set[\ell] \qquad
+      \T :: \forall \rho : \Gamma \leq \Delta,~\Gamma \der b : \Size.~
+        \sty \Gamma {T(\rho,b)} \ell \\
+    }{\sty \Delta {T'} \ell}
+\]
+Then $\sttmd \Delta t {T'} {\T'}$ iff $\Delta \der t : T'$ and
+forall $\rho : \Gamma \leq \Delta$ and $\bb :: \Gamma \der b : \Size$ and  $\resurrect\Gamma \der a : \Size$,
+\begin{enumerate}
+\item $\tcal :: \sttmd \Gamma {t \ann a} {T(\rho,b)} {\T(\rho,\bb)}$, and
+\item $\sttmeqd \Gamma {t \ann a} {t \ann {a'}} {T(\rho,b)} {\T(\rho,\bb)} {\tcal}$ for all $\resurrect\Gamma \der a' : \Size$.
+\end{enumerate}
+
+\nextcase
+\[
+  \T' ::
+  \rul{\Delta \der T' \evalsto \piT U T : \Set[\ell] \qquad \cext \Delta U \der T : \Set[\ell] \\ \relax
+      \U :: \forall \rho : \Gamma \leq \Delta.~ \sty{\Gamma}{U\rho} {\ell} \\ \relax
+      \T :: \forall \rho : \Gamma \leq \Delta,~\sttmd \Gamma u {U\rho} {\U(\rho)}.~
+        \sty \Gamma {T(\rho,u)} \ell \\
+      \forall \rho : \Gamma \leq \Delta,~ \uu :: \sttmd \Gamma u {U\rho} {\U(\rho)}.~
+        \sttmeqd \Gamma u {u'} {U\rho} {\U(\rho)} {\uu} \implies
+        \steqd \Gamma {T(\rho,u)} {T(\rho,u')} \ell {\T(\rho,\uu)}
+    }{\sty \Delta {T'} \ell}
+\]
+Then $\sttmd \Delta t {T'} {\T'}$ iff $\Delta \der t : T'$ and
+forall $\rho : \Gamma \leq \Delta$ and $\uu :: \sttmd \Gamma u {U\rho} {\U(\rho)}$,
+\begin{enumerate}
+\item $\tcal :: \sttmd \Gamma {t\rho\,u} {T(\rho,u)} {\T(\rho,\uu)}$, and
+\item $\sttmeqd \Gamma {t\rho\,u} {t\rho\,u'} {T(\rho,u)} {\T(\rho,\uu)} {\tcal}$ for all
+  $\sttmeqd \Gamma u {u'} {U\rho} {\U(\rho)} {\uu}$.
+\end{enumerate}
+
+\end{caselist}
+
 \newpage
+
+\section{Metatheory}
+\label{sec:meta}
+
+\begin{lemma}[Typing of constructed data]
+\label{lem:invdata} \bla
+If\/ $\Gamma \der c : \Nat b$ then $\resurrect\Gamma \der b : \Size$ and $b > 0$.
+\end{lemma}
 

### 2017-02-11 prototyping/size-irrelevance/Evaluation.hs
[ Sit ] started on prototype implementation for ICFP 17.

@@ -1,0 +1,9 @@
+-- | Substitution and weak head evaluation
+
+module Evaluation
+
+import Internal
+
+
+
+class Subst a where

### 2017-02-11 prototyping/size-irrelevance/Internal.hs

@@ -1,0 +1,98 @@
+{-# LANGUAGE GeneralizedNewtypeDeriving #-}
+
+-- | Internal Syntax.
+
+module Internal where
+
+-- | Variables are de Bruijn indices.
+
+newtype Index = Index { dbIndex :: Int }
+  deriving (Eq, Ord, Show, Enum, Num)
+
+-- | Size expressions.
+
+data Size
+  = SVar Index Integer
+    -- ^ @i + k@: Size variable plus offset.
+  | SConst Integer
+    -- ^ @k@:     Constant finite size.
+  | SInfty
+    -- ^ @oo@:    Infinity size.
+  deriving (Eq, Ord, Show)
+
+type Level = Size
+
+-- | Terms/types
+
+type Type = Term
+
+data Term
+  = -- | Universe with level.
+    Type Level
+  | -- | Type of sizes (internal use only).
+    Size
+  | -- | Sized natural number type.
+    Nat Size
+  | -- | Zero constructor.
+    Zero
+  | -- | Successor constructor.
+    Suc Term
+  | -- ^ (Dependent) function type.
+    Pi (Dom Type) (Abs Term)
+  | -- ^ Lambda abstraction
+    Lam ArgInfo (Abs Term)
+  | -- ^ Application (neutral).
+    Var Index Elims
+  deriving (Eq, Ord, Show)
+
+-- | Eliminations.
+
+type Elims = [ Elim ]
+
+data Elim
+  = -- | Function application.
+    Apply (Arg Term)
+  | -- | Case distinction
+    Case
+    { caseReturn :: Type -- ^ @T : Nat (b + 1) -> Setω@
+    , caseZero   :: Term -- ^ @tz : T zero@
+    , caseSuc    :: Term -- ^ @ts : (t : Nat b) -> T (suc t)@
+    }
+  | -- | Recursion
+    Fix
+    { fixReturn :: Term
+      -- ^ @T : ..(i : Size) -> Nat i -> Setω@
+    , fixBody   :: Term
+      -- ^ @t : .(i : Size) (f : (x : Nat i) -> T i x) (x : Nat (i + 1)) -> T (i + 1) x@
+    }
+  deriving (Eq, Ord, Show)
+
+-- | Abstraction.
+
+type AbsName = String
+
+data Abs a
+  = -- | Actual abstraction (body contains one more index).
+    Abs   { absName :: AbsName, absBody :: a }
[…]

### 2017-02-11 prototyping/size-irrelevance/Sit.hs

@@ -1,0 +1,45 @@
+{-# LANGUAGE LambdaCase #-}
+
+import System.Environment (getArgs)
+import System.Exit (exitFailure)
+
+import Data.Foldable
+
+import Sit.Abs
+import Sit.ErrM
+import Sit.Lex
+import Sit.Par
+import Sit.Print
+
+import TypeChecker
+
+main :: IO ()
+main = getArgs >>= \case
+  [file] -> check =<< readFile file
+  _ -> usage
+
+usage :: IO ()
+usage = do
+  putStrLn "Usage: Sit <SourceFile>"
+  exitFailure
+
+-- | Handle error by failing hard.
+
+failOnErr :: String -> Err a -> IO a
+failOnErr msg = \case
+  Ok a    -> return a
+  Bad err -> exitMsg $ unlines [ msg , err ]
+
+exitMsg :: String -> IO a
+exitMsg msg = do
+  putStrLn msg
+  exitFailure
+
+-- | Run the type checker on file contents
+check :: String -> IO ()
+check txt = do
+  Prg decls <- failOnErr "PARSE ERROR" $ pPrg $ myLexer txt
+  putStrLn "Parsed the following declarations"
+  forM_ decls $ \ d -> do
+    putStrLn $ printTree d
+  either (\ err -> exitMsg $ unlines [ "TYPE ERROR" , err ]) return $ typeCheck decls

### 2017-02-11 prototyping/size-irrelevance/TypeChecker.hs

@@ -1,0 +1,82 @@
+{-# LANGUAGE LambdaCase #-}
+{-# LANGUAGE TupleSections #-}
+
+module TypeChecker where
+
+import Control.Applicative
+import Control.Monad
+import Control.Monad.Except
+import Control.Monad.Reader
+import Control.Monad.State
+
+import Data.Functor
+import Data.Maybe
+import Data.Map (Map)
+import qualified Data.Map as Map
+import qualified Data.Set as Set
+
+import Debug.Trace
+
+import qualified Sit.Abs as A
+import Sit.Print
+import Sit.ErrM
+
+import Internal
+
+-- | Type errors are just strings.
+type TypeError = String
+
+-- | The type checking monad
+type Check = ReaderT Env (StateT St (Except TypeError))
+
+-- | Local context
+
+data Env = Env
+  { envCxt :: Cxt
+  }
+
+type Cxt = [ (Id, Type) ]
+
+type Id = String
+
+-- | Global state
+
+data St = St
+  { stSig :: Sig
+  }
+
+type Sig = Map Id Function
+
+data Function = Function
+  { funType :: Type
+  , funVal  :: Term
+  }
+
+typeCheck :: [A.Decl] -> Either String ()
+typeCheck decls = runExcept (evalStateT (runReaderT (checkDecls decls) initEnv) initSt)
+  where
+  initEnv = Env { envCxt = [] }
+  initSt  = St  { stSig  = Map.empty }
+
+checkDecls :: [A.Decl] -> Check ()
+checkDecls = mapM_ checkDecl
+
+checkDecl :: A.Decl -> Check ()
+checkDecl = \case
+  A.Open{} -> return ()
+  A.Sig x a -> checkSig x a
+  A.Def x e -> checkDef x e
+
+checkSig :: A.Ident -> A.Exp -> Check ()
+checkSig x a = traceCheck (A.Sig x a) $ do
+  nyi "checking a type signature"
+
+checkDef :: A.Ident -> A.Exp -> Check ()
+checkDef x e = traceCheck (A.Def x e) $ do
+  nyi "checking a  definition"
+
+traceCheck :: Print a => a -> b -> b
+traceCheck a = trace $ "Checking " ++ printTree a
+
+nyi :: String -> Check a
+nyi = throwError . ("Not yet implemented: " ++)

### 2017-02-11 prototyping/size-irrelevance/Evaluation.hs
[ Sit ] Started on evaluation, before giving up on size expressions.

@@ -1,9 +1,70 @@
--- | Substitution and weak head evaluation
+{-# LANGUAGE CPP #-}
+{-# LANGUAGE LambdaCase #-}
+{-# LANGUAGE TypeSynonymInstances, FlexibleInstances, MultiParamTypeClasses #-}
 
-module Evaluation
+-- | Values and weak head evaluation
 
+module Evaluation where
+
+import Control.Applicative
+import Control.Monad
+
+import Sit.Abs (Ident)
 import Internal
 
+import Impossible
+#include "undefined.h"
 
+-- | Generic values are de Bruijn levels.
+type VGen   = Int
 
-class Subst a where
+type VSize  = Size' VGen
+type VLevel = VSize
+
+type VType  = Val
+type VElim  = Elim' VGen
+type VElims = [VElim]
+
+data Val
+  = VType VLevel
+  | VSize
+  | VNat VSize
+  | VZero
+  | VSuc Val
+  | VPi (Dom VType) VClos
+  | VLam VClos
+  | -- | Neutrals.
+    VUp VType VNe
+  | -- | Type annotation for readback
+    VDown VType Val
+
+data VNe
+  = VNe VLevel VElims
+
+data VClos = VClos
+  { closBody :: Abs Term
+  , closEnv  :: Env
+  }
+
+-- | An environment maps de Bruijn indices to values.
+--   The first entry in the list is the binding for index 0.
+
+type Env = [Val]
+
+-- | Evaluation monad.
+
+class (Functor m, Applicative m, Monad m) => MonadEval m where
+  getDef :: Ident -> m Val
+
+-- | Evaluation.
+
+class Evaluate a b where
+  eval :: MonadEval m => a -> Env -> m b
+
+instance Evaluate Size VSize where
+  eval a rho = case a of
+    SVar i k ->  __IMPOSSIBLE__
+
+instance Evaluate Term Val where
+  eval t rho = case t of
+    _ -> __IMPOSSIBLE__

### 2017-02-11 prototyping/size-irrelevance/Internal.hs

@@ -1,8 +1,16 @@
 {-# LANGUAGE GeneralizedNewtypeDeriving #-}
+{-# LANGUAGE DeriveFunctor #-}
+{-# LANGUAGE DeriveFoldable #-}
+{-# LANGUAGE DeriveTraversable #-}
 
 -- | Internal Syntax.
 
 module Internal where
+
+import Data.Foldable (Foldable)
+import Data.Traversable (Traversable)
+
+import Sit.Abs (Ident)
 
 -- | Variables are de Bruijn indices.
 
@@ -11,15 +19,16 @@ newtype Index = Index { dbIndex :: Int }
 
 -- | Size expressions.
 
-data Size
-  = SVar Index Integer
+data Size' a
+  = SVar a Integer
     -- ^ @i + k@: Size variable plus offset.
   | SConst Integer
     -- ^ @k@:     Constant finite size.
   | SInfty
     -- ^ @oo@:    Infinity size.
-  deriving (Eq, Ord, Show)
+  deriving (Eq, Ord, Show, Functor, Foldable, Traversable)
 
+type Size = Size' Index
 type Level = Size
 
 -- | Terms/types
@@ -43,29 +52,32 @@ data Term
     Lam ArgInfo (Abs Term)
   | -- ^ Application (neutral).
     Var Index Elims
+  | -- ^ Function call
+    Def Ident Elims
   deriving (Eq, Ord, Show)
 
 -- | Eliminations.
 
 type Elims = [ Elim ]
+type Elim  = Elim' Term
 
-data Elim
+data Elim' a
   = -- | Function application.
     Apply (Arg Term)
   | -- | Case distinction
     Case
-    { caseReturn :: Type -- ^ @T : Nat (b + 1) -> Setω@
-    , caseZero   :: Term -- ^ @tz : T zero@
-    , caseSuc    :: Term -- ^ @ts : (t : Nat b) -> T (suc t)@
+    { caseReturn :: a -- ^ @T : Nat (b + 1) -> Setω@
+    , caseZero   :: a -- ^ @tz : T zero@
+    , caseSuc    :: a -- ^ @ts : (t : Nat b) -> T (suc t)@
     }
   | -- | Recursion
     Fix
-    { fixReturn :: Term
+    { fixReturn :: a
       -- ^ @T : ..(i : Size) -> Nat i -> Setω@
-    , fixBody   :: Term
+    , fixBody   :: a
[…]

### 2017-02-11 prototyping/size-irrelevance/Substitute.hs

@@ -1,0 +1,157 @@
+{-# LANGUAGE CPP #-}
+{-# LANGUAGE LambdaCase #-}
+{-# LANGUAGE TypeSynonymInstances, FlexibleInstances #-}
+
+-- | Substitution and weak head evaluation
+
+module Evaluation where
+
+import Internal
+
+import Impossible
+#include "undefined.h"
+
+-- | Substitutions are lists of terms or size expressions
+--   or variables that can denote either.
+
+type Subst = [SubstEntry]
+
+data SubstEntry
+  = STerm  Term
+  | SSize  Size
+  | SIndex Index
+
+sizeEntry :: SubstEntry -> Size
+sizeEntry = \case
+  SSize a  -> a
+  SIndex i -> SVar i 0
+  STerm{}  -> __IMPOSSIBLE__
+
+termEntry :: SubstEntry -> Term
+termEntry = \case
+  STerm t  -> t
+  SIndex i -> Var i []
+  SSize{}  -> __IMPOSSIBLE__
+
+-- | Weakening substitution @Γ.Δ ⊢ wkS |Δ| : Γ@
+
+wkS :: Int -> Subst
+wkS n = map (SIndex . Index) [n,n+1..]
+
+-- | Identity substitution @Γ ⊢ idS : Γ@.
+
+idS :: Subst
+idS = wkS 0
+
+-- | Composing substitution
+--   @
+--      Γ₁ ⊢ τ : Γ₂    Γ₂ ⊢ σ : Γ₃
+--      -------------------------
+--      Γ₁ ⊢ compS τ σ : Γ₃
+--   @
+
+compS :: Subst -> Subst -> Subst
+compS = subst
+
+-- | Extending a substitution
+--   @
+--      Γ ⊢ σ : Δ    Δ ⊢ T    Γ ⊢ t : Tσ
+--      --------------------------------
+--      Γ ⊢ consS t σ : Δ.T
+--   @
+
+consS :: SubstEntry -> Subst -> Subst
+consS = (:)
+
+-- | Lifting a substitution under a binder.
+--   @
+--      Γ ⊢ σ : Δ      Δ ⊢ T
+--      --------------------
+--      Γ.Tσ ⊢ liftS σ : Δ.T
+--   @
+
+liftS :: Subst -> Subst
+liftS s = consS (SIndex 0) $ weakS s
+
+-- | Weakening a substitution.
+--
+--   @
[…]

### 2017-02-11 prototyping/size-irrelevance/TypeChecker.hs

@@ -27,11 +27,11 @@ import Internal
 type TypeError = String
 
 -- | The type checking monad
-type Check = ReaderT Env (StateT St (Except TypeError))
+type Check = ReaderT TCEnv (StateT TCSt (Except TypeError))
 
 -- | Local context
 
-data Env = Env
+data TCEnv = TCEnv
   { envCxt :: Cxt
   }
 
@@ -41,7 +41,7 @@ type Id = String
 
 -- | Global state
 
-data St = St
+data TCSt = TCSt
   { stSig :: Sig
   }
 
@@ -55,8 +55,8 @@ data Function = Function
 typeCheck :: [A.Decl] -> Either String ()
 typeCheck decls = runExcept (evalStateT (runReaderT (checkDecls decls) initEnv) initSt)
   where
-  initEnv = Env { envCxt = [] }
-  initSt  = St  { stSig  = Map.empty }
+  initEnv = TCEnv { envCxt = [] }
+  initSt  = TCSt  { stSig  = Map.empty }
 
 checkDecls :: [A.Decl] -> Check ()
 checkDecls = mapM_ checkDecl

### 2017-02-11 prototyping/size-irrelevance/Evaluation.hs
[ Sit ] Before adding size annotations to zero and suc.

@@ -1,6 +1,8 @@
 {-# LANGUAGE CPP #-}
 {-# LANGUAGE LambdaCase #-}
+{-# LANGUAGE ScopedTypeVariables #-}
 {-# LANGUAGE TypeSynonymInstances, FlexibleInstances, MultiParamTypeClasses #-}
+-- {-# LANGUAGE FunctionalDependencies, UndecidableInstances #-}
 
 -- | Values and weak head evaluation
 
@@ -8,6 +10,9 @@ module Evaluation where
 
 import Control.Applicative
 import Control.Monad
+import Control.Monad.Reader
+
+import Data.Traversable (traverse)
 
 import Sit.Abs (Ident)
 import Internal
@@ -18,11 +23,11 @@ import Impossible
 -- | Generic values are de Bruijn levels.
 type VGen   = Int
 
-type VSize  = Size' VGen
+type VSize  = Val
 type VLevel = VSize
 
 type VType  = Val
-type VElim  = Elim' VGen
+type VElim  = Elim' Val
 type VElims = [VElim]
 
 data Val
@@ -31,6 +36,7 @@ data Val
   | VNat VSize
   | VZero
   | VSuc Val
+  | VInfty
   | VPi (Dom VType) VClos
   | VLam VClos
   | -- | Neutrals.
@@ -58,13 +64,58 @@ class (Functor m, Applicative m, Monad m) => MonadEval m where
 
 -- | Evaluation.
 
-class Evaluate a b where
-  eval :: MonadEval m => a -> Env -> m b
+evaluate :: MonadEval m => Term -> Env -> m Val
+evaluate t rho = runReaderT (eval t) rho
 
-instance Evaluate Size VSize where
-  eval a rho = case a of
-    SVar i k ->  __IMPOSSIBLE__
+class Evaluate a b where -- -- | a -> b where
+  eval :: MonadEval m => a -> ReaderT Env m b
+
+instance Evaluate Index Val where
+  eval (Index i) = (!! i) <$> ask
 
 instance Evaluate Term Val where
-  eval t rho = case t of
-    _ -> __IMPOSSIBLE__
+  eval t = case t of
+    Type l   -> VType <$> eval l
+    Nat a    -> VNat <$> eval a
+    Size     -> pure VSize
+    Infty    -> pure VInfty
+    Zero     -> pure VZero
+    Suc t    -> VSuc <$> eval t
+    Pi u t   -> liftA2 VPi (eval u) (eval t)
[…]

### 2017-02-11 prototyping/size-irrelevance/Internal.hs

@@ -19,16 +19,7 @@ newtype Index = Index { dbIndex :: Int }
 
 -- | Size expressions.
 
-data Size' a
-  = SVar a Integer
-    -- ^ @i + k@: Size variable plus offset.
-  | SConst Integer
-    -- ^ @k@:     Constant finite size.
-  | SInfty
-    -- ^ @oo@:    Infinity size.
-  deriving (Eq, Ord, Show, Functor, Foldable, Traversable)
-
-type Size = Size' Index
+type Size  = Term
 type Level = Size
 
 -- | Terms/types
@@ -42,10 +33,12 @@ data Term
     Size
   | -- | Sized natural number type.
     Nat Size
-  | -- | Zero constructor.
+  | -- | Zero constructor / size.
     Zero
-  | -- | Successor constructor.
+  | -- | Successor constructor / size.
     Suc Term
+  | -- | Infinity size.
+    Infty
   | -- ^ (Dependent) function type.
     Pi (Dom Type) (Abs Term)
   | -- ^ Lambda abstraction
@@ -63,7 +56,7 @@ type Elim  = Elim' Term
 
 data Elim' a
   = -- | Function application.
-    Apply (Arg Term)
+    Apply (Arg a)
   | -- | Case distinction
     Case
     { caseReturn :: a -- ^ @T : Nat (b + 1) -> Setω@
@@ -108,3 +101,11 @@ data Relevance
   | ShapeIrr
   | Irrelevant
   deriving (Eq, Ord, Show)
+
+-- * Smart constructor.
+
+var :: Index -> Term
+var i = Var i []
+
+defaultArg :: a -> Arg a
+defaultArg = Arg Relevant

### 2017-02-11 prototyping/size-irrelevance/Substitute.hs

@@ -11,32 +11,14 @@ import Internal
 import Impossible
 #include "undefined.h"
 
--- | Substitutions are lists of terms or size expressions
---   or variables that can denote either.
+-- | Substitutions are lists of terms.
 
-type Subst = [SubstEntry]
-
-data SubstEntry
-  = STerm  Term
-  | SSize  Size
-  | SIndex Index
-
-sizeEntry :: SubstEntry -> Size
-sizeEntry = \case
-  SSize a  -> a
-  SIndex i -> SVar i 0
-  STerm{}  -> __IMPOSSIBLE__
-
-termEntry :: SubstEntry -> Term
-termEntry = \case
-  STerm t  -> t
-  SIndex i -> Var i []
-  SSize{}  -> __IMPOSSIBLE__
+type Subst = [Term]
 
 -- | Weakening substitution @Γ.Δ ⊢ wkS |Δ| : Γ@
 
 wkS :: Int -> Subst
-wkS n = map (SIndex . Index) [n,n+1..]
+wkS n = map (var . Index) [n,n+1..]
 
 -- | Identity substitution @Γ ⊢ idS : Γ@.
 
@@ -60,7 +42,7 @@ compS = subst
 --      Γ ⊢ consS t σ : Δ.T
 --   @
 
-consS :: SubstEntry -> Subst -> Subst
+consS :: Term -> Subst -> Subst
 consS = (:)
 
 -- | Lifting a substitution under a binder.
@@ -71,7 +53,7 @@ consS = (:)
 --   @
 
 liftS :: Subst -> Subst
-liftS s = consS (SIndex 0) $ weakS s
+liftS s = consS (var 0) $ weakS s
 
 -- | Weakening a substitution.
 --
@@ -86,7 +68,7 @@ weakS = compS (wkS 1)
 
 -- | Looking up an entry in a substitution.
 
-lookupS :: Subst -> Index -> SubstEntry
+lookupS :: Subst -> Index -> Term
 lookupS s i =  s !! dbIndex i
 
 -- | Substitution for various syntactic categories.
@@ -102,28 +84,25 @@ class Substitute a where
   substApply :: a -> Subst -> Elims -> a
   substApply t s es = subst s t `applyE` es
 
-instance Substitute SubstEntry where
-  subst s = \case
-    STerm  t -> STerm $ subst s t
-    SSize  t -> SSize $ subst s t
-    SIndex i -> lookupS s i
-
-  applyE t [] = t
[…]

### 2017-02-11 prototyping/size-irrelevance/Evaluation.hs
[ Sit ] started on readback

@@ -34,14 +34,16 @@ data Val
   = VType VLevel
   | VSize
   | VNat VSize
-  | VZero
-  | VSuc Val
+  | VZero VSize
+  | VSuc VSize Val
   | VInfty
   | VPi (Dom VType) VClos
   | VLam VClos
+  | -- | @\ x -> x e@ for internal use in fix.
+    VElimBy VElim
   | -- | Neutrals.
     VUp VType VNe
-  | -- | Type annotation for readback
+  | -- | Type annotation for readback (normal form).
     VDown VType Val
 
 data VNe
@@ -74,23 +76,23 @@ instance Evaluate Index Val where
   eval (Index i) = (!! i) <$> ask
 
 instance Evaluate Term Val where
-  eval t = case t of
+  eval = \case
     Type l   -> VType <$> eval l
     Nat a    -> VNat <$> eval a
     Size     -> pure VSize
     Infty    -> pure VInfty
-    Zero     -> pure VZero
-    Suc t    -> VSuc <$> eval t
+    Zero a   -> VZero <$> eval a
+    Suc a t  -> liftA2 VSuc (eval a) (eval t)
     Pi u t   -> liftA2 VPi (eval u) (eval t)
     Lam ai t -> VLam <$> eval t
     Var x es -> do
       h   <- eval x
       ves <- mapM eval es
-      lift $ apply h ves
+      lift $ applyE h ves
     Def f es -> do
       h   <- lift $ getDef f
       ves <- mapM eval es
-      lift $ apply h ves
+      lift $ applyE h ves
 
 instance Evaluate (Abs Term) VClos where
   eval t = VClos t <$> ask
@@ -104,18 +106,94 @@ instance Evaluate a b => Evaluate (Dom a) (Dom b) where
 instance Evaluate a b => Evaluate (Elim' a) (Elim' b) where
   eval = traverse eval
 
-apply :: MonadEval m => Val -> VElims -> m Val
-apply v []       = return v
-apply v (e : es) = (`apply` es) =<<
+applyE :: MonadEval m => Val -> VElims -> m Val
+applyE v []       = return v
+applyE v (e : es) = (`applyE` es) =<<
   case (v, e) of
-    (VLam cl , Apply u   ) -> applyClos cl $ unArg u
-    (VZero   , Case _ u _) -> return u
[…]

### 2017-02-11 prototyping/size-irrelevance/Internal.hs

@@ -33,10 +33,10 @@ data Term
     Size
   | -- | Sized natural number type.
     Nat Size
-  | -- | Zero constructor / size.
-    Zero
-  | -- | Successor constructor / size.
-    Suc Term
+  | -- | Zero constructor, or zero size (then @Size@ is ignored).
+    Zero Size
+  | -- | Successor constructor, or successor size (then @Size@ is ignored).
+    Suc Size Term
   | -- | Infinity size.
     Infty
   | -- ^ (Dependent) function type.
@@ -109,3 +109,13 @@ var i = Var i []
 
 defaultArg :: a -> Arg a
 defaultArg = Arg Relevant
+
+-- | Zero size.
+
+sZero :: Term
+sZero = Zero Infty
+
+-- | Successor size.
+
+sSuc  :: Term -> Term
+sSuc  = Suc Infty

### 2017-02-13 prototyping/size-irrelevance/Evaluation.hs
[ Sit ] First version of readback type-checks.

@@ -23,6 +23,7 @@ import Impossible
 -- | Generic values are de Bruijn levels.
 type VGen   = Int
 
+-- | We have just a single type of values, including size values.
 type VSize  = Val
 type VLevel = VSize
 
@@ -47,7 +48,7 @@ data Val
     VDown VType Val
 
 data VNe
-  = VNe VLevel VElims
+  = VNe VGen VElims
 
 data VClos = VClos
   { closBody :: Abs Term
@@ -151,7 +152,7 @@ class Readback a b where
   readback :: MonadEval m => a -> ReaderT Int m b
 
 instance Readback VGen Index where
-  readback k = Index (- 1 - n) <$> ask
+  readback k = Index . (\ n -> n - (k + 1)) <$> ask
 
 instance Readback Val Term where
   readback = \case
@@ -160,12 +161,24 @@ instance Readback Val Term where
 
     VDown (VPi a b) d -> do
       v0 <- VUp (unDom a) . (`VNe` []) <$> ask
-      c <- applyClos b v0
-      Lam (domInfo a) <$> do
+      c <- lift $ applyClos b v0
+      Lam (domInfo a) . Abs "x" <$> do
         local succ . readback . VDown c =<< do
-          apply d $ Arg (domInfo a) v0
+          lift $ apply d $ Arg (domInfo a) v0
 
-    VDown (VUp _  ) (VUp _ n) -> readbackNe n
+    VDown (VUp _ _) (VUp _ n) -> readbackNe n
+
+instance Readback a b => Readback [a] [b] where
+  readback = traverse readback
+
+instance Readback a b => Readback (Dom a) (Dom b) where
+  readback = traverse readback
+
+instance Readback a b => Readback (Arg a) (Arg b) where
+  readback = traverse readback
+
+instance Readback a b => Readback (Elim' a) (Elim' b) where
+  readback = traverse readback
 
 readbackType :: MonadEval m => Val -> ReaderT Int m Term
 readbackType = \case
@@ -173,27 +186,30 @@ readbackType = \case
   VType a -> Type <$> readbackSize a
   VNat a  -> Nat  <$> readbackSize a
   VPi a b -> do
-    u  <- traverse readbackType u
+    u  <- traverse readbackType a
     v0 <- VUp (unDom a) . (`VNe` []) <$> ask
[…]

### 2017-02-13 prototyping/size-irrelevance/Evaluation.hs
[ Sit ] Implementing checking of Pi.

@@ -14,11 +14,12 @@ import Control.Monad.Reader
 
 import Data.Traversable (traverse)
 
-import Sit.Abs (Ident)
 import Internal
 
 import Impossible
 #include "undefined.h"
+
+-- * Values
 
 -- | Generic values are de Bruijn levels.
 type VGen   = Int
@@ -39,9 +40,13 @@ data Val
   | VSuc VSize Val
   | VInfty
   | VPi (Dom VType) VClos
-  | VLam VClos
+  -- Functions
+  | -- | Lambda abstraction
+    VLam VClos
   | -- | @\ x -> x e@ for internal use in fix.
     VElimBy VElim
+  -- -- | -- | Constant function
+  -- --   VConst Val
   | -- | Neutrals.
     VUp VType VNe
   | -- | Type annotation for readback (normal form).
@@ -55,6 +60,89 @@ data VClos = VClos
   , closEnv  :: Env
   }
 
+-- | Variable
+
+vVar :: VType -> VGen -> Val
+vVar t x = VUp t $ VNe x []
+
+-- * Size arithmetic
+
+-- | Zero size.
+
+vsZero :: VSize
+vsZero = VZero VInfty
+
+-- | Successor size.
+
+vsSuc :: VSize -> VSize
+vsSuc = VSuc VInfty
+
+-- | Variable size.
+
+vsVar :: VGen -> VSize
+vsVar = vVar VSize
+
+-- | Size increment.
+
+vsPlus :: Int -> VSize -> VSize
+vsPlus n v = iterate vsSuc v !! n
+
+-- | Constant size.
+
+vsConst :: Int -> VSize
+vsConst n = vsPlus n vsZero
+
+-- | View a value as a size expression.
+
+data SizeView
+  = SVConst Int
+    -- ^ @n@
+  | SVVar VGen Int
+    -- ^ @i + n@
+  | SVInfty
+    -- ^ @oo@
+
+-- | Successor size on view.
+
+svSuc :: SizeView -> SizeView
+svSuc = \case
+  SVConst n -> SVConst $ succ n
+  SVVar x n -> SVVar x $ succ n
+  SVInfty   -> SVInfty
+
+-- | View a value as a size expression.
+
+sizeView :: Val -> Maybe SizeView
+sizeView = \case
+  VZero _              -> return $ SVConst 0
+  VSuc _ v             -> svSuc <$> sizeView v
[…]

### 2017-02-13 prototyping/size-irrelevance/Internal.hs

@@ -10,7 +10,9 @@ module Internal where
 import Data.Foldable (Foldable)
 import Data.Traversable (Traversable)
 
-import Sit.Abs (Ident)
+-- | Definition names are strings.
+
+type Id = String
 
 -- | Variables are de Bruijn indices.
 
@@ -46,7 +48,7 @@ data Term
   | -- ^ Application (neutral).
     Var Index Elims
   | -- ^ Function call
-    Def Ident Elims
+    Def Id Elims
   deriving (Eq, Ord, Show)
 
 -- | Eliminations.
@@ -109,6 +111,9 @@ var i = Var i []
 
 defaultArg :: a -> Arg a
 defaultArg = Arg Relevant
+
+defaultDom :: a -> Dom a
+defaultDom = Dom Relevant
 
 -- | Zero size.
 

### 2017-02-13 prototyping/size-irrelevance/TypeChecker.hs

@@ -1,9 +1,13 @@
+{-# LANGUAGE CPP #-}
 {-# LANGUAGE LambdaCase #-}
 {-# LANGUAGE TupleSections #-}
+{-# LANGUAGE TemplateHaskell #-}
+{-# LANGUAGE TypeSynonymInstances, FlexibleInstances, MultiParamTypeClasses #-}
 
 module TypeChecker where
 
 import Control.Applicative
+import Control.Lens hiding (Level)
 import Control.Monad
 import Control.Monad.Except
 import Control.Monad.Reader
@@ -22,41 +26,44 @@ import Sit.Print
 import Sit.ErrM
 
 import Internal
+import Evaluation
+
+import Impossible
+#include "undefined.h"
 
 -- | Type errors are just strings.
 type TypeError = String
 
--- | The type checking monad
-type Check = ReaderT TCEnv (StateT TCSt (Except TypeError))
-
 -- | Local context
 
+type Cxt = [ (Id, Dom VType) ]
+
 data TCEnv = TCEnv
-  { envCxt :: Cxt
+  { _envCxt :: Cxt  -- ^ Typing context.
+  , _envEnv :: Env  -- ^ Default environment.
   }
 
-type Cxt = [ (Id, Type) ]
-
-type Id = String
+makeLenses ''TCEnv
 
 -- | Global state
 
 data TCSt = TCSt
-  { stSig :: Sig
+  { _stTySigs :: Map Id VType
+  , _stDefs   :: Map Id Val
   }
 
-type Sig = Map Id Function
+makeLenses ''TCSt
 
-data Function = Function
-  { funType :: Type
-  , funVal  :: Term
-  }
+-- | The type checking monad
+type Check = ReaderT TCEnv (StateT TCSt (Except TypeError))
+
+-- * Type checker
 
 typeCheck :: [A.Decl] -> Either String ()
 typeCheck decls = runExcept (evalStateT (runReaderT (checkDecls decls) initEnv) initSt)
   where
-  initEnv = TCEnv { envCxt = [] }
-  initSt  = TCSt  { stSig  = Map.empty }
+  initEnv = TCEnv { _envCxt   = []        , _envEnv = []        }
+  initSt  = TCSt  { _stTySigs = Map.empty , _stDefs = Map.empty }
 
 checkDecls :: [A.Decl] -> Check ()
 checkDecls = mapM_ checkDecl
@@ -68,15 +75,147 @@ checkDecl = \case
   A.Def x e -> checkDef x e
 
[…]

### 2017-02-13 prototyping/size-irrelevance/Evaluation.hs
[ Sit ] maxSize can be a total function

@@ -12,6 +12,7 @@ import Control.Applicative
 import Control.Monad
 import Control.Monad.Reader
 
+import Data.Maybe
 import Data.Traversable (traverse)
 
 import Internal
@@ -126,20 +127,17 @@ unSizeView = \case
   SVConst n -> vsConst n
   SVVar x n -> vsPlus n $ vsVar x
 
--- | Compute the maximum of two sizes, if it exists.
+-- | Compute the maximum of two sizes.
 
-maxSize :: VSize -> VSize -> Maybe VSize
-maxSize v1 v2 = do
-  s1 <- sizeView v1
-  s2 <- sizeView v2
-  case (s1, s2) of
-    (SVInfty, _) -> return $ VInfty
-    (_, SVInfty) -> return $ VInfty
-    (SVConst n, SVConst m)          -> return $ unSizeView $ SVConst $ max n m
-    (SVVar x n, SVVar y m) | x == y -> return $ unSizeView $ SVVar x $ max n m
-    (SVConst n, SVVar y m) | n <= m -> return $ unSizeView $ SVVar y m
-    (SVVar x n, SVConst m) | n >= m -> return $ unSizeView $ SVVar x n
-    _ -> Nothing
+maxSize :: VSize -> VSize -> VSize
+maxSize v1 v2 =
+  case ( fromMaybe __IMPOSSIBLE__ $ sizeView v1
+       , fromMaybe __IMPOSSIBLE__ $ sizeView v2) of
+    (SVConst n, SVConst m)          -> unSizeView $ SVConst $ max n m
+    (SVVar x n, SVVar y m) | x == y -> unSizeView $ SVVar x $ max n m
+    (SVConst n, SVVar y m) | n <= m -> unSizeView $ SVVar y m
+    (SVVar x n, SVConst m) | n >= m -> unSizeView $ SVVar x n
+    _ -> VInfty
 
 -- * Evaluation
 

### 2017-02-13 prototyping/size-irrelevance/TypeChecker.hs

@@ -122,8 +122,7 @@ inferType = \case
   e@(A.Arrow a b) -> do
     (u, l1) <- inferType a
     (t, l2) <- inferType b
-    l <- maxLevel e l1 l2
-    return (Pi (defaultDom u) (NoAbs "_" t) , l)
+    return (Pi (defaultDom u) (NoAbs "_" t) , maxSize l1 l2)
 
   e@(A.Pi xs a b) -> do
     (u, l1) <- inferType a
@@ -141,9 +140,9 @@ checkLevel = \case
   e@(A.Var x)    -> checkExp e VSize
   e -> throwError $ "Not a valid level expression: " ++ printTree e
 
-maxLevel :: A.Exp -> VLevel -> VLevel -> Check VLevel
-maxLevel e l1 l2 = maybe failure return $ maxSize l1 l2
-  where failure = throwError $ "Cannot assign a universe level to type " ++ printTree e
+-- maxLevel :: A.Exp -> VLevel -> VLevel -> Check VLevel
+-- maxLevel e l1 l2 = maybe failure return $ maxSize l1 l2
+--   where failure = throwError $ "Cannot assign a universe level to type " ++ printTree e
 
 checkExp :: A.Exp -> VType -> Check Term
 checkExp = \case

### 2017-02-13 prototyping/size-irrelevance/Abstract.hs
[ Sit ] Checking of function types done (but untested).

@@ -1,0 +1,35 @@
+{-# LANGUAGE LambdaCase #-}
+
+-- | Auxiliary functions for abstract syntax.
+
+module Abstract where
+
+import Sit.Abs as A
+
+introduction :: Exp -> Bool
+introduction = \case
+  Var{}    -> False
+  App{}    -> False
+  Case{}   -> False
+  Int{}    -> True
+  Infty    -> True
+  Nat      -> True
+  Set      -> True
+  Set1     -> True
+  Set2     -> True
+  Zero     -> True
+  Suc      -> True
+  Fix      -> True
+  LZero    -> True
+  LSuc     -> True
+  Lam{}    -> True
+  Forall{} -> True
+  Pi{}     -> True
+  Arrow{}  -> True
+  Plus{}   -> True
+  ELam{}   -> True
+
+fromIdU :: A.IdU -> String
+fromIdU = \case
+  A.Id (A.Ident x) -> x
+  A.Under -> "_"

### 2017-02-13 prototyping/size-irrelevance/TypeChecker.hs

@@ -25,6 +25,7 @@ import qualified Sit.Abs as A
 import Sit.Print
 import Sit.ErrM
 
+import Abstract as A
 import Internal
 import Evaluation
 
@@ -74,6 +75,8 @@ checkDecl = \case
   A.Sig x a -> checkSig x a
   A.Def x e -> checkDef x e
 
+-- | Check a type signature.
+
 checkSig :: A.Ident -> A.Exp -> Check ()
 checkSig x0@(A.Ident x) a = traceCheck (A.Sig x0 a) $ do
 
@@ -86,6 +89,8 @@ checkSig x0@(A.Ident x) a = traceCheck (A.Sig x0 a) $ do
   t <- checkType a
   addTySig x =<< evaluate t
 
+-- | Check a definition.
+
 checkDef :: A.Ident -> A.Exp -> Check ()
 checkDef x0@(A.Ident x) e = traceCheck (A.Def x0 e) $ do
 
@@ -94,7 +99,7 @@ checkDef x0@(A.Ident x) e = traceCheck (A.Def x0 e) $ do
   t <- maybe noSig return =<< lookupTySig x
 
   -- Check that x has not yet a definition
-  mv <- lookupTySig x
+  mv <- lookupDef x
   unless (isNothing mv) $
     throwError $ "Duplicate definition of " ++ x
 
@@ -102,36 +107,96 @@ checkDef x0@(A.Ident x) e = traceCheck (A.Def x0 e) $ do
   v <- checkExp e t
   addDef x =<< evaluate v
 
+-- | Check well-formedness of a type.
+
 checkType :: A.Exp -> Check Type
 checkType e = fst <$> inferType e
 
--- | Check that something is a type and infer its universe level
+-- | Check that something is a type and infer its universe level.
+
 inferType :: A.Exp -> Check (Type, VLevel)
-inferType = \case
+inferType e = do
+  let invalidType = throwError $ "Not a valid type expression: " ++ printTree e
+  case e of
 
-  -- Universe
-  A.Set  -> return (Type sZero, vsConst 1)
-  A.Set1 -> return (Type $ sSuc sZero, vsConst 2)
-  A.Set2 -> return (Type $ sSuc $ sSuc sZero, vsConst 3)
-  A.App A.Set l -> do
-    a <- checkLevel l
-    v <- evaluate a
-    return (Type a, vsSuc v)
+    -- Universes
 
-  -- Function types
-  e@(A.Arrow a b) -> do
[…]

### 2017-02-13 prototyping/size-irrelevance/Abstract.hs
[ Sit ] Checking lambdas.

@@ -6,6 +6,22 @@ module Abstract where
 
 import Sit.Abs as A
 
+-- | Can this expression only denote a type?
+
+mustBeType :: Exp -> Bool
+mustBeType = \case
+  Nat      -> True
+  Set      -> True
+  Set1     -> True
+  Set2     -> True
+  Forall{} -> True
+  Pi{}     -> True
+  Arrow{}  -> True
+  App f _  -> mustBeType f
+  _ -> False
+
+-- | Is this expression an introduction?
+
 introduction :: Exp -> Bool
 introduction = \case
   Var{}    -> False

### 2017-02-13 prototyping/size-irrelevance/Evaluation.hs

@@ -52,14 +52,17 @@ data Val
     VUp VType VNe
   | -- | Type annotation for readback (normal form).
     VDown VType Val
+  deriving (Show)
 
 data VNe
   = VNe VGen VElims
+  deriving (Show)
 
 data VClos = VClos
   { closBody :: Abs Term
   , closEnv  :: Env
   }
+  deriving (Show)
 
 -- | Variable
 

### 2017-02-13 prototyping/size-irrelevance/Sit.hs

@@ -39,7 +39,7 @@ exitMsg msg = do
 check :: String -> IO ()
 check txt = do
   Prg decls <- failOnErr "PARSE ERROR" $ pPrg $ myLexer txt
-  putStrLn "Parsed the following declarations"
-  forM_ decls $ \ d -> do
-    putStrLn $ printTree d
+  -- putStrLn "Parsed the following declarations"
+  -- forM_ decls $ \ d -> do
+  --   putStrLn $ printTree d
   either (\ err -> exitMsg $ unlines [ "TYPE ERROR" , err ]) return $ typeCheck decls

### 2017-02-13 prototyping/size-irrelevance/TypeChecker.hs

@@ -210,20 +210,38 @@ checkLevel = \case
 --   where failure = throwError $ "Cannot assign a universe level to type " ++ printTree e
 
 checkExp :: A.Exp -> VType -> Check Term
-checkExp = \case
-  _ -> __IMPOSSIBLE__
-
+checkExp e0 t = do
+  case e0 of
+    A.Lam []     e -> checkExp e t
+    A.Lam (x:xs) e -> do
+      case t of
+        VPi dom cl -> addContext (x, dom) $ do
+          t' <- applyClosure cl =<< lastVal
+          u  <- checkExp (A.Lam xs e) t'
+          return $ Lam (domInfo dom) $ Abs (fromIdU x) u
+        _ -> throwError $ "Lambda abstraction expects function type, but got " ++ show t
+    e -> do
+      (u, ti) <- inferExp e
+      coerce u ti t
+    -- e -> nyi $ "checking " ++ printTree e
 
 -- | Precondition: @not (A.introduction e)@.
 
 inferExp :: A.Exp -> Check (Term, VType)
 inferExp = \case
+
+  e | mustBeType e -> do
+    (t, l) <- inferType e
+    return (t, VType l)
+
   A.Var x -> do
     (u, Dom r t) <- inferVar x
     if r == Relevant then return (u,t) else
       throwError $ "Illegal reference to variable: " ++ printTree x
+
   A.App f e -> nyi "application"
   A.Case{}  -> nyi "case"
+
   _ -> __IMPOSSIBLE__
 
 -- | Infer type of a variable
@@ -234,6 +252,10 @@ inferVar (A.Ident x) = do
     Nothing     -> throwError $ "Variable not in scope: " ++ x
     Just (i, t) -> return (var $ Index i, t)
 
+-- | Coercion / subtype checking.
+
+coerce :: Term -> VType -> VType -> Check Term
+coerce u ti tc = nyi "Coercion"
 
 -- | Type checker auxiliary functions.
 
@@ -257,14 +279,22 @@ addTySig x t = stTySigs %= Map.insert x t
 addDef :: Id -> Val -> Check ()
 addDef x v = stDefs %= Map.insert x v
 
+-- * Invoking evaluation
+
+instance MonadEval (Reader (Map Id Val)) where
+  getDef x = fromMaybe __IMPOSSIBLE__ . Map.lookup x <$> ask
+
 evaluate :: Term -> Check Val
 evaluate t = do
   sig   <- use stDefs
[…]

### 2017-02-13 prototyping/size-irrelevance/Evaluation.hs
[ Sit ] Checking application.

@@ -52,17 +52,17 @@ data Val
     VUp VType VNe
   | -- | Type annotation for readback (normal form).
     VDown VType Val
-  deriving (Show)
+  deriving (Eq, Show)
 
 data VNe
   = VNe VGen VElims
-  deriving (Show)
+  deriving (Eq, Show)
 
 data VClos = VClos
   { closBody :: Abs Term
   , closEnv  :: Env
   }
-  deriving (Show)
+  deriving (Eq, Show)
 
 -- | Variable
 
@@ -255,9 +255,9 @@ instance Readback Val Term where
     VDown (VPi a b) d -> do
       v0 <- vVar (unDom a) <$> ask
       c <- lift $ applyClos b v0
-      Lam (domInfo a) . Abs "x" <$> do
+      Lam (_domInfo a) . Abs "x" <$> do
         local succ . readback . VDown c =<< do
-          lift $ apply d $ Arg (domInfo a) v0
+          lift $ apply d $ Arg (_domInfo a) v0
 
     VDown (VUp _ _) (VUp _ n) -> readbackNe n
 

### 2017-02-13 prototyping/size-irrelevance/Internal.hs

@@ -2,10 +2,13 @@
 {-# LANGUAGE DeriveFunctor #-}
 {-# LANGUAGE DeriveFoldable #-}
 {-# LANGUAGE DeriveTraversable #-}
+{-# LANGUAGE TemplateHaskell #-}
 
 -- | Internal Syntax.
 
 module Internal where
+
+import Control.Lens hiding (Level, Index)
 
 import Data.Foldable (Foldable)
 import Data.Traversable (Traversable)
@@ -87,7 +90,7 @@ data Abs a
 
 -- | Function domain decoration.
 
-data Dom a = Dom { domInfo :: ArgInfo, unDom :: a }
+data Dom a = Dom { _domInfo :: ArgInfo, unDom :: a }
   deriving (Eq, Ord, Show, Functor, Foldable, Traversable)
 
 -- | Argument decoration.
@@ -109,6 +112,12 @@ data Relevance
 var :: Index -> Term
 var i = Var i []
 
+app :: Term -> Arg Term -> Maybe Term
+app t u = case t of
+  Var x es -> Just $ Var x $ es ++ [ Apply u ]
+  Def f es -> Just $ Def f $ es ++ [ Apply u ]
+  _ -> Nothing
+
 defaultArg :: a -> Arg a
 defaultArg = Arg Relevant
 
@@ -124,3 +133,6 @@ sZero = Zero Infty
 
 sSuc  :: Term -> Term
 sSuc  = Suc Infty
+
+
+makeLenses ''Dom

### 2017-02-13 prototyping/size-irrelevance/TypeChecker.hs

@@ -119,20 +119,20 @@ inferType e = do
   let invalidType = throwError $ "Not a valid type expression: " ++ printTree e
   case e of
 
-    -- Universes
+    -- Universes (shape irrelevant)
 
     A.Set  -> return (Type sZero, vsConst 1)
     A.Set1 -> return (Type $ sSuc sZero, vsConst 2)
     A.Set2 -> return (Type $ sSuc $ sSuc sZero, vsConst 3)
     A.App A.Set l -> do
-      a <- checkLevel l
+      a <- resurrect ShapeIrr $ checkLevel l
       v <- evaluate a
       return (Type a, vsSuc v)
 
-    -- Natural number type
+    -- Natural number type (shape irrelevant)
 
     A.App A.Nat s -> do
-      a <- checkSize s
+      a <- resurrect ShapeIrr $ checkSize s
       v <- evaluate a
       return (Nat a, vsZero)
 
@@ -218,7 +218,7 @@ checkExp e0 t = do
         VPi dom cl -> addContext (x, dom) $ do
           t' <- applyClosure cl =<< lastVal
           u  <- checkExp (A.Lam xs e) t'
-          return $ Lam (domInfo dom) $ Abs (fromIdU x) u
+          return $ Lam (_domInfo dom) $ Abs (fromIdU x) u
         _ -> throwError $ "Lambda abstraction expects function type, but got " ++ show t
     e -> do
       (u, ti) <- inferExp e
@@ -235,27 +235,42 @@ inferExp = \case
     return (t, VType l)
 
   A.Var x -> do
-    (u, Dom r t) <- inferVar x
+    (u, Dom r t) <- inferId x
     if r == Relevant then return (u,t) else
       throwError $ "Illegal reference to variable: " ++ printTree x
 
-  A.App f e -> nyi "application"
+  e0@(A.App f e) -> do
+    (tf, t) <- inferExp f
+    case t of
+      VPi (Dom r tdom) cl -> do
+        te <- resurrect r $ checkExp e tdom
+        v  <- evaluate te
+        (fromMaybe __IMPOSSIBLE__ $ app tf (Arg r te),) <$> applyClosure cl v
+      _ -> throwError $ "Function type expected in application " ++ printTree e0
+             ++ " ; but found type" ++ show t
+
+
   A.Case{}  -> nyi "case"
 
   _ -> __IMPOSSIBLE__
 
 -- | Infer type of a variable
 
[…]

### 2017-02-14 prototyping/size-irrelevance/Abstract.hs
[ Sit ] started type checking of fix

@@ -1,10 +1,28 @@
 {-# LANGUAGE LambdaCase #-}
 
+{-# OPTIONS_GHC -fwarn-incomplete-patterns #-}
+
 -- | Auxiliary functions for abstract syntax.
 
 module Abstract where
 
+import Control.Lens
+
 import Sit.Abs as A
+
+-- data AppView
+--   = FixV [Exp]
+--   | SetV [Exp]
+--   | VarV [Exp]
+--   | ZeroV [Exp]
+--   | SucV  [Exp]
+
+-- | Gather applications to expose head.
+
+appView :: Exp -> (Exp, [Exp])
+appView = \case
+  App f e -> over _2 (++ [e]) $ appView f
+  e -> (e, [])
 
 -- | Can this expression only denote a type?
 
@@ -29,6 +47,7 @@ introduction = \case
   Case{}   -> False
   Int{}    -> True
   Infty    -> True
+  Size     -> True
   Nat      -> True
   Set      -> True
   Set1     -> True
@@ -44,6 +63,8 @@ introduction = \case
   Arrow{}  -> True
   Plus{}   -> True
   ELam{}   -> True
+
+-- | Convert "identifier-or-underscore" to string.
 
 fromIdU :: A.IdU -> String
 fromIdU = \case

### 2017-02-14 prototyping/size-irrelevance/Evaluation.hs

@@ -105,6 +105,7 @@ data SizeView
     -- ^ @i + n@
   | SVInfty
     -- ^ @oo@
+  deriving (Eq, Show)
 
 -- | Successor size on view.
 
@@ -306,3 +307,23 @@ readbackSize = \case
   VSuc _ a -> sSuc <$> readbackSize a
   VUp VSize (VNe x []) -> var <$> readback x
   _ -> __IMPOSSIBLE__
+
+-- * Comparison
+
+-- | Size values are partially ordered
+
+cmpSizes :: VSize -> VSize -> Maybe Ordering
+cmpSizes v1 v2 = do
+  s1 <- sizeView v1
+  s2 <- sizeView v2
+  case (s1, s2) of
+    (a,b) | a == b -> return EQ
+    (SVInfty, _) -> return GT
+    (_, SVInfty) -> return LT
+    (SVConst n, SVConst m) -> return $ compare n m
+    (SVVar x n, SVVar y m) | x == y -> return $ compare n m
+    (SVConst n, SVVar y m) | n <= m -> __IMPOSSIBLE__  -- Here, LT is too strong.
+    _ -> __IMPOSSIBLE__  -- TODO
+
+leqSize :: VSize -> VSize -> Bool
+leqSize a b = maxSize a b == b

### 2017-02-14 prototyping/size-irrelevance/TypeChecker.hs

@@ -10,6 +10,7 @@ import Control.Applicative
 import Control.Lens hiding (Level)
 import Control.Monad
 import Control.Monad.Except
+import Control.Monad.Identity
 import Control.Monad.Reader
 import Control.Monad.State
 
@@ -119,6 +120,13 @@ inferType e = do
   let invalidType = throwError $ "Not a valid type expression: " ++ printTree e
   case e of
 
+    -- Size type (internal use only).
+
+    -- Each universe is closed under size quantification.
+    -- Thus, we place Size in Set0.
+
+    A.Size -> return (Size, vsConst 0)
+
     -- Universes (shape irrelevant)
 
     A.Set  -> return (Type sZero, vsConst 1)
@@ -193,6 +201,7 @@ inferPiType x dom cont = do
 
 checkSize :: A.Exp -> Check Size
 checkSize = \case
+  A.Infty        -> return Infty
   A.LZero        -> return $ sZero
   A.App A.LSuc e -> sSuc <$> checkSize e
   e@(A.Var x)    -> checkExp e VSize
@@ -234,6 +243,17 @@ inferExp = \case
     (t, l) <- inferType e
     return (t, VType l)
 
+  e | (A.Fix, es) <- appView e -> do
+    case es of
+      [ et , ef , en ] -> do
+        tT <- checkExp et fixKind
+        tF <- fixType tT
+        tf <- checkExp ef =<< evaluate tF
+        (tn, a) <- inferNat en
+        -- return $ App tn [ Fix tT tf ]
+        nyi "checking fix -- need non-normal terms"
+      _ -> throwError $ "fix expects exactly 3 arguments: " ++ printTree e
+
   A.Var x -> do
     (u, Dom r t) <- inferId x
     if r == Relevant then return (u,t) else
@@ -254,6 +274,19 @@ inferExp = \case
 
   _ -> __IMPOSSIBLE__
 
+-- | @fixKind = ..(i : Size) -> Nat i -> Setω@
+
+fixKind :: VType
+fixKind = evaluateClosed $
+  Pi (Dom ShapeIrr Size) $ Abs "i" $
+    Pi (Dom Relevant (Nat $ var 0)) $ Abs "x" $
+      Type Infty
+
+-- | Construct the type of the functional for fix.
+
+fixType :: Term -> Check Term
+fixType t = nyi "fixType -- need renaming"
+
[…]

### 2017-02-15 prototyping/size-irrelevance/Evaluation.hs
[ Sit ] tc implemented, needs debugging.

@@ -172,8 +172,8 @@ instance Evaluate Term Val where
     Nat a    -> VNat <$> eval a
     Size     -> pure VSize
     Infty    -> pure VInfty
-    Zero a   -> VZero <$> eval a
-    Suc a t  -> liftA2 VSuc (eval a) (eval t)
+    Zero a   -> VZero <$> eval (unArg a)
+    Suc a t  -> liftA2 VSuc (eval $ unArg a) (eval t)
     Pi u t   -> liftA2 VPi (eval u) (eval t)
     -- Lam ai (NoAbs x t) -> VConst <$> eval t
     Lam ai t -> VLam <$> eval t
@@ -222,6 +222,7 @@ apply v arg@(Arg ai u) = case v of
   VUp (VPi a b) (VNe x es) -> do
     t' <- applyClos b u
     return $ VUp t' $ VNe x $ es ++ [ Apply $ Arg ai $ VDown (unDom a) u ]
+  _ -> error $ "NYI: apply  " ++ show v ++ "  to  " ++ show u
 
 -- | Apply a closure to a value.
 
@@ -238,7 +239,7 @@ unfoldFix t f a v = applyEs f $ map (Apply . defaultArg) [ a , VElimBy (Fix t f)
 -- | Eliminate a neutral natural number.
 
 elimNeNat :: MonadEval m => VSize -> VNe -> VElim -> m Val
-elimNeNat a n e = __IMPOSSIBLE__
+elimNeNat a n e = error $ "NYI: elimNeNat"
 
 -- * Readback
 
@@ -292,8 +293,8 @@ readbackType = \case
 
 readbackNat  :: MonadEval m => Val -> ReaderT Int m Term
 readbackNat = \case
-  VZero a        -> Zero <$> readbackSize a
-  VSuc a t       -> liftA2 Suc (readbackSize a) (readbackNat t)
+  VZero a        -> zero <$> readbackSize a
+  VSuc a t       -> liftA2 suc (readbackSize a) (readbackNat t)
   VUp (VNat _) n -> readbackNe n
   _ -> __IMPOSSIBLE__
 
@@ -329,3 +330,12 @@ cmpSizes v1 v2 = do
 
 leqSize :: VSize -> VSize -> Bool
 leqSize a b = maxSize a b == b
+
+-- | Compute predecessor size, if possible.
+sizePred :: VSize -> Maybe VSize
+sizePred v = do
+  sizeView v >>= \case
[…]

### 2017-02-15 prototyping/size-irrelevance/Internal.hs

@@ -39,9 +39,9 @@ data Term
   | -- | Sized natural number type.
     Nat Size
   | -- | Zero constructor, or zero size (then @Size@ is ignored).
-    Zero Size
+    Zero (Arg Size)
   | -- | Successor constructor, or successor size (then @Size@ is ignored).
-    Suc Size Term
+    Suc (Arg Size) Term
   | -- | Infinity size.
     Infty
   | -- ^ (Dependent) function type.
@@ -98,7 +98,11 @@ data Dom a = Dom { _domInfo :: ArgInfo, unDom :: a }
 -- | Argument decoration.
 
 data Arg a = Arg { argInfo :: ArgInfo, unArg :: a }
-  deriving (Eq, Ord, Show, Functor, Foldable, Traversable)
+  deriving (Ord, Show, Functor, Foldable, Traversable)
+
+instance Eq a => Eq (Arg a) where
+   Arg Irrelevant _ == Arg Irrelevant _ = True
+   Arg r a == Arg r' a' = r == r' && a == a'
 
 type ArgInfo = Relevance
 
@@ -110,6 +114,12 @@ data Relevance
   deriving (Eq, Ord, Show)
 
 -- * Smart constructor.
+
+zero :: Size -> Term
+zero = Zero . Arg Irrelevant
+
+suc :: Size -> Term -> Term
+suc = Suc . Arg Irrelevant
 
 -- var :: Index -> Term
 -- var i = Var i []
@@ -129,12 +139,12 @@ defaultDom = Dom Relevant
 -- | Zero size.
 
 sZero :: Term
-sZero = Zero Infty
+sZero = zero Infty
 
 -- | Successor size.
 
 sSuc  :: Term -> Term
-sSuc  = Suc Infty
+sSuc  = suc Infty
 
 
 makeLenses ''Dom

### 2017-02-15 prototyping/size-irrelevance/TypeChecker.hs

@@ -234,29 +234,55 @@ checkExp e0 t = do
           return $ Lam (_domInfo dom) $ Abs (fromIdU x) u
         _ -> throwError $ "Lambda abstraction expects function type, but got " ++ show t
 
-    A.ELam az ez as en es -> do
-       case t of
-         VPi (Dom r (VNat b)) cl -> do
-           nyi $ "checking " ++ printTree e0
-         _ -> throwError $ "Extended lambda is function from Nat _, but here it got type " ++ show t
+    e@(A.ELam jz ez js x es) -> do
+      case t of
+        VPi (Dom r (VNat b)) cl -> do
+          tt <- reifyType t
+          -- Make sure that b is a successor size
+          let failNotSuc = throwError $
+                "Splitting Nat is only possible at successor size, when checking " ++ printTree e
+          a <- maybe failNotSuc return $ sizePred b
+          tz <- checkExp ez =<< applyClosure cl (VZero a)
+          addContext (x, Dom Relevant $ VNat a) $ do
+            ts <- checkExp es =<< applyClosure cl =<< do VSuc a <$> lastVal
+            return $ Lam Relevant $ Abs "x" $ App (Var 0) $ raise 1 $
+              Case tt tz ts
+
+        _ -> throwError $ "Extended lambda is function from Nat _, but here it got type " ++ show t
 
     e -> do
       (u, ti) <- inferExp e
       coerce u ti t
     -- e -> nyi $ "checking " ++ printTree e
 
--- | Precondition: @not (A.introduction e)@.
+-- | Infers neutrals, natural numbers, types.
 
 inferExp :: A.Exp -> Check (Term, VType)
-inferExp = \case
+inferExp e0 = case (e0, appView e0) of
 
-  e | mustBeType e -> do
+  (e,_) | mustBeType e -> do
     (t, l) <- inferType e
     return (t, VType l)
 
-  e | (A.Fix, es) <- appView e -> do
+  (e, (A.Zero, es)) -> do
     case es of
-      [ et , ef , en ] -> do
+      [ ea ] -> do
+        a <- resurrect Irrelevant $ checkSize ea
+        (zero a ,) . VNat . vsSuc <$> evaluate a
+      _ -> throwError $ "zero expects exactly 1 argument: " ++ printTree e
+
[…]

### 2017-02-16 prototyping/size-irrelevance/Internal.hs
[ Sit ] testing reduction rules for plus.

@@ -146,5 +146,9 @@ sZero = zero Infty
 sSuc  :: Term -> Term
 sSuc  = suc Infty
 
+-- | Size increment.
+
+sPlus :: Term -> Integer -> Term
+sPlus t n = iterate sSuc t !! fromInteger n
 
 makeLenses ''Dom

### 2017-02-16 prototyping/size-irrelevance/TypeChecker.hs

@@ -73,7 +73,8 @@ checkDecls = mapM_ checkDecl
 
 checkDecl :: A.Decl -> Check ()
 checkDecl = \case
-  A.Open{} -> return ()
+  A.Blank{} -> return ()
+  A.Open{}  -> return ()
   A.Sig x a -> checkSig x a
   A.Def x e -> checkDef x e
 
@@ -292,10 +293,12 @@ inferExp e0 = case (e0, appView e0) of
         (App tn $ Fix tT tf,) <$> applyArgs vT [ Arg ShapeIrr a , Arg Relevant vn ]
       _ -> throwError $ "fix expects at least 3 arguments: " ++ printTree e
 
-  (A.Var x, _) -> do
-    (u, Dom r t) <- inferId x
-    if r == Relevant then return (u,t) else
-      throwError $ "Illegal reference to variable: " ++ printTree x
+  (A.Plus x k, _) -> do
+    (u, t) <- inferId x
+    subType t VSize
+    return (sPlus u k, t)
+
+  (A.Var x, _) -> inferId x
 
   (e0@(A.App f e), _) -> do
     (tf, t) <- inferExp f
@@ -337,14 +340,17 @@ fixType t =
 
 -- | Infer type of a variable
 
-inferId :: A.Ident -> Check (Term, Dom VType)
+inferId :: A.Ident -> Check (Term, VType)
 inferId (A.Ident x) = do
   (lookupCxt x <$> asks _envCxt) >>= \case
-    Just (i, t) -> return (Var $ Index i, t)
+    Just (i, Dom r t)
+      | r == Relevant -> return (Var $ Index i, t)
+      | otherwise     -> throwError $ "Illegal reference to " ++ show r ++ " variable: " ++ printTree x
+
     Nothing     -> do
       (Map.lookup x <$> use stTySigs) >>= \case
         Nothing -> throwError $ "Identifier not in scope: " ++ x
-        Just t  -> return (Def x, defaultDom t)
+        Just t  -> return (Def x, t)
 
 inferNat :: A.Exp -> Check (Term, VSize)
 inferNat e = do

### 2017-02-16 prototyping/size-irrelevance/TypeChecker.hs
[ Sit ] removed size variable from zero and suc pattern

@@ -235,7 +235,7 @@ checkExp e0 t = do
           return $ Lam (_domInfo dom) $ Abs (fromIdU x) u
         _ -> throwError $ "Lambda abstraction expects function type, but got " ++ show t
 
-    e@(A.ELam jz ez js x es) -> do
+    e@(A.ELam ez x es) -> do
       case t of
         VPi (Dom r (VNat b)) cl -> do
           tt <- reifyType t

### 2017-02-16 prototyping/size-irrelevance/Evaluation.hs
[ Sit ] Added forgotten abstraction in suc-branch of ELam

@@ -234,7 +234,11 @@ applyClos (VClos b rho) u = case b of
 -- | Unfold a fixed-point.
 
 unfoldFix :: MonadEval m => VType -> Val -> VSize -> Val -> m Val
-unfoldFix t f a v = applyEs f $ map (Apply . defaultArg) [ a , VElimBy (Fix t f) , v ]
+unfoldFix t f a v = applyEs f $ map Apply
+  [ Arg Irrelevant a
+  , defaultArg $ VElimBy $ Fix t f
+  , defaultArg v
+  ]
 
 -- | Eliminate a neutral natural number.
 

### 2017-02-16 prototyping/size-irrelevance/TypeChecker.hs

@@ -90,6 +90,7 @@ checkSig x0@(A.Ident x) a = traceCheck (A.Sig x0 a) $ do
 
   -- Check type and add to signature
   t <- checkType a
+  -- traceM $ "Adding  " ++ show x ++ "  of type  " ++ show t
   addTySig x =<< evaluate t
 
 -- | Check a definition.
@@ -108,6 +109,7 @@ checkDef x0@(A.Ident x) e = traceCheck (A.Def x0 e) $ do
 
   -- Check definition and add to signature
   v <- checkExp e t
+  -- traceM $ "Adding  " ++ show x ++ "  of value  " ++ show v
   addDef x =<< evaluate v
 
 -- | Check well-formedness of a type.
@@ -244,10 +246,11 @@ checkExp e0 t = do
                 "Splitting Nat is only possible at successor size, when checking " ++ printTree e
           a <- maybe failNotSuc return $ sizePred b
           tz <- checkExp ez =<< applyClosure cl (VZero a)
-          addContext (x, Dom Relevant $ VNat a) $ do
-            ts <- checkExp es =<< applyClosure cl =<< do VSuc a <$> lastVal
-            return $ Lam Relevant $ Abs "x" $ App (Var 0) $ raise 1 $
-              Case tt tz ts
+          ts <- Lam Relevant . Abs (A.fromIdU x) <$> do
+            addContext (x, Dom Relevant $ VNat a) $ do
+              checkExp es =<< applyClosure cl =<< do VSuc a <$> lastVal
+          return $ Lam Relevant $ Abs "x" $ App (Var 0) $ raise 1 $
+            Case tt tz ts
 
         _ -> throwError $ "Extended lambda is function from Nat _, but here it got type " ++ show t
 
@@ -283,15 +286,23 @@ inferExp e0 = case (e0, appView e0) of
 
   (e, (A.Fix, es)) -> do
     case es of
-      (et : ef : en : es') -> do
+      (et : ef : en : []) -> do
+        -- Check the motive of elimination
         tT <- checkExp et fixKind
-        vT <- evaluate tT
-        let tF = fixType tT
-        tf <- checkExp ef =<< evaluate tF
+        -- Check the functional
+        tf <- checkExp ef =<< evaluate (fixType tT)
+        -- Check the argument
         (tn, a) <- inferNat en
+        -- Compute the type of the elimination
+        vT <- evaluate tT
[…]

### 2017-02-16 prototyping/size-irrelevance/Evaluation.hs
[ Sit ] Reification of fix.

@@ -1,6 +1,7 @@
 {-# LANGUAGE CPP #-}
 {-# LANGUAGE LambdaCase #-}
 {-# LANGUAGE ScopedTypeVariables #-}
+{-# LANGUAGE TemplateHaskell #-}
 {-# LANGUAGE TypeSynonymInstances, FlexibleInstances, MultiParamTypeClasses #-}
 -- {-# LANGUAGE FunctionalDependencies, UndecidableInstances #-}
 
@@ -9,6 +10,7 @@
 module Evaluation where
 
 import Control.Applicative
+import Control.Lens hiding (Index, Level)
 import Control.Monad
 import Control.Monad.Reader
 
@@ -16,6 +18,7 @@ import Data.Maybe
 import Data.Traversable (traverse)
 
 import Internal
+import Substitute
 
 import Impossible
 #include "undefined.h"
@@ -54,8 +57,10 @@ data Val
     VDown VType Val
   deriving (Eq, Show)
 
-data VNe
-  = VNe VGen VElims
+data VNe = VNe
+  { _neVar :: VGen
+  , _neElims :: VElims
+  }
   deriving (Eq, Show)
 
 data VClos = VClos
@@ -63,6 +68,13 @@ data VClos = VClos
   , closEnv  :: Env
   }
   deriving (Eq, Show)
+
+-- | An environment maps de Bruijn indices to values.
+--   The first entry in the list is the binding for index 0.
+
+type Env = [Val]
+
+makeLenses ''VNe
 
 -- | Variable
 
@@ -145,15 +157,16 @@ maxSize v1 v2 =
 
 -- * Evaluation
 
--- | An environment maps de Bruijn indices to values.
---   The first entry in the list is the binding for index 0.
-
-type Env = [Val]
-
 -- | Evaluation monad.
 
 class (Functor m, Applicative m, Monad m) => MonadEval m where
   getDef :: Id -> m Val
+
+instance MonadEval Identity where
+  getDef x = __IMPOSSIBLE__
+
+evaluateClosed :: Term -> Val
+evaluateClosed t = runIdentity $ evalIn t []
 
 -- | Evaluation.
 
@@ -206,8 +219,8 @@ applyE v e =
     (_        , Apply u   ) -> apply v u
     (VZero _  , Case _ u _) -> return u
     (VSuc _ n , Case _ _ f) -> apply f $ defaultArg n
[…]

### 2017-02-16 prototyping/size-irrelevance/Internal.hs

@@ -74,6 +74,9 @@ data Elim' a
     Fix
     { fixReturn :: a
       -- ^ @T : ..(i : Size) -> Nat i -> Setω@
+    , fixTyBody :: a
+      -- ^ Type of fixBody.  Stored here for convenience, must be
+      -- @.(i : Size) (f : (x : Nat i) -> T i x) (x : Nat (i + 1)) -> T (i + 1) x@.
     , fixBody   :: a
       -- ^ @t : .(i : Size) (f : (x : Nat i) -> T i x) (x : Nat (i + 1)) -> T (i + 1) x@
     }
@@ -113,6 +116,8 @@ data Relevance
   | Irrelevant
   deriving (Eq, Ord, Show)
 
+makeLenses ''Dom
+
 -- * Smart constructor.
 
 zero :: Size -> Term
@@ -151,4 +156,10 @@ sSuc  = suc Infty
 sPlus :: Term -> Integer -> Term
 sPlus t n = iterate sSuc t !! fromInteger n
 
-makeLenses ''Dom
+-- | @fixKind = ..(i : Size) -> Nat i -> Setω@
+
+fixKind :: Type
+fixKind =
+  Pi (Dom ShapeIrr Size) $ Abs "i" $
+    Pi (Dom Relevant (Nat $ Var 0)) $ Abs "x" $
+      Type Infty

### 2017-02-16 prototyping/size-irrelevance/Substitute.hs

@@ -182,3 +182,20 @@ instance Apply (Abs Term) where
 
 
 -}
+
+
+
+-- | Construct the type of the functional for fix.
+--
+--   @fixType t = .(i : Size) -> ((x : Nat i) -> T i x) -> (x : Nat (i + 1)) -> T (i + 1) x
+
+fixType :: Term -> Term
+fixType t =
+  Pi (Dom Irrelevant Size) $ Abs "i" $
+    Pi (Dom Relevant $ f $ Var 0) $ NoAbs "_" $
+      f $ sSuc $ Var 0
+  where
+  f a = Pi (Dom Relevant (Nat a)) $ Abs "x" $
+          raise 2 t
+            `App` Apply (Arg ShapeIrr $ raise 1 a)
+            `App` Apply (Arg Relevant $ Var 0)

### 2017-02-16 prototyping/size-irrelevance/TypeChecker.hs

@@ -288,9 +288,10 @@ inferExp e0 = case (e0, appView e0) of
     case es of
       (et : ef : en : []) -> do
         -- Check the motive of elimination
-        tT <- checkExp et fixKind
+        tT <- checkExp et fixKindV
         -- Check the functional
-        tf <- checkExp ef =<< evaluate (fixType tT)
+        let tF = fixType tT
+        tf <- checkExp ef =<< evaluate tF
         -- Check the argument
         (tn, a) <- inferNat en
         -- Compute the type of the elimination
@@ -298,7 +299,7 @@ inferExp e0 = case (e0, appView e0) of
         vn <- evaluate tn
         ve <- applyArgs vT [ Arg ShapeIrr a , Arg Relevant vn ]
         -- Return as postfix application
-        return (App tn $ Fix tT tf, ve)
+        return (App tn $ Fix tT tF tf, ve)
 
       _ -> throwError $ "fix expects exactly 3 arguments: " ++ printTree e
 
@@ -325,29 +326,6 @@ inferExp e0 = case (e0, appView e0) of
   (A.Case{}, _)  -> nyi "case"
 
   (e, _) -> nyi $ "inferring type of " ++ printTree e
-
--- | @fixKind = ..(i : Size) -> Nat i -> Setω@
-
-fixKind :: VType
-fixKind = evaluateClosed $
-  Pi (Dom ShapeIrr Size) $ Abs "i" $
-    Pi (Dom Relevant (Nat $ Var 0)) $ Abs "x" $
-      Type Infty
-
--- | Construct the type of the functional for fix.
---
---   @fixType t = .(i : Size) -> ((x : Nat i) -> T i x) -> (x : Nat (i + 1)) -> T (i + 1) x
-
-fixType :: Term -> Term
-fixType t =
-  Pi (Dom Irrelevant Size) $ Abs "i" $
-    Pi (Dom Relevant $ f $ Var 0) $ NoAbs "_" $
-      f $ sSuc $ Var 0
-  where
-  f a = Pi (Dom Relevant (Nat a)) $ Abs "x" $
-          raise 2 t
-            `App` Apply (Arg ShapeIrr $ raise 1 a)
-            `App` Apply (Arg Relevant $ Var 0)
 
 -- | Infer type of a variable
 
@@ -400,12 +378,6 @@ addDef :: Id -> Val -> Check ()
 addDef x v = stDefs %= Map.insert x v
 
[…]

### 2017-02-16 prototyping/size-irrelevance/Evaluation.hs
[ Sit ] first successful run.  Reify case still broken.

@@ -17,6 +17,8 @@ import Control.Monad.Reader
 import Data.Maybe
 import Data.Traversable (traverse)
 
+import Debug.Trace
+
 import Internal
 import Substitute
 
@@ -235,7 +237,9 @@ apply v arg@(Arg ai u) = case v of
   VUp (VPi a b) (VNe x es) -> do
     t' <- applyClos b u
     return $ VUp t' $ VNe x $ es ++ [ Apply $ Arg ai $ VDown (unDom a) u ]
-  _ -> error $ "NYI: apply  " ++ show v ++ "  to  " ++ show u
+  _ -> do
+    traceM $ "apply  " ++ show v ++ "  to  " ++ show u
+    __IMPOSSIBLE__
 
 -- | Apply a closure to a value.
 
@@ -259,7 +263,18 @@ unfoldFix t tf f a v = applyEs f $ map Apply
 elimNeNat :: MonadEval m => VSize -> VNe -> VElim -> m Val
 elimNeNat a n e = case e of
   Apply{} -> __IMPOSSIBLE__
-  Case t u f -> error "NYI: elimNeNat Case"
+
+  Case t u f -> do
+    -- Compute the type of the result of the elimination application
+    tr <- apply t $ Arg Relevant $ VUp (VNat a) n
+    -- Compute the type of the zero branch
+    tz <- apply t $ Arg Relevant u
+    -- Compute the type of the suc branch
+    ts <- return t -- TODO: must be (x : Nat a) -> t (suc a x)
+    -- Assemble the elimination
+    let e = Case (VDown (VType VInfty) t) (VDown tz u) (VDown ts f)
+    -- Assemble the result
+    return $ VUp tr $ over neElims (++ [e]) n
 
   Fix t tf f -> do
     -- Compute the type of the result of the elimination application

### 2017-02-18 prototyping/size-irrelevance/Evaluation.hs
[ Sit ] partial admissibility test

@@ -218,12 +218,12 @@ applyEs v (e : es) = applyE v e >>= (`applyEs` es)
 applyE :: MonadEval m => Val -> VElim -> m Val
 applyE v e =
   case (v, e) of
-    (_        , Apply u   ) -> apply v u
-    (VZero _  , Case _ u _) -> return u
-    (VSuc _ n , Case _ _ f) -> apply f $ defaultArg n
-    (VZero a  , Fix t tf f) -> unfoldFix t tf f a v -- apply f $ e : map (Apply . defaultArg) [ v , VZero ]
-    (VSuc a n , Fix t tf f) -> unfoldFix t tf f a v
-    (VUp (VNat a) n , _)    -> elimNeNat a n e
+    (_        , Apply u     ) -> apply v u
+    (VZero _  , Case _ u _ _) -> return u
+    (VSuc _ n , Case _ _ _ f) -> apply f $ defaultArg n
+    (VZero a  , Fix t tf f  ) -> unfoldFix t tf f a v -- apply f $ e : map (Apply . defaultArg) [ v , VZero ]
+    (VSuc a n , Fix t tf f  ) -> unfoldFix t tf f a v
+    (VUp (VNat a) n , _)      -> elimNeNat a n e
     _ -> __IMPOSSIBLE__
 
 -- | Apply a function to an argument.
@@ -264,7 +264,7 @@ elimNeNat :: MonadEval m => VSize -> VNe -> VElim -> m Val
 elimNeNat a n e = case e of
   Apply{} -> __IMPOSSIBLE__
 
-  Case t u f -> do
+  Case t u tf f -> do
     -- Compute the type of the result of the elimination application
     tr <- apply t $ Arg Relevant $ VUp (VNat a) n
     -- Compute the type of the zero branch
@@ -272,7 +272,7 @@ elimNeNat a n e = case e of
     -- Compute the type of the suc branch
     ts <- return t -- TODO: must be (x : Nat a) -> t (suc a x)
     -- Assemble the elimination
-    let e = Case (VDown (VType VInfty) t) (VDown tz u) (VDown ts f)
+    let e = Case (VDown (VType VInfty) t) (VDown tz u) (VDown (VType VInfty) tf) (VDown tf f)
     -- Assemble the result
     return $ VUp tr $ over neElims (++ [e]) n
 
@@ -302,6 +302,7 @@ instance Readback VGen Index where
 
 instance Readback Val Term where
   readback = \case
[…]

### 2017-02-18 prototyping/size-irrelevance/Internal.hs

@@ -68,6 +68,8 @@ data Elim' a
     Case
     { caseReturn :: a -- ^ @T : Nat (b + 1) -> Setω@
     , caseZero   :: a -- ^ @tz : T zero@
+    , caseTySuc  :: a -- ^ Type of @caseSuc@.  Stored here for convenience, must be
+                      --        @(t : Nat b) -> T (suc t)@
     , caseSuc    :: a -- ^ @ts : (t : Nat b) -> T (suc t)@
     }
   | -- | Recursion
@@ -75,7 +77,7 @@ data Elim' a
     { fixReturn :: a
       -- ^ @T : ..(i : Size) -> Nat i -> Setω@
     , fixTyBody :: a
-      -- ^ Type of fixBody.  Stored here for convenience, must be
+      -- ^ Type of @fixBody@.  Stored here for convenience, must be
       -- @.(i : Size) (f : (x : Nat i) -> T i x) (x : Nat (i + 1)) -> T (i + 1) x@.
     , fixBody   :: a
       -- ^ @t : .(i : Size) (f : (x : Nat i) -> T i x) (x : Nat (i + 1)) -> T (i + 1) x@

### 2017-02-18 prototyping/size-irrelevance/TypeChecker.hs

@@ -204,12 +204,13 @@ inferPiType x dom cont = do
     return ( Pi (dom $> u) $ Abs (fromIdU x) t , l )
 
 checkSize :: A.Exp -> Check Size
-checkSize = \case
-  A.Infty        -> return Infty
-  A.LZero        -> return $ sZero
-  A.App A.LSuc e -> sSuc <$> checkSize e
-  e@(A.Var x)    -> checkExp e VSize
-  e -> throwError $ "Not a valid size expression: " ++ printTree e
+checkSize e = checkExp e VSize
+-- checkSize = \case
+--   A.Infty        -> return Infty
+--   A.LZero        -> return $ sZero
+--   A.App A.LSuc e -> sSuc <$> checkSize e
+--   e@(A.Var x)    -> checkExp e VSize
+--   e -> throwError $ "Not a valid size expression: " ++ printTree e
 
 checkLevel :: A.Exp -> Check Level
 checkLevel = \case
@@ -237,20 +238,29 @@ checkExp e0 t = do
           return $ Lam (_domInfo dom) $ Abs (fromIdU x) u
         _ -> throwError $ "Lambda abstraction expects function type, but got " ++ show t
 
-    e@(A.ELam ez x es) -> do
+    e@(A.ELam ez x0 es) -> do
       case t of
         VPi (Dom r (VNat b)) cl -> do
+          let x = A.fromIdU x0
+          unless (r == Relevant) $ throwError $
+            "Extended lambda constructs relevant function: " ++ printTree e
+          -- Infer the type of the case expression
           tt <- reifyType t
           -- Make sure that b is a successor size
-          let failNotSuc = throwError $
-                "Splitting Nat is only possible at successor size, when checking " ++ printTree e
-          a <- maybe failNotSuc return $ sizePred b
+          -- let failNotSuc = throwError $ "Splitting Nat is only possible at successor size, when checking " ++ printTree e
+          -- a  <- maybe failNotSuc return $ sizePred b
+          let a = fromMaybe __IMPOSSIBLE__ $ sizePred b
+          ta <- reifySize a
           tz <- checkExp ez =<< applyClosure cl (VZero a)
-          ts <- Lam Relevant . Abs (A.fromIdU x) <$> do
+          (ts0, tS0) <-
             addContext (x, Dom Relevant $ VNat a) $ do
[…]

### 2017-02-18 nbe/icfp17.tex
[ ICFP 17 ] Started on reification.  Before simplifying types in typing.

--- 2017-02-05 nbe/icfp17.tex
@@
 \Data & \ni & c
-   & ::= & \tzero \mid \suc a t & \mbox{constructed data} \\
+   & ::= & \zero a \mid \suc a t & \mbox{constructed data} \\
 \Ne & \ni & n
@@
 We write $\sigma$ and $\tau$ for substitutions, which are lists of terms, and $t \sigma$ for the application of substitution $\sigma$ to term $t$.
-We write $\slift^n_m$ for the substitution $(v_{n+m-1},\dots,v_{n+1},v_n)$ which applies to a term with $m$ free indices and increases all of them by $n$.
+We write $\slift^k_m$ for the substitution $(\ind{k+m-1},\dots,\ind{k+1},\ind k)$ which applies to a term with $m$ free indices and increases all of them by $k$.
 We write $\slift_m$ for the lifting $\slift^1_m$ and $\sid_m$ for the identity substitution $ \slift^0_m$.
@@
 \\[2ex] % \qquad
-  \ru{\resurrect\Gamma \der b : \Size
-    }{\Gamma \der \tzero : \Nat (b + 1)}
+  \ru{\Gamma \der a : \Size \qquad
+      \resurrect\Gamma \der b : \Size
+    }{\Gamma \der \zero{a} : \Nat (b + 1)}
 \qquad
@@
   \ru{\Gamma \der u : \Nat (b + 1) \qquad
-      \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
-      \Gamma \der t_z : T \, \tzero \qquad
-      \Gamma \der a : \Size \qquad
-      \Gamma \der t_s : (n : \Nat b) \to T \, (\suc a n)
+      \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \\
+      \Gamma \der a_z,a_s : \Size \qquad
+      \Gamma \der t_z : T \, (\zero{a_z}) \qquad
+      \Gamma \der t_s : (t : \Nat b) \to T \, (\suc{a_s} t)
     }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
@@
       \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
-      \Gamma \der t_z : T\,\tzero \qquad
-      \Gamma \der a : \Size \qquad
-      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc a n)
-    }{\Gamma \der \tzero\,\case[\ell] T {t_z} {t_s} = t_z : T\,\tzero}
+      \Gamma \der a_z,a_s : \Size \qquad
+      \Gamma \der t_z : T\,\zero{a_z} \qquad
+      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc{a_s}n)
+    }{\Gamma \der (\zero{a_z})\,\case[\ell] T {t_z} {t_s} = t_z : T\,\zero{a_z}}
 \\[2ex]
@@
       \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
-      \Gamma \der t_z : T\,\tzero \qquad
-      \Gamma \der a : \Size \qquad
-      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc a n)
-    }{\Gamma \der (\suc a t)\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,(\suc a t)}
+      \Gamma \der a_z,a_s : \Size \qquad
+      \Gamma \der t_z : T\,\zero{a_z} \qquad
+      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc{a_s}n)
+    }{\Gamma \der (\suc{a_s} t)\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,(\suc{a_s} t)}
 \\[2ex]
@@
      }{\Gamma \der c\,\fix[\ell] T t = t\ann a (\lambda x. x\,\fix[\ell] T t)\, c : T \,b \, c
-     }%{u \in \{\tzero, \suc a t'\}}
+     }%{c \in \{\zero a, \suc a t'\}}
 \end{gather*}
@@
     }{\Gamma \der \epiT \Size T = \epiT \Size T' : s}
-\qquad
+\\[2ex]
   \ru{%\der \Gamma \qquad
@@
 %     }{\Gamma \der t\,e : T'}
-\qquad
+\\[2ex]
   \ru{\Gamma \der t = t' : \piT \Size T \qquad
@@
     }{\Gamma \der t\,a = t'\,a : T[a]}
-\\[2ex] % \qquad
-  \ru{\resurrect\Gamma \der b : \Size
-    }{\Gamma \der \tzero = \tzero : \Nat (b + 1)}
-\qquad
-  \ru{\Gamma \der a, a' : \Size \qquad
-      \Gamma \der t = t' : \Nat b
-    }{\Gamma \der \suc a t = \suc {a'} t' : \Nat (b + 1)}
 \qquad
@@
     }{\Gamma \der t \ann a = t' \ann{a'} : T[b]}
+\\[2ex] % \qquad
+  \ru{\Gamma \der a, a' : \Size \qquad
+      \resurrect\Gamma \der b : \Size
+    }{\Gamma \der \zero a = \zero {a'}: \Nat (b + 1)}
+\qquad
+  \ru{\Gamma \der a, a' : \Size \qquad
+      \Gamma \der t = t' : \Nat b
+    }{\Gamma \der \suc a t = \suc {a'} t' : \Nat (b + 1)}
 \\[2ex]
   \ru{\Gamma \der u = u' : \Nat (a + 1) \qquad
-      \Gamma \der T = T' : \Nat (a + 1) \to \Set[\ell] \qquad
-      \Gamma \der t_z = t_z' : T \, \tzero \qquad
+      \Gamma \der T = T' : \Nat (a + 1) \to \Set[\ell] \\
+      \Gamma \der t_z = t_z' : T \, (\zero a) \qquad
       \Gamma \der t_s = t_s' : (n : \Nat a) \to T \, (\suc a n)
@@
 \item Symmetry:  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then $\sueq \Gamma {\sigma'} {\sigma} {\tau} \Delta$.
-\item Transitivity:  If\/ $\sueq \Gamma {\sigma_1} {\sigma_2} {\tau} \Delta$ and $\sueq \Gamma {\sigma_2} {\sigma_3} {\tau} \Delta$ then $\sueq \Gamma {\sigma_1} {\sigma-3} {\tau} \Delta$.
+\item Transitivity:  If\/ $\sueq \Gamma {\sigma_1} {\sigma_2} {\tau} \Delta$ and $\sueq \Gamma {\sigma_2} {\sigma_3} {\tau} \Delta$ then $\sueq \Gamma {\sigma_1} {\sigma_3} {\tau} \Delta$.
 \item Functionality: Let $\sueq \Gamma \sigma {\sigma'} \tau \Delta$.
@@
 
+\begin{lemma}[Inversion of typing]
+\label{lem:invty} \bla
+\begin{enumerate}
+\item TODO
+\end{enumerate}
+\end{lemma}
+
 \section{Semantics}
 \label{sec:sem}
+
+We extend the syntax by de Bruijn levels $\lev k$
+to be used as generic values (unknowns) and type annotations
+$\up T t$ and $\down T t$ for lazy realisations of the
+reflection and reification operations of NbE.
+\[
+\begin{array}{lllrl@{\qquad}l}
+%  \Exp  & \ni & t & ::= & \dots \mid \lev k \mid \up T t \mid \down T t \\
+  \Ne   & \ni & n & ::= & \dots \mid \lev k \\
+  \Whnf & \ni & w,d,f,g,A,B,F,G & ::= & \dots \mid  \up T n \mid \down T t \\
+\end{array}
+\]
+
+\subsection{Operational semantics}
 
@@
   \ru{t_z \evalsto w
-    }{\tzero @ \casett \evalsto w}
+    }{(\zero a) @ \casett \evalsto w}
 \qquad
@@
 \qquad
-  \ru{t\,\ann\infty\,(\lambda x.\, x@ \fixtt)\,c \evalsto w
+  \ru{t\,\ann a\,(\lambda x.\, x@ \fixtt)\,c \evalsto w
      }{c @ \fixtt \evalsto w
-     }%{w \in \{ \tzero, \suc a u \}}
+     }{c\in \{ \zero a, \suc a u \}}
+\end{gather*}
+The rules added for NbE deal with elimination of delayed reflection:
+\begin{gather*}
+  \ru{T' \evalsto \piT U T
+    }{(\up {T'} n) @ u \evalsto \up{T[u]} (n \, \down U u) }
+\qquad
+  \ru{T' \evalsto \piT \Size T
+    }{(\up {T'} n) @ a \evalsto \up{T[a]} (n \, a)}
+\qquad
+  \ru{T' \evalsto \forallT T
+    }{(\up {T'} n) @ \ann a \evalsto \up{T[a]} (n \ann a)}
+%   \ru{T \evalsto \piT A B \qquad
+%       \up{B[u]} (n \, \down A u) \evalsto w
+%     }{(\up T n) @ u \evalsto w}
+% \qquad
+%   \ru{T \evalsto \piT \Size B \qquad
+%       \up{B[u]} (n \, a) \evalsto w
+%     }{(\up T n) @ a \evalsto w}
+% \qquad
+%   \ru{T \evalsto \forallT B \qquad
+%       \up{B[a]} (n \ann a) \evalsto w
+%     }{(\up T n) @ \ann a \evalsto w}
+\\[2ex]
+  \ru{T' \evalsto \Nat b \qquad
+    }{(\up {T'} n) @ \case[\ell] T {t_z} {t_s} \evalsto
+      \up{T\,(\up{\Nat \infty} n)} (n\,\case[\ell] {(\down {\Set[\infty]} T)} {(\down {T\,{\zero\infty}} t_z)} {(t_s)})
+    }
 \end{gather*}
 
+We write \fbox{$t \approx t'$} iff $t$ and $t'$ are syntactically equal if we ignore sizes $\ann a$ in irrelevant positions.
+For example, $\suc a \tzero \approx \suc 0 \tzero$, but $\Nat a \not\approx \Nat 0$.
+
 \newpage
+
+\section{Metatheory}
+\label{sec:meta}
+
+\begin{lemma}[Typing of constructed data]
+\label{lem:invdata} \bla
+If\/ $\Gamma \der c : \Nat b$ then $\resurrect\Gamma \der b : \Size$ and $b > 0$.
+\end{lemma}
 

### 2017-02-19 nbe/icfp17.tex
[ ICFP 17 ] Reification rules, o as offset, N for Up terms.

@@
     & ::= & t \mid a \mid \ann a \mid \case[\ell] T {t_1} t_2 \mid \fix[\ell] T t & \mbox{eliminations} \\
-\SizeExp & \ni & a,b & ::= & \infty \mid k \mid \ind i + k ~(k \in \NN) & \mbox{size expressions} \\
+\SizeExp & \ni & a,b & ::= & \infty \mid o \mid \ind i + o ~(o \in \NN) & \mbox{size expressions} \\
 \Cxt & \ni & \Gamma, \Delta & ::= & \cempty \mid \erext[\noterased] \Gamma T \mid \erext \Gamma \Size \\
@@
 
-Size increment \fbox{$a + k$} for $k \in \NN$ extends addition by $\infty + k = \infty$ and $(\ind i + k') + k = \ind i + (k' + k)$.
-Size comparison \fbox{$a \leq b$} holds if either $b = \infty$ or  $k \leq k'$ where either $a = k$ and $b = k'$ or $a \in \{ k, \ind i + k \}$ and $b = \ind i + k'$.
+Size increment \fbox{$a + o'$} for $o' \in \NN$ extends addition by $\infty + o' = \infty$ and $(\ind i + o) + o' = \ind i + (o + o')$.
+Size comparison \fbox{$a \leq b$} holds if either $b = \infty$ or  $o \leq o'$ where either $a = o$ and $b = o'$ or $a \in \{ o, \ind i + o \}$ and $b = \ind i + o'$.
 
@@
 \\[2ex]
-  \ru{\Gamma \der u : \Nat (b + 1) \qquad
-      \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \\
-      \Gamma \der a_z,a_s : \Size \qquad
-      \Gamma \der t_z : T \, (\zero{a_z}) \qquad
-      \Gamma \der t_s : (t : \Nat b) \to T \, (\suc{a_s} t)
+  \ru{\Gamma \der u : \Nat (a + 1) \qquad
+      \Gamma \der T : \Nat (a + 1) \to \Set[\ell] \qquad
+      \Gamma \der t_z : T \, (\zero{a}) \qquad
+      \Gamma \der t_s : (t : \Nat a) \to T \, (\suc{a} t)
     }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
 \\[2ex]
-  \ru{\Gamma \der u : \Nat b \qquad
+  \ru{\Gamma \der u : \Nat a \qquad
       \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
       \Gamma \der t : \FixT\, T
-     }{\Gamma \der u\,\fix[\ell] T t : T \,b \, u}
+     }{\Gamma \der u\,\fix[\ell] T t : T \,a \, u}
 \end{gather*}
@@
       \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
-      \Gamma \der a_z,a_s : \Size \qquad
-      \Gamma \der t_z : T\,\zero{a_z} \qquad
-      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc{a_s}n)
-    }{\Gamma \der (\zero{a_z})\,\case[\ell] T {t_z} {t_s} = t_z : T\,\zero{a_z}}
+      \Gamma \der a : \Size \qquad
+      \Gamma \der t_z : T\,\zero{b} \qquad
+      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc{b}n)
+    }{\Gamma \der (\zero{a})\,\case[\ell] T {t_z} {t_s} = t_z : T\,\zero{a_z}}
 \\[2ex]
@@
       \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
-      \Gamma \der a_z,a_s : \Size \qquad
-      \Gamma \der t_z : T\,\zero{a_z} \qquad
-      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc{a_s}n)
-    }{\Gamma \der (\suc{a_s} t)\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,(\suc{a_s} t)}
+      \Gamma \der a : \Size \qquad
+      \Gamma \der t_z : T\,\zero{b} \qquad
+      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc{b}n)
+    }{\Gamma \der (\suc{a} t)\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,(\suc{a} t)}
 \\[2ex]
@@
   \Ne   & \ni & n & ::= & \dots \mid \lev k \\
-  \Whnf & \ni & w,d,f,g,A,B,F,G & ::= & \dots \mid  \up T n \mid \down T t \\
+  \tUp  & \ni & N & ::= & \up T n \\
+  \Whnf & \ni & w & ::= & \dots \mid N \mid \down T t \\
+%  \Whnf & \ni & w,d,f,g,A,B,F,G & ::= & \dots \mid  \up T n \mid \down T t \\
 \end{array}
@@
     }{(\suc a t) @ \casett \evalsto w}
-\qquad
+\\[2ex]
   \ru{t\,\ann a\,(\lambda x.\, x@ \fixtt)\,c \evalsto w
@@
     }{(\up {T'} n) @ \case[\ell] T {t_z} {t_s} \evalsto
-      \up{T\,(\up{\Nat \infty} n)} (n\,\case[\ell] {(\down {\Set[\infty]} T)} {(\down {T\,{\zero\infty}} t_z)} {(t_s)})
+      \up{T\,(\up{\Nat \infty} n)}
+         n\,\case[\ell]
+             T %% {(\down {\Set[\infty]} T)}
+             {(\down {T\,{\zero\infty}} t_z)}
+             {(\down {\funS x {\Nat \infty} T\,(\suc \infty x)} t_s)}
     }
+\\[2ex]
+  \ru{T' \evalsto \Nat b \qquad
+    }{(\up {T'} n) @ \fix[\ell] T t \evalsto
+      n\,\fix[\ell] {(\down {\FixK} T)} {(\down {\FixT\, T} t)}
+    }
+\end{gather*}
+
+\subsection{Read back}
+\label{sec:read}
+
+Judgements.
+\[
+\begin{array}{l@{\qquad}l}
+   \R k t  v     & \mbox{} \\
+   \RTy k T  V  & \mbox{} \\
+   \RNat k t  v & \mbox{} \\
+   \RNe k t  m  & \mbox{} \\
+\end{array}
+\]
+\fbox{$\R k t v$}
+\begin{gather*}
+  \ru{t \evalsto \down{U} T \qquad
+      U \evalsto s \\
+      \RTy k T V
+    }{\R k t V}
+\qquad
+  \ru{t \evalsto \down{U} u \qquad
+      U \evalsto \Nat a \\
+      \RNat k u v
+    }{\R k t v}
+\qquad
+  \ru{t \evalsto \down{U} u \qquad
+      U \evalsto \up {T'} {t'} \\
+      \RNe k u v
+    }{\R k t v}
+\\[2ex]
+  \ru{t \evalsto \down{U} f \qquad
+      U \evalsto \piT A B \qquad
+      \R{k+1} {\down{B[\up A \lev k]} (f\,\up A \lev k)} v
+    }{\R k t {\lambda v}}
+\\[2ex]
+  \ru{t \evalsto \down{U} f \qquad
+      U \evalsto \piT \Size B \qquad
+      \R{k+1} {\down{B[\lev k]} (f\,\lev k)} v
+    }{\R k t {\lambda v}}
+\qquad
+  \ru{t \evalsto \down{U} f \qquad
+      U \evalsto \forallT B \qquad
+      \R{k+1} {\down{B[\lev k]} (f\,\ann{\lev k})} v
+    }{\R k t {\lambda v}}
+\end{gather*}
+
+\fbox{$\RNat k t v $}
+\begin{gather*}
+  \ru{t \evalsto n \qquad
+      \RNe k n m
+    }{\RNat k t m}
+\qquad
+  \ru{t \evalsto \zero{a_v} \qquad
+      \RSize k {a_v} a
+    }{\RNat k t {\zero a}}
+\qquad
+  \ru{t \evalsto \suc{a_v}u \qquad
+      \RSize k {a_v} a \qquad
+      \RNat k u v
+    }{\RNat k t {\suc a v}}
+\end{gather*}
+
+\fbox{$\RSize k {a_v} a$}
+\begin{gather*}
+  \ru{
+    }{\RSize k \infty \infty}
+\qquad
+  \ru{
+    }{\RSize k o o}
+\qquad
+  \ru{
+    }{\RSize k {\lev j + o} {\ind{k-1-j} + o}}
+\end{gather*}
+
+\fbox{$\RNe k t m$} and \fbox{$\RElim k e e^v$}
+\begin{gather*}
+%   \ru{t \evalsto \lev j
+%     }{\RNe k t {\ind{k-1-j}}}
+% \qquad
+  \ru{t \evalsto \lev j\,\vec e \qquad
+      \RElim k {e_i} {e^v_i} \mforall i
+    }{\RNe k t {\ind{k-1-j}\,\vec e^v}}
+\qquad
+% \end{gather*}
+% \begin{gather*}
+  \ru{\R k t v
+    }{\RElim k t v}
+\qquad
+  \ru{\RSize k a b
+    }{\RElim k a b}
+\qquad
+  \ru{\RSize k a b
+    }{\RElim k {\ann a} {\ann b}}
+\\[2ex]
+  \ru{\RTy k T V \qquad
+      \R k {t_z} {v_z} \qquad
+      \R k {t_s} {v_s}
+    }{\RElim k {(\case[\ell] T {t_z} {t_s})} {\case[\ell] V {v_z} {v_s}}}
+\qquad
+  \ru{\RTy k T V \qquad
+      \R k t v
+    }{\RElim k {(\fix[\ell] T t)} {\fix[\ell] V v}}
+\end{gather*}
+\fbox{$\RTy k T V$}
+\begin{gather*}
+  \ru{T \evalsto \Set[\ell]
+    }{\RTy k T {\Set[\ell]}}
+\qquad
+  \ru{T \evalsto \Nat a \qquad
+      \RSize k a b
+    }{\RTy k T {\Nat b}}
+\qquad
+  \ru{T \evalsto \up{U}u \qquad
+      \RNe k u m
+    }{\RTy k T m}
+\\[2ex]
+  \ru{T \evalsto \piT A B \qquad
+      \RTy k A {V_a} \qquad
+      \RTy {k+1} B {V_b}
+    }{\RTy k T {\piT {V_a} V_b}}
+\qquad
+  \ru{T \evalsto \epiT \Size B \qquad
+      \RTy {k+1} B {V_b}
+    }{\RTy k T {\epiT \Size V_b}}
 \end{gather*}
@@
 For example, $\suc a \tzero \approx \suc 0 \tzero$, but $\Nat a \not\approx \Nat 0$.
+
+\subsection{Partial equivalence relations}
+\label{sec:per}
+
+A type $T$ will be interpreted as a partial equivalence relation (PER) $\A$ on terms, \ie, a relation which is symmetric and transitive.
+The domain $\dom(\A)$ of the relation can be thought of the set of terms which define the extension of the type; on $\dom(\A)$ the relation $\A$ is in fact an equivalence relation.
+
+The PERs $\NE$ and $\NF$ charaterize (neutral) normalizing terms.
+\[
+\begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
+  \fbox{$t = t' \in \NE$}
+    & \defas &
+    \RNe k t m &\mand& \RNe k {t'} {m'} &\mand& m \approx m' &\mforall k \\
+  \fbox{$t = t' \in \NF$}
+    & \defas &
+    \R k t v &\mand& \R k {t'} {v'} &\mand& v \approx v' &\mforall k \\
+\end{array}
+  % \ru{\RNe k t m \mand \RNe k {t'} {m'} \mand m \approx m' \mforall k
+  %   }{t = t' \in \NE}
+  % \ru{t \evalsto \up N n \qquad t' \evalsto \up{N'}{n'} \qquad n \approx n'
+  %   }{t = t' \in \NE}
+\]
+All PERs $\A$ we consider are sandwiched $\NE \subseteq \A \subseteq \NF$ between $\NE$ and $\NF$.
+
+\subsection{PER model}
+\label{sec:permode}
+
+By induction on $\ell \in \NN$ we define the PER family
+$\_ = \_ \in \Set[\ell]$ of types
+together with the extension $\EL[\ell] T$ (for $T = T' \in \Set[\ell]$)
+which is PER of values of type $A$.
+The rules for \fbox{$T = T' \in \SET[\ell]$} are:
+\[
+\begin{array}{l@{\qquad}l}
+  \ru{T \evalsto N \qquad T' \evalsto N'
+    }{T = T' \in \SET[\ell]}
+  & \EL[\ell](T) = \NE
+\\[3ex]
+  \ru{T \evalsto \Nat\,a \qquad T' \evalsto \Nat\,a
+    }{T = T' \in \SET[\ell]}
+  & \EL[\ell](T) = \NAT(a)
+\\[3ex]
+  \rux{T \evalsto \Set[\ell'] \qquad T' \evalsto \Set[\ell']
+     }{T = T' \in \SET[\ell]
+     }{\ell'{<}\ell}
+  & \EL[\ell](T) = \SET[\ell']
+\\[3ex]
+  \rul{T \evalsto \piT A B \qquad
+      T' \evalsto \piT{A'}{B'} \\
+      A = A' \in \SET[\ell] \qquad
+      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
+    }{T = T' \in \SET[\ell]}
+  & \EL[\ell](T) = \PI(\EL[\ell](A),\;u \mapsto \EL[\ell](B[u]))
+\\[3ex]
+  \rul{T \evalsto \piT \Size B \quad
+      T' \evalsto \piT{\Size}{B'} \quad
+      B[a] = B'[a] \in \SET[\ell] \mforall a \in \SIZE
+    }{T = T' \in \SET[\ell]}
+  & \EL[\ell](T) = \PI(\SIZE,\; a \mapsto \EL[\ell](B[a]))
+\\[3ex]
+  \rul{T \evalsto \forallT B \quad
+      T' \evalsto \forallT{B'} \quad
+      B[a] = B'[a] \in \SET[\ell] \mforall a \in \SIZE
+    }{T = T' \in \SET[\ell]}
+  & \EL[\ell](T) = \FORALL(a \mapsto \EL[\ell](B[a]))
+\end{array}
+\]
 

### 2017-02-20 nbe/icfp17.tex
[ ICFP 17 ] Fixed some calls to Rne.
[ ICFP 17 ] Some cases of the fundamental theorem.

@@
   \tUp  & \ni & N & ::= & \up T n \\
-  \Whnf & \ni & w & ::= & \dots \mid N \mid \down T t \\
+  \Whnf & \ni & w & ::= & \dots \mid N \\
+  \tDown & \ni & d & ::= & \down T t \\
+  \Exp  & \ni & t & ::= & \dots \mid d \\
 %  \Whnf & \ni & w,d,f,g,A,B,F,G & ::= & \dots \mid  \up T n \mid \down T t \\
@@
 \begin{array}{l@{\qquad}l}
-   \R k t  v     & \mbox{} \\
+   \R k d  v     & \mbox{} \\
    \RTy k T  V  & \mbox{} \\
@@
 \]
-\fbox{$\R k t v$}
-\begin{gather*}
-  \ru{t \evalsto \down{U} T \qquad
-      U \evalsto s \\
+\fbox{$\R k d v$}
+\begin{gather*}
+  \ru{U \evalsto s \qquad
       \RTy k T V
-    }{\R k t V}
-\qquad
-  \ru{t \evalsto \down{U} u \qquad
-      U \evalsto \Nat a \\
+    }{\R k {\down{U} T} V}
+\qquad
+  \ru{U \evalsto \Nat a \qquad
       \RNat k u v
-    }{\R k t v}
-\qquad
-  \ru{t \evalsto \down{U} u \qquad
-      U \evalsto \up {T'} {t'} \\
-      \RNe k u v
-    }{\R k t v}
-\\[2ex]
-  \ru{t \evalsto \down{U} f \qquad
-      U \evalsto \piT A B \qquad
+    }{\R k {\down{U} u} v}
+\qquad
+  \ru{U \evalsto N \qquad
+      u \evalsto \up{T} n \qquad
+      \RNe k n m
+    }{\R k {\down{U} u} m}
+\\[2ex]
+  \ru{U \evalsto \piT A B \qquad
       \R{k+1} {\down{B[\up A \lev k]} (f\,\up A \lev k)} v
-    }{\R k t {\lambda v}}
-\\[2ex]
-  \ru{t \evalsto \down{U} f \qquad
-      U \evalsto \piT \Size B \qquad
+    }{\R k {\down{U} f} {\lambda v}}
+\\[2ex]
+  \ru{U \evalsto \piT \Size B \qquad
       \R{k+1} {\down{B[\lev k]} (f\,\lev k)} v
-    }{\R k t {\lambda v}}
-\qquad
-  \ru{t \evalsto \down{U} f \qquad
-      U \evalsto \forallT B \qquad
+    }{\R k {\down{U} f} {\lambda v}}
+\qquad
+  \ru{U \evalsto \forallT B \qquad
       \R{k+1} {\down{B[\lev k]} (f\,\ann{\lev k})} v
-    }{\R k t {\lambda v}}
+    }{\R k {\down{U} f} {\lambda v}}
 \end{gather*}
@@
 \begin{gather*}
-  \ru{t \evalsto n \qquad
+  \ru{t \evalsto \up {\Nat a} n \qquad
       \RNe k n m
@@
   \ru{
-    }{\RSize k {\lev j + o} {\ind{k-1-j} + o}}
-\end{gather*}
-
-\fbox{$\RNe k t m$} and \fbox{$\RElim k e e^v$}
+    }{\RSize k {\lev j + o} {\ind{k\dotminus(1+j)} + o}}
+\end{gather*}
+
+\fbox{$\RNe k n m$} and \fbox{$\RElim k e e^v$}
 \begin{gather*}
@@
 % \qquad
-  \ru{t \evalsto \lev j\,\vec e \qquad
+  \ru{%t \evalsto \lev j\,\vec e \qquad
       \RElim k {e_i} {e^v_i} \mforall i
-    }{\RNe k t {\ind{k-1-j}\,\vec e^v}}
+    }{\RNe k {\lev j\,\vec e} {\ind{k \monus (1+j)}\,\vec e^v}}
 \qquad
@@
 % \begin{gather*}
-  \ru{\R k t v
-    }{\RElim k t v}
+  \ru{\R k d v
+    }{\RElim k d v}
 \qquad
@@
   \ru{\RTy k T V \qquad
-      \R k {t_z} {v_z} \qquad
-      \R k {t_s} {v_s}
-    }{\RElim k {(\case[\ell] T {t_z} {t_s})} {\case[\ell] V {v_z} {v_s}}}
+      \R k {d_z} {v_z} \qquad
+      \R k {d_s} {v_s}
+    }{\RElim k {(\case[\ell] T {d_z} {d_s})} {\case[\ell] V {v_z} {v_s}}}
 \qquad
   \ru{\RTy k T V \qquad
-      \R k t v
-    }{\RElim k {(\fix[\ell] T t)} {\fix[\ell] V v}}
+      \R k d v
+    }{\RElim k {(\fix[\ell] T d)} {\fix[\ell] V v}}
 \end{gather*}
@@
 \qquad
-  \ru{T \evalsto \up{U}u \qquad
-      \RNe k u m
+  \ru{T \evalsto \up{U}n \qquad
+      \RNe k n m
     }{\RTy k T m}
@@
 
-The PERs $\NE$ and $\NF$ charaterize (neutral) normalizing terms.
+The PERs $\NE$ characterizes neutral reifyable terms.
+% \[
+% \begin{array}{lll}%{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
+%   \fbox{$\up T n = \up {T'} {n'} \in \NE$}
+%     & \defas &
+%     T \evalsto N \mand T' \evalsto N' \mand
+%     \mforall k \mboth
+%     \RNe k n m \mand \RNe k {n'} {m'} \mand m \approx m'
+%     .\\
+% \end{array}
+% \]
+The PERs $\NE$ and $\NF$ characterize (neutral) normalizing terms.
 \[
 \begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
-  \fbox{$t = t' \in \NE$}
+  \fbox{$n = n' \in \NE$}
     & \defas &
-    \RNe k t m &\mand& \RNe k {t'} {m'} &\mand& m \approx m' &\mforall k \\
-  \fbox{$t = t' \in \NF$}
+    \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m \approx m' &\mforall k \\
+  \fbox{$d = d' \in \NF$}
     & \defas &
-    \R k t v &\mand& \R k {t'} {v'} &\mand& v \approx v' &\mforall k \\
+    \R k d v &\mand& \R k {d'} {v'} &\mand& v \approx v' &\mforall k \\
 \end{array}
@@
 \]
-All PERs $\A$ we consider are sandwiched $\NE \subseteq \A \subseteq \NF$ between $\NE$ and $\NF$.
+%All PERs $\A$ we consider are sandwiched $\NE \subseteq \A \subseteq \NF$ between $\NE$ and $\NF$.
+
+\begin{lemma}[Closure properties of $\NE$ and $\NF$]
+  \bla
+  \begin{enumerate}
+  \item $\lev k = \lev k \in \NE$.
+  \item If $n = n' \in \NE$ and $d = d' \in \NF$ then $n\,d = n'\,d' in \NE$.
+%  \item If $d = d' \in \NF$
+  \end{enumerate}
+\end{lemma}
+
+\[
+\fbox{$\NEE$} \defas \{ (\up T n, \up{T'}{n'})
+  \msuchthat T \evalsto N \mand T' \evalsto N' \mand n = n' \in \NE \}
+\]
+\fbox{$\NAT(a)$}
+\begin{gather*}
+  \ru{t \evalsto \zero a \qquad
+      t' \evalsto \zero{a'}
+    }{t = t' \in \NAT(b+1)}
+\qquad
+  \ru{t \evalsto \suc a u \qquad
+      t' \evalsto \suc {a'} {u'} \qquad
+      u = u' \in \NAT(b)
+    }{t = t' \in \NAT(b+1)}
+\\[2ex]
+  \ru{t \evalsto \up{\Nat a} n \qquad
+      t' \evalsto \up{\Nat a'} n' \qquad
+      n = n' \in \NE
+    }{t = t' \in \NAT(b)}
+\end{gather*}
+
+\fbox{$\SIZE$} is a discrete PER of size values:
+\begin{gather*}
+  \ru{
+    }{\infty = \infty \in \SIZE}
+\qquad
+  \ru{
+    }{o = o \in \SIZE}
+\qquad
+  \ru{
+    }{\lev k + o = \lev k + o \in \SIZE}
+\end{gather*}
+
+Let $\A$ be a PER and $\F$ a family of PERs over $\A$ such that
+$\F(u) = \F(u')$ whenever $u = u' \in \A$.  We define
+\[
+  \fbox{$\PIAF$} \defas \{(t,t') \mid t\,u = t'\,u' \in \F(u) \mforall u = u' \in \A \}
+  .
+\]
+This works also for $\A = \SIZE$.
+For a family $\F$ over $\SIZE$ we also have the irrelevant function space
+\[
+  \fbox{$\FORALL\,\F$} \defas
+  \{(t,t') \mid t \ann a = t' \ann{a'} \in \F(b) \mforall a,a',b \in \SIZE \}
+  .
+\]
 
@@
     }{T = T' \in \SET[\ell]}
-  & \EL[\ell](T) = \NE
+  & \EL[\ell](T) = \NEE
 \\[3ex]
@@
 \]
-
-\newpage
-
-\section{Metatheory}
-\label{sec:meta}
-
-\begin{lemma}[Typing of constructed data]
-\label{lem:invdata} \bla
-If\/ $\Gamma \der c : \Nat b$ then $\resurrect\Gamma \der b : \Size$ and $b > 0$.
+All PERs involved here are closed under weak head expansion.
+
+The PER model has the usual properties, one of them being that $\EL[\ell](T)$ does not depend $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$.
+Thus, we simply write $t = t' \in \EL(T)$ or even $t = t' \in T$.
+
+\subsection{Type shapes}
+
+Reflection and reification perform $\eta$-expansion such that we arrive at an $\eta$-long $\beta$-normal form.
+To perform the $\eta$-expansion, not the precise type is needed, just the approximate shape, in particular, whether it is a function type (do expand) or a base type (do not expand).
+For the logical framework, the shape of a dependent type is just its underlying simple type \cite{harperPfenning:equivalenceLF}.
+However, in the presence of universes and large eliminations, there is no underlying simple type.
+Of course, we can take a type as its own shape, but we want at least that $\Nat a$ and $\Nat b$ have the same shape even for different $a, b$.
+Thus, we define a relation $T \shape S$ to express that $S$ is a possible shape of type $T$.
+
+\fbox{$T \shape S$} for $T \in \SET[\ell]$.
+\begin{gather*}
+  \ru{T \evalsto \Nat a \qquad
+      S \evalsto \Nat b
+    }{T \shape S}
+\qquad
+  \ru{T \evalsto N \qquad
+      S \evalsto N'
+    }{T \shape S}
+\qquad
+   \ru{T \evalsto \Set[\ell] \qquad
+       S \evalsto \Set[\ell]
+    }{T \shape S}
+\\[2ex]
+  \ru{T \evalsto \piT A B \qquad
+      S \evalsto \piT{A'}{B'} \qquad
+      A \shape A' \qquad
+      B[u] \shape B'[u'] \mforall u = u' \in A
+    }{T \shape S}
+\\[2ex]
+  \ru{T \evalsto \piT \Size B \qquad
+      S \evalsto \piT{\Size}{B'} \qquad
+      B[a] \shape B'[a] \mforall a \in \SIZE
+    }{T \shape S}
+\\[2ex]
+  \ru{T \evalsto \forallT B \qquad
+      S \evalsto \forallT{B'} \qquad
+      B[a] \shape B'[a'] \mforall a,a' \in \SIZE
+    }{T \shape S}
+\end{gather*}
+% The last rule is the crucial one
+\begin{lemma}[Reflexivity for shapes]
+  If $T = T' \in \SET[\ell]$ then $T \shape T'$.
 \end{lemma}
+
+\begin{theorem}[Reification and reflection]
+Let $T \in \SET[\ell]$ and $T \shape S_1$ and $T \shape S_2$.
+\begin{enumerate}
+\item If $n_1 = n_2 \in \NE$ then $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$.
+\item If $t_1 = t_2 \in T$ then $\down{S_1}{t_1} = \down{S_2}{t_2} \in \NF$.
+\end{enumerate}
+\end{theorem}
+\begin{proof}
+By induction on $T \in \SET[\ell]$ and cases on $T \shape S_1$ and $T \shape S_2$.
+\begin{caselist}
+
+\nextcase $T \evalsto \piT A B$
+\begin{gather*}
+  \ru{A \in \SET[\ell] \qquad
+      B[u] = B[u'] \in \SET[\ell] \mforall u = u' \in A
+    }{T \in \SET[\ell]}
+\\[2ex]
+  \ru{S_1 \evalsto \piT{A_1}{B_1} \qquad
+      A \shape A_1 \qquad
+      B[u] \shape B_1[u'] \mforall u = u' \in A
+    }{T \shape S_1}
+\\[2ex]
+  \ru{S_2 \evalsto \piT{A_2}{B_2} \qquad
+      A \shape A_2 \qquad
+      B[u] \shape B_2[u'] \mforall u = u' \in A
+    }{T \shape S_2}
+\end{gather*}
+\begin{enumerate}
+\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $u_1 = u_2 \in A$.
+  Let $d_i = \down{A_i}{u_i}$.
+  By induction hypothesis, $d_1 = d_2 \in \NF$.
+  Thus, $n_1\,d_1 = n_2\,d_2 \in \NE$, and
+  by induction hypothesis,  $\up{B_1[u_1]}(n_1\,d_1) = \up{B_2[u_2]}(n_2\,d_2) \in B[u_1]$.
+  With $(\up{S_i}{n_i})\,u_i \evalsto \up{B_i[u_i]}(n_i\,d_i)$ we are done by definition of $\EL(T)$.
+
+\item We assume $k \in \NN$ and show $\R k {\down{S_i}{t_i}} \lambda v_i$ for some normal forms
+$v_1,v_2$.  Let $u_i = \up{A_i} \lev k$.  Note that $u_1 = u_2 \in A$ by induction hypothesis.
+It is sufficient to show
+$\R{k+1}{\down{B_i[u_i]}{(t_i\,u_i)}} {v_i}$.
+By definition of $\EL(T)$ we have $t_1\,u_1 = t_2\,u_2 \in B[u_1]$, thus, by induction hypothesis,
+$\down{B_1[u_1]}(t_1\,u_1) = \down{B_2[u_2]}(t_2\,u_2) \in \NF$, which delivers $v_1$ and $v_2$ for $k+1$.
+\end{enumerate}
+
+\nextcase $T \evalsto \forallT B$
+\begin{gather*}
+  \ru{B[a] = B[a] \in \SET[\ell] \mforall a \in \SIZE
+    }{T \in \SET[\ell]}
+\\[2ex]
+  \ru{S_1 \evalsto \forallT{B_1} \qquad
+      B[a] \shape B_1[a'] \mforall a, a' \in \SIZE
+    }{T \shape S_1}
+\\[2ex]
+  \ru{S_2 \evalsto \forallT{B_2} \qquad
+      B[a] \shape B_2[a'] \mforall a, a' \in \SIZE
+    }{T \shape S_2}
+\end{gather*}
+\begin{enumerate}
+\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $a_1,a_2 \in \SIZE$.
+  Since $n_1 \ann{a_1} = n_2 \ann{a_2} \in \NE$, we obtain
+  $\up{B_1[a_1]}{(n_1 \ann{a_1})} = \up{B_2[a_2]}{(n_2 \ann{a_2})} \in B[a_1]$
+  by induction hypothesis.
+  Thus, $(\up{S_1}{n_1}) \ann{a_1} = (\up{S_2}{n_2}) \ann{a_2} \in B[a_1]$
+  by weak head expansion, which entails the goal by definition of $\EL(T)$.
+\item Assume $k \in \NN$ and note that $\lev k = \lev k \in \SIZE$, hence,
+  $t_1\,\ann{\lev k} = t_2\,\ann{\lev k} \in B[\lev k]$.  Thus, by induction hypothesis,
+  $\R{k+1}{\down{B_i[\lev k]}(t_i\,\ann{\lev k})}{v_i}$, and finally
+  $\R k {\down{S_i} t_i} {\lambda v_i}$ by definition of read back.
+
+\end{enumerate}
+
+\end{caselist}
+\end{proof}
+
+\begin{corollary}
+  If\/ $T \in \SET[\ell]$ and $t = t' \in T$ then $\down T t = \down T {t'} \in \NF$.
+\end{corollary}
+
+\subsection{Fundamental Theorem}
+\label{sec:fund}
+
+PER of substitutions
+\fbox{$\eta = \eta' \eeq \rho \in \Gamma$}.
+\begin{gather*}
+  \ru{
+    }{\sempty = \sempty \eeq \sempty \in \cempty}
+\qquad
+  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad
+      T\rho \in \SET[\ell] \qquad
+      u = u' = t \in T \rho
+    }{\sext \eta u = \sext{\eta'}{u'} \eeq \sext{\rho}{t} \in \cext \Gamma T}
+\\[2ex]
+  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad a \in \SIZE
+    }{\sext \eta a = \sext{\eta'}{a} \eeq \sext{\rho}{a} \in \cext \Gamma \Size}
+\qquad
+  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad a,a',b \in \SIZE
+    }{\sext \eta a = \sext{\eta'}{a'} \eeq \sext{\rho}{b} \in \erext \Gamma \Size}
+\end{gather*}
+We write $\rho \in \Gamma$ for $\rho = \rho \eeq \rho \in \Gamma$.
+\begin{lemma}[Resurrection]
+  \label{lem:resenv}
+  If $\eta = \eta' \eeq \rho \in \Gamma$ then $\rho \in \resurrect\Gamma$.
+\end{lemma}
+
+Semantic judgements:
+\[
+\begin{array}{l@{\quad}l@{\quad}l}
+  \models () & \defas & \mtrue \\
+  \models \eext \Gamma \Size & \defas & \models \Gamma \\
+  \models \cext \Gamma T     & \defas & \models \Gamma \mand \resurrect\Gamma \models T
+\\[2ex]
+  \Gamma \models s & \defas & \models \Gamma \\
+%  \Gamma \models s & \defas & \mtrue \\
+  \Gamma \models T & \defas & \Gamma \models T = T \\
+  \Gamma \models T = T' & \defas & \Gamma \models T = T' : s \mforsome s %\Set[\ell] \mforsome \ell
+\\[2ex]
+  \Gamma \models t : T & \defas & \Gamma \models t = t : T \\
+  \Gamma \models t = t' : T & \defas & %\models \Gamma \mand
+    % \resurrect\Gamma \models T \mand
+    \models \cext \Gamma T \mand
+    t \eta = t' \eta' \in T \rho \mforall \eta = \eta' \eeq \rho \in \Gamma
+\\[2ex]
+  \Gamma \models \sigma : \Delta & \defas & \Gamma \models \sigma = \sigma \eeq \sigma : \Delta \\
+  \Gamma \models \sigma = \sigma' \eeq \tau : \Delta & \defas &
+    \models \Gamma \mand \models \Delta \mand
+    \sigma \eta = \sigma' \eta' \eeq \tau \rho \in \Delta
+    \mforall \eta = \eta' \eeq \rho \in \Gamma
+\end{array}
+\]
+
+\begin{theorem}[Fundamental theorem]
+  \bla
+  \begin{enumerate}
+  \item
+  If $\der \Gamma$ then $\models \Gamma$.
+  \item
+  If $\Gamma \der J$ then % $\models \Gamma$ and
+  $\Gamma \models J$
+  \end{enumerate}
+\end{theorem}
+\begin{proof}
+Simultaneously, by induction on the derivation.
+
+\begin{caselist}
+\nextcase $\forall$-introduction.
+\[
+  \ru{\erext \Gamma \Size \der t = t' : T
+    }{\Gamma \der \lambda t = \lambda t' : \forall T}
+\]
+First $\models \Gamma$ follows from the induction hypothesis $\models {\erext \Gamma \Size}$.
+To show $\resurrect\Gamma \models \forall T$,
+assume $\eta = \eta' \eeq \rho \in \resurrect \Gamma$
+and show $(\forall T)\eta = (\forall T)\eta' \in \SET[\ell]$ for some $\ell$.
+To this end, assume $a \in \SIZE$ and show $T(\eta,a) = T(\eta',a) \in \SET[\ell]$.
+Note that $(\eta,a) = (\eta',a) \eeq (\rho,a) \in
+\resurrect {\erext \Gamma \Size} = \cext {\resurrect\Gamma}\Size$,
+thus, we can instantiate the induction hypothesis and obtain our goal.
+
+For the main goal,
+assume $\eta = \eta' \eeq \rho \in \Gamma$ and
+show $(\lambda t)\eta = (\lambda t')\eta' \in (\forall T)\rho$.
+To this end, assume arbitrary $a,a',b \in \SIZE$ and show
+$t(\eta,a) = t'(\eta',a') \in T(\rho,b)$.
+Since $(\eta,a) = (\eta',a') \eeq (\rho,b) \in \erext \Gamma \Size$,
+we conclude by induction hypothesis.
+
+\nextcase $\forall$-elimination.
+\[
+  \ru{\Gamma \der t = t' : \forall T \qquad
+      \Gamma \der a, a' : \Size \qquad
+      \resurrect\Gamma \der b : \Size
+    }{\Gamma \der t \ann a = t' \ann {a'} : T[b]}
+\]
+First, $\models \Gamma$ follows by induction hypothesis.
+For goal $\resurrect \Gamma \models T[b]$,
+assume $\eta = \eta' \eeq \rho \in \resurrect\Gamma$ and show
+$T[b]\eta = T[b]\eta' \in \SET[\ell]$ for some $\ell$.
+By induction hypothesis,  $b\eta = b\eta' \in \SIZE$.
+% , thus,
+% $(\eta,b\eta) = (\eta',b\eta') \eeq (\rho,b\eta) \in \cext{\resurrect\Gamma}\Size$
+By another induction hypothesis, $(\forall T)\eta = (\forall T)\eta' \in \SET[\ell]$,
+which by definition entails our goal $T(\eta, b\eta) = T(\eta', b\eta') \in \SET[\ell]$.
+
+For the remaining main goal,
+assume $\eta = \eta' \eeq \rho \in \Gamma$ and
+show $t \ann a \eta = t' \ann{a'} \eta' \in T[b]\rho$.
+By definition of substitution, it suffices to show
+$t \eta \ann{a \eta} = t' \eta' \ann{a' \eta'} \in T(\rho, b \rho)$.
+By induction hypothesis,
+$t \eta = t' \eta' : (\forall T) \rho$, thus, by definition of this PER,
+$t \eta \ann {a_1} = t' \eta' \ann{a_2} \in T(\rho, b')$ for any size values $a_1$, $a_2$, and $b'$.
+We conclude by choosing $a_1 = a \eta$ and $a_2 = a' \eta'$ and $b' = b \rho$.
+
+We now argue that this choice is possible,
+namely that $a \eta, a'\eta, b\rho \in \SIZE$.
+Note that the induction hypothesis gives us
+$a \eta  = a  \eta' \in \SIZE$ and
+$a' \eta = a' \eta' \in \SIZE$.
+By resurrection (Lemma~\ref{lem:resenv}) we have $\rho \in \resurrect\Gamma$,
+thus, by induction hypothesis, $b \rho \in \SIZE$.
+\end{caselist}
+\end{proof}
+
+\section{Completeness of NbE}
+\label{sec:compl}
+
+% \begin{lemma}[Typing of constructed data]
+% \label{lem:invdata} \bla
+% If\/ $\Gamma \der c : \Nat b$ then $\resurrect\Gamma \der b : \Size$ and $b > 0$.
+% \end{lemma}
 

### 2017-02-20 nbe/icfp17.tex
[ ICFP 17 ] Overview over NbE.

@@
 %\documentclass[acmlarge]{acmart}\settopmatter{}
+
+\usepackage{mathtools} %mathrlap
+\usepackage[all]{xy}
 
@@
 \]
+
+\begin{figure}
+$$
+\xymatrix@R=12ex@C=12ex{
+\mathrlap{\mbox{Semantics ($\beta$)}} &
+  &                f \in \D   \ar@[red][d]_{\down A}
+\\
+\mathrlap{\mbox{Semantics ($\beta\eta$)}} &
+  &                d \in \DNf \ar@[red][d]_{\tR_k{}}
+  &                n \in \DNe \ar[d]_{\tRne_k} \ar@[blue][ul]_{\up A}
+  & \ar@[blue][l]_{\var{}} j \in \Level \ar@{<->}[d]_{k\dotminus(1+\_)}
+\\
+\mathrlap{\mbox{Syntax}} &
+t \in \Exp \ar@[red]@(u,dl)[uur]^{\_\color{blue}\eta} %\ar@/^/[uur]^{\den\_}
+  & \ar[l]_{\supseteq}  v \in \Nf
+  & \ar[l]_{\supseteq}  m \in \Ne
+  & \ar[l]_{\ind{}}     i \in \Index
+  &
+}
+$$
+\caption{Type-assignment NbE in locally nameless
+    style.\label{fig:typedchart}}
+\end{figure}
+
+Figure~\ref{fig:typedchart}, adapted from \citet{abel:habil} shows the syntactic categories and main operations involved in NbE in what we call \emph{locally nameless style}.
+The red path $\Exp \to \D \to \Dnf \to \Nf$ decomposes $\beta\eta$-normalization into three steps.
+\begin{enumerate}
+\item
+First, we close the term $t$ with an environment $\eta$ that maps the free de Bruijn indices of $t$ to reflected de Bruijn levels.
+Reflection of de Bruijn levels follows the blue path $\Level \to \Dne \to \D$: Levels embed via constructor $\var{}$ into semantic neutrals $\DNe$ which are labelled with their type $A \in \D$ to become and element $\up A \lev j \in \D$.
+
+\item
+Then, we label value $t \eta \in \D$ with its type $A$ to obtain $\down A t \eta \in \Dnf$.
+\item
+Finally, \emph{read back} %% $\tR_k$
+$\tR_k \down{A} t\eta$ produces a long normal form $v \in \Nf$, converting de Bruijn levels back to indices.
+Herein, $k$ should be the length of the context the original term $t$ lived in.
+If this is the case, each de Bruijn level encountered during read back is below $k$ and can be safely converted to a de Bruijn index.
+\end{enumerate}
 
@@
 
+Note that $T \in \SET[\ell]$ and $T \shape S$ do not imply $S \in \SET[\ell]$.
+Type shapes are not well-defined types in general.
+For instance, assume a term $F : \Nat 0 \to \Set[0]$ which diverges if applied to a successor term.
+Then $T := \funT x {\Nat 0} {F\,x}$ is a welldefined type; we have $T \in \SET[0]$.
+Now consider $S := \funT x {\Nat \infty} {F\,x}$.
+We have $T \shape S$, but not $S$ is not well-defined; $S \not\in \SET[0]$.
+% $\funT x {\Nat 0} {F\,x} \shape , but the latter one is not a well-defined type.
+
 \begin{theorem}[Reification and reflection]
@@
 
-\section{Completeness of NbE}
+\subsection{Completeness of NbE}
 \label{sec:compl}
+
+\section{Soundness  of NbE}
+\label{sec:compl}
+
+In this section, we show that NbE is sound for judgemental equality, i.e., that same normal form implies definitional equality.
+The proof follows previous work on NbE \cite{abelCoquandDybjer:lics07,abel:habil}.
+
+\[
+  \fbox{$\LRSize \Gamma a b$} \defas
+  \forall \sigma : \Gamma' \leq \Gamma.\
+  \exists a'.\
+  \RSize {\Gamma'} b a' \mand
+  \Gamma' \der a\sigma = a' : \Size
+\]
+
+\fbox{$\LR \Gamma t {T'} f {A'}$} defined by induction on ${A'} \in \SET[\ell]$.
+\begin{caselist}
+
+\nextcase $A' \evalsto \piT A B$.  Then $\LR \Gamma t {T'} f {A'}$ if all of the following:
+\[
+\begin{array}{l}
+\Gamma \der T' = \piT U T \mforsome U, T\\
+\LRT \Gamma U A \\
+\forall \sigma : \Gamma' \leq \Gamma.\ \
+\LR {\Gamma'} u {U\sigma} a A \implies \LR {\Gamma'} {t\sigma\,u} {T(\sigma,u)} {f\,a} {B[a]}
+\end{array}
+\]
+
+\nextcase $A' \evalsto \forallT B$.  Then $\LR \Gamma t {T'} f {A'}$ if all of the following:
+\[
+\begin{array}{l}
+\Gamma \der T' = \forallT T \mforsome T\\
+\forall \sigma : \Gamma' \leq \Gamma.\ \
+\LRSize {\Gamma'} u a \implies \LR {\Gamma'} {t\sigma\,u} {T(\sigma,u)} {f\,a} {B[a]}
+\end{array}
+\]
+
+\end{caselist}
 

### 2017-02-20 nbe/icfp17.tex
[ ICFP 17 ] Logical relation.

@@
 \qquad
-  \ru{t[a] \evalsto w
-    }{(\lambda t) @ a \evalsto w}
-\qquad
-  \ru{t[a] \evalsto w
-    }{(\lambda t) @ \ann a \evalsto w}
+  \ru{t[\alpha] \evalsto w
+    }{(\lambda t) @ \alpha \evalsto w}
+\qquad
+  \ru{t[\alpha] \evalsto w
+    }{(\lambda t) @ \ann \alpha \evalsto w}
 \\[2ex]
   \ru{t_z \evalsto w
-    }{(\zero a) @ \casett \evalsto w}
+    }{(\zero \alpha) @ \casett \evalsto w}
 \qquad
   \ru{t_s\, t \evalsto w
-    }{(\suc a t) @ \casett \evalsto w}
-\\[2ex]
-  \ru{t\,\ann a\,(\lambda x.\, x@ \fixtt)\,c \evalsto w
+    }{(\suc \alpha t) @ \casett \evalsto w}
+\\[2ex]
+  \ru{t\,\ann \alpha\,(\lambda x.\, x@ \fixtt)\,c \evalsto w
      }{c @ \fixtt \evalsto w
-     }{c\in \{ \zero a, \suc a u \}}
+     }{c\in \{ \zero \alpha, \suc \alpha u \}}
 \end{gather*}
@@
   \ru{T' \evalsto \piT \Size T
-    }{(\up {T'} n) @ a \evalsto \up{T[a]} (n \, a)}
+    }{(\up {T'} n) @ \alpha \evalsto \up{T[\alpha]} (n \, \alpha)}
 \qquad
   \ru{T' \evalsto \forallT T
-    }{(\up {T'} n) @ \ann a \evalsto \up{T[a]} (n \ann a)}
+    }{(\up {T'} n) @ \ann \alpha \evalsto \up{T[\alpha]} (n \ann \alpha)}
 %   \ru{T \evalsto \piT A B \qquad
@@
 %   \ru{T \evalsto \piT \Size B \qquad
-%       \up{B[u]} (n \, a) \evalsto w
-%     }{(\up T n) @ a \evalsto w}
+%       \up{B[u]} (n \, \alpha) \evalsto w
+%     }{(\up T n) @ \alpha \evalsto w}
 % \qquad
 %   \ru{T \evalsto \forallT B \qquad
-%       \up{B[a]} (n \ann a) \evalsto w
-%     }{(\up T n) @ \ann a \evalsto w}
+%       \up{B[\alpha]} (n \ann \alpha) \evalsto w
+%     }{(\up T n) @ \ann \alpha \evalsto w}
 \\[2ex]
@@
 \qquad
-  \ru{U \evalsto \Nat a \qquad
+  \ru{U \evalsto \Nat \alpha \qquad
       \RNat k u v
@@
 \begin{gather*}
-  \ru{t \evalsto \up {\Nat a} n \qquad
+  \ru{t \evalsto \up {\Nat \alpha} n \qquad
       \RNe k n m
@@
 \qquad
-  \ru{t \evalsto \zero{a_v} \qquad
-      \RSize k {a_v} a
+  \ru{t \evalsto \zero{\alpha_v} \qquad
+      \RSize k {\alpha_v} a
     }{\RNat k t {\zero a}}
 \qquad
-  \ru{t \evalsto \suc{a_v}u \qquad
-      \RSize k {a_v} a \qquad
+  \ru{t \evalsto \suc{\alpha_v}u \qquad
+      \RSize k {\alpha_v} a \qquad
       \RNat k u v
@@
 
-\fbox{$\RSize k {a_v} a$}
+\fbox{$\RSize k {\alpha_v} a$}
 \begin{gather*}
@@
 \qquad
-  \ru{\RSize k a b
-    }{\RElim k a b}
-\qquad
-  \ru{\RSize k a b
-    }{\RElim k {\ann a} {\ann b}}
+  \ru{\RSize k \alpha b
+    }{\RElim k \alpha b}
+\qquad
+  \ru{\RSize k \alpha b
+    }{\RElim k {\ann \alpha} {\ann b}}
 \\[2ex]
@@
 \qquad
-  \ru{T \evalsto \Nat a \qquad
-      \RSize k a b
+  \ru{T \evalsto \Nat \alpha \qquad
+      \RSize k \alpha b
     }{\RTy k T {\Nat b}}
@@
 \]
-\fbox{$\NAT(a)$}
-\begin{gather*}
-  \ru{t \evalsto \zero a \qquad
+\fbox{$\NAT(\alpha)$}
+\begin{gather*}
+  \ru{t \evalsto \zero \alpha \qquad
       t' \evalsto \zero{a'}
@@
 \qquad
-  \ru{t \evalsto \suc a u \qquad
+  \ru{t \evalsto \suc \alpha u \qquad
       t' \evalsto \suc {a'} {u'} \qquad
@@
 \\[2ex]
-  \ru{t \evalsto \up{\Nat a} n \qquad
+  \ru{t \evalsto \up{\Nat \alpha} n \qquad
       t' \evalsto \up{\Nat a'} n' \qquad
@@
   \fbox{$\FORALL\,\F$} \defas
-  \{(t,t') \mid t \ann a = t' \ann{a'} \in \F(b) \mforall a,a',b \in \SIZE \}
+  \{(t,t') \mid t \ann \alpha = t' \ann{a'} \in \F(b) \mforall \alpha,a',b \in \SIZE \}
   .
@@
 \\[3ex]
-  \ru{T \evalsto \Nat\,a \qquad T' \evalsto \Nat\,a
+  \ru{T \evalsto \Nat\,\alpha \qquad T' \evalsto \Nat\,\alpha
     }{T = T' \in \SET[\ell]}
-  & \EL[\ell](T) = \NAT(a)
+  & \EL[\ell](T) = \NAT(\alpha)
 \\[3ex]
@@
       T' \evalsto \piT{\Size}{B'} \quad
-      B[a] = B'[a] \in \SET[\ell] \mforall a \in \SIZE
+      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
     }{T = T' \in \SET[\ell]}
-  & \EL[\ell](T) = \PI(\SIZE,\; a \mapsto \EL[\ell](B[a]))
+  & \EL[\ell](T) = \PI(\SIZE,\; \alpha \mapsto \EL[\ell](B[\alpha]))
 \\[3ex]
@@
       T' \evalsto \forallT{B'} \quad
-      B[a] = B'[a] \in \SET[\ell] \mforall a \in \SIZE
+      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
     }{T = T' \in \SET[\ell]}
-  & \EL[\ell](T) = \FORALL(a \mapsto \EL[\ell](B[a]))
+  & \EL[\ell](T) = \FORALL(\alpha \mapsto \EL[\ell](B[\alpha]))
 \end{array}
@@
 However, in the presence of universes and large eliminations, there is no underlying simple type.
-Of course, we can take a type as its own shape, but we want at least that $\Nat a$ and $\Nat b$ have the same shape even for different $a, b$.
+Of course, we can take a type as its own shape, but we want at least that $\Nat \alpha$ and $\Nat b$ have the same shape even for different $\alpha, b$.
 Thus, we define a relation $T \shape S$ to express that $S$ is a possible shape of type $T$.
@@
 \begin{gather*}
-  \ru{T \evalsto \Nat a \qquad
+  \ru{T \evalsto \Nat \alpha \qquad
       S \evalsto \Nat b
@@
       S \evalsto \piT{\Size}{B'} \qquad
-      B[a] \shape B'[a] \mforall a \in \SIZE
+      B[\alpha] \shape B'[\alpha] \mforall \alpha \in \SIZE
     }{T \shape S}
@@
       S \evalsto \forallT{B'} \qquad
-      B[a] \shape B'[a'] \mforall a,a' \in \SIZE
+      B[\alpha] \shape B'[a'] \mforall \alpha,a' \in \SIZE
     }{T \shape S}
@@
 \begin{gather*}
-  \ru{B[a] = B[a] \in \SET[\ell] \mforall a \in \SIZE
+  \ru{B[\alpha] = B[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
     }{T \in \SET[\ell]}
@@
   \ru{S_1 \evalsto \forallT{B_1} \qquad
-      B[a] \shape B_1[a'] \mforall a, a' \in \SIZE
+      B[\alpha] \shape B_1[a'] \mforall \alpha, a' \in \SIZE
     }{T \shape S_1}
@@
   \ru{S_2 \evalsto \forallT{B_2} \qquad
-      B[a] \shape B_2[a'] \mforall a, a' \in \SIZE
+      B[\alpha] \shape B_2[a'] \mforall \alpha, a' \in \SIZE
     }{T \shape S_2}
@@
 \begin{enumerate}
-\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $a_1,a_2 \in \SIZE$.
-  Since $n_1 \ann{a_1} = n_2 \ann{a_2} \in \NE$, we obtain
-  $\up{B_1[a_1]}{(n_1 \ann{a_1})} = \up{B_2[a_2]}{(n_2 \ann{a_2})} \in B[a_1]$
+\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $\alpha_1,\alpha_2 \in \SIZE$.
+  Since $n_1 \ann{\alpha_1} = n_2 \ann{\alpha_2} \in \NE$, we obtain
+  $\up{B_1[\alpha_1]}{(n_1 \ann{\alpha_1})} = \up{B_2[\alpha_2]}{(n_2 \ann{\alpha_2})} \in B[\alpha_1]$
   by induction hypothesis.
-  Thus, $(\up{S_1}{n_1}) \ann{a_1} = (\up{S_2}{n_2}) \ann{a_2} \in B[a_1]$
+  Thus, $(\up{S_1}{n_1}) \ann{\alpha_1} = (\up{S_2}{n_2}) \ann{\alpha_2} \in B[\alpha_1]$
   by weak head expansion, which entails the goal by definition of $\EL(T)$.
@@
 \\[2ex]
-  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad a \in \SIZE
-    }{\sext \eta a = \sext{\eta'}{a} \eeq \sext{\rho}{a} \in \cext \Gamma \Size}
-\qquad
-  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad a,a',b \in \SIZE
-    }{\sext \eta a = \sext{\eta'}{a'} \eeq \sext{\rho}{b} \in \erext \Gamma \Size}
+  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad \alpha \in \SIZE
+    }{\sext \eta \alpha = \sext{\eta'}{\alpha} \eeq \sext{\rho}{\alpha} \in \cext \Gamma \Size}
+\qquad
+  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad \alpha,a',b \in \SIZE
+    }{\sext \eta \alpha = \sext{\eta'}{a'} \eeq \sext{\rho}{b} \in \erext \Gamma \Size}
 \end{gather*}
@@
 and show $(\forall T)\eta = (\forall T)\eta' \in \SET[\ell]$ for some $\ell$.
-To this end, assume $a \in \SIZE$ and show $T(\eta,a) = T(\eta',a) \in \SET[\ell]$.
-Note that $(\eta,a) = (\eta',a) \eeq (\rho,a) \in
+To this end, assume $\alpha \in \SIZE$ and show $T(\eta,\alpha) = T(\eta',\alpha) \in \SET[\ell]$.
+Note that $(\eta,\alpha) = (\eta',\alpha) \eeq (\rho,\alpha) \in
 \resurrect {\erext \Gamma \Size} = \cext {\resurrect\Gamma}\Size$,
@@
 show $(\lambda t)\eta = (\lambda t')\eta' \in (\forall T)\rho$.
-To this end, assume arbitrary $a,a',b \in \SIZE$ and show
-$t(\eta,a) = t'(\eta',a') \in T(\rho,b)$.
-Since $(\eta,a) = (\eta',a') \eeq (\rho,b) \in \erext \Gamma \Size$,
+To this end, assume arbitrary $\alpha,a',b \in \SIZE$ and show
+$t(\eta,\alpha) = t'(\eta',a') \in T(\rho,b)$.
+Since $(\eta,\alpha) = (\eta',a') \eeq (\rho,b) \in \erext \Gamma \Size$,
 we conclude by induction hypothesis.
@@
 $t \eta = t' \eta' : (\forall T) \rho$, thus, by definition of this PER,
-$t \eta \ann {a_1} = t' \eta' \ann{a_2} \in T(\rho, b')$ for any size values $a_1$, $a_2$, and $b'$.
-We conclude by choosing $a_1 = a \eta$ and $a_2 = a' \eta'$ and $b' = b \rho$.
+$t \eta \ann {\alpha_1} = t' \eta' \ann{\alpha_2} \in T(\rho, b')$ for any size values $\alpha_1$, $\alpha_2$, and $b'$.
+We conclude by choosing $\alpha_1 = a \eta$ and $\alpha_2 = a' \eta'$ and $b' = b \rho$.
 
@@
 \[
-  \fbox{$\LRSize \Gamma a b$} \defas
-  \forall \sigma : \Gamma' \leq \Gamma.\
-  \exists a'.\
-  \RSize {\Gamma'} b a' \mand
-  \Gamma' \der a\sigma = a' : \Size
-\]
-
-\fbox{$\LR \Gamma t {T'} f {A'}$} defined by induction on ${A'} \in \SET[\ell]$.
+  \fbox{$\LRSize \Gamma a \alpha$} \defas
+  \forall \sigma : \Gamma' \leq \Gamma.\ \
+  \RSize {\Gamma'} \alpha {a\sigma}
+  % \exists a'.\
+  % \RSize {\Gamma'} \alpha a' \mand
+  % \Gamma' \der a\sigma = a' : \Size
+\]
+
+\fbox{$\LRT \Gamma {T'} {A'} \ell$} and \fbox{$\LR \Gamma t {T'} f {A'}$} defined by induction on ${A'} \in \SET[\ell]$.
+
 \begin{caselist}
 
-\nextcase $A' \evalsto \piT A B$.  Then $\LR \Gamma t {T'} f {A'}$ if all of the following:
+\nextcase $A' \evalsto N$ neutral.
+
+  \noindent
+  $\LRT{\Gamma}{T'}{A'}\ell$ iff
+  $\forall \sigma : \Gamma' \leq \Gamma.\ \ \exists M.\ \ \RTy{\Gamma'}{A'}M \mand \Gamma' \der T'\sigma = M : \Set[\ell]$.
+
+  \noindent
+  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff
+  $\forall \sigma : \Gamma' \leq \Gamma.\ \
+   \exists m. \ \
+   \R {\Gamma'} {\down {A'} f} m \mand
+   \Gamma' \der t\sigma = m : T'\sigma$.
+
+\nextcase $A' \evalsto \Nat \alpha$.
+
+  \noindent
+  $\LRT{\Gamma}{T'}{A'}\ell$ iff $\Gamma \der T' = \Nat a$ for some $a$ and $\LRSize \Gamma a \alpha$.
+
+  \noindent
+  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff $\resurrect\Gamma \der T' = \Nat a$ for some $a$ and
+  $\LRSize {\resurrect\Gamma} a \alpha$ and \\
+  $\forall \sigma : \Gamma' \leq \Gamma.\ \
+   \exists v. \ \
+   \R {\Gamma'} {\down {A'} f} v \mand
+   \Gamma' \der t\sigma = v : \Nat\,a\sigma$.
+
+\nextcase $A' \evalsto \Set[\ell']$.
+
+  \noindent
+  $\LRT {\Gamma} {T'} {A'} \ell$ iff $\Gamma \der T' = \Set[\ell'] : \Set[\ell]$.
+
+  \noindent
+  $\LR \Gamma U {T'} B {A'}$ iff
+  $\resurrect\Gamma \der T' = \Set[\ell'] : \Set[\ell]$ and $\LRT \Gamma U B {\ell'}$.
+
+\nextcase $A' \evalsto \piT A B$.
+
+\noindent
+$\LRT \Gamma {T'} {A'} \ell$ iff
 \[
 \begin{array}{l}
-\Gamma \der T' = \piT U T \mforsome U, T\\
-\LRT \Gamma U A \\
+\Gamma \der T' = \piT U T : \Set[\ell] \mforsome U, T \mand \LRT \Gamma U A \ell \mand \\
+\forall \sigma : \Gamma' \leq \Gamma.\ \
+\LR {\Gamma'} u {U\sigma} a A \implies \LRT {\Gamma'} {T(\sigma,u)} {B[a]} \ell
+. \\
+\end{array}
+\]
+$\LR \Gamma t {T'} f {A'}$ iff
+\[
+\begin{array}{l}
+\resurrect\Gamma \der T' = \piT U T : \Set[\ell] \mforsome U, T \mand \LRT {\resurrect\Gamma} U A \ell \mand \\
 \forall \sigma : \Gamma' \leq \Gamma.\ \
 \LR {\Gamma'} u {U\sigma} a A \implies \LR {\Gamma'} {t\sigma\,u} {T(\sigma,u)} {f\,a} {B[a]}
+.
 \end{array}
@@
 
-\nextcase $A' \evalsto \forallT B$.  Then $\LR \Gamma t {T'} f {A'}$ if all of the following:
+\nextcase $A' \evalsto \piT \Size B$.
+
+\noindent
+$\LRT \Gamma {T'} {A'} \ell$ iff
 \[
 \begin{array}{l}
-\Gamma \der T' = \forallT T \mforsome T\\
+\Gamma \der T' = \piT \Size T : \Set[\ell] \mforsome T\\
 \forall \sigma : \Gamma' \leq \Gamma.\ \
-\LRSize {\Gamma'} u a \implies \LR {\Gamma'} {t\sigma\,u} {T(\sigma,u)} {f\,a} {B[a]}
+\LRSize {\Gamma'} a \alpha
+\implies \LRT {\Gamma'} {T(\sigma,a)} {B[\alpha]} \ell
+.
 \end{array}
 \]
+$\LR \Gamma t {T'} f {A'}$ iff
+\[
+\begin{array}{l}
+\resurrect\Gamma \der T' = \piT \Size T : \Set[\ell] \mforsome T\\
+\forall \sigma : \Gamma' \leq \Gamma.\ \
+\LRSize {\Gamma'} a \alpha
+\implies \LR {\Gamma'} {t\sigma\,a} {T(\sigma,a)} {f\,\alpha} {B[\alpha]}
+.
+\end{array}
+\]
+
+\nextcase $A' \evalsto \forallT B$.
+
+\noindent
+$\LRT \Gamma {T'} {A'} \ell$ iff
+\[
+\begin{array}{l}
+\Gamma \der T' = \forallT T : \Set[\ell] \mforsome T\\
+\forall \sigma : \Gamma' \leq \Gamma,\
+{\Gamma'} \der b : \Size,\
+\beta \in \SIZE. \ \
+\LRSize {\Gamma'} b \beta
+\implies \LRT {\Gamma'} {T(\sigma,b)} {B[\beta]} \ell
+.
+\end{array}
+\]
+$\LR \Gamma t {T'} f {A'}$ iff
+\[
+\begin{array}{l}
+\resurrect\Gamma \der T' = \forallT T : \Set[\ell] \mforsome T\\
+\forall \sigma : \Gamma' \leq \Gamma,\
+\Gamma' \der a : \Size,\
+\resurrect{\Gamma'} \der b : \Size,\
+\alpha,\beta \in \SIZE. \ \\ \qquad
+\LRSize {\resurrect{\Gamma'}} b \beta
+\implies \LR {\Gamma'} {t\sigma\,\ann a} {T(\sigma,b)} {f\,\ann\alpha} {B[\beta]}
+.
+\end{array}
+\]
 
 \end{caselist}
+
+\begin{theorem}[Into and out of the logical relation]
+  \bla
+  Let $A \in \Set[\ell]$.
+  \begin{enumerate}
+
+  \item $\LR \Gamma t T {\up A n} A$ if for all $\sigma : \Gamma' \leq \Gamma$ there is an $m$ such that
+     $\RNe{\Gamma'}n m$ and $\Gamma' \der t\sigma = m : T$.
+
+  \item If\/ $\LRT{\Gamma} T A \ell$ then for all $\sigma : \Gamma' \leq \Gamma$ there is some normal form $V$ such that
+  $\Rty{\Gamma'} A V$ and $\Gamma' \der T\sigma = V : \Set[\ell]$.
+
+  \item If\/ $\LR \Gamma t T f A$ then for all $\sigma : \Gamma' \leq \Gamma$ there is some normal form $v$ such that
+  $\R{\Gamma'} {\down A f} v$ and $\Gamma' \der t\sigma = v : T\sigma$.
+
+  \end{enumerate}
+\end{theorem}
+
+Logical relation for substitutions.
+\fbox{$\LRSub \Gamma \sigma \tau \Delta \eta \rho$}.
+\begin{gather*}
+  \ru{\der \Gamma
+    }{\LRSub \Gamma {\sempty} {\sempty} {\cempty} {\sempty} {\sempty}}
+\\[2ex]
+  \ru{\LRSub \Gamma \sigma \tau \Delta \eta \rho \qquad
+      \resurrect \Delta \der T \qquad
+      \Gamma \der u = t : T\tau \qquad
+      \LR \Gamma t {T \tau} f {T \rho} \qquad
+      f = g \in T\rho
+    }{\LRSub \Gamma {(\sigma,u)} {(\tau,t)} {\cext \Delta T} {(\eta,f)} {(\rho,q)}}
+\\[2ex]
+  \ru{\LRSub \Gamma \sigma \tau \Delta \eta \rho \qquad
+      \Gamma \der a : \Size \qquad
+      \LRSize \Gamma a \alpha
+    }{\LRSub \Gamma {(\sigma,a)} {(\tau,a)} {\cext \Delta \Size} {(\eta,\alpha)} {(\rho,\alpha)}}
+\\[2ex]
+  \ru{\LRSub \Gamma \sigma \tau \Delta \eta \rho \qquad
+      \Gamma \der a, b : \Size \qquad
+      \alpha,\beta \in \SIZE \qquad
+      \LRSize \Gamma b \beta
+    }{\LRSub \Gamma {(\sigma,a)} {(\tau,b)} {\erext \Delta \Size} {(\eta,\alpha)} {(\rho,\beta)}}
+\end{gather*}
 
@@
 % \end{lemma}
+
+Validity judgements.
+\[
+\begin{array}{l@{\quad}l@{\quad}l}
+  \valid () & \defas & \mtrue \\
+  \valid \eext \Gamma \Size & \defas & \valid \Gamma \\
+  \valid \cext \Gamma T     & \defas & \valid \Gamma \mand \resurrect\Gamma \valid T
+\\[2ex]
+  \Gamma \valid s & \defas & \valid \Gamma \\
+%  \Gamma \valid s & \defas & \mtrue \\
+  \Gamma \valid T & \defas & \Gamma \valid T = T \\
+  \Gamma \valid T = T' & \defas & \Gamma \valid T = T' : s \mforsome s %\Set[\ell] \mforsome \ell
+\\[2ex]
+  \Gamma \valid t : T & \defas & \Gamma \valid t = t : T \\
+  \Gamma \valid t = t' : T & \defas & %\valid \Gamma \mand
+    % \resurrect\Gamma \valid T \mand
+    \valid \cext \Gamma T \mand
+    \LR {\Gamma'} {t \sigma} {T \tau} {t' \eta} {T \rho}
+    \mforall \LRSub {\Gamma'} \sigma \tau \Gamma \eta \rho
+\\[2ex]
+  \Gamma \valid \sigma : \Delta & \defas & \Gamma \valid \sigma = \sigma \eeq \sigma : \Delta \\
+  \Gamma \valid \sigma_1 = \sigma_2 \eeq \tau : \Delta & \defas &
+    \valid \Gamma \mand \valid \Delta \mand
+    \LRSub {\Gamma'} {\sigma_1 \sigma'} {\tau \tau'} \Delta {\sigma_2 \eta'} {\tau \rho'}
+    \mforall \LRSub {\Gamma'} {\sigma'} {\tau'} \Gamma {\eta'} {\rho'}
+\end{array}
+\]
+
+\begin{theorem}[Fundamental theorem]
+  \bla
+  \begin{enumerate}
+  \item
+  If $\der \Gamma$ then $\valid \Gamma$.
+  \item
+  If $\Gamma \der J$ then % $\valid \Gamma$ and
+  $\Gamma \valid J$
+  \end{enumerate}
+\end{theorem}
+\begin{proof}
+Simultaneously, by induction on the derivation.
+
+\begin{caselist}
+\nextcase $\forall$-introduction.
+\[
+  \ru{\erext \Gamma \Size \der t = t' : T
+    }{\Gamma \der \lambda t = \lambda t' : \forall T}
+\]
+First $\valid \Gamma$ follows from the induction hypothesis $\valid {\erext \Gamma \Size}$.
+To show $\resurrect\Gamma \valid \forall T$,
+assume $\eta = \eta' \eeq \rho \in \resurrect \Gamma$
+and show $(\forall T)\eta = (\forall T)\eta' \in \SET[\ell]$ for some $\ell$.
+To this end, assume $\alpha \in \SIZE$ and show $T(\eta,\alpha) = T(\eta',\alpha) \in \SET[\ell]$.
+Note that $(\eta,\alpha) = (\eta',\alpha) \eeq (\rho,\alpha) \in
+\resurrect {\erext \Gamma \Size} = \cext {\resurrect\Gamma}\Size$,
+thus, we can instantiate the induction hypothesis and obtain our goal.
+
+For the main goal,
+assume $\eta = \eta' \eeq \rho \in \Gamma$ and
+show $(\lambda t)\eta = (\lambda t')\eta' \in (\forall T)\rho$.
+To this end, assume arbitrary $\alpha,a',b \in \SIZE$ and show
+$t(\eta,\alpha) = t'(\eta',a') \in T(\rho,b)$.
+Since $(\eta,\alpha) = (\eta',a') \eeq (\rho,b) \in \erext \Gamma \Size$,
+we conclude by induction hypothesis.
+
+\nextcase $\forall$-elimination.
+\[
+  \ru{\Gamma \der t = t' : \forall T \qquad
+      \Gamma \der a, a' : \Size \qquad
+      \resurrect\Gamma \der b : \Size
+    }{\Gamma \der t \ann a = t' \ann {a'} : T[b]}
+\]
+First, $\valid \Gamma$ follows by induction hypothesis.
+For goal $\resurrect \Gamma \valid T[b]$,
+assume $\eta = \eta' \eeq \rho \in \resurrect\Gamma$ and show
+$T[b]\eta = T[b]\eta' \in \SET[\ell]$ for some $\ell$.
+By induction hypothesis,  $b\eta = b\eta' \in \SIZE$.
+% , thus,
+% $(\eta,b\eta) = (\eta',b\eta') \eeq (\rho,b\eta) \in \cext{\resurrect\Gamma}\Size$
+By another induction hypothesis, $(\forall T)\eta = (\forall T)\eta' \in \SET[\ell]$,
+which by definition entails our goal $T(\eta, b\eta) = T(\eta', b\eta') \in \SET[\ell]$.
+
+For the remaining main goal,
+assume $\eta = \eta' \eeq \rho \in \Gamma$ and
+show $t \ann a \eta = t' \ann{a'} \eta' \in T[b]\rho$.
+By definition of substitution, it suffices to show
+$t \eta \ann{a \eta} = t' \eta' \ann{a' \eta'} \in T(\rho, b \rho)$.
+By induction hypothesis,
+$t \eta = t' \eta' : (\forall T) \rho$, thus, by definition of this PER,
+$t \eta \ann {\alpha_1} = t' \eta' \ann{\alpha_2} \in T(\rho, b')$ for any size values $\alpha_1$, $\alpha_2$, and $b'$.
+We conclude by choosing $\alpha_1 = a \eta$ and $\alpha_2 = a' \eta'$ and $b' = b \rho$.
+
+We now argue that this choice is possible,
+namely that $a \eta, a'\eta, b\rho \in \SIZE$.
+Note that the induction hypothesis gives us
+$a \eta  = a  \eta' \in \SIZE$ and
+$a' \eta = a' \eta' \in \SIZE$.
+By resurrection (Lemma~\ref{lem:resenv}) we have $\rho \in \resurrect\Gamma$,
+thus, by induction hypothesis, $b \rho \in \SIZE$.
+\end{caselist}
+\end{proof}
 

### 2017-02-20 nbe/icfp17.tex
[ ICFP 17] Second fund. thm.

@@
 \begin{enumerate}
-\item If\/ $\der \Gamma$ then $\der \resurrect\Gamma$ and $\resurrect \Gamma \der \sid : \Gamma$.
+\item $\der \Gamma$ iff $\der \resurrect\Gamma$;  then $\resurrect \Gamma \der \sid : \Gamma$.
   % Consequently, if\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
@@
 \item If\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
-\item If\/ $\Gamma \der J : \resurrect \Delta$ and $\der \Delta$ then $\Gamma \der J : \Delta$.
-\item If\/ $\Gamma \der \sigma : \Delta$ then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
+\item If\/ $\Gamma \der J : \resurrect \Delta$ % and $\der \Delta$
+  then $\Gamma \der J : \Delta$.
+\item If\/ $\Gamma \der \sigma : \Delta$ then $\Gamma \der \sigma : \resurrect\Delta$.
+% then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
 \end{enumerate}
@@
 
+\begin{lemma}[Properties of the logical relation for substitution]
+  \label{lem:wklrsub}
+  Let $\LRSub \Gamma \sigma \tau \Delta \eta \rho$.
+  \begin{enumerate}
+  \item Weakening:
+  If $\sigma' : \Gamma' \leq \Gamma$ then
+  $\LRSub{\Gamma'}{\sigma\sigma'}{\tau\sigma'}{\Delta}\eta\rho$.
+  \item Resurrection:
+  Let $\LRSub {\Gamma} \tau \tau {\resurrect\Delta} \rho \rho$.
+  \end{enumerate}
+\end{lemma}
+
 % \begin{lemma}[Typing of constructed data]
@@
 \begin{array}{l@{\quad}l@{\quad}l}
-  \valid () & \defas & \mtrue \\
-  \valid \eext \Gamma \Size & \defas & \valid \Gamma \\
-  \valid \cext \Gamma T     & \defas & \valid \Gamma \mand \resurrect\Gamma \valid T
-\\[2ex]
-  \Gamma \valid s & \defas & \valid \Gamma \\
-%  \Gamma \valid s & \defas & \mtrue \\
-  \Gamma \valid T & \defas & \Gamma \valid T = T \\
-  \Gamma \valid T = T' & \defas & \Gamma \valid T = T' : s \mforsome s %\Set[\ell] \mforsome \ell
-\\[2ex]
-  \Gamma \valid t : T & \defas & \Gamma \valid t = t : T \\
-  \Gamma \valid t = t' : T & \defas & %\valid \Gamma \mand
+  \Gamma \valid t : T & \defas & %\valid \Gamma \mand
     % \resurrect\Gamma \valid T \mand
     \valid \cext \Gamma T \mand
-    \LR {\Gamma'} {t \sigma} {T \tau} {t' \eta} {T \rho}
+    \LR {\Gamma'} {t \sigma} {T \tau} {t \eta} {T \rho}
     \mforall \LRSub {\Gamma'} \sigma \tau \Gamma \eta \rho
-\\[2ex]
-  \Gamma \valid \sigma : \Delta & \defas & \Gamma \valid \sigma = \sigma \eeq \sigma : \Delta \\
-  \Gamma \valid \sigma_1 = \sigma_2 \eeq \tau : \Delta & \defas &
-    \valid \Gamma \mand \valid \Delta \mand
-    \LRSub {\Gamma'} {\sigma_1 \sigma'} {\tau \tau'} \Delta {\sigma_2 \eta'} {\tau \rho'}
+\\[1ex]
+  \Gamma \valid \sigma : \Delta & \defas &
+    \LRSub {\Gamma'} {\sigma \sigma'} {\sigma \tau'} \Delta {\sigma \eta'} {\sigma \rho'}
     \mforall \LRSub {\Gamma'} {\sigma'} {\tau'} \Gamma {\eta'} {\rho'}
@@
 
-\begin{theorem}[Fundamental theorem]
+\begin{theorem}[Fundamental theorem of typing]
   \bla
@@
   \item
-  If $\der \Gamma$ then $\valid \Gamma$.
+  If $\Gamma \der t : T$ then $\Gamma \valid t : T$.
   \item
-  If $\Gamma \der J$ then % $\valid \Gamma$ and
-  $\Gamma \valid J$
+  If $\Gamma \der \sigma : \Delta$ then
+  $\Gamma \valid \sigma : \Delta$
   \end{enumerate}
@@
 \begin{proof}
-Simultaneously, by induction on the derivation.
+Each by induction on the derivation.
 
 \begin{caselist}
+
 \nextcase $\forall$-introduction.
 \[
-  \ru{\erext \Gamma \Size \der t = t' : T
-    }{\Gamma \der \lambda t = \lambda t' : \forall T}
-\]
-First $\valid \Gamma$ follows from the induction hypothesis $\valid {\erext \Gamma \Size}$.
-To show $\resurrect\Gamma \valid \forall T$,
-assume $\eta = \eta' \eeq \rho \in \resurrect \Gamma$
-and show $(\forall T)\eta = (\forall T)\eta' \in \SET[\ell]$ for some $\ell$.
-To this end, assume $\alpha \in \SIZE$ and show $T(\eta,\alpha) = T(\eta',\alpha) \in \SET[\ell]$.
-Note that $(\eta,\alpha) = (\eta',\alpha) \eeq (\rho,\alpha) \in
-\resurrect {\erext \Gamma \Size} = \cext {\resurrect\Gamma}\Size$,
-thus, we can instantiate the induction hypothesis and obtain our goal.
-
-For the main goal,
-assume $\eta = \eta' \eeq \rho \in \Gamma$ and
-show $(\lambda t)\eta = (\lambda t')\eta' \in (\forall T)\rho$.
-To this end, assume arbitrary $\alpha,a',b \in \SIZE$ and show
-$t(\eta,\alpha) = t'(\eta',a') \in T(\rho,b)$.
-Since $(\eta,\alpha) = (\eta',a') \eeq (\rho,b) \in \erext \Gamma \Size$,
-we conclude by induction hypothesis.
+  \ru{\erext \Gamma \Size \der t : T
+    }{\Gamma \der \lambda t : \forall T}
+\]
+Assume $\LRSub {\Delta} {\sigma} {\tau} \Gamma {\eta} {\rho}$ and show
+$\LR \Delta {(\lambda t)\sigma} {(\forall T)\tau} {(\lambda t)\eta} {(\forall T)\rho}$.
+To this end, assume a weakening $\sigma' : \Delta' \leq \Delta$ and
+size expressions $\Delta' \der a : \Size$ and
+$\resurrect{\Delta'} \der b : \Size$ and size values $\alpha,\beta$ with
+$\LRSize {\Delta'} b \beta$ and show
+$\LR {\Delta'} {t(\sigma\sigma',a)} {T(\tau\sigma',b)} {t(\eta,\alpha)} {T(\rho,\beta)}$.
+This follows from the induction hypothesis, since
+$\LRSub {\Delta'} {(\sigma\sigma',a)} {(\tau\sigma',b)} {\erext \Gamma \Size} {(\eta,\alpha)} {(\rho,\beta)}$
+by Lemma~\ref{lem:wklrsub} and substitution extension.
 
@@
 \[
-  \ru{\Gamma \der t = t' : \forall T \qquad
-      \Gamma \der a, a' : \Size \qquad
-      \resurrect\Gamma \der b : \Size
-    }{\Gamma \der t \ann a = t' \ann {a'} : T[b]}
-\]
-First, $\valid \Gamma$ follows by induction hypothesis.
-For goal $\resurrect \Gamma \valid T[b]$,
-assume $\eta = \eta' \eeq \rho \in \resurrect\Gamma$ and show
-$T[b]\eta = T[b]\eta' \in \SET[\ell]$ for some $\ell$.
-By induction hypothesis,  $b\eta = b\eta' \in \SIZE$.
-% , thus,
-% $(\eta,b\eta) = (\eta',b\eta') \eeq (\rho,b\eta) \in \cext{\resurrect\Gamma}\Size$
-By another induction hypothesis, $(\forall T)\eta = (\forall T)\eta' \in \SET[\ell]$,
-which by definition entails our goal $T(\eta, b\eta) = T(\eta', b\eta') \in \SET[\ell]$.
-
-For the remaining main goal,
-assume $\eta = \eta' \eeq \rho \in \Gamma$ and
-show $t \ann a \eta = t' \ann{a'} \eta' \in T[b]\rho$.
-By definition of substitution, it suffices to show
-$t \eta \ann{a \eta} = t' \eta' \ann{a' \eta'} \in T(\rho, b \rho)$.
-By induction hypothesis,
-$t \eta = t' \eta' : (\forall T) \rho$, thus, by definition of this PER,
-$t \eta \ann {\alpha_1} = t' \eta' \ann{\alpha_2} \in T(\rho, b')$ for any size values $\alpha_1$, $\alpha_2$, and $b'$.
-We conclude by choosing $\alpha_1 = a \eta$ and $\alpha_2 = a' \eta'$ and $b' = b \rho$.
-
-We now argue that this choice is possible,
-namely that $a \eta, a'\eta, b\rho \in \SIZE$.
-Note that the induction hypothesis gives us
-$a \eta  = a  \eta' \in \SIZE$ and
-$a' \eta = a' \eta' \in \SIZE$.
-By resurrection (Lemma~\ref{lem:resenv}) we have $\rho \in \resurrect\Gamma$,
-thus, by induction hypothesis, $b \rho \in \SIZE$.
+  \ru{\Gamma \der t : \forall T \qquad \Gamma \der a : \Size \qquad \resurrect \Gamma \der b : \Size
+    }{\Gamma \der t \ann a : T[b]}
+\]
+Assume $\LRSub {\Delta} {\sigma} {\tau} \Gamma {\eta} {\rho}$ and show
+$\LR \Delta {t \ann a \sigma} {T[b]\tau} {t \ann a \eta} {T[b]\rho}$.
+By induction hypothesis we have $\LR \Delta {t \sigma} {(\forall T)\tau} {t \eta} {(\forall T) \rho}$.
+It suffices to show
+\begin{enumerate}
+\item $\Delta \der a\sigma : \Size$
+%\item $\resurrect\Delta \der b\tau : \Size$.
+ and $a\eta \in \SIZE$: follows by induction hypothesis on $\Gamma \der a : \Size$.
+%\item $b\rho \in \SIZE$.
+\item $\LRSize {\resurrect \Delta} {b\tau} {b\rho}$: follows from induction hypothesis on $\resurrect \Gamma \der b : \Size$ via $\LRSub {\resurrect \Delta} \tau \tau {\resurrect\Gamma} \rho \rho$ (from Lemma~\ref{lem:wklrsub}).
+\end{enumerate}
+
 \end{caselist}

### 2017-02-21 nbe/icfp17.tex
[ ICFP 17 ] Proved subsumption.

@@
      }{\Gamma \der u\,\fix[\ell] T t : T \,a \, u}
+\qquad
+  \ru{\Gamma \der t : T \qquad
+      \resurrect\Gamma \der T \leq T'
+     }{\Gamma \der t : T'}
 \end{gather*}
@@
      }{\Gamma \der u\,\fix[\ell] T t = u'\,\fix[\ell] {T'} t' : T \, a \, u}
+\qquad
+  \ru{\Gamma \der t = t' : T \qquad
+      \resurrect\Gamma \der T \leq T'
+     }{\Gamma \der t = t' : T'}
 \end{gather*}
@@
 
-\fbox{$\Gamma \der T \leq T'$} (implies $\resurrect \Gamma \der T,T'$)
+\fbox{$\Gamma \der T \leq T'$} (implies $\Gamma \der T,T'$)
+% (implies $\resurrect \Gamma \der T,T'$)
 \begin{gather*}
@@
     }{\Gamma \der \piT U T \leq \piT {U'} {T'}}
+\qquad
+  \ru{\cext \Gamma {\Size} \der T \leq T'
+    }{\Gamma \der \epiT \Size T \leq \epiT {\Size} {T'}}
 \qquad
@@
     }{\Gamma \der \sext \tau a : \eext \Delta \Size}
+\end{gather*}
+
+Weakenings \fbox{$\xi : \Gamma \leq \Delta$}  (implies $\der \Gamma$ and $\der \Delta$ and $\Gamma \der \xi : \Delta$).
+\begin{gather*}
+  \ru{\der \Gamma
+    }{\sempty : \Gamma \leq \cempty}
+\qquad
+  \ru{\xi : \Gamma \leq \Delta \qquad
+      \resurrect\Gamma \der T
+    }{\xi \slift : \cext \Gamma T \leq \Delta}
+\qquad
+  \ru{\xi : \Gamma \leq \Delta \qquad
+      \resurrect \Delta \der U \qquad
+      \resurrect\Gamma \der T \leq U \xi
+    }{(\xi \slift, \ind 0) : \cext \Gamma T \leq \cext \Delta U}
 \end{gather*}
@@
 \\[2ex]
-  \ru{T' \evalsto \Nat b \qquad
+  \ru{T' \evalsto \Nat \beta \qquad
     }{(\up {T'} n) @ \case[\ell] T {t_z} {t_s} \evalsto
@@
 \\[2ex]
-  \ru{T' \evalsto \Nat b \qquad
+  \ru{T' \evalsto \Nat \beta \qquad
     }{(\up {T'} n) @ \fix[\ell] T t \evalsto
@@
 \qquad
-  \ru{t \evalsto \zero{\alpha_v} \qquad
-      \RSize k {\alpha_v} a
+  \ru{t \evalsto \zero{\alpha} \qquad
+      \RSize k {\alpha} a
     }{\RNat k t {\zero a}}
 \qquad
-  \ru{t \evalsto \suc{\alpha_v}u \qquad
-      \RSize k {\alpha_v} a \qquad
+  \ru{t \evalsto \suc{\alpha}u \qquad
+      \RSize k {\alpha} a \qquad
       \RNat k u v
@@
 
-\fbox{$\RSize k {\alpha_v} a$}
+\fbox{$\RSize k {\alpha} a$}
 \begin{gather*}
@@
 A type $T$ will be interpreted as a partial equivalence relation (PER) $\A$ on terms, \ie, a relation which is symmetric and transitive.
-The domain $\dom(\A)$ of the relation can be thought of the set of terms which define the extension of the type; on $\dom(\A)$ the relation $\A$ is in fact an equivalence relation.
+The domain $\dom(\A)$ of the relation can be thought of the set of terms which define the extension of the type; on $\dom(\A) = \{a \mid \exists a'.\ (a,a') \in \A \}$ the relation $\A$ is in fact an equivalence relation.
+We write $a = a' \in \A$ for relatedness in $\A$ and $a \in \A$ if $a \in \dom(\A)$.
 
@@
   \item $\lev k = \lev k \in \NE$.
-  \item If $n = n' \in \NE$ and $d = d' \in \NF$ then $n\,d = n'\,d' in \NE$.
+  \item If $n = n' \in \NE$ and $d = d' \in \NF$ then $n\,d = n'\,d' \in \NE$.
 %  \item If $d = d' \in \NF$
@@
   \ru{t \evalsto \zero \alpha \qquad
-      t' \evalsto \zero{a'}
-    }{t = t' \in \NAT(b+1)}
+      t' \evalsto \zero{\alpha'}
+    }{t = t' \in \NAT(\beta+1)}
 \qquad
   \ru{t \evalsto \suc \alpha u \qquad
-      t' \evalsto \suc {a'} {u'} \qquad
-      u = u' \in \NAT(b)
-    }{t = t' \in \NAT(b+1)}
+      t' \evalsto \suc {\alpha'} {u'} \qquad
+      u = u' \in \NAT(\beta)
+    }{t = t' \in \NAT(\beta+1)}
 \\[2ex]
   \ru{t \evalsto \up{\Nat \alpha} n \qquad
-      t' \evalsto \up{\Nat a'} n' \qquad
+      t' \evalsto \up{\Nat \alpha'} n' \qquad
       n = n' \in \NE
-    }{t = t' \in \NAT(b)}
+    }{t = t' \in \NAT(\beta)}
 \end{gather*}
@@
   \fbox{$\FORALL\,\F$} \defas
-  \{(t,t') \mid t \ann \alpha = t' \ann{a'} \in \F(b) \mforall \alpha,a',b \in \SIZE \}
+  \{(t,t') \mid t \ann \alpha = t' \ann{\alpha'} \in \F(\beta) \mforall \alpha,\alpha',\beta \in \SIZE \}
   .
 \]
+
+\begin{lemma}[PER Subsumption]
+\label{lem:subsump} \hfill
+\begin{enumerate}
+\item If $\alpha \leq \beta$ then $\NAT(\alpha) \subseteq \NAT(\beta)$.
+\item If $\F(\alpha) \leq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
+\item IF $\A' \subseteq \A$ and $\F(u) \subseteq \F'(u)$ for all $u \in \A'$, then $\PIAF \subseteq \PI\,\A'\,\F'$.
+\end{enumerate}
+\end{lemma}
 
@@
 \]
-All PERs involved here are closed under weak head expansion.
-
-The PER model has the usual properties, one of them being that $\EL[\ell](T)$ does not depend $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$.
+All relations involved here are closed under weak head expansion.
+\begin{lemma}[Well-definedness]
+  Let $\DD :: T_1 = T_2 \in \SET[\ell]$.
+  \begin{enumerate}
+  \item Symmetry: $T_2 = T_1 \in \SET[\ell]$.
+  \item Transitivity: If $T_2 = T_3 \in \SET[\ell]$ then $T_1 = T_3 \in \SET[\ell]$.
+  \item Extension is PER: $\EL[\ell](T_1) = \EL[\ell](T_2)$ and ``both'' are PERs.
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+  Simultaneously by induction on $\DD$.
+\end{proof}
+
+% The PER model has the usual properties, one of them being that $\EL[\ell](T)$ does not depend $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$.
 Thus, we simply write $t = t' \in \EL(T)$ or even $t = t' \in T$.
+\begin{lemma}[Derivation independence of extension]
+\label{lem:indep}
+  If $\DD_1 :: T = T_1 \in \SET[\ell_1]$
+  and $\DD_2 :: T_2 = T \in \SET[\ell_2]$ then $\EL[\ell_1](T) = \EL[\ell_2](T)$.
+\end{lemma}
+\begin{proof}
+  By induction on $\DD_1$ and cases on $\DD_2$.
+\end{proof}
+Sine $\EL[\ell](T)$ does not depend $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$,  we may simply write $t = t' \in \EL(T)$ or even $t = t' \in T$.
+
+\begin{lemma}[Universe subsumption]
+  \label{lem:usubsump}
+  If $\ell \leq \ell'$ then $\SET[\ell] \subseteq \SET[\ell']$.
+\end{lemma}
+\begin{proof}
+  Prove $T = T' \in \SET[\ell']$ by induction on $T = T' \in \SET[\ell]$.
+  For the base types this is direct, let us look at the function space.
+\[
+  \rul{T \evalsto \piT A B \qquad
+      T' \evalsto \piT{A'}{B'} \qquad
+      A = A' \in \SET[\ell] \qquad
+      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
+    }{T = T' \in \SET[\ell]}
+\]
+  By induction hypothesis, $A = A' \in \SET[\ell']$, and we have $\EL[\ell'](A) = \EL[\ell](A)$ by Lemma~\ref{lem:indep}.
+Assuming $u = u' \in \EL(A)$, we get $B[u] = B'[u'] \in \SET[\ell']$ by induction hypothesis on $B[u] = B'[u'] \in \SET[\ell]$.
+\end{proof}
 
@@
 However, in the presence of universes and large eliminations, there is no underlying simple type.
-Of course, we can take a type as its own shape, but we want at least that $\Nat \alpha$ and $\Nat b$ have the same shape even for different $\alpha, b$.
+Of course, we can take a type as its own shape, but we want at least that $\Nat \alpha$ and $\Nat \beta$ have the same shape even for different $\alpha, \beta$.
 Thus, we define a relation $T \shape S$ to express that $S$ is a possible shape of type $T$.
@@
   \ru{T \evalsto \Nat \alpha \qquad
-      S \evalsto \Nat b
+      S \evalsto \Nat \beta
     }{T \shape S}
@@
       S \evalsto \forallT{B'} \qquad
-      B[\alpha] \shape B'[a'] \mforall \alpha,a' \in \SIZE
+      B[\alpha] \shape B'[\alpha'] \mforall \alpha,\alpha' \in \SIZE
     }{T \shape S}
@@
   \ru{S_1 \evalsto \forallT{B_1} \qquad
-      B[\alpha] \shape B_1[a'] \mforall \alpha, a' \in \SIZE
+      B[\alpha] \shape B_1[\alpha'] \mforall \alpha, \alpha' \in \SIZE
     }{T \shape S_1}
@@
   \ru{S_2 \evalsto \forallT{B_2} \qquad
-      B[\alpha] \shape B_2[a'] \mforall \alpha, a' \in \SIZE
+      B[\alpha] \shape B_2[\alpha'] \mforall \alpha, \alpha' \in \SIZE
     }{T \shape S_2}
@@
 \qquad
-  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad \alpha,a',b \in \SIZE
-    }{\sext \eta \alpha = \sext{\eta'}{a'} \eeq \sext{\rho}{b} \in \erext \Gamma \Size}
+  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad \alpha,\alpha',\beta \in \SIZE
+    }{\sext \eta \alpha = \sext{\eta'}{\alpha'} \eeq \sext{\rho}{\beta} \in \erext \Gamma \Size}
 \end{gather*}
@@
   \Gamma \models T = T' & \defas & \Gamma \models T = T' : s \mforsome s %\Set[\ell] \mforsome \ell
+\\[2ex]
+  \Gamma \models T \leq T' & \defas & \Gamma \models T \mand \Gamma \models T' \mand
+    \EL(T\eta) \subseteq \EL(T'\eta') \mforall \eta = \eta' \eeq \rho \in \Gamma
 \\[2ex]
@@
 Simultaneously, by induction on the derivation.
+The soundness of subtyping and the conversion rule is due to subsumption (lemmata \ref{lem:subsump} and \ref{lem:usubsump}).
 
@@
 show $(\lambda t)\eta = (\lambda t')\eta' \in (\forall T)\rho$.
-To this end, assume arbitrary $\alpha,a',b \in \SIZE$ and show
-$t(\eta,\alpha) = t'(\eta',a') \in T(\rho,b)$.
-Since $(\eta,\alpha) = (\eta',a') \eeq (\rho,b) \in \erext \Gamma \Size$,
+To this end, assume arbitrary $\alpha,\alpha',\beta \in \SIZE$ and show
+$t(\eta,\alpha) = t'(\eta',\alpha') \in T(\rho,\beta)$.
+Since $(\eta,\alpha) = (\eta',\alpha') \eeq (\rho,\beta) \in \erext \Gamma \Size$,
 we conclude by induction hypothesis.

### 2017-02-21 nbe/icfp17.tex
[ ICFP 17 ] Properties of shape (mostly failing ones).

@@
   \Gamma \der T \leq T' & \mbox{in context $\Gamma$, type $T$ is a subtype of $T'$} \\
+  \Gamma \der T : \Adm \ell & \mbox{in context $\Gamma$ type constructor $T$ is admissible for recursion} \\
   \suty \Gamma \sigma \tau \Delta & \mbox{$\sigma$/$\tau$ is a valid term/type-level substitution for $\Delta$} \\
@@
     & \mbox{in context $\Gamma$, types $T$ and $T'$ are equal} \\
-  \Gamma \der a \leq b : \Size & \defas & \multicolumn 2 {l@{\qquad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
+  \Gamma \der a = b : \Size & \defas & \multicolumn 2 {l@{\qquad}} {\Gamma \der a : \Size \mand a = b}
+    & \mbox{sizes $a$ and $b$ are well-formed and identical} \\
+  \Gamma \der a \leq b : \Size & \defas &
+     \multicolumn 2 {l@{\qquad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
+%     \multicolumn 2 {l@{\qquad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
     & \mbox{sizes $a$ and $b$ are well-formed and comparable} \\
@@
 \\[2ex] % \qquad
-  \ru{\Gamma \der a : \Size \qquad
-      \resurrect\Gamma \der b : \Size
+  \ru{%\Gamma \der a : \Size \qquad
+      \resurrect\Gamma \der a, b : \Size
     }{\Gamma \der \zero{a} : \Nat (b + 1)}
 \qquad
-  \ru{\Gamma \der a : \Size \qquad
+  \ru{\resurrect\Gamma \der a : \Size \qquad
       \Gamma \der t : \Nat b
@@
   \ru{\Gamma \der t : \forallT T \qquad
-      \Gamma \der a : \Size \qquad
-      \resurrect\Gamma \der b : \Size
+      %\Gamma \der a : \Size \qquad
+      \resurrect\Gamma \der a,b : \Size
     }{\Gamma \der t \ann a : T[b]}
@@
   \ru{U \evalsto N \qquad
-      u \evalsto \up{T} n \qquad
-      \RNe k n m
+      \RUp k u m
+      % u \evalsto \up{T} n \qquad \RNe k n m
     }{\R k {\down{U} u} m}
@@
 
+\fbox{$\RUp k t m$}\ \  Read back neutrals under annotation, which is ignored.
+\begin{gather*}
+  \ru{t \evalsto \up T n \qquad
+      \RNe k n m
+    }{\RUp k t m}
+\end{gather*}
+
 \fbox{$\RNat k t v $}
 \begin{gather*}
-  \ru{t \evalsto \up {\Nat \alpha} n \qquad
-      \RNe k n m
+  \ru{% t \evalsto \up {\Nat \alpha} n \qquad
+      %t \evalsto \up {T} n \qquad \RNe k n m
+      \RUp k t m
     }{\RNat k t m}
@@
 \qquad
-  \ru{T \evalsto \up{U}n \qquad
-      \RNe k n m
+  \ru{% T \evalsto \up{U}n \qquad \RNe k n m
+      \RUp k T m
     }{\RTy k T m}
@@
   \ru{T \evalsto \epiT \Size B \qquad
-      \RTy {k+1} B {V_b}
-    }{\RTy k T {\epiT \Size V_b}}
+      \RTy {k+1} B {V}
+    }{\RTy k T {\epiT \Size V}}
 \end{gather*}
@@
 
-\[
-\fbox{$\NEE$} \defas \{ (\up T n, \up{T'}{n'})
-  \msuchthat T \evalsto N \mand T' \evalsto N' \mand n = n' \in \NE \}
+Neutrals in the value world $\D$.
+\[
+\fbox{$t = t' \in \NEE$} \defas
+  t \evalsto \up T n \mand t' \evalsto \up{T'}{n'} \mand
+  n = n' \in \NE
+% \fbox{$\NEE$} \defas \{ (\up T n, \up{T'}{n'})
+%   % \msuchthat T \evalsto N \mand T' \evalsto N' \mand
+%   \mid n = n' \in \NE \}
 \]
@@
 \begin{gather*}
+  \ru{% t \evalsto \up{\Nat \alpha} n \qquad
+      % t' \evalsto \up{\Nat \alpha'} n' \qquad
+      % n = n' \in \NE
+      t = t' \in \NEE
+    }{t = t' \in \NAT(\beta)}
+\qquad
   \ru{t \evalsto \zero \alpha \qquad
@@
     }{t = t' \in \NAT(\beta+1)}
-\\[2ex]
-  \ru{t \evalsto \up{\Nat \alpha} n \qquad
-      t' \evalsto \up{\Nat \alpha'} n' \qquad
-      n = n' \in \NE
-    }{t = t' \in \NAT(\beta)}
 \end{gather*}
@@
 
-\begin{lemma}[PER Subsumption]
-\label{lem:subsump} \hfill
-\begin{enumerate}
-\item If $\alpha \leq \beta$ then $\NAT(\alpha) \subseteq \NAT(\beta)$.
-\item If $\F(\alpha) \leq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
-\item IF $\A' \subseteq \A$ and $\F(u) \subseteq \F'(u)$ for all $u \in \A'$, then $\PIAF \subseteq \PI\,\A'\,\F'$.
-\end{enumerate}
-\end{lemma}
+% \begin{lemma}[PER Subsumption]
+% \label{lem:subsump} \hfill
+% \begin{enumerate}
+% \item If $\alpha \leq \beta$ then $\NAT(\alpha) \subseteq \NAT(\beta)$.
+% \item If $\F(\alpha) \leq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
+% \item IF $\A' \subseteq \A$ and $\F(u) \subseteq \F'(u)$ for all $u \in \A'$, then $\PIAF \subseteq \PI\,\A'\,\F'$.
+% \end{enumerate}
+% \end{lemma}
 
@@
 \begin{array}{l@{\qquad}l}
-  \ru{T \evalsto N \qquad T' \evalsto N'
+  \ru{% T \evalsto \up{\Set[\ell']} n \qquad
+      % T' \evalsto \up{\Set[\ell'']} n' \qquad
+      % n = n' \in \NE
+      T = T' \in \NEE
     }{T = T' \in \SET[\ell]}
@@
   \item Transitivity: If $T_2 = T_3 \in \SET[\ell]$ then $T_1 = T_3 \in \SET[\ell]$.
-  \item Extension is PER: $\EL[\ell](T_1) = \EL[\ell](T_2)$ and ``both'' are PERs.
+  \item Extension: $\EL[\ell](T_1) = \EL[\ell](T_2)$ and ``both'' are PERs.
   \end{enumerate}
@@
 
-\begin{lemma}[Universe subsumption]
-  \label{lem:usubsump}
-  If $\ell \leq \ell'$ then $\SET[\ell] \subseteq \SET[\ell']$.
+\subsection{Subtyping}
+\label{sec:subty}
+
+The semantic types (PERs) admit subsumption:
+
+\begin{lemma}[Subsumption]
+\label{lem:subsump} \hfill
+\begin{enumerate}
+\item \label{it:natsub} If $\alpha \leq \beta$ then $\NAT(\alpha) \subseteq \NAT(\beta)$.
+\item \label{it:allsub} If $\F(\alpha) \leq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
+\item \label{it:pisub}  If $\A' \subseteq \A$ and $\F(u) \subseteq \F'(u)$ for all $u \in \A'$, then $\PIAF \subseteq \PI\,\A'\,\F'$.
+\item \label{it:usub}  If $\ell \leq \ell'$ then $\SET[\ell] \subseteq \SET[\ell']$.
+\end{enumerate}
 \end{lemma}
 \begin{proof}
-  Prove $T = T' \in \SET[\ell']$ by induction on $T = T' \in \SET[\ell]$.
+  Propositions (\ref{it:natsub}--\ref{it:pisub}) are clear.
+  For (\ref{it:usub}), prove $T = T' \in \SET[\ell']$ by induction on $T = T' \in \SET[\ell]$.
   For the base types this is direct, let us look at the function space.
@@
 \end{proof}
+We can define subtyping of type values \fbox{$T \leq T'$} by induction on $T \in \SET[\ell]$ and $T' \in \SET[\ell']$.
+Simultaneously, we need to prove correctness, namely that $T \leq T'$ implies $\EL(T) \subseteq \EL(T')$.  The correctness follows from Lemma~\ref{lem:subsump} and we do not spell it out here.
+\begin{gather*}
+  \ru{T = T' \in \NEE
+    }{T \leq T'}
+\qquad
+  \ru{T \evalsto \Nat \alpha \qquad
+      T' \evalsto \Nat \alpha' \qquad
+      \alpha \leq \alpha'
+    }{T \leq T'}
+\qquad
+  \ru{T \evalsto \Set[\ell_0] \qquad
+      T' \evalsto \Set[\ell_0'] \qquad
+      \ell_0 \leq \ell_0'
+    }{T \leq T'}
+\\[2ex]
+  \ru{T \evalsto \piT A B \qquad
+      T' \evalsto \piT {A'}{B'} \qquad
+      A' \leq A \qquad
+      B[u] \leq B'[u'] \mforall u = u' \in A'
+    }{T \leq T'}
+\\[2ex]
+  \ru{T \evalsto \epiT \Size B \qquad
+      T' \evalsto \epiT {\Size}{B'} \qquad
+      B[\alpha] \leq B'[\alpha] \mforall \alpha \in \SIZE
+    }{T \leq T'}
+\end{gather*}
+\begin{lemma}[Subtyping is preorder]
+\label{lem:preord}
+\bla
+\begin{enumerate}
+\item If\/ $T = T' \in \SET[\ell]$ then $T \leq T'$.
+\item If\/ $T_1 \leq T_2$ and $T_2 \leq T_3$ then $T_1 \leq T_3$.
+\end{enumerate}
+\end{lemma}
+
+% \begin{lemma}[Universe subsumption]
+%   \label{lem:usubsump}
+%   If $\ell \leq \ell'$ then $\SET[\ell] \subseteq \SET[\ell']$.
+% \end{lemma}
+% \begin{proof}
+%   Prove $T = T' \in \SET[\ell']$ by induction on $T = T' \in \SET[\ell]$.
+%   For the base types this is direct, let us look at the function space.
+% \[
+%   \rul{T \evalsto \piT A B \qquad
+%       T' \evalsto \piT{A'}{B'} \qquad
+%       A = A' \in \SET[\ell] \qquad
+%       B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
+%     }{T = T' \in \SET[\ell]}
+% \]
+%   By induction hypothesis, $A = A' \in \SET[\ell']$, and we have $\EL[\ell'](A) = \EL[\ell](A)$ by Lemma~\ref{lem:indep}.
+Assuming $u = u' \in \EL(A)$, we get $B[u] = B'[u'] \in \SET[\ell']$ by induction hypothesis on $B[u] = B'[u'] \in \SET[\ell]$.
+% \end{proof}
 
 \subsection{Type shapes}
+\label{sec:shape}
 
@@
 
-\fbox{$T \shape S$} for $T \in \SET[\ell]$.
+\fbox{$T \shape S$} for $T \in \SET[\ell]$.  We call $T$ the \emph{template} and $S$ one of its possible \emph{shapes}.
 \begin{gather*}
@@
 % The last rule is the crucial one
-\begin{lemma}[Reflexivity for shapes]
-  If $T = T' \in \SET[\ell]$ then $T \shape T'$.
-\end{lemma}
 
@@
 
-\begin{theorem}[Reification and reflection]
+\begin{lemma}[Types are their own shapes]
+  If\/ $T = T' \in \SET[\ell]$ then $T \shape T'$.
+\end{lemma}
+
+\begin{lemma}[Equal types make equally good templates]
+  If\/ $T = T' \in \SET[\ell]$ and $T' \shape S$ then $T \shape S$.
+\end{lemma}
+
+However, templates are not closed under subtyping in either direction because of contravariance.
+
+Further, it is not true that equal types make equally good shapes.  We do not have that
+$T \shape S$ and $S = S' \in \SET[\ell]$ imply $T \shape S'$.  This property fails for function types.
+Given $\piT U T \shape \piT R S$ and $\piT R S = \piT {R'}{S'} \in \SET[\ell]$ we would need to show that $T[u] \shape S'[u']$ for all $u = u' \in U$, but we only have $S[u] = S'[u'] \in \SET[\ell]$ for all $u = u' \in R$, thus the induction does not go through.
+The fact that $U \shape R$ does not give us a handle on their inhabitants, we would need $U \leq R$.
+It is possible to construct an actual counterexample, using
+$\piT {R'} S' =  \funT x {\Nat 0} {F\,x}$ from above and $\piT R S = \funT x {\Nat 0} {G\,x}$
+such that $G$ is defined on all of $\Nat \infty$ but agrees with $F$ only on $x \in \Nat 0$.
+Then $\piT U T = \funT x {\Nat \infty} {G\,x}$ gives the desired counterexample.
+
+% Note that while such examples exist in the semantics, they probably do not exist in the syntax (where no well-typed term is diverging).  However, it is not clear to us how to define
+
+% %% WRONG
+% \begin{lemma}[Subtypes are valid templates]
+%   Let $T \in \SET[\ell]$ and $T' \in \SET[\ell']$.
+%   If\/ $T \leq T'$ and $T' \shape S'$ then $T \shape S'$.
+% \end{lemma}
+% \begin{proof}
+% By induction on the derivations.
+% \begin{caselist}
+
+% \nextcase $T \evalsto \piT A B$ and $T' \evalsto \piT{A'}{B'}$ and $T' \evalsto \piT R S$.
+% \\
+% We have $A \shape R$ by induction hypothesis on $A' \leq A$ and $A' \shape R$.  No!
+% \end{caselist}
+% \end{proof}
+
+Shapes are used to direct $\eta$-expansion when we reflect neutrals into semantic types and reify semantic values to long normal forms.
+The following theorem is the heart of our technical development.
+
+\begin{theorem}[Reflection and reification]
 Let $T \in \SET[\ell]$ and $T \shape S_1$ and $T \shape S_2$.
@@
 \begin{gather*}
-  \ru{B[\alpha] = B[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
+  \ru{B[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
     }{T \in \SET[\ell]}
@@
   \Gamma \models T \leq T' & \defas & \Gamma \models T \mand \Gamma \models T' \mand
-    \EL(T\eta) \subseteq \EL(T'\eta') \mforall \eta = \eta' \eeq \rho \in \Gamma
+    T\eta \leq T'\eta'
+    % \EL(T\eta) \subseteq \EL(T'\eta')
+    \mforall \eta = \eta' \eeq \rho \in \Gamma
 \\[2ex]
@@
 Simultaneously, by induction on the derivation.
-The soundness of subtyping and the conversion rule is due to subsumption (lemmata \ref{lem:subsump} and \ref{lem:usubsump}).
+% The soundness of subtyping and the conversion rule is due to subsumption (lemmata \ref{lem:subsump} and \ref{lem:usubsump}).
 

### 2017-02-21 nbe/icfp17.tex
[ ICFP 17 ] Working on "Into the LR".

@@
 \qquad
+  \rux{\der \Gamma}{\Gamma \der o : \Size}{o \in \NN}
+\qquad
   \ru{\der \Gamma \qquad \Gamma(i) = \erann[\noterased]\Size
@@
   \ru{\erext \Gamma \Size \der t : T \qquad
-      \Gamma \der a : \Size \qquad
-      \resurrect \Gamma \der b : \Size
+      % \Gamma \der a : \Size \qquad
+      \resurrect \Gamma \der a, b : \Size
     }{\Gamma \der (\lambda t)\,\ann a = t[a] : T[b]}
 \\[2ex]
-  \ru{\resurrect\Gamma \der b : \Size \qquad
+  \ru{\resurrect\Gamma \der a, b : \Size \qquad
       \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
-      \Gamma \der a : \Size \qquad
+      % \Gamma \der a : \Size \qquad
       \Gamma \der t_z : T\,\zero{b} \qquad
       \Gamma \der t_s : (n : \Nat b) \to T\,(\suc{b}n)
-    }{\Gamma \der (\zero{a})\,\case[\ell] T {t_z} {t_s} = t_z : T\,\zero{a_z}}
-\\[2ex]
-  \ru{\Gamma \der t : \Nat b \qquad
+    }{\Gamma \der (\zero{a})\,\case[\ell] T {t_z} {t_s} = t_z : T\,\zero{b}}
+\\[2ex]
+  \ru{\resurrect\Gamma \der a : \Size \qquad
+      \Gamma \der t : \Nat b \qquad
       \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
-      \Gamma \der a : \Size \qquad
       \Gamma \der t_z : T\,\zero{b} \qquad
       \Gamma \der t_s : (n : \Nat b) \to T\,(\suc{b}n)
-    }{\Gamma \der (\suc{a} t)\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,(\suc{a} t)}
+    }{\Gamma \der (\suc{a} t)\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,(\suc{b} t)}
 \\[2ex]
   \ru{\Gamma \der c : \Nat b \qquad
-      \Gamma \der a : \Size \qquad
+      \resurrect\Gamma \der a : \Size \qquad
       \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
@@
 \qquad
-  \ru{\Gamma \der t : \forallT T \qquad \erext \Gamma \Size \der a : \Size
+  \ru{\Gamma \der t : \forallT T \qquad
+      \cext {\resurrect\Gamma} \Size \der a : \Size
+      %\erext \Gamma \Size \der a : \Size
     }{\Gamma \der t = \lambda i. t \ann a : \forallT T}
@@
   \ru{\Gamma \der t = t' : \forallT T \qquad
-      \Gamma \der a, a' : \Size \qquad
-      \resurrect\Gamma \der b : \Size
+      % \Gamma \der a, a' : \Size \qquad
+      \resurrect\Gamma \der a,a',b : \Size
     }{\Gamma \der t \ann a = t' \ann{a'} : T[b]}
 \\[2ex] % \qquad
-  \ru{\Gamma \der a, a' : \Size \qquad
-      \resurrect\Gamma \der b : \Size
+  \ru{%\Gamma \der a, a' : \Size \qquad
+      \resurrect\Gamma \der a, a', b : \Size
     }{\Gamma \der \zero a = \zero {a'}: \Nat (b + 1)}
 \qquad
-  \ru{\Gamma \der a, a' : \Size \qquad
+  \ru{\resurrect \Gamma \der a, a' : \Size \qquad
       \Gamma \der t = t' : \Nat b
@@
 (implies $\der \Gamma$ and $\der \Delta$
+ %% and $\Gamma \der \tau : \resurrect \Delta$  %NO
  [and $\resurrect\Gamma \der \tau : \Delta$]
@@
       \Gamma \der a : \Size
-    }{\Gamma \der \sext \tau a : \eext \Delta \Size}
-\end{gather*}
-
-Weakenings \fbox{$\xi : \Gamma \leq \Delta$}  (implies $\der \Gamma$ and $\der \Delta$ and $\Gamma \der \xi : \Delta$).
-\begin{gather*}
-  \ru{\der \Gamma
-    }{\sempty : \Gamma \leq \cempty}
-\qquad
-  \ru{\xi : \Gamma \leq \Delta \qquad
-      \resurrect\Gamma \der T
-    }{\xi \slift : \cext \Gamma T \leq \Delta}
-\qquad
-  \ru{\xi : \Gamma \leq \Delta \qquad
-      \resurrect \Delta \der U \qquad
-      \resurrect\Gamma \der T \leq U \xi
-    }{(\xi \slift, \ind 0) : \cext \Gamma T \leq \cext \Delta U}
-\end{gather*}
+    }{\Gamma \der \sext \tau a : \cext \Delta \Size}
+\qquad
+  \ru{\Gamma \der \tau : \Delta \qquad
+      \resurrect\Gamma \der a : \Size
+    }{\Gamma \der \sext \tau a : \erext \Delta \Size}
+\end{gather*}
+
+Liftings: $\fbox{$\xi : \Gamma \leq \Delta$}
+  \defas \Gamma \der \xi : \Delta \mand \xi = \slift^k_m \mforsome k,m$
+%  \defas \xi = \slift^k_m \mwith \Gamma \der \xi : \Delta$
+(implies $\der \Gamma$ and $\der \Delta$).
+% (implies $\der \Gamma$ and $\der \Delta$ and $\Gamma \der \xi : \Delta$).
+% \begin{gather*}
+%   \ru{\der \Gamma
+%     }{\sempty : \Gamma \leq \cempty}
+% \qquad
+%   \ru{\xi : \Gamma \leq \Delta \qquad
+%       \resurrect\Gamma \der T
+%     }{\xi \slift : \cext \Gamma T \leq \Delta}
+% \qquad
+%   \ru{\xi : \Gamma \leq \Delta \qquad
+%       \resurrect \Delta \der U \qquad
+%       \resurrect\Gamma \der T \leq U \xi
+%     }{(\xi \slift, \ind 0) : \cext \Gamma T \leq \cext \Delta U}
+% \end{gather*}
 
@@
   \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
-      \Gamma \der a, a', b : T \tau
+      \resurrect\Gamma \der a, a', b : T \tau
     }{\sueq \Gamma {\sext \sigma a } {\sext {\sigma'} {a'}} {\sext \tau b} {\erext \Delta \Size}}
@@
   then $\Gamma \der J : \Delta$.
-\item If\/ $\Gamma \der \sigma : \Delta$ then $\Gamma \der \sigma : \resurrect\Delta$.
-% then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
+\item If\/ $\Gamma \der \sigma : \Delta$ % then $\Gamma \der \sigma : \resurrect\Delta$.
+  then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
 \end{enumerate}
@@
 \end{lemma}
+\begin{corollary}[Partial resurrection for substitution equality]
+  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then
+  $\sueq {\resurrect \Gamma} \tau \tau \tau {\resurrect \Delta}$.
+\end{corollary}
 
@@
 \begin{theorem}[Reflection and reification]
+\label{thm:rere}
 Let $T \in \SET[\ell]$ and $T \shape S_1$ and $T \shape S_2$.
@@
 \begin{corollary}
-  If\/ $T \in \SET[\ell]$ and $t = t' \in T$ then $\down T t = \down T {t'} \in \NF$.
+  \label{cor:rere}
+  Let\/ $T \in \SET[\ell]$.
+  \begin{enumerate}
+  \item If $n = n' \in \NE$ then $\up T n = \up T n' \in T$.
+  \item If $t = t' \in T$ then $\down T t = \down T {t'} \in \NF$.
+  \end{enumerate}
 \end{corollary}
@@
 \begin{theorem}[Fundamental theorem]
+  \label{thm:fund}
   \bla
@@
 
+We write $\var\Gamma$ for $\var{|\Gamma|}$.
+We define the identity environment $\rho_\Gamma$ by induction on $\Gamma$.
+We set $\rho_{\cempty} = \sempty$ and
+\[
+\begin{array}{lll}
+  \rho_{\cext \Gamma T} & = & (\rho_\Gamma, \up{T\rho_\Gamma} \var\Gamma) \\
+  \rho_{\eext \Gamma \Size} & = & (\rho_\Gamma, \var\Gamma) \\
+\end{array}
+\]
+\begin{lemma}[Identity environment]
+  If $\der \Gamma$ then
+  $\rho_\Gamma \in \Gamma$.
+  % $\rho_\Gamma = \rho_\Gamma \eeq \rho_\Gamma \in \Gamma$
+\end{lemma}
+\begin{proof}
+By induction on $\der \Gamma$.
+\begin{caselist}
+
+\nextcase
+\[
+  \ru{\der \Gamma \qquad \resurrect\Gamma \der T
+    }{\der \cext\Gamma T}
+\]
+By induction hypothesis $\rho_\Gamma \in \Gamma$.
+By resurrection (Lemma~\ref{lem:resenv}) $\rho_\Gamma \in \resurrect\Gamma$.
+By the fundamental theorem (Thm.~\ref{thm:fund}) we have $A := T\rho_\Gamma \in \SET[\ell]$ for some $\ell$.
+By reflection (Cor.~\ref{cor:rere}) it follows that $\up A \var\Gamma \in A$, thus
+$(\rho_\Gamma, \up A \var\Gamma) \in \cext \Gamma T$.
+
+\nextcase
+\[
+  \ru{\der \Gamma
+    }{\der \eext\Gamma \Size}
+\]
+By induction hypothesis $\rho_\Gamma \in \Gamma$.  Since $\var\Gamma \in \SIZE$ we conclude
+$(\rho_\Gamma, \var\Gamma) \in \eext\Gamma \Size)$.
+\end{caselist}
+\end{proof}
+
+We can now define the normalization relation
+$\nbe \Gamma T t \evalsto v \iff \R {|\Gamma|} {\down{T \rho_\Gamma} (t \rho_\Gamma)} v$.
+
+\begin{theorem}[Completeness of NbE]
+  If\/ $\Gamma \der t = t' : T$ then there are normal forms $v \approx v'$ such that
+  $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v'$.
+\end{theorem}
+\begin{proof}
+  By the fundamental theorem, $T \rho_\Gamma \in \SET[\ell]$ for some $\ell$ and
+  $t \rhog = t' \rhog \in T\rhog$.  By reification (Cor.~\ref{cor:rere}) we have
+  $\down{T\rhog} (t\rhog) = \down{T\rhog}(t'\rhog) \in \NF$ which implies the theorem
+  by read back with $k = |\Gamma|$.
+\end{proof}
+
 \section{Soundness  of NbE}
@@
   \fbox{$\LRSize \Gamma a \alpha$} \defas
-  \forall \sigma : \Gamma' \leq \Gamma.\ \
-  \RSize {\Gamma'} \alpha {a\sigma}
+  \forall \xi : \Gamma' \leq \Gamma.\ \
+  \RSize {\Gamma'} \alpha {a\xi}
   % \exists a'.\
   % \RSize {\Gamma'} \alpha a' \mand
-  % \Gamma' \der a\sigma = a' : \Size
-\]
-
-\fbox{$\LRT \Gamma {T'} {A'} \ell$} and \fbox{$\LR \Gamma t {T'} f {A'}$} defined by induction on ${A'} \in \SET[\ell]$.
+  % \Gamma' \der a\xi = a' : \Size
+\]
+
+We simultaneously define
+\fbox{$\LRT \Gamma {T'} {A'} \ell$} for $\Gamma \der T' : \Set[\ell]$
+and \fbox{$\LR \Gamma t {T'} f {A'}$} for $\Gamma \der t : T'$ and $f \in A'$
+defined by induction on ${A'} \in \SET[\ell]$.
 
@@
 
+\vspace{2ex}
 \nextcase $A' \evalsto N$ neutral.
@@
   $\LRT{\Gamma}{T'}{A'}\ell$ iff
-  $\forall \sigma : \Gamma' \leq \Gamma.\ \ \exists M.\ \ \RTy{\Gamma'}{A'}M \mand \Gamma' \der T'\sigma = M : \Set[\ell]$.
+  $\forall \xi : \Gamma' \leq \Gamma.\ \ \exists M.\ \ \RTy{\Gamma'}{A'}M \mand \Gamma' \der T'\xi = M : \Set[\ell]$.
 
@@
   $\LR{\Gamma}{t}{T'}{f}{A'}$ iff
-  $\forall \sigma : \Gamma' \leq \Gamma.\ \
+  $\forall \xi : \Gamma' \leq \Gamma.\ \
    \exists m. \ \
    \R {\Gamma'} {\down {A'} f} m \mand
-   \Gamma' \der t\sigma = m : T'\sigma$.
-
+   \Gamma' \der t\xi = m : T'\xi$.
+
+\vspace{2ex}
 \nextcase $A' \evalsto \Nat \alpha$.
@@
   $\LRSize {\resurrect\Gamma} a \alpha$ and \\
-  $\forall \sigma : \Gamma' \leq \Gamma.\ \
+  $\forall \xi : \Gamma' \leq \Gamma.\ \
    \exists v. \ \
    \R {\Gamma'} {\down {A'} f} v \mand
-   \Gamma' \der t\sigma = v : \Nat\,a\sigma$.
-
+   \Gamma' \der t\xi = v : \Nat\,a\xi$.
+
+\vspace{2ex}
 \nextcase $A' \evalsto \Set[\ell']$.
@@
 
+\vspace{2ex}
 \nextcase $A' \evalsto \piT A B$.
@@
 \Gamma \der T' = \piT U T : \Set[\ell] \mforsome U, T \mand \LRT \Gamma U A \ell \mand \\
-\forall \sigma : \Gamma' \leq \Gamma.\ \
-\LR {\Gamma'} u {U\sigma} a A \implies \LRT {\Gamma'} {T(\sigma,u)} {B[a]} \ell
+\forall \xi : \Gamma' \leq \Gamma.\ \
+\LR {\Gamma'} u {U\xi} a A \implies \LRT {\Gamma'} {T(\xi,u)} {B[a]} \ell
 . \\
@@
 \resurrect\Gamma \der T' = \piT U T : \Set[\ell] \mforsome U, T \mand \LRT {\resurrect\Gamma} U A \ell \mand \\
-\forall \sigma : \Gamma' \leq \Gamma.\ \
-\LR {\Gamma'} u {U\sigma} a A \implies \LR {\Gamma'} {t\sigma\,u} {T(\sigma,u)} {f\,a} {B[a]}
+\forall \xi : \Gamma' \leq \Gamma.\ \
+\LR {\Gamma'} u {U\xi} a A \implies \LR {\Gamma'} {t\xi\,u} {T(\xi,u)} {f\,a} {B[a]}
 .
@@
 
+\vspace{2ex}
 \nextcase $A' \evalsto \piT \Size B$.
@@
 \Gamma \der T' = \piT \Size T : \Set[\ell] \mforsome T\\
-\forall \sigma : \Gamma' \leq \Gamma.\ \
+\forall \xi : \Gamma' \leq \Gamma.\ \
 \LRSize {\Gamma'} a \alpha
-\implies \LRT {\Gamma'} {T(\sigma,a)} {B[\alpha]} \ell
+\implies \LRT {\Gamma'} {T(\xi,a)} {B[\alpha]} \ell
 .
@@
 \resurrect\Gamma \der T' = \piT \Size T : \Set[\ell] \mforsome T\\
-\forall \sigma : \Gamma' \leq \Gamma.\ \
+\forall \xi : \Gamma' \leq \Gamma.\ \
 \LRSize {\Gamma'} a \alpha
-\implies \LR {\Gamma'} {t\sigma\,a} {T(\sigma,a)} {f\,\alpha} {B[\alpha]}
+\implies \LR {\Gamma'} {t\xi\,a} {T(\xi,a)} {f\,\alpha} {B[\alpha]}
 .
@@
 
+\vspace{2ex}
 \nextcase $A' \evalsto \forallT B$.
@@
 \Gamma \der T' = \forallT T : \Set[\ell] \mforsome T\\
-\forall \sigma : \Gamma' \leq \Gamma,\
+\forall \xi : \Gamma' \leq \Gamma,\
 {\Gamma'} \der b : \Size,\
@@
 \LRSize {\Gamma'} b \beta
-\implies \LRT {\Gamma'} {T(\sigma,b)} {B[\beta]} \ell
+\implies \LRT {\Gamma'} {T(\xi,b)} {B[\beta]} \ell
 .
@@
 \resurrect\Gamma \der T' = \forallT T : \Set[\ell] \mforsome T\\
-\forall \sigma : \Gamma' \leq \Gamma,\
-\Gamma' \der a : \Size,\
-\resurrect{\Gamma'} \der b : \Size,\
+\forall \xi : \Gamma' \leq \Gamma,\
+%\Gamma' \der a : \Size,\
+\resurrect{\Gamma'} \der a,b : \Size,\
 \alpha,\beta \in \SIZE. \ \\ \qquad
 \LRSize {\resurrect{\Gamma'}} b \beta
-\implies \LR {\Gamma'} {t\sigma\,\ann a} {T(\sigma,b)} {f\,\ann\alpha} {B[\beta]}
+\implies \LR {\Gamma'} {t\xi\,\ann a} {T(\xi,b)} {f\,\ann\alpha} {B[\beta]}
 .
@@
 
+\begin{lemma}[Weakening]
+  Let $\xi : \Gamma' \leq \Gamma$.
+  \begin{enumerate}
+  \item If\/ $\LRSize \Gamma a \alpha$ then $\LRSize{\Gamma'}{a\xi}\alpha$.
+  \item If\/ $\LRT {\Gamma} T A \ell$ then $\LRT{\Gamma'}{T\xi}A \ell$.
+  \item If\/ $\LR \Gamma t T f A$ then $\LR{\Gamma'}{t\xi}{T\xi} f A$.
+  \end{enumerate}
+\end{lemma}
+
 \begin{theorem}[Into and out of the logical relation]
+  \label{thm:into}
   \bla
-  Let $A \in \Set[\ell]$.
-  \begin{enumerate}
-
-  \item $\LR \Gamma t T {\up A n} A$ if for all $\sigma : \Gamma' \leq \Gamma$ there is an $m$ such that
-     $\RNe{\Gamma'}n m$ and $\Gamma' \der t\sigma = m : T$.
-
-  \item If\/ $\LRT{\Gamma} T A \ell$ then for all $\sigma : \Gamma' \leq \Gamma$ there is some normal form $V$ such that
-  $\Rty{\Gamma'} A V$ and $\Gamma' \der T\sigma = V : \Set[\ell]$.
-
-  \item If\/ $\LR \Gamma t T f A$ then for all $\sigma : \Gamma' \leq \Gamma$ there is some normal form $v$ such that
-  $\R{\Gamma'} {\down A f} v$ and $\Gamma' \der t\sigma = v : T\sigma$.
+  Let $A \in \SET[\ell]$.
+  \begin{enumerate}
+
+  \item \label{it:into}
+     $\LR \Gamma t T {\up A n} A$ if for all $\xi : \Gamma' \leq \Gamma$ there is an $m$ such that
+     $\RNe{\Gamma'}n m$ and\/ $\Gamma' \der t\xi = m : T$.
+
+  \item \label{it:outty}
+     If\/ $\LRT{\Gamma} T A \ell$ then for all $\xi : \Gamma' \leq \Gamma$
+     there is some normal form $V$ such that
+     $\Rty{\Gamma'} A V$ and $\Gamma' \der T\xi = V : \Set[\ell]$.
+
+  \item \label{it:out}
+    If\/ $\LR \Gamma t T f A$ then for all $\xi : \Gamma' \leq \Gamma$
+    there is some normal form $v$ such that
+    $\R{\Gamma'} {\down A f} v$ and $\Gamma' \der t\xi = v : T\xi$.
 
@@
 \end{theorem}
+
+\begin{proof}
+We prove the propositions for $A'$ and $T'$ (instead of $A$ and $T$)
+simultaneously by induction on $A' \in \SET[\ell]$.
+
+\begin{caselist}
+
+\nextcase $A' \evalsto \piT A B$ and $\resurrect\Gamma \der T' = \piT U T$.
+\begin{enumerate}
+\item To demonstrate $\LR \Gamma t {T'} {\up {A'} n} {A'}$ we assume
+$\xi : \Gamma' \leq \Gamma$ and $\LR{\Gamma'}{u}{U\xi} a A$ and show
+$\LR{\Gamma'} {(t\xi\,u)} {T(\xi,u)} {\up{B[a]}(n\,\down A a)} {B[a]}$.
+This follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$ if
+we can prove that for all $\xi' : \Gamma'' \leq \Gamma'$ there is a neutral normal form $m$
+such that $\RNe{\Gamma''} {(n\,\down A a)} m$ and $\Gamma'' \der (t\xi\,u)\xi' = m : T(\xi,u)\xi'$.
+Note that the last goal can be rewritten to $\Gamma'' \der t\xi\xi'\,(u\xi') = m : T(\xi\xi',u\xi')$.
+We know that $\RNe{\Gamma''} n {m_0}$ with $\Gamma'' \der t\xi\xi' = m_0 : T'\xi\xi'$.
+Thus, it is sufficient to show that there is some normal form $v$ such that
+$\R{\Gamma''} {\down A a} v$ and $\Gamma'' \der u\xi' = v : U\xi\xi'$.
+However, this follows by induction hypothesis (\ref{it:out}) on
+assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
+\end{enumerate}
+\end{caselist}
+\end{proof}
+
+\begin{corollary}[Fresh variable]
+  \label{cor:fresh}
+  If $\der \cext \Gamma T$ and $A \in \SET[\ell]$
+  then $\LR {\cext \Gamma T} {\ind 0} {(T \slift)} {(\up A \var\Gamma)} A$.
+\end{corollary}
+\begin{proof}
+  Assume $\xi : \Gamma' \leq \cext \Gamma T$.
+  Note that for the length
+  $i = |\Gamma'| - |\cext \Gamma T|$ of the context extension
+  we have $\ind i = \ind 0 \xi$, thus
+  $\Gamma' \der \ind 0 \xi = \ind i : T \slift \xi$.
+  Since $\RNe{\Gamma'} {\var\Gamma} {\ind i}$,
+  the goal follows now from Thm.~\ref{thm:into} part (\ref{it:into}).
+\end{proof}
 
@@
   \ru{\LRSub \Gamma \sigma \tau \Delta \eta \rho \qquad
-      \Gamma \der a, b : \Size \qquad
+      \resurrect \Gamma \der a, b : \Size \qquad
       \alpha,\beta \in \SIZE \qquad
-      \LRSize \Gamma b \beta
+      \LRSize {\resurrect\Gamma} b \beta
     }{\LRSub \Gamma {(\sigma,a)} {(\tau,b)} {\erext \Delta \Size} {(\eta,\alpha)} {(\rho,\beta)}}
 \end{gather*}
+We write \fbox{$\LRS \Gamma \tau \Delta \rho$} for $\LRSub \Gamma \tau \tau \Delta \rho \rho$.
 
@@
   \label{lem:wklrsub}
-  Let $\LRSub \Gamma \sigma \tau \Delta \eta \rho$.
-  \begin{enumerate}
+  Let $\LRSub \Gamma \sigma \tau \Delta \eta \rho$. Then:
+  \begin{enumerate}
+  \item Welltypedness: $\Gamma \der \sigma : \Delta$
+   and $\Gamma \der \sigma = \sigma \eeq \tau : \Delta$
+   [which in turn implies $\Gamma \der \tau : \Delta$].
   \item Weakening:
-  If $\sigma' : \Gamma' \leq \Gamma$ then
-  $\LRSub{\Gamma'}{\sigma\sigma'}{\tau\sigma'}{\Delta}\eta\rho$.
+  If $\xi : \Gamma' \leq \Gamma$ then
+  $\LRSub{\Gamma'}{\sigma\xi}{\tau\xi}{\Delta}\eta\rho$.
   \item Resurrection:
-  Let $\LRSub {\Gamma} \tau \tau {\resurrect\Delta} \rho \rho$.
+  $\LRS {\resurrect\Gamma} \tau {\resurrect\Delta} \rho$.
+  % $\LRSub {\resurrect\Gamma} \tau \tau {\resurrect\Delta} \rho \rho$.
+  \item \label{it:sizesub} Size substitution:
+  If $\resurrect\Delta \der a : \Size$ then $a\eta \in \SIZE$.
   \end{enumerate}
 \end{lemma}
+\begin{proof}
+  For part (\ref{it:sizesub}), the only interesting case $a = \ind i + o$ can be proved by observing that $\eta(i) \in \SIZE$.
+\end{proof}
 
@@
 \\[1ex]
-  \Gamma \valid \sigma : \Delta & \defas &
-    \LRSub {\Gamma'} {\sigma \sigma'} {\sigma \tau'} \Delta {\sigma \eta'} {\sigma \rho'}
-    \mforall \LRSub {\Gamma'} {\sigma'} {\tau'} \Gamma {\eta'} {\rho'}
+  \Gamma \valid \sigma_0 : \Delta & \defas &
+    \LRSub {\Gamma'} {\sigma_0 \sigma} {\sigma_0 \tau} \Delta {\sigma_0 \eta} {\sigma_0 \rho}
+    \mforall \LRSub {\Gamma'} {\sigma} {\tau} \Gamma {\eta} {\rho}
 \end{array}
@@
 $\LR \Delta {(\lambda t)\sigma} {(\forall T)\tau} {(\lambda t)\eta} {(\forall T)\rho}$.
-To this end, assume a weakening $\sigma' : \Delta' \leq \Delta$ and
-size expressions $\Delta' \der a : \Size$ and
-$\resurrect{\Delta'} \der b : \Size$ and size values $\alpha,\beta$ with
-$\LRSize {\Delta'} b \beta$ and show
-$\LR {\Delta'} {t(\sigma\sigma',a)} {T(\tau\sigma',b)} {t(\eta,\alpha)} {T(\rho,\beta)}$.
+To this end, assume a weakening $\xi : \Delta' \leq \Delta$ and
+size expressions % $\Delta' \der a : \Size$ and
+$\resurrect{\Delta'} \der a, b : \Size$ and size values $\alpha,\beta$ with
+$\LRSize {\resurrect{\Delta'}} b \beta$ and show
+$\LR {\Delta'} {(\lambda t)\sigma\xi \ann a} {T(\tau \slift, \ind 0)(\xi,b)} {(\lambda t)\eta \ann \alpha} {T(\rho\slift, \ind 0)[\beta]}$.
+It suffices to show (weak head reduction, substitution laws) that
+$\LR {\Delta'} {t(\sigma\xi,a)} {T(\tau\xi,b)} {t(\eta,\alpha)} {T(\rho,\beta)}$.
 This follows from the induction hypothesis, since
-$\LRSub {\Delta'} {(\sigma\sigma',a)} {(\tau\sigma',b)} {\erext \Gamma \Size} {(\eta,\alpha)} {(\rho,\beta)}$
+$\LRSub {\Delta'} {(\sigma\xi,a)} {(\tau\xi,b)} {\erext \Gamma \Size} {(\eta,\alpha)} {(\rho,\beta)}$
 by Lemma~\ref{lem:wklrsub} and substitution extension.
@@
 \[
-  \ru{\Gamma \der t : \forall T \qquad \Gamma \der a : \Size \qquad \resurrect \Gamma \der b : \Size
+  \ru{\Gamma \der t : \forall T \qquad % \Gamma \der a : \Size \qquad
+      \resurrect \Gamma \der a,b : \Size
     }{\Gamma \der t \ann a : T[b]}
@@
 \begin{enumerate}
-\item $\Delta \der a\sigma : \Size$
+\item $\resurrect\Delta \der a\sigma : \Size$: follows from $\resurrect\Gamma \der a : \Size$
+  via $\Delta \der \sigma : \Gamma$ and $\resurrect\Delta \der \sigma : \resurrect\Gamma$
+  and substitution.
 %\item $\resurrect\Delta \der b\tau : \Size$.
- and $a\eta \in \SIZE$: follows by induction hypothesis on $\Gamma \der a : \Size$.
+\item $a\eta \in \SIZE$: follows from $\resurrect\Gamma \der a : \Size$
+  by Lemma~\ref{lem:wklrsub}, part (\ref{it:sizesub}).
 %\item $b\rho \in \SIZE$.
@@
 \end{enumerate}
+
+\end{caselist}
+\end{proof}
+
+\begin{lemma}[Identity environment]
+  If $\der \Gamma$ then
+  $\LRS \Gamma \sid \Gamma \rhog$.
+\end{lemma}
+\begin{proof}
+  By induction on $\der \Gamma$.
+\begin{caselist}
+
+\nextcase
+\[
+  \ru{\der \Gamma \qquad \resurrect\Gamma \der T
+    }{\der \cext \Gamma T}
+\]
+Let $f = \up{T\rhog} \var\Gamma$.
+We have to show $\LRS{\cext \Gamma T} \sid {\cext\Gamma T} {(\rhog, f)}$.
+Note that $\sid = (\slift,\ind0)$, thus it remains to show that
+$\LRS {\cext \Gamma T} {\slift} \Gamma \rhog$ and
+$\LR {\cext \Gamma T} {\ind0} {T\slift} f {T\rhog}$.
+The first goal follows by weakening from the induction hypothesis
+$\LRS \Gamma \sid \Gamma \rhog$.
+%Further, by resurrection we get
+%$\LRS {\cext \Gamma T} {\slift} \Gamma \rhog$, and by resurrection
+%$\LRS {\cext{\resurrect\Gamma}T}{\slift}{\resurrect\Gamma}\rhog$.
+Since $\rhog \in \resurrect\Gamma$, we get $T \rhog \in \SET[\ell]$ for some $\ell$
+by the first fundamental theorem (Thm.~\ref{thm:fund}).
+Thus, the second goal follows by Cor.~\ref{cor:fresh}.
+% Further note that for all $\Gamma' = \cext {\cext \Gamma T} {\Delta}$ and
+% $\xi : \Gamma' \leq \cext \Gamma T$ we have that
+% $\RNe{\Gamma'} {\var\Gamma} {\ind{|\Gamma'|\monus|\Gamma|}} = {\ind{|\Delta|}}$ and
+% $\Gamma' \der \ind 0 \xi = \ind{|\Delta|} : T\xi$.
+% This allows us to apply Thm.~\ref{thm:into} to get $\ind0$ into the logical relation as
+% $\LR {\cext \Gamma T} {\ind0} T f {T\rhog}$.
+
+\nextcase
 

### 2017-02-23 prototyping/size-irrelevance/TypeChecker.hs
[ Sit ] Simple admissibility check (monotonicity).

@@ -490,7 +490,7 @@ resurrect = \case
       ShapeIrr -> Relevant
       r -> r
 
--- * Subtyping
+-- * Subtyping and type equality
 
 subType :: Val -> Val -> Check ()
 subType ti tc = do
@@ -509,9 +509,6 @@ subType ti tc = do
         subType b1 b2
     _ -> equalType ti tc
 
--- subDom :: Dom Val -> Dom Val -> Check ()
--- subDom (Dom r1 t1) (Dom r2 t2) = do
-
 equalType :: Val -> Val -> Check ()
 equalType v v' = do
   t  <- reifyType v
@@ -519,11 +516,34 @@ equalType v v' = do
   unless (t == t') $
     throwError $ "Inferred type " ++ show t ++ " is not equal to expected type " ++ show t'
 
+-- * Admissibility check for the type of @fix@.
 
--- * Continuity check
+-- | A simple positivity check.
+--
+--   For the type constructor T of fix we check that
+--   @
+--     i : ..Size, x : Nat i |- T i x <= T oo x
+--   @
+--   This does not introduce a new concept an is sound, but excludes
+--   @
+--     min : forall .i -> Nat i -> Nat i -> Nat i
+--   @
 
 admissible :: Val -> Check ()
 admissible v = do
+  k <- length <$> asks _envCxt
+  addContext ("i", VSize) $ do
+    va <- lastVal
+    addContext ("x", VNat $ va) $ do
+      u  <- lastVal
+      t1  <- applyArgs v [ Arg ShapeIrr va, Arg Relevant u]
+      t2  <- applyArgs v [ Arg ShapeIrr VInfty, Arg Relevant u]
+      subType t1 t2
+
+-- | Semi-continuity check (to be completed)
+
+admissibleSemi :: Val -> Check ()
+admissibleSemi v = do
   k <- length <$> asks _envCxt
   addContext ("i", VSize) $ do
     va <- lastVal

### 2017-02-23 nbe/icfp17.tex
[ ICFP 17 ] Erase size annotations during read back.

@@
 \qquad
-  \ru{\RSize k \alpha b
-    }{\RElim k {\ann \alpha} {\ann b}}
+  \ru{
+    }{\RElim k {\ann \alpha} {\ann \infty}}
+  % \ru{\RSize k \alpha b
+  %   }{\RElim k {\ann \alpha} {\ann b}}
 \\[2ex]
@@
 \qquad
-  \ru{\RTy k T V \qquad
+  \ru{\R k D V \qquad
       \R k d v
-    }{\RElim k {(\fix[\ell] T d)} {\fix[\ell] V v}}
+    }{\RElim k {(\fix[\ell] D d)} {\fix[\ell] V v}}
 \end{gather*}
@@
 
-We write \fbox{$t \approx t'$} iff $t$ and $t'$ are syntactically equal if we ignore sizes $\ann a$ in irrelevant positions.
+% We write \fbox{$t \approx t'$} iff $t$ and $t'$ are syntactically equal if we ignore sizes $\ann a$ in irrelevant positions.
 For example, $\suc a \tzero \approx \suc 0 \tzero$, but $\Nat a \not\approx \Nat 0$.
@@
 %     \mforall k \mboth
-%     \RNe k n m \mand \RNe k {n'} {m'} \mand m \approx m'
+%     \RNe k n m \mand \RNe k {n'} {m'} \mand m \nfeq m'
 %     .\\
@@
     & \defas &
-    \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m \approx m' &\mforall k \\
+    \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m \nfeq m' &\mforall k \\
   \fbox{$d = d' \in \NF$}
     & \defas &
-    \R k d v &\mand& \R k {d'} {v'} &\mand& v \approx v' &\mforall k \\
+    \R k d v &\mand& \R k {d'} {v'} &\mand& v \nfeq v' &\mforall k \\
+  \fbox{$e = e' \in \ELIM$}
+    & \defas &
+    \RElim k e e_v &\mand& \R k {e'} {e_v'} &\mand& e_v \nfeq e_v' &\mforall k \\
+  \fbox{$A = A' \in \TY$}
+    & \defas &
+    \RTy k A V &\mand& \RTy k {A'} {V'} &\mand& V \nfeq V' &\mforall k \\
 \end{array}
-  % \ru{\RNe k t m \mand \RNe k {t'} {m'} \mand m \approx m' \mforall k
+  % \ru{\RNe k t m \mand \RNe k {t'} {m'} \mand m \nfeq m' \mforall k
   %   }{t = t' \in \NE}
-  % \ru{t \evalsto \up N n \qquad t' \evalsto \up{N'}{n'} \qquad n \approx n'
+  % \ru{t \evalsto \up N n \qquad t' \evalsto \up{N'}{n'} \qquad n \nfeq n'
   %   }{t = t' \in \NE}
@@
 
-\begin{lemma}[Closure properties of $\NE$ and $\NF$]
+\begin{lemma}[Closure properties of $\NE$]
+  \label{lem:closne}
   \bla
@@
   \item $\lev k = \lev k \in \NE$.
-  \item If $n = n' \in \NE$ and $d = d' \in \NF$ then $n\,d = n'\,d' \in \NE$.
-%  \item If $d = d' \in \NF$
+  \item If $n = n' \in \NE$ and $e = e' \in \ELIM$ then $n\,e = n'\,e' \in \NE$.
+  \end{enumerate}
+\end{lemma}
+
+\begin{lemma}[Closure properties of $\ELIM$]
+  \label{lem:closelim}
+  \bla
+  \begin{enumerate}
+  \item If $d = d' \in \NF$ then $d = d' \in \ELIM$.
+  \item If $\alpha \in \SIZE$ then $\alpha = \alpha \in \ELIM$.
+  \item $\ann \alpha = \ann{\alpha'} \in \ELIM$.
+  \item If $A = A' \in \TY$ and $d_z = d_z' \in \NF$ and $d_s = d_s' \in \NF$
+        then $\case[\ell] A {d_z}{d_s} = \case[\ell]{A'}{d_z'}{d_s'} \in \ELIM$.
+  \item If $D = D' \in \NF$ and $d = d' \in \NF$ then $\fix[\ell] D d = \fix[\ell]{D'}{d'} \in \ELIM$.
   \end{enumerate}
@@
   By induction hypothesis, $d_1 = d_2 \in \NF$.
-  Thus, $n_1\,d_1 = n_2\,d_2 \in \NE$, and
+  Thus, $n_1\,d_1 = n_2\,d_2 \in \NE$ by Lemma~\ref{lem:closne}, and
   by induction hypothesis,  $\up{B_1[u_1]}(n_1\,d_1) = \up{B_2[u_2]}(n_2\,d_2) \in B[u_1]$.
@@
 \item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $\alpha_1,\alpha_2 \in \SIZE$.
-  Since $n_1 \ann{\alpha_1} = n_2 \ann{\alpha_2} \in \NE$, we obtain
+  Since $n_1 \ann{\alpha_1} = n_2 \ann{\alpha_2} \in \NE$ by Lemma~\ref{lem:closne}, we obtain
   $\up{B_1[\alpha_1]}{(n_1 \ann{\alpha_1})} = \up{B_2[\alpha_2]}{(n_2 \ann{\alpha_2})} \in B[\alpha_1]$
@@
 \begin{theorem}[Completeness of NbE]
-  If\/ $\Gamma \der t = t' : T$ then there are normal forms $v \approx v'$ such that
+  If\/ $\Gamma \der t = t' : T$ then there are normal forms $v \nfeq v'$ such that
   $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v'$.

### 2017-02-23 nbe/icfp17.tex
[ ICFP 17 ] Pi-injectivity, subject reduction, algorithmic subtyping.

@@
 \begin{enumerate}
-\item TODO
+\item If $\Gamma \der \lambda t : T'$ then either $\Gamma \der \piT U T \leq T'$ and
+  $\cext \Gamma U \der t : T$ for some $U,T$ or $\Gamma \der \epiT\Size T \leq T'$ and
+  $\eext \Gamma \Size \der t : T$ for some $T$.
 \end{enumerate}
@@
 We can now define the normalization relation
-$\nbe \Gamma T t \evalsto v \iff \R {|\Gamma|} {\down{T \rho_\Gamma} (t \rho_\Gamma)} v$.
+$\fbox{$\nbe \Gamma T t \evalsto v$} \iff \R {\Gamma} {\down{T \rho_\Gamma} (t \rho_\Gamma)} v$.
+Whenever $\nbe \Gamma T t \evalsto v$, we may write $\nbe \Gamma T t$ for $v$.
 
@@
 \begin{lemma}[Identity environment]
+  \label{lem:idenv}
   If $\der \Gamma$ then
@@
 
+\begin{corollary}[Soundness of NbE]
+  \label{cor:soundnbe} \bla
+  \begin{enumerate}
+  \item \label{it:sound}
+  If\/ $\Gamma \der t : T$ then $\Gamma \der t = \nbe \Gamma T t : T$.
+  \item
+  If\/ $\Gamma \der t,t' : T$ and $\nbe \Gamma T t \equiv \nbe \Gamma T {t'}$ then
+  $\Gamma \der t = t' : T$.
+  \end{enumerate}
+\end{corollary}
+\begin{proof}
+  \bla
+  \begin{enumerate}
+
+  \item For the identity environment $\LRS \Gamma \sid \Gamma \rhog$ (Lemm~\ref{lem:idenv})
+  the Fundamental Theorem for Typing gives $\LR \Gamma t T {t\rhog} {T\rhog}$.
+  This implies $\R \Gamma {\down{(T\rhog)} (t\rhog)} v$ for some normal form $v$ and
+  $\Gamma \der t = v : T$ by Thm.~\ref{thm:into}.
+
+  \item By transitivity of judgmental equality from (\ref{it:sound}).
+     Here we use that the computed normal forms are identical because
+     we erase all size annotations $\ann\alpha$ to $\ann\infty$ during readback.
+  \end{enumerate}
+\end{proof}
+
+\begin{corollary}[Decidability of judgemental equality]
+  If\/ $\Gamma \der t,t' : T$ then the test whether $\nbe\Gamma T t \equiv \nbe\Gamma T{t'}$
+  terminates and
+  decides $\Gamma \der t = t' : T$.
+\end{corollary}
+\begin{proof}
+  Follows directly from soundness (which includes termination) and completeness of NbE.
+\end{proof}
+
+For normalization of types, we introduce \fbox{$\Nbe \Gamma T$}
+via the relation $\Nbe \Gamma T \evalsto V \defas \RTy \Gamma {T\rhog} V$.
+Analogously, we show that the test $\Nbe \Gamma T \equiv \Nbe \Gamma T'$
+decides type equality $\Gamma \der T = T'$.
+
+\begin{corollary}[Function type injectivity]
+  \label{cor:piinj}
+  \bla
+  \begin{enumerate}
+  \item \label{it:piinj}
+    If $\Gamma \der \piT U T = \piT{U'}{T'} : s$ then
+    $\Gamma \der U = U' : s$ and
+    $\cext \Gamma U \der T = T' : s$.
+  \item \label{it:allinj}
+    If $\Gamma \der \epiT \Size T = \piT{\Size}{T'} : s$ then
+    $\cext \Gamma \Size \der T = T' : s$.
+  \end{enumerate}
+\end{corollary}
+\begin{proof}\bla
+We prove statement (\ref{it:piinj}), the second statement follows analogously.
+By the fundamental theorem,
+$(\piT U T)\rhog = (\piT{U'}{T'})\rhog \in s$ which by definition means
+$U\rhog = U'\rhog \in s$ and $T(\rhog,u) = T'(\rhog,u') \in s$ for all $u = u' \in s$.
+By reification (Thm.~\ref{thm:rere}),
+$\RTy\Gamma {U\rhog} V$ and $\RTy\Gamma {U'\rhg} V$ for some normal form
+$V$.
+Since by inversion we have $\Gamma \der U,U' : s$ we get $\Gamma \der U = U' : s$ by soundness of NbE.
+Now, choosing $u := u' := \up U \var\Gamma \in U$, obtain
+$T\rhogu = T' \rhogu \in s$, which analogously gives us $\cext\Gamma U \der T = T' : s$.
+\end{proof}
+
+The injectivity of function types allows a stronger inversion theorem for redexes.
+
+\begin{lemma}[Strong inversion of redexes]
+  \label{lem:stronginv}
+  \bla
+  \begin{enumerate}
+  \item   If $\Gamma \der (\lambda t)\,u : T'$ then
+  $\cext \Gamma U \der t : T$ and $\Gamma \der u : U$ and $\Gamma \der T[u] \leq T'$
+  for some $U,T$.
+  \item   If $\Gamma \der (\lambda t)\,a : T'$ then
+  $\cext \Gamma \Size \der t : T$ and $\Gamma \der a : \Size$ and $\Gamma \der T[a] \leq T'$
+  for some $U,T$.
+  \item   If $\Gamma \der (\lambda t)\,\ann a : T'$ then
+  $\erext \Gamma \Size \der t : T$ and $\resurrect \Gamma \der a : \Size$ and $\Gamma \der T[a] \leq T'$
+  for some $U,T$.
+  \end{enumerate}
+\end{lemma}
+
+\begin{theorem}[Subject reduction]
+  \label{thm:sr}
+  If $\Gamma \der t : T$ and $t \evalsto w$ then $\Gamma \der t = w : T$.
+\end{theorem}
+\begin{proof}
+  By induction on $t \evalsto w$, using strong inversion on $\Gamma \der t : T$.
+\end{proof}
+
+\section{Type Checking}
+\label{sec:tycheck}
+
+Algorithmic subtyping \fbox{$\Gamma \der T \asub T'$}.
+\begin{gather*}
+  \ru{T \evalsto n \qquad
+      T' \evalsto n' \qquad
+      \Nbe \Gamma n \equiv \Nbe \Gamma n'
+    }{\Gamma \der T \asub T'}
+\\[2ex]
+  \rux{T \evalsto \Set[\ell] \qquad
+      T' \evalsto \Set[\ell']
+    }{\Gamma \der T \asub T'
+    }{\ell \leq \ell'}
+\qquad
+  \ru{T \evalsto \Nat a \qquad
+      T' \evalsto \Nat a'
+    }{\Gamma \der T \asub T'
+    }{a \leq a'}
+\\[2ex]
+  \ru{T_1' \evalsto \piT{U_1}{T_1} \qquad
+      T_2' \evalsto \piT{U_2}{T_2} \qquad
+      \Gamma \der U_2 \asub U_1 \qquad
+      \cext \Gamma {U_2} \der T_1 \asub T_2
+    }{\Gamma \der T_1' \asub T_2'}
+\\[2ex]
+  \ru{T_1' \evalsto \epiT\Size{T_1} \qquad
+      T_2' \evalsto \epiT\Size{T_2} \qquad
+      \cext \Gamma {\Size} \der T_1 \asub T_2
+    }{\Gamma \der T_1' \asub T_2'}
+\end{gather*}
+
+\begin{lemma}[Soundness of algorithmic subtyping]
+  If $\Gamma \der T \asub T'$ then $\Gamma \der T \leq T'$.
+\end{lemma}
+\begin{proof}
+  By induction on $\Gamma \der T \asub T'$, using soundness of NbE and subject reduction.
+\end{proof}
+
 \section{Conclusions}

### 2017-02-23 nbe/icfp17.tex
[ ICFP 17 ] Subtyping properties for completeness of type checking (WIP).

@@
       \Gamma \der a : \Size
-    }{\Gamma \der \Nat a : s}
+    }{\Gamma \der \Nat a : \Set[0]}
 \\[2ex]
@@
       \Gamma \der t_z : T \, (\zero{a}) \qquad
-      \Gamma \der t_s : (t : \Nat a) \to T \, (\suc{a} t)
+      \Gamma \der t_s : (x : \Nat a) \to T \, (\suc{a} x)
     }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
@@
       \Gamma \der t_z : T\,\zero{b} \qquad
-      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc{b}n)
+      \Gamma \der t_s : (x : \Nat b) \to T\,(\suc{b}x)
     }{\Gamma \der (\zero{a})\,\case[\ell] T {t_z} {t_s} = t_z : T\,\zero{b}}
@@
       \Gamma \der t_z : T\,\zero{b} \qquad
-      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc{b}n)
+      \Gamma \der t_s : (x : \Nat b) \to T\,(\suc{b}x)
     }{\Gamma \der (\suc{a} t)\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,(\suc{b} t)}
@@
       %\erext \Gamma \Size \der a : \Size
-    }{\Gamma \der t = \lambda i. t \ann a : \forallT T}
+    }{\Gamma \der t = \lambda x. t \ann a : \forallT T}
 \end{gather*}
@@
       \cext \Gamma U \der T = T' : s
-    }{\Gamma \der \piT U T = \piT{U'}T': s}
+    }{\Gamma \der \piT U T = \piT{U'}{T'}: s}
 \qquad
@@
       \Gamma \der a : \Size
-    }{\Gamma \der \Nat a = \Nat a : s}
+    }{\Gamma \der \Nat a = \Nat a : \Set[0]}
 \\[2ex]
@@
       \Gamma \der t_z = t_z' : T \, (\zero a) \qquad
-      \Gamma \der t_s = t_s' : (n : \Nat a) \to T \, (\suc a n)
+      \Gamma \der t_s = t_s' : (x : \Nat a) \to T \, (\suc a x)
     }{\Gamma \der u\,\case[\ell] T {t_z} t_s = u'\,\case[\ell] {T'} {t_z'} t_s' : T \, u}
@@
 \begin{enumerate}
-\item If $\Gamma \der \lambda t : T'$ then either $\Gamma \der \piT U T \leq T'$ and
-  $\cext \Gamma U \der t : T$ for some $U,T$ or $\Gamma \der \epiT\Size T \leq T'$ and
-  $\eext \Gamma \Size \der t : T$ for some $T$.
-\end{enumerate}
-\end{lemma}
+\item If $\Gamma \der \Nat a : T'$ then $\Gamma \der a : \Size$ and $\resurrect\Gamma \der \Set[0] \leq T'$.
+\item If $\Gamma \der \Set[\ell] : T'$ then $\resurrect\Gamma \der \Set[\ell+1] \leq T'$.
+\item If $\Gamma \der \piT U T : T'$ then $\Gamma \der U : s$ and $\cext \Gamma U \der T : s$ and $\resurrect\Gamma \der s \leq T'$ for some $s$.
+\item If $\Gamma \der \epiT \Size T : T'$ then $\cext\Gamma \Size \der T : s$ and $\resurrect\Gamma \der s \leq T'$.
+\item If $\Gamma \der \ind i : T'$ then $\Gamma(i) = \erann[\noterased] T$ and $\resurrect\Gamma \der T \leq T'$ for some $T$.
+\item If $\Gamma \der \lambda t : T'$ then either
+  $\cext \Gamma U \der t : T$
+  and
+  $\resurrect\Gamma \der \piT U T \leq T'$
+  for some $U,T$ or
+  $\eext \Gamma \Size \der t : T$
+  and
+  $\resurrect\Gamma \der \epiT\Size T \leq T'$
+  for some $T$.
+\item If $\Gamma \der t\,u : T'$ then $\Gamma \der t : \piT U T$ and $\Gamma \der u : U$ and $\resurrect\Gamma \der T[u] \leq T'$ for some $U$, $T$.
+\item If $\Gamma \der t\,a : T'$ then $\Gamma \der t : \piT \Size T$ and $\Gamma \der a : \Size$ and $\resurrect\Gamma \der T[a] \leq T'$ for some $T$.
+\item If $\Gamma \der t\,\ann a : T'$ then $\Gamma \der t : \forallT T$ and $\resurrect\Gamma \der a : \Size$ and $\resurrect\Gamma \der T[a] \leq T'$ for some $T$.
+\item If $\Gamma \der \zero a : T'$ then $\resurrect \Gamma \der a,b : \Size$ and $\resurrect\Gamma \der \Nat (b + 1) \leq T'$ for some $b$.
+\item If $\Gamma \der \suc a t : T'$ then $\resurrect \Gamma \der a,b : \Size$ and $\Gamma \der t : \Nat b$ and $\resurrect\Gamma \der \Nat(b+1) \leq T'$.
+\item If $\Gamma \der u \case[\ell] T {t_z} {t_s} : T'$ then $\Gamma \der u : \Nat (a+1)$ and $\Gamma \der T : \Nat(a+1) \to \Set[\ell]$ and $\Gamma \der t_z : T\,(\zero a)$ and $\Gamma \der t_s : \funT x {\Nat a} {T\,(\suc a x)}$ and $\resurrect\Gamma \der T\,u \leq T'$ for some $a$.
+\item If $\Gamma \der u \fix[\ell] T t : T'$ then $\Gamma \der u : \Nat a$ and $\Gamma \der T : \FixK\,\ell$ and $\Gamma \der t : \FixT\,T$ and $\resurrect\Gamma \der T\,a\,u \leq T'$.
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+  Each by induction on the typing derivation, gathering applications of the conversion rule via transitivity of subtyping.
+\end{proof}
 
@@
 \item \label{it:natsub} If $\alpha \leq \beta$ then $\NAT(\alpha) \subseteq \NAT(\beta)$.
-\item \label{it:allsub} If $\F(\alpha) \leq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
+\item \label{it:allsub} If $\F(\alpha) \subseteq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
 \item \label{it:pisub}  If $\A' \subseteq \A$ and $\F(u) \subseteq \F'(u)$ for all $u \in \A'$, then $\PIAF \subseteq \PI\,\A'\,\F'$.
@@
 \end{proof}
-We can define subtyping of type values \fbox{$T \leq T'$} by induction on $T \in \SET[\ell]$ and $T' \in \SET[\ell']$.
-Simultaneously, we need to prove correctness, namely that $T \leq T'$ implies $\EL(T) \subseteq \EL(T')$.  The correctness follows from Lemma~\ref{lem:subsump} and we do not spell it out here.
+We can define subtyping of type values \fbox{$T \leq T' \in \TYPE$} by induction on $T \in \SET[\ell]$ and $T' \in \SET[\ell']$.
+Simultaneously, we need to prove correctness, namely that $T \leq T' in \TYPE$ implies $\EL(T) \subseteq \EL(T')$.
+The correctness follows from Lemma~\ref{lem:subsump} and we do not spell it out here.
 \begin{gather*}
   \ru{T = T' \in \NEE
-    }{T \leq T'}
+    }{T \leq T' \in \TYPE}
 \qquad
@@
       \alpha \leq \alpha'
-    }{T \leq T'}
+    }{T \leq T' \in \TYPE}
 \qquad
@@
       \ell_0 \leq \ell_0'
-    }{T \leq T'}
+    }{T \leq T' \in \TYPE}
 \\[2ex]
@@
       T' \evalsto \piT {A'}{B'} \qquad
-      A' \leq A \qquad
-      B[u] \leq B'[u'] \mforall u = u' \in A'
-    }{T \leq T'}
+      A' \leq A \in \TYPE \qquad
+      B[u] \leq B'[u'] \in \TYPE \mforall u = u' \in A'
+    }{T \leq T' \in \TYPE}
 \\[2ex]
@@
       T' \evalsto \epiT {\Size}{B'} \qquad
-      B[\alpha] \leq B'[\alpha] \mforall \alpha \in \SIZE
-    }{T \leq T'}
+      B[\alpha] \leq B'[\alpha] \in \TYPE \mforall \alpha \in \SIZE
+    }{T \leq T' \in \TYPE}
 \end{gather*}
@@
 \begin{enumerate}
-\item If\/ $T = T' \in \SET[\ell]$ then $T \leq T'$.
-\item If\/ $T_1 \leq T_2$ and $T_2 \leq T_3$ then $T_1 \leq T_3$.
+\item If\/ $T = T' \in \SET[\ell]$ then $T \leq T' \in \TYPE$.
+\item If\/ $T_1 \leq T_2 \in \TYPE$ and $T_2 \leq T_3 \in \TYPE$ then $T_1 \leq T_3 \in \TYPE$.
 \end{enumerate}
@@
 Given $\piT U T \shape \piT R S$ and $\piT R S = \piT {R'}{S'} \in \SET[\ell]$ we would need to show that $T[u] \shape S'[u']$ for all $u = u' \in U$, but we only have $S[u] = S'[u'] \in \SET[\ell]$ for all $u = u' \in R$, thus the induction does not go through.
-The fact that $U \shape R$ does not give us a handle on their inhabitants, we would need $U \leq R$.
+The fact that $U \shape R$ does not give us a handle on their inhabitants, we would need $U \leq R \in \TYPE$.
 It is possible to construct an actual counterexample, using
@@
   \Gamma \models T \leq T' & \defas & \Gamma \models T \mand \Gamma \models T' \mand
-    T\eta \leq T'\eta'
+    T\eta \leq T'\eta' \in \TYPE
     % \EL(T\eta) \subseteq \EL(T'\eta')
@@
   \item If\/ $\LR \Gamma t T f A$ then $\LR{\Gamma'}{t\xi}{T\xi} f A$.
+  \end{enumerate}
+\end{lemma}
+\begin{lemma}[Compatibility with definitional equality]
+  Let $\Gamma \der T = T'$.
+  \begin{enumerate}
+  \item If\/ $\LRT \Gamma T A \ell$ then $\LRT \Gamma {T'} A \ell$.
+  \item If\/ $\LR \Gamma t T f A$ and $\Gamma \der t = t' : T$ then $\LR \Gamma {t'}{T'} f A$.
   \end{enumerate}
@@
 \end{proof}
+\begin{corollary}[One-to-one]
+  If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} A {\ell'}$ then $\Gamma \der T = T'$.
+\end{corollary}
 
@@
 \end{proof}
+
+\begin{lemma}[\citet{fridlenderPagano:tlca13}]
+   If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} {A'} {\ell'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \leq T'$.
+\end{lemma}
+
+\begin{lemma}[\citet{fridlenderPagano:tlca13}]
+   If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} {A'} {\ell'}$ and $A \leq A' \in \TYPE$ then
+   $\LR \Gamma t T f A$ implies $\LR \Gamma t {T'} f {A'}$.
+\end{lemma}
 
@@
 By reification (Thm.~\ref{thm:rere}),
-$\RTy\Gamma {U\rhog} V$ and $\RTy\Gamma {U'\rhg} V$ for some normal form
+$\RTy\Gamma {U\rhog} V$ and $\RTy\Gamma {U'\rhog} V$ for some normal form
 $V$.
 Since by inversion we have $\Gamma \der U,U' : s$ we get $\Gamma \der U = U' : s$ by soundness of NbE.
-Now, choosing $u := u' := \up U \var\Gamma \in U$, obtain
+Now, choosing $u := u' := \up {U\rhog} \var\Gamma \in U$, we obtain
 $T\rhogu = T' \rhogu \in s$, which analogously gives us $\cext\Gamma U \der T = T' : s$.
@@
 
-The injectivity of function types allows a stronger inversion theorem for redexes.
+% \section{Algorithmic Subtyping}
+% \label{sec:algosub}
+
+\begin{lemma}[Inversion of subtyping]
+
+\end{lemma}
+
+\begin{lemma}[Strong inversion of redexes]
+  \label{lem:invabs}
+  \bla
+  \begin{enumerate}
+  \item   If $\Gamma \der \lambda t : \piT U T$ then
+  $\cext \Gamma U \der t : T$.
+  \item   If $\Gamma \der \lambda t : \epiT \Size T$ then
+  $\eext \Gamma \Size \der t : T$.
+  \end{enumerate}
+\end{lemma}
 
@@
   $\cext \Gamma \Size \der t : T$ and $\Gamma \der a : \Size$ and $\Gamma \der T[a] \leq T'$
-  for some $U,T$.
+  for some $T$.
   \item   If $\Gamma \der (\lambda t)\,\ann a : T'$ then
   $\erext \Gamma \Size \der t : T$ and $\resurrect \Gamma \der a : \Size$ and $\Gamma \der T[a] \leq T'$
-  for some $U,T$.
+  for some $T$.
   \end{enumerate}
@@
   \label{thm:sr}
-  If $\Gamma \der t : T$ and $t \evalsto w$ then $\Gamma \der t = w : T$.
+  \bla
+  \begin{enumerate}
+  \item
+  If $\DD :: \Gamma \der t : T$ and $\DD' :: t \evalsto w$ then $\Gamma \der t = w : T$.
+  \item
+  If $\DD :: \Gamma \der w\, e : T$ and $\DD' :: w @ e \evalsto w'$ then $\Gamma \der w\,e = w : T$.
+  \end{enumerate}
 \end{theorem}
 \begin{proof}
-  By induction on $t \evalsto w$, using strong inversion on $\Gamma \der t : T$.
+  Simultaneously by induction on $\DD'$
+  % By induction on $t \evalsto w$,
+  using inversion (Lemma~\ref{lem:invty}) and strong inversion (Lemma~\ref{lem:stronginv})
+  on the typing derivation $\DD$. % $\Gamma \der t : T$.
 \end{proof}
@@
 \begin{lemma}[Soundness of algorithmic subtyping]
+  \label{lem:asubsound}
   If $\Gamma \der T \asub T'$ then $\Gamma \der T \leq T'$.
@@
 \begin{proof}
-  By induction on $\Gamma \der T \asub T'$, using soundness of NbE and subject reduction.
+  By induction on $\Gamma \der T \asub T'$, using soundness of NbE and subject reduction (Thm~\ref{thm:sr}).
+\end{proof}
+
+Bidirectional typing of normal forms.
+
+Checking \fbox{$\Gamma \der t \jchk T$}.
+\begin{gather*}
+  \ru{T' \evalsto s \qquad
+      \Gamma \der a : \Size
+    }{\Gamma \der \Nat a \jchk T'}
+\qquad
+  \ru{T' \evalsto \Set[\ell'] \qquad \ell < \ell'
+    }{\Gamma \der \Set[\ell] \jchk T'}
+\\[2ex]
+  \ru{T' \evalsto s \qquad
+      \Gamma \der U \jchk s \qquad
+      \cext \Gamma U \der T \jchk s
+    }{\Gamma \der \piT U T \jchk T'}
+\qquad
+  \ru{T' \evalsto s \qquad
+      \cext \Gamma \Size \der T \jchk s
+    }{\Gamma \der \epiT \Size T \jchk T'}
+\\[2ex]
+  \ru{T' \evalsto \piT U T \qquad
+      \cext \Gamma U \der t \jchk T
+    }{\Gamma \der \lambda t \jchk T'}
+\qquad
+  \ru{T' \evalsto \epiT \Size T \qquad
+      \eext \Gamma \Size \der t \jchk T
+    }{\Gamma \der \lambda t \jchk T'}
+\\[2ex]
+  \ru{T' \evalsto \Nat b \qquad
+      \resurrect \Gamma \der a+1 \leq b : \Size
+    }{\Gamma \der \zero a \jchk T'}
+\qquad
+  \ru{T' \evalsto \Nat b \qquad
+      \resurrect \Gamma \der a+1 \leq b : \Size \qquad
+      \Gamma \der t \jchk \Nat a
+    }{\Gamma \der \suc a t \jchk T'}
+\\[2ex]
+  \ru{\Gamma \der t \jinf T \qquad \Gamma \der T \asub T'
+    }{\Gamma \der t \jchk T'}
+\end{gather*}
+
+Inference \fbox{$\Gamma \der t \jinf T$}.
+\begin{gather*}
+  \ru{\Gamma(i) = \erann[\noterased] T
+    }{\Gamma \der \ind i \jinf T}
+\qquad
+  \ru{\Gamma \der t \jinf T' \qquad
+      T' \evalsto \piT U T \qquad
+      \Gamma \der u \jchk U
+    }{\Gamma \der t\, u \jinf T[u]}
+\\[2ex]
+  \ru{\Gamma \der t \jinf T' \qquad
+      T' \evalsto \piT \Size T \qquad
+      \Gamma \der a : \Size
+    }{\Gamma \der t\,a \jinf T[a]}
+\qquad
+  \ru{\Gamma \der t \jinf T' \qquad
+      T' \evalsto \erpiT \Size T \qquad
+      \resurrect\Gamma \der a : \Size
+    }{\Gamma \der t\,\ann a \jinf T[a]}
+\\[2ex]
+  \ru{\Gamma \der u \jinf \Nat (a+1) \qquad
+      \Gamma \der T \jchk \Nat (a+1) \to \Set[\ell] \qquad
+      \Gamma \der t_z \jchk T\,(\zero a) \qquad
+      \Gamma \der t_s \jchk \funT x {\Nat a} {T\,(\suc a x)}
+    }{\Gamma \der u\, \case[\ell] T {t_z} {t_s} \jinf T\,u}
+\\[2ex]
+  \ru{\Gamma \der u \jinf \Nat a \qquad
+      \Gamma \der T \jchk \FixK\,\ell \qquad
+      \Gamma \der t \jchk \FixT\,T
+    }{\Gamma \der u\, \fix[\ell] T t \jinf T\,a\,u}
+\end{gather*}
+
+We could go beyond normal forms, by adding inference rules for the $\Nat$-constructors:
+\begin{gather*}
+  \ru{\resurrect \Gamma \der a : \Size
+    }{\Gamma \der \zero a \jinf \Nat(a+1)}
+\qquad
+  \ru{\resurrect \Gamma \der a : \Size \qquad
+      \Gamma \der t \jchk \Nat a
+    }{\Gamma \der \suc a t \jinf \Nat(a+1)}
+\end{gather*}
+
+\begin{theorem}[Soundness of type checking]
+  Let $\der \Gamma$.
+  \begin{enumerate}
+  \item If $\resurrect\Gamma \der T$ and $\DD :: \Gamma \der t \jchk T$ then $\Gamma \der t : T$.
+  \item If $\DD :: \Gamma \der t \jinf T$ then $\resurrect\Gamma \der T$ and $\Gamma \der t : T$.
+  \end{enumerate}
+\end{theorem}
+\begin{proof}
+  Simultaneously by induction on $\DD$, using subject reduction (Thm.~\ref{thm:sr})
+  and soundness of algorithmic subtyping (Lemma~\ref{lem:asubsound}).
+\end{proof}
+
+\begin{lemma}[Weak head reduction of subtypes]
+\label{lem:whdsub}
+Let $\DD :: \Gamma \der T \leq T'$.
+\begin{enumerate}
+\item If $T' \evalsto \Nat a'$ then $T \evalsto \Nat a$ and $\Gamma \der a \leq a' : \Size$.
+\item If $T' \evalsto \Set[\ell']$ then $T \evalsto \Set[\ell]$ and $\ell \leq \ell'$.
+\item If $T' \evalsto \piT{A'}{B'}$ then $T \evalsto \piT {A} {B}$ and
+      $\Gamma \der A' \leq A$ and $\cext \Gamma {A'} \der B \leq B'$.
+\item If $T' \evalsto \epiT \Size {B'}$ and $T \evalsto \epiT \Size B$ and $\cext \Gamma \Size \der B \leq B'$.
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+  By induction on $\DD$, using standardization of weak head evaluation \ref{lem:stdwhd}.
+\end{proof}
+\begin{lemma}[Weak head reduction of supertypes]
+\label{lem:whdsuper}
+Let $\DD :: \Gamma \der T \leq T'$.
+\begin{enumerate}
+\item If $T \evalsto \Nat a$ then $T' \evalsto \Nat a'$ and $\Gamma \der a \leq a' : \Size$.
+\item If $T \evalsto \Set[\ell]$ then $T' \evalsto \Set[\ell']$ and $\ell \leq \ell'$.
+\item If $T \evalsto \piT A B$ then $T' \evalsto \piT {A'} {B'}$ and
+      $\Gamma \der A' \leq A$ and $\cext \Gamma {A'} \der B \leq B'$.
+\item If $T \evalsto \epiT \Size B$ and $T' \evalsto \epiT \Size B'$ and $\cext \Gamma \Size \der B \leq B'$.
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+  By induction on $\DD$, using standardization of weak head evaluation \ref{lem:stdwhd}.
+\end{proof}
+\begin{lemma}[Subsumption for type checking]
+  \label{lem:subtc}
+  Let $\sid : \Gamma' \leq \Gamma$.
+  \begin{enumerate}
+  \item If $\DD :: \Gamma \der t \jchk T$ and $\Gamma \der T \leq T'$ then $\Gamma' \der t \jchk T'$.
+  \item If $\DD :: \Gamma \der t \jinf T$ then $\Gamma' \der t \jinf T'$ and $\Gamma' \der T \leq T'$.
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+  Simultaneously by induction on $\DD$, using lemmata \ref{lem:whdsub} and \ref{lem:whdsuper}
+  and soundness and completeness of algorithmic subtyping.
+\end{proof}
+\begin{theorem}[Completeness of type checking for normal terms]
+\bla
+\begin{enumerate}
+\item If $\DD :: \Gamma \ders v : T$ then $\Gamma \der v \jchk T$.
+\item If $\DD :: \Gamma \ders m : T$ then $\Gamma \der m \jinf U$ and $\Gamma \der U \leq T$.
+\end{enumerate}
+\end{theorem}
+\begin{proof}
+Simultaneously by induction on $\DD$, using (strong) inversion and Lemma~\ref{lem:subtc}.
 \end{proof}

### 2017-02-23 nbe/icfp17.tex
[ ICFP 17 ] Some refinement of the Fridlender/Pagano theorems.

@@
 \begin{corollary}[One-to-one]
+\bla
+% \begin{enumerate}
+% \item
+% If $\LRSize \Gamma a \alpha$ and $\LRSize \Gamma {a'} \alpha$ then $a = a'$.
+% \item
 If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} A {\ell'}$ then $\Gamma \der T = T'$.
+%\end{enumerate}
 \end{corollary}
@@
 
-\begin{lemma}[\citet{fridlenderPagano:tlca13}]
+\begin{lemma}[Semantic subtyping implies judgmental subtyping \cite{fridlenderPagano:tlca13}]
+\bla
+\begin{enumerate}
+\item
+   If $\LRSize \Gamma a \alpha$ and $\LRSize \Gamma {b} \beta$ and $\alpha \leq \beta$ then $a \leq b$.
+\item
    If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} {A'} {\ell'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \leq T'$.
-\end{lemma}
-
-\begin{lemma}[\citet{fridlenderPagano:tlca13}]
+\end{enumerate}
+\end{lemma}
+
+\begin{lemma}[Subsumption for the logical relation \cite{fridlenderPagano:tlca13}]
    If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} {A'} {\ell'}$ and $A \leq A' \in \TYPE$ then

### 2017-02-24 nbe/icfp17.tex
[ ICFP 17 ] Inversion of subtyping.

@@
 %\documentclass[acmlarge]{acmart}\settopmatter{}
+
+\usepackage[right]{showlabels}
+\renewcommand{\showlabelfont}{\small\ttfamily\color{gray}}
 
@@
 \label{lem:sub}\bla
+\begin{enumerate}
+\item
   If\/ $\Gamma \der \sigma : \Delta$ and $\Delta \der J$ then $\Gamma \der J\sigma$.
+\item
+  If\/ $\Gamma \der \sigma = \sigma' \eeq tau : \Delta$ and $\Delta \der t : T$ then
+  $\Gamma \der t\sigma : T\tau$ and $\Gamma \der t\sigma' : T\tau$.
+\end{enumerate}
 \end{lemma}
@@
 \begin{corollary}[One-to-one]
+\label{cor:oneone}
 \bla
-% \begin{enumerate}
-% \item
-% If $\LRSize \Gamma a \alpha$ and $\LRSize \Gamma {a'} \alpha$ then $a = a'$.
-% \item
+\begin{enumerate}
+\item
+If $\LRSize \Gamma a \alpha$ and $\LRSize \Gamma {a'} \alpha$ then $a = a'$.
+\item
 If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} A {\ell'}$ then $\Gamma \der T = T'$.
-%\end{enumerate}
+\end{enumerate}
 \end{corollary}
@@
 \begin{lemma}[Semantic subtyping implies judgmental subtyping \cite{fridlenderPagano:tlca13}]
+\label{lem:lrsub}
 \bla
@@
 \begin{lemma}[Subsumption for the logical relation \cite{fridlenderPagano:tlca13}]
+\label{lem:lrsump}
    If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} {A'} {\ell'}$ and $A \leq A' \in \TYPE$ then
@@
 \begin{theorem}[Fundamental theorem of typing]
+\label{thm:fundty}
 \bla
@@
 
-\begin{corollary}[Function type injectivity]
+\begin{corollary}[Type constructor injectivity]
 \label{cor:piinj}
@@
 \begin{enumerate}
+\item \label{it:setinj}
+  If $\Gamma \der \Set[\ell] = \Set[\ell'] : s$ then $\ell = \ell'$.
+\item \label{it:natinj}
+  If $\Gamma \der \Nat a = \Nat b : s$ then $a = b$.
 \item \label{it:piinj}
@@
 \end{corollary}
-\begin{proof}\bla
-We prove statement (\ref{it:piinj}), the second statement follows analogously.
+\begin{proof}
+\bla
+Statement (\ref{it:setinj}) follows by inversion on  $\Set[\ell] = \Set[\ell'] \in s$, which is a direct consequence of the fundamental theorem (Thm.~\ref{thm:fund}).
+
+For (\ref{it:natinj}), observe that $\Nat (a\rhog) = \Nat (b\rhog) \in s$
+by the fundamental theorem (Thm.~\ref{thm:fund}), which by definition implies $a\rhog = b\rhog$. % \in \SIZE$.
+By the fundamental theorem of typing (Thm.~\ref{thm:fundty}), $\LRTS \Gamma {\Nat a} {\Nat (a\rhog)} s$
+which by definition implies $\LRSize \Gamma a {a\rhog}$.  Analogously, we get $\LRSize \Gamma b {b\rhog}$.
+By Cor.~\ref{cor:oneone}, $a = b$.
+
+Last, we prove statement (\ref{it:piinj}), the last statement follows analogously.
 By the fundamental theorem,
@@
 
+\begin{lemma}[Type constructor discrimination]
+  \label{lem:discr}
+  Different type constructors are not related by subtyping.  For instance,
+  $\Gamma \der \Nat a \leq \piT U T$ and
+  $\Gamma \der \erpiT \Size T \leq \piT \Size T'$ are both impossible.
+%  There are $4\times4 = 16$ statements of that sort.
+  % The following judgments are \emph{not} derivable.
+  % \begin{enumerate}
+  % \item $\Gamma \der \Nat a \leq \Set[\ell]$ and $\Gamma \der \
+  % \end{enumerate}
+\end{lemma}
+\begin{proof}
+  Follows directly by the fundamental theorem (Thm~\ref{thm:fund}) and inversion on semantic subtyping.
+\end{proof}
 % \section{Algorithmic Subtyping}
@@
 
-\begin{lemma}[Inversion of subtyping]
-
-\end{lemma}
-
-\begin{lemma}[Strong inversion of redexes]
+\begin{lemma}[Inversion of subtyping \cite{fridlenderPagano:tlca13}]
+\label{lem:invsubty}
+\bla
+\begin{enumerate}
+\item \label{it:setinvsub}
+  If $\Gamma \der \Set[\ell] \leq \Set[\ell'] : s$ then $\ell \leq \ell'$.
+\item \label{it:natinvsub}
+  If $\Gamma \der \Nat a \leq \Nat b : s$ then $a \leq b$.
+\item \label{it:piinvsub}
+  If $\Gamma \der \piT U T \leq \piT{U'}{T'} : s$ then
+  $\Gamma \der U' \leq U : s$ and
+  $\cext \Gamma {U'} \der T \leq T' : s$.
+\item \label{it:allinvsub}
+  If $\Gamma \der \epiT \Size T \leq \piT{\Size}{T'} : s$ then
+  $\cext \Gamma \Size \der T \leq T' : s$.
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+Similar to the proof of Cor.~\ref{cor:piinj}, using Lemma~\ref{lem:lrsub}.
+We recapitulate the proof for (\ref{it:piinvsub}).
+By the fundamental theorems, we get
+$\RG \Gamma {\piT U T} {(\piT U T)\rhog}$ and
+$\RG \Gamma {\piT {U'} {T'}} {(\piT {U'} {T'})\rhog}$ and
+${(\piT U T)\rhog} \leq {(\piT {U'} {T'})\rhog} \in \TYPE$.
+By inversion, first
+$\RG \Gamma U {U\rhog}$ and
+$\RG \Gamma {U'} {U'\rhog}$ and
+$U'\rhog \leq U\rhog \in \TYPE$ which imply $\Gamma \der U' \leq U$ by Lemma~\ref{lem:lrsub}.
+Recall that
+$\rhogup = (\rhog, u)$ where
+$u = \up{U'\rhog} \var\Gamma \in U'\rhog$ and
+$\LR {\cext \Gamma U'} {\ind0} {U'\slift} {u} {U'\rhog}$ by Cor.~\ref{cor:fresh}.
+From
+$\LRTS {\cext \Gamma U'} T {T(\rhogup)} s$ and
+$\LRTS {\cext \Gamma U'} {T'} {T'(\rhogup)} s$ and
+$T(\rhogup) \leq T'(\rhogup) \in s$, again by Lemma~\ref{lem:lrsub},
+$\cext \Gamma {U'} \der T \leq T' : s$.
+\end{proof}
+
+\begin{lemma}[Strong inversion for abstraction]
   \label{lem:invabs}
@@
 \end{lemma}
+\begin{proof}
+  By inversion of typing (Lem~\ref{lem:invty}), type constructor discrimination \label{lem:discr}
+  and inversion of subtyping (Lem~\ref{lem:invsubty}).
+
+  For instance, inversion on $\Gamma \der \lambda t : \piT U T$ gives us $\cext \Gamma {U'} \der t : {T'}$ with $\Gamma \der \piT{U'}{T'} \leq \piT U T$ (the other case, $\epiT \Size T$ is excluded by discrimination).
+By inversion of subtyping, $\Gamma \der U \leq U'$ and $\cext \Gamma {U} \der T' \leq T$.
+Since $\sid : \cext \Gamma U \leq \cext \Gamma {U'}$, we have $\cext \Gamma U \der t : T'$ by weakening, and our goal follows by subsumption.
+\end{proof}
 
@@
   for some $T$.
-  \item   If $\Gamma \der (\lambda t)\,\ann a : T'$ then
+  \item \label{it:sinvall}
+  If $\Gamma \der (\lambda t)\,\ann a : T'$ then
   $\erext \Gamma \Size \der t : T$ and $\resurrect \Gamma \der a : \Size$ and $\Gamma \der T[a] \leq T'$
@@
 \end{lemma}
+\begin{proof}
+  For (\ref{it:sinvall}), from $\Gamma \der (\lambda t) \ann a : T'$ we get $\Gamma \der \lambda t : \forall T$ and $\resurrect \Gamma \der a,b : \Size$ and $\Gamma \der T[b] \leq T'$.
+Strong inversion for abstraction gives us $\erext \Gamma \Size \der t : T$.
+Since $\Gamma \der [a] = [a] \eeq [b] : \erext \Gamma \Size$,  the substitution lemma yields
+$\Gamma \der t[a] : T[b]$, thus $\Gamma \der t[a] : T'$ by subsumption.
+\end{proof}
 

### 2017-02-24 nbe/icfp17.tex
[ ICFP 17 ] nfeq not needed since nfs are identical.

@@
     & \defas &
-    \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m \nfeq m' &\mforall k \\
+    \RNe k n m &\mand& \RNe k {n'} {m} &\mforall k \\
   \fbox{$d = d' \in \NF$}
     & \defas &
-    \R k d v &\mand& \R k {d'} {v'} &\mand& v \nfeq v' &\mforall k \\
+    \R k d v &\mand& \R k {d'} {v} &\mforall k \\
   \fbox{$e = e' \in \ELIM$}
     & \defas &
-    \RElim k e e_v &\mand& \R k {e'} {e_v'} &\mand& e_v \nfeq e_v' &\mforall k \\
+    \RElim k e e_v &\mand& \R k {e'} {e_v} &\mforall k \\
   \fbox{$A = A' \in \TY$}
     & \defas &
-    \RTy k A V &\mand& \RTy k {A'} {V'} &\mand& V \nfeq V' &\mforall k \\
+    \RTy k A V &\mand& \RTy k {A'} {V} &\mforall k \\
 \end{array}
+% \begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
+%   \fbox{$n = n' \in \NE$}
+%     & \defas &
+%     \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m \nfeq m' &\mforall k \\
+%   \fbox{$d = d' \in \NF$}
+%     & \defas &
+%     \R k d v &\mand& \R k {d'} {v'} &\mand& v \nfeq v' &\mforall k \\
+%   \fbox{$e = e' \in \ELIM$}
+%     & \defas &
+%     \RElim k e e_v &\mand& \R k {e'} {e_v'} &\mand& e_v \nfeq e_v' &\mforall k \\
+%   \fbox{$A = A' \in \TY$}
+%     & \defas &
+%     \RTy k A V &\mand& \RTy k {A'} {V'} &\mand& V \nfeq V' &\mforall k \\
+% \end{array}
   % \ru{\RNe k t m \mand \RNe k {t'} {m'} \mand m \nfeq m' \mforall k
@@
 \begin{theorem}[Completeness of NbE]
-  If\/ $\Gamma \der t = t' : T$ then there are normal forms $v \nfeq v'$ such that
-  $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v'$.
+  If\/ $\Gamma \der t = t' : T$ then there is a normal form $v$ such that
+  $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v$.
+  % If\/ $\Gamma \der t = t' : T$ then there are normal forms $v \nfeq v'$ such that
+  % $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v'$.
 \end{theorem}
@@
 In this section, we show that NbE is sound for judgemental equality, i.e., that same normal form implies definitional equality.
-The proof follows previous work on NbE \cite{abelCoquandDybjer:lics07,abel:habil}.
+The proof follows \citet{abelCoquandDybjer:lics07,abel:habil}.
+However, the logical relation defined in the following will also yield a weak head normalization theorem.
 

### 2017-02-24 nbe/icfp17.tex
[ ICFP 17 ] Refactored the logical relation.

@@
 However, the logical relation defined in the following will also yield a weak head normalization theorem.
-
-\[
-  \fbox{$\LRSize \Gamma a \alpha$} \defas
-  \forall \xi : \Gamma' \leq \Gamma.\ \
-  \RSize {\Gamma'} \alpha {a\xi}
-  % \exists a'.\
-  % \RSize {\Gamma'} \alpha a' \mand
-  % \Gamma' \der a\xi = a' : \Size
-\]
+\begin{alignat*}{8}
+& \Gamma \der a && \doteq \tRsize\, \alpha
+  && \quad\defas\quad
+  && \forall \xi : \Gamma' \leq \Gamma.\ \
+  && \mathrlap{ \RSize {\Gamma'} \alpha {a\xi} }
+\\[-0.6ex]
+& \Gamma \der T && \doteq \tRty\, A  : s
+  && \quad\defas\quad
+  && \forall \xi : \Gamma' \leq \Gamma.\ \
+  && \exists V.\
+  && \RTy{\Gamma'} A V
+  && ~\mand~ \Gamma' \der T\xi = V : s
+\\[-0.6ex]
+& \Gamma \der t && \doteq \tR\, d  : T
+  && \quad\defas\quad
+  && \forall \xi : \Gamma' \leq \Gamma.\ \
+  && \exists v.\
+  && \R{\Gamma'} d v
+  && ~\mand~ \Gamma' \der t\xi = v : T\xi
+\\[-0.6ex]
+& \Gamma \der t && \doteq \tRne\, n  : T
+  && \quad\defas\quad
+  && \forall \xi : \Gamma' \leq \Gamma.\ \
+  && \exists m.\
+  && \RNe{\Gamma'} n m
+  && ~\mand~ \Gamma' \der t\xi = m : T\xi
+\end{alignat*}
+% \[
+% \begin{array}{lll}
+% \Gamma \der a \doteq \tRsize\, \alpha
+%   & \defas &
+%   \forall \xi : \Gamma' \leq \Gamma.\ \ \RSize {\Gamma'} \alpha {a\xi}
+% \\
+% \Gamma \der T \doteq \tRty\, A : s
+%   & \defas &
+%   \forall \xi : \Gamma' \leq \Gamma.\ \
+%   \exists V.\
+%   \RTy{\Gamma'} A V \mand
+%   \Gamma' \der T\xi = V : s
+% \\
+% \Gamma \der t \doteq \tR\, d : T
+%   & \defas &
+%   \forall \xi : \Gamma' \leq \Gamma.\ \
+%   \exists v.\
+%   \R{\Gamma'} d v \mand
+%   \Gamma' \der t\xi = v : T\xi
+% \\
+% \Gamma \der t \doteq \tRne\, n : T
+%   & \defas &
+%   \forall \xi : \Gamma' \leq \Gamma.\ \
+%   \exists m.\
+%   \RNe{\Gamma'} n m \mand
+%   \Gamma' \der t\xi = m : T\xi
+% \end{array}
+% \]
+% \[
+%   \fbox{$\LRSize \Gamma a \alpha$} \defas
+%   \forall \xi : \Gamma' \leq \Gamma.\ \
+%   \RSize {\Gamma'} \alpha {a\xi}
+%   % \exists a'.\
+%   % \RSize {\Gamma'} \alpha a' \mand
+%   % \Gamma' \der a\xi = a' : \Size
+% \]
+By definitions, these relations are closed under subsumption and
+weakening, e.g., if $\RD \Gamma t d T$ and $\Gamma \der T \leq T'$ then $\RD \Gamma t d T'$,
+and if $\xi : \Gamma \leq \Gamma'$ then $\RD{\Gamma'}{t\xi}d{T\xi}$.
+
+\begin{lemma}[Closure properties for neutrals]
+\label{lem:rclosne}
+\bla
+\begin{enumerate}
+\item If\/ $\RNE \Gamma t n \piT U T$ and $\RD \Gamma u d U$ then $\RNE \Gamma {t\,u} {n\,d} {T[u]}$.
+\item If\/ $\RNE \Gamma t n \piT \Size T$ and $\RSIZE \Gamma a \alpha$ then $\RNE \Gamma {t\,a} {n\,\alpha} {T[a]}$.
+\item If\/ $\RNE \Gamma t n \forallT T$ and $\RSIZE {\resurrect\Gamma} a \alpha$ then $\RNE \Gamma {t\,\ann a} {n\,\ann \alpha} {T[a]}$.
+\end{enumerate}
+\end{lemma}
 
 We simultaneously define
-\fbox{$\LRT \Gamma {T'} {A'} \ell$} for $\Gamma \der T' : \Set[\ell]$
+\fbox{$\LRTS \Gamma {T'} {A'} s$} for $\Gamma \der T' : s$
 and \fbox{$\LR \Gamma t {T'} f {A'}$} for $\Gamma \der t : T'$ and $f \in A'$
-defined by induction on ${A'} \in \SET[\ell]$.
+by induction on ${A'} \in s$.
 
@@
   \noindent
-  $\LRT{\Gamma}{T'}{A'}\ell$ iff
-  $\forall \xi : \Gamma' \leq \Gamma.\ \ \exists M.\ \ \RTy{\Gamma'}{A'}M \mand \Gamma' \der T'\xi = M : \Set[\ell]$.
+  $\LRTS{\Gamma}{T'}{A'} s$ iff
+  $\Gamma \der T' \doteq \tRty\,A' : s$
+  %$\forall \xi : \Gamma' \leq \Gamma.\ \ \exists M.\ \ \RTy{\Gamma'}{A'}M \mand \Gamma' \der T'\xi = M : \Set[\ell]$.
 
@@
   $\LR{\Gamma}{t}{T'}{f}{A'}$ iff
-  $\forall \xi : \Gamma' \leq \Gamma.\ \
-   \exists m. \ \
-   \R {\Gamma'} {\down {A'} f} m \mand
-   \Gamma' \der t\xi = m : T'\xi$.
+  $\Gamma \der t \doteq \tR \, \down {A'} f : T'$
+  % $\forall \xi : \Gamma' \leq \Gamma.\ \
+  %  \exists m. \ \
+  %  \R {\Gamma'} {\down {A'} f} m \mand
+  %  \Gamma' \der t\xi = m : T'\xi$.
 
@@
   $\LR{\Gamma}{t}{T'}{f}{A'}$ iff $\resurrect\Gamma \der T' = \Nat a$ for some $a$ and
-  $\LRSize {\resurrect\Gamma} a \alpha$ and \\
-  $\forall \xi : \Gamma' \leq \Gamma.\ \
-   \exists v. \ \
-   \R {\Gamma'} {\down {A'} f} v \mand
-   \Gamma' \der t\xi = v : \Nat\,a\xi$.
+  $\LRSize {\resurrect\Gamma} a \alpha$ and
+  $\Gamma \der t \doteq \tR\,\down{A'} f : \Nat a$
+  % \\
+  % $\forall \xi : \Gamma' \leq \Gamma.\ \
+  %  \exists v. \ \
+  %  \R {\Gamma'} {\down {A'} f} v \mand
+  %  \Gamma' \der t\xi = v : \Nat\,a\xi$.
 
@@
 \bla
-  Let $A \in \SET[\ell]$.
+Let $A \in s$.
 \begin{enumerate}
@@
 \item \label{it:into}
-     $\LR \Gamma t T {\up A n} A$ if for all $\xi : \Gamma' \leq \Gamma$ there is an $m$ such that
-     $\RNe{\Gamma'}n m$ and\/ $\Gamma' \der t\xi = m : T$.
+   If\/ $\Gamma \der t \doteq \tRne\,n : T$ then
+   $\LR \Gamma t T {\up A n} A$.
 
 \item \label{it:outty}
-     If\/ $\LRT{\Gamma} T A \ell$ then for all $\xi : \Gamma' \leq \Gamma$
-     there is some normal form $V$ such that
-     $\Rty{\Gamma'} A V$ and $\Gamma' \der T\xi = V : \Set[\ell]$.
+   If\/ $\LRTS{\Gamma} T A s$ then
+   $\Gamma \der T \doteq \tRty\,A : s$
 
 \item \label{it:out}
-    If\/ $\LR \Gamma t T f A$ then for all $\xi : \Gamma' \leq \Gamma$
-    there is some normal form $v$ such that
-    $\R{\Gamma'} {\down A f} v$ and $\Gamma' \der t\xi = v : T\xi$.
+  If\/ $\LR \Gamma t T f A$ then
+  $\Gamma \der t \doteq \tR\,\down A f : T$.
+
+% \item \label{it:into}
+%    $\LR \Gamma t T {\up A n} A$ if for all $\xi : \Gamma' \leq \Gamma$ there is an $m$ such that
+%    $\RNe{\Gamma'}n m$ and\/ $\Gamma' \der t\xi = m : T$.
+
+% \item \label{it:outty}
+%    If\/ $\LRT{\Gamma} T A \ell$ then for all $\xi : \Gamma' \leq \Gamma$
+%    there is some normal form $V$ such that
+%    $\Rty{\Gamma'} A V$ and $\Gamma' \der T\xi = V : \Set[\ell]$.
+
+% \item \label{it:out}
+%   If\/ $\LR \Gamma t T f A$ then for all $\xi : \Gamma' \leq \Gamma$
+%   there is some normal form $v$ such that
+%   $\R{\Gamma'} {\down A f} v$ and $\Gamma' \der t\xi = v : T\xi$.
 
@@
 \begin{enumerate}
-\item To demonstrate $\LR \Gamma t {T'} {\up {A'} n} {A'}$ we assume
+\item
+The premise is, after type conversion, $\Gamma \der t \doteq \tRne\, n : \piT U T$.
+To demonstrate $\LR \Gamma t {T'} {\up {A'} n} {A'}$ we assume
 $\xi : \Gamma' \leq \Gamma$ and $\LR{\Gamma'}{u}{U\xi} a A$ and show
 $\LR{\Gamma'} {(t\xi\,u)} {T(\xi,u)} {\up{B[a]}(n\,\down A a)} {B[a]}$.
-This follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$ if
-we can prove that for all $\xi' : \Gamma'' \leq \Gamma'$ there is a neutral normal form $m$
-such that $\RNe{\Gamma''} {(n\,\down A a)} m$ and $\Gamma'' \der (t\xi\,u)\xi' = m : T(\xi,u)\xi'$.
-Note that the last goal can be rewritten to $\Gamma'' \der t\xi\xi'\,(u\xi') = m : T(\xi\xi',u\xi')$.
-We know that $\RNe{\Gamma''} n {m_0}$ with $\Gamma'' \der t\xi\xi' = m_0 : T'\xi\xi'$.
-Thus, it is sufficient to show that there is some normal form $v$ such that
-$\R{\Gamma''} {\down A a} v$ and $\Gamma'' \der u\xi' = v : U\xi\xi'$.
-However, this follows by induction hypothesis (\ref{it:out}) on
-assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
+By induction hypothesis (\ref{it:out}) we have $\RD {\Gamma'} u {\down A a} {U\xi}$, and
+together with the weakened premise $\Gamma' \der t\xi \doteq \tRne\, n : (\piT U T) \xi$ we get
+$\Gamma' \der t\xi\, u \doteq \tRne\, (n\,\down A a) : T(\xi,u)$.
+The goal follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$.
+% This follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$ if
+% we can prove that for all $\xi' : \Gamma'' \leq \Gamma'$ there is a neutral normal form $m$
+% such that $\RNe{\Gamma''} {(n\,\down A a)} m$ and $\Gamma'' \der (t\xi\,u)\xi' = m : T(\xi,u)\xi'$.
+% Note that the last goal can be rewritten to $\Gamma'' \der t\xi\xi'\,(u\xi') = m : T(\xi\xi',u\xi')$.
+% We know that $\RNe{\Gamma''} n {m_0}$ with $\Gamma'' \der t\xi\xi' = m_0 : T'\xi\xi'$.
+% Thus, it is sufficient to show that there is some normal form $v$ such that
+% $\R{\Gamma''} {\down A a} v$ and $\Gamma'' \der u\xi' = v : U\xi\xi'$.
+% However, this follows by induction hypothesis (\ref{it:out}) on
+% assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
 \end{enumerate}
@@
 \end{proof}
+
 \begin{corollary}[One-to-one]

### 2017-02-24 nbe/icfp17.tex
[ ICFP 17 ] Correctness of algorithmic subtyping depending on whnormalization.

@@
 
+\begin{corollary}[Type weak head normalization]
+  If $\Gamma \der T$ then $T \evalsto W$ for some $W$.
+\end{corollary}
+\begin{proof}
+  By the fundamental theorem of typing, $\RG \Gamma T {T\rhog}$,
+  which implies $T \evalsto W$ by definition of the logical relation..
+\end{proof}
+
 \begin{corollary}[Type constructor injectivity]
@@
 
-\section{Type Checking}
-\label{sec:tycheck}
+\section{Algorithmic subtyping}
+\label{sec:asub}
 
@@
 \end{proof}
+
+\begin{lemma}[Semantic subtyping implies algorithmic subtyping]
+\label{lem:lrasub}
+\bla
+   If\/ $\LRTS \Gamma T A s$ and $\LRTS \Gamma {T'} {A'} {s'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \asub T'$.
+\end{lemma}
+\begin{proof}
+By induction on $A_1' \in s_1$ and $A_2' \in s_2$ and cases on the inductive relations
+$\LRTS \Gamma {T_1'} {A_1'} {s_1}$ and $\LRTS \Gamma {T_2'} {A_2'} {s_2}$ and $A_1' \leq A_2' \in \TYPE$
+we prove $\Gamma \der T_1' \asub T_2'$.
+
+\begin{caselist}
+
+\nextcase $A_i' \evalsto N_i$ and $T_i' \evalsto n_i$ and $N_1 = N_2 \in \NE$ and $\RTY \Gamma {n_i} {N_i} {s_i}$.
+By Cor.~\ref{cor:oneone} $\Gamma \der n_1 = n_2$, thus, by completeness of NbE, $\Nbe\Gamma n_1 \equiv \Nbe\Gamma n_2$, which entails $\Gamma \der T_1' \asub T_2'$.
+
+\nextcase $A_i' \evalsto \piT {A_i} {B_i}$ and  $T_i' \evalsto \piT{U_i}{T_i}$.
+From $\LRTS \Gamma {U_i} {A_i} {s_i}$ and $A_2 \leq A_1 \in \TYPE$ we get $\Gamma \der U_2 \asub U_1$
+by induction hypothesis on $A_i \in s_i$.
+Let $a := {\up{A_2} \var\Gamma}$.  With $\LR {\cext \Gamma {U_2}} {\ind 0} {(U_2\slift)} a {A_2}$ we get
+$\LRTS {\cext \Gamma {U_2}} {T_i} {B_i[a]} s$ and $B_1[a] \leq B_2[a] \in \TYPE$.
+Thus by induction hypothesis on $B_i[a] \in s_i$ we obtain $\cext \Gamma {U_2} \der T_1 \leq T_2$, together $\Gamma \der T_1' \asub T_2'$.
+
+\end{caselist}
+\end{proof}
+
+\begin{corollary}[Completeness of algorithmic subtyping]
+\label{cor:complasub}
+If $\Gamma \der T \leq T'$ then $\Gamma \der T \asub T'$.
+\end{corollary}
+\begin{proof}
+  By the fundamental theorems $\RG \Gamma T {T\rhog}$ and $\RG \Gamma {T'} {T'\rhog}$ and $T\rhog \leq T'\rhog \in \TYPE$.
+  By Lemma~\ref{lem:lrasub}, $\Gamma \der T \asub T'$.
+\end{proof}
+
+Since the algorithmic subtyping relation is equivalent to the declarative one, we can freely swap one relation for the other.
+
+\begin{lemma}[Termination of algorithmic subtyping]
+  \label{lem:termasub}
+  If $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ then the query $\Gamma \der T \asub T'$ terminates.
+\end{lemma}
+\begin{proof}
+  By induction on $A \in s$ and $A' \in s'$ and cases on $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$.
+\end{proof}
+
+\begin{theorem}[Decidability of subtyping]
+  If $\Gamma \der T,T'$, then $\Gamma \der T \leq T'$ is decided by the query $\Gamma \der T \asub T'$.
+\end{theorem}
+\begin{proof}
+  By the fundamental theorem of typing, $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$,
+  thus, the query  $\Gamma \der T \asub T'$ terminates by Lemma~\ref{lem:termasub}.
+  If successfully, then $\Gamma \der T \leq T'$ by soundness of algorithmic equality.
+  If unsuccessfully, then $\Gamma \der T \leq T'$ is impossible by completeness of algorithmic equality.
+\end{proof}
+
+\section{Type Checking}
+\label{sec:tycheck}
 

### 2017-02-24 nbe/icfp17.tex
[ ICFP 17 ] Decidability of type checking.

@@
 
+In this section, we show that type checking for normal forms is decidable, and succeeds for those which can be type via the restricted rule for size polymorphism elimination:
+\[
+  \ru{\Gamma \ders t : \forall T \qquad
+      \resurrect\Gamma \der a : \Size
+    }{\Gamma \ders t\, \ann a : T[a]}
+\]
+We will refer to the restricted typing judgement as \fbox{$\Gamma \ders t : T$}, and obviously,
+if $\Gamma \ders t : T$ then $\Gamma \der t : T$.
+
 Bidirectional typing of normal forms.
@@
   \begin{enumerate}
-  \item If $\resurrect\Gamma \der T$ and $\DD :: \Gamma \der t \jchk T$ then $\Gamma \der t : T$.
-  \item If $\DD :: \Gamma \der t \jinf T$ then $\resurrect\Gamma \der T$ and $\Gamma \der t : T$.
+  \item If $\resurrect\Gamma \der T$ and $\DD :: \Gamma \der t \jchk T$ then $\Gamma \ders t : T$.
+  \item If $\DD :: \Gamma \der t \jinf T$ then $\resurrect\Gamma \der T$ and $\Gamma \ders t : T$.
   \end{enumerate}
@@
 \label{lem:whdsub}
-Let $\DD :: \Gamma \der T \leq T'$.
-\begin{enumerate}
-\item If $T' \evalsto \Nat a'$ then $T \evalsto \Nat a$ and $\Gamma \der a \leq a' : \Size$.
-\item If $T' \evalsto \Set[\ell']$ then $T \evalsto \Set[\ell]$ and $\ell \leq \ell'$.
+Let $\DD :: \Gamma \der T \asub T'$.
+\begin{enumerate}
+\item If $T' \evalsto \Nat a'$ then $T \evalsto \Nat a$ and $\Gamma \der a \asub a' : \Size$.
+\item If $T' \evalsto \Set[\ell']$ then $T \evalsto \Set[\ell]$ and $\ell \asub \ell'$.
 \item If $T' \evalsto \piT{A'}{B'}$ then $T \evalsto \piT {A} {B}$ and
-      $\Gamma \der A' \leq A$ and $\cext \Gamma {A'} \der B \leq B'$.
-\item If $T' \evalsto \epiT \Size {B'}$ and $T \evalsto \epiT \Size B$ and $\cext \Gamma \Size \der B \leq B'$.
-\end{enumerate}
-\end{lemma}
-\begin{proof}
-  By induction on $\DD$, using standardization of weak head evaluation \ref{lem:stdwhd}.
+      $\Gamma \der A' \asub A$ and $\cext \Gamma {A'} \der B \asub B'$.
+\item If $T' \evalsto \epiT \Size {B'}$ and $T \evalsto \epiT \Size B$ and $\cext \Gamma \Size \der B \asub B'$.
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+  By cases on $\DD$, since weak head evaluation is deterministic.
 \end{proof}
@@
 \label{lem:whdsuper}
-Let $\DD :: \Gamma \der T \leq T'$.
-\begin{enumerate}
-\item If $T \evalsto \Nat a$ then $T' \evalsto \Nat a'$ and $\Gamma \der a \leq a' : \Size$.
-\item If $T \evalsto \Set[\ell]$ then $T' \evalsto \Set[\ell']$ and $\ell \leq \ell'$.
+Let $\DD :: \Gamma \der T \asub T'$.
+\begin{enumerate}
+\item If $T \evalsto \Nat a$ then $T' \evalsto \Nat a'$ and $\Gamma \der a \asub a' : \Size$.
+\item If $T \evalsto \Set[\ell]$ then $T' \evalsto \Set[\ell']$ and $\ell \asub \ell'$.
 \item If $T \evalsto \piT A B$ then $T' \evalsto \piT {A'} {B'}$ and
-      $\Gamma \der A' \leq A$ and $\cext \Gamma {A'} \der B \leq B'$.
-\item If $T \evalsto \epiT \Size B$ and $T' \evalsto \epiT \Size B'$ and $\cext \Gamma \Size \der B \leq B'$.
-\end{enumerate}
-\end{lemma}
-\begin{proof}
-  By induction on $\DD$, using standardization of weak head evaluation \ref{lem:stdwhd}.
+      $\Gamma \der A' \asub A$ and $\cext \Gamma {A'} \der B \asub B'$.
+\item If $T \evalsto \epiT \Size B$ and $T' \evalsto \epiT \Size B'$ and $\cext \Gamma \Size \der B \asub B'$.
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+  By cases on $\DD$, since weak head evaluation is deterministic.
 \end{proof}
@@
   \begin{enumerate}
-  \item If $\DD :: \Gamma \der t \jchk T$ and $\Gamma \der T \leq T'$ then $\Gamma' \der t \jchk T'$.
-  \item If $\DD :: \Gamma \der t \jinf T$ then $\Gamma' \der t \jinf T'$ and $\Gamma' \der T \leq T'$.
-  \end{enumerate}
-\end{lemma}
-\begin{proof}
-  Simultaneously by induction on $\DD$, using lemmata \ref{lem:whdsub} and \ref{lem:whdsuper}
+  \item If $\DD :: \Gamma \der t \jchk T$ and $\resurrect\Gamma \der T \leq T'$ then $\Gamma' \der t \jchk T'$.
+  \item If $\DD :: \Gamma \der t \jinf T$ then $\Gamma' \der t \jinf T'$ and $\resurrect\Gamma' \der T \leq T'$.
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+  Simultaneously by induction on $\DD$, using % weak head normalization of types,
+  lemmata \ref{lem:whdsub} and \ref{lem:whdsuper},
   and soundness and completeness of algorithmic subtyping.
@@
 \item If $\DD :: \Gamma \ders v : T$ then $\Gamma \der v \jchk T$.
-\item If $\DD :: \Gamma \ders m : T$ then $\Gamma \der m \jinf U$ and $\Gamma \der U \leq T$.
+\item If $\DD :: \Gamma \ders m : T$ then $\Gamma \der m \jinf U$ and $\resurrect\Gamma \der U \leq T$.
 \end{enumerate}
@@
 \end{proof}
+
+\begin{lemma}[Termination of type checking]
+  Let $\der \Gamma$.
+  \begin{enumerate}
+  \item The query $\Gamma \der t \jinf \mbox{?}$ terminates.
+  \item If $\resurrect\Gamma \der T$ then the query $\Gamma \der t \jchk T$ terminates.
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+  By induction on $t$, using type weak head normalization and soundness of type checking, to maintain well-formedness of types.
+  And, of course, decidability of subtyping.
+\end{proof}
+
+\begin{theorem}[Decidability of type checking for normal terms]
+  Let $\der \Gamma$ and $\resurrect \Gamma \der T$.
+  Then $\Gamma \ders v : T$ is decided by $\Gamma \der v \jchk T$.
+\end{theorem}
 

### 2017-02-25 nbe/icfp17.tex
[ ICFP 17 ] Added weak head evaluation to the logical relation.
[ ICFP 17 ] Substitution laws and Adm predicate.

@@
 \SizeExp & \ni & a,b & ::= & \infty \mid o \mid \ind i + o ~(o \in \NN) & \mbox{size expressions} \\
-\Cxt & \ni & \Gamma, \Delta & ::= & \cempty \mid \erext[\noterased] \Gamma T \mid \erext \Gamma \Size \\
+\Cxt & \ni & \Gamma, \Delta & ::= & \cempty \mid \erext[\noterased] \Gamma T \mid \erext \Gamma \Size
+   & \mbox{contexts} \\
+\Subst & \ni & \eta,\rho,\sigma,\tau,\xi & ::= & \sempty \mid (\sigma,t)
+  & \mbox{substitutions} \\
 \end{array}
 \]
+
+%We write $\sigma$ and $\tau$ for substitutions, which are lists of terms, and
+We write $t \sigma$ for the application of substitution $\sigma$ to term $t$ which is defined as usual.
+We write $\slift^k_m$ for the substitution $(\ind{k+m-1},\dots,\ind{k+1},\ind k)$ which applies to a term with $m$ free indices and increases all of them by $k$.
+We write $\slift_m$ for the lifting $\slift^1_m$ and $\sid_m$ for the identity substitution $ \slift^0_m$.
+In general, we refer to liftings by letter $\xi$.
+The substitution $[u]_m = (\sid_m, u)$ replaces free index $\ind 0$ by term $u$ and decrements all other free indices by $1$. We drop subscript $m$ when clear from the context.
+Substitution composition $\sigma\tau$ is the pointwise application of substitution $\tau$ to the list of terms $\sigma$.  In the proofs to follow, we freely use the following identities:
+\begin{gather*}
+  t\sid \equiv t
+\quad
+  (t\sigma)\tau \equiv t(\sigma\tau)
+\qquad
+  \sigma \sid \equiv \sigma
+\quad
+  \sid \tau \equiv \tau
+\quad
+  (\rho\sigma)\tau \equiv \rho(\sigma\tau)
+\\
+  \ind 0 (\sigma,t) \equiv t
+\quad
+  \slift(\sigma,t) \equiv \sigma
+\qquad
+  [t]\sigma \equiv (\sigma, t\sigma)
+\quad
+  \slift[t] \equiv \sid
+%  (\sigma\slift)[t] \equiv \sigma
+\end{gather*}
+
 We omit the ``$\noterased$''-markers from $\Gamma$ and $\Pi$ by default.  We write $\forallT T$ for $\erpiT \Size T$.
-For concrete function types, we may use a (named dependent) function type notation as syntactic sugar for the corresponding de Bruijn representation.
+For concrete function types, we may use a named dependent function type notation as syntactic sugar for the corresponding de Bruijn representation.
 For instance, $\funT i \Size \Nat i \to \Set[\ell]$ is sugar for $\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.
@@
 (t \slift)\,\ind 0$ if $t$ is outside the scope of $x$.
+We may also use names when we construct concrete contexts, for instance, if $T$ is well-formed in context $\Gamma$, we may write $T\,i\,x$ in context $\Gamma.i\of\Size.x\of\Nat i$ to mean $T\slift^2\,\ind 1\,\ind 0$ in context
+$\Gamma.\Size.\Nat \ind 0$.
 
@@
 them by ``$\noterased$''-markers).  Note that, trivially, resurrection is idempotent: $\resurrecttwice\Gamma = \resurrect{\Gamma}$.
-
-We write $\sigma$ and $\tau$ for substitutions, which are lists of terms, and $t \sigma$ for the application of substitution $\sigma$ to term $t$.
-We write $\slift^k_m$ for the substitution $(\ind{k+m-1},\dots,\ind{k+1},\ind k)$ which applies to a term with $m$ free indices and increases all of them by $k$.
-We write $\slift_m$ for the lifting $\slift^1_m$ and $\sid_m$ for the identity substitution $ \slift^0_m$.
-The substitution $[u]_m = (\sid_m, u)$ replaces free index $\ind 0$ by term $u$ and decrements all other free indices by $1$.
-We drop subscript $m$ when clear from the context.
 
@@
   \Gamma \der T \leq T' & \mbox{in context $\Gamma$, type $T$ is a subtype of $T'$} \\
-  \Gamma \der T : \Adm \ell & \mbox{in context $\Gamma$ type constructor $T$ is admissible for recursion} \\
+%  \Gamma \der T : \Adm \ell & \mbox{in context $\Gamma$ type constructor $T$ is admissible for recursion} \\
   \suty \Gamma \sigma \tau \Delta & \mbox{$\sigma$/$\tau$ is a valid term/type-level substitution for $\Delta$} \\
@@
 \[
-\begin{array}{lll@{~}l@{\qquad}l}
+\begin{array}{l@{~~}l@{~~}l@{~}l@{\qquad}l}
   \Gamma \der T & \defas & \Gamma \der T : s & \mforsome s
@@
     & \mbox{in context $\Gamma$, types $T$ and $T'$ are equal} \\
-  \Gamma \der a = b : \Size & \defas & \multicolumn 2 {l@{\qquad}} {\Gamma \der a : \Size \mand a = b}
+  \Gamma \der a = b : \Size & \defas &
+    \multicolumn 2 {@{}l@{\qquad}} {\Gamma \der a : \Size \mand a = b}
     & \mbox{sizes $a$ and $b$ are well-formed and identical} \\
   \Gamma \der a \leq b : \Size & \defas &
-     \multicolumn 2 {l@{\qquad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
+     \multicolumn 2 {@{}l@{\qquad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
 %     \multicolumn 2 {l@{\qquad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
@@
 %    & \mbox{in context $\Gamma$, well-formed size $a$ is less or equal than well-formed size $b$} \\
+  \Gamma \der T : \Adm \ell & \defas &
+     \multicolumn 2 {@{}l@{\qquad}}
+       {\Gamma \der T : \FixK \mand \Gamma.i\of\Size.x\of\Nat i \der T\,i\,x \leq T\,\infty\,x}
+    & \mbox{% in context $\Gamma$
+            type constructor $T$ is admissible for recursion} \\
 \end{array}
@@
   \ru{\Gamma \der u : \Nat a \qquad
-      \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
+      \Gamma \der T : \Adm\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
       \Gamma \der t : \FixT\, T
@@
       \resurrect\Gamma \der a : \Size \qquad
-      \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
+      \Gamma \der T : \Adm\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
       \Gamma \der t : \FixT\,T %% \allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x
@@
   \ru{\Gamma \der u = u' : \Nat a \qquad
-      \Gamma \der T = T' : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
+      \Gamma \der T = T' : \Adm\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
       \Gamma \der t = t' : \FixT\,T \qquad %% \allT i \Size (\funT x {\Nat i} T \,i\, x) \to \funT x {\Nat (i + 1)} T \,{i+1}\, x
@@
 \item If $\Gamma \der u \case[\ell] T {t_z} {t_s} : T'$ then $\Gamma \der u : \Nat (a+1)$ and $\Gamma \der T : \Nat(a+1) \to \Set[\ell]$ and $\Gamma \der t_z : T\,(\zero a)$ and $\Gamma \der t_s : \funT x {\Nat a} {T\,(\suc a x)}$ and $\resurrect\Gamma \der T\,u \leq T'$ for some $a$.
-\item If $\Gamma \der u \fix[\ell] T t : T'$ then $\Gamma \der u : \Nat a$ and $\Gamma \der T : \FixK\,\ell$ and $\Gamma \der t : \FixT\,T$ and $\resurrect\Gamma \der T\,a\,u \leq T'$.
+\item If $\Gamma \der u \fix[\ell] T t : T'$ then $\Gamma \der u : \Nat a$ and $\Gamma \der T : \Adm\,\ell$ and $\Gamma \der t : \FixT\,T$ and $\resurrect\Gamma \der T\,a\,u \leq T'$.
 \end{enumerate}
@@
 
+Let \fbox{$\Gamma \der T \evalsto W : s$} denote the conjunction of $T \evalsto W$ and $\Gamma \der T = W : s$.
 We simultaneously define
@@
   $\LRTS{\Gamma}{T'}{A'} s$ iff
+  $\Gamma \der T' \evalsto n : s$ for some neutral $n$ and
   $\Gamma \der T' \doteq \tRty\,A' : s$
@@
   \noindent
-  $\LRT{\Gamma}{T'}{A'}\ell$ iff $\Gamma \der T' = \Nat a$ for some $a$ and $\LRSize \Gamma a \alpha$.
+  $\LRTS{\Gamma}{T'}{A'} s$ iff $\Gamma \der T' \evalsto \Nat a : s$ for some $a$ and $\LRSize \Gamma a \alpha$.
 
   \noindent
-  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff $\resurrect\Gamma \der T' = \Nat a$ for some $a$ and
+  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff $\resurrect\Gamma \der T' \evalsto \Nat a$ for some $a$ and
   $\LRSize {\resurrect\Gamma} a \alpha$ and
@@
   \noindent
-  $\LRT {\Gamma} {T'} {A'} \ell$ iff $\Gamma \der T' = \Set[\ell'] : \Set[\ell]$.
+  $\LRTS {\Gamma} {T'} {A'} s$ iff $\Gamma \der T' \evalsto \Set[\ell'] : s$.
 
@@
   $\LR \Gamma U {T'} B {A'}$ iff
-  $\resurrect\Gamma \der T' = \Set[\ell'] : \Set[\ell]$ and $\LRT \Gamma U B {\ell'}$.
+  $\resurrect\Gamma \der T' \evalsto \Set[\ell'] : s$ and $\LRT \Gamma U B {\ell'}$.
 
@@
 \noindent
-$\LRT \Gamma {T'} {A'} \ell$ iff
+$\LRTS \Gamma {T'} {A'} s$ iff
 \[
 \begin{array}{l}
-\Gamma \der T' = \piT U T : \Set[\ell] \mforsome U, T \mand \LRT \Gamma U A \ell \mand \\
+\Gamma \der T' \evalsto \piT U T : s \mforsome U, T \mand \LRTS \Gamma U A s \mand \\
 \forall \xi : \Gamma' \leq \Gamma.\ \
-\LR {\Gamma'} u {U\xi} a A \implies \LRT {\Gamma'} {T(\xi,u)} {B[a]} \ell
+\LR {\Gamma'} u {U\xi} a A \implies \LRTS {\Gamma'} {T(\xi,u)} {B[a]} s
 . \\
@@
 \begin{array}{l}
-\resurrect\Gamma \der T' = \piT U T : \Set[\ell] \mforsome U, T \mand \LRT {\resurrect\Gamma} U A \ell \mand \\
+\resurrect\Gamma \der T' \evalsto \piT U T : s \mforsome U, T \mand \LRTS {\resurrect\Gamma} U A s \mand \\
 \forall \xi : \Gamma' \leq \Gamma.\ \
@@
 \noindent
-$\LRT \Gamma {T'} {A'} \ell$ iff
+$\LRTS \Gamma {T'} {A'} s$ iff
 \[
 \begin{array}{l}
-\Gamma \der T' = \piT \Size T : \Set[\ell] \mforsome T\\
+\Gamma \der T' \evalsto \piT \Size T : s \mforsome T\\
 \forall \xi : \Gamma' \leq \Gamma.\ \
 \LRSize {\Gamma'} a \alpha
-\implies \LRT {\Gamma'} {T(\xi,a)} {B[\alpha]} \ell
+\implies \LRTS {\Gamma'} {T(\xi,a)} {B[\alpha]} s
 .
@@
 \begin{array}{l}
-\resurrect\Gamma \der T' = \piT \Size T : \Set[\ell] \mforsome T\\
+\resurrect\Gamma \der T' \evalsto \piT \Size T : s \mforsome T\\
 \forall \xi : \Gamma' \leq \Gamma.\ \
@@
 \noindent
-$\LRT \Gamma {T'} {A'} \ell$ iff
+$\LRTS \Gamma {T'} {A'} s$ iff
 \[
 \begin{array}{l}
-\Gamma \der T' = \forallT T : \Set[\ell] \mforsome T\\
+\Gamma \der T' \evalsto \forallT T : s \mforsome T\\
 \forall \xi : \Gamma' \leq \Gamma,\
@@
 \LRSize {\Gamma'} b \beta
-\implies \LRT {\Gamma'} {T(\xi,b)} {B[\beta]} \ell
+\implies \LRTS {\Gamma'} {T(\xi,b)} {B[\beta]} s
 .
@@
 \begin{array}{l}
-\resurrect\Gamma \der T' = \forallT T : \Set[\ell] \mforsome T\\
+\resurrect\Gamma \der T' \evalsto \forallT T : s \mforsome T\\
 \forall \xi : \Gamma' \leq \Gamma,\
@@
 \]
-
 \end{caselist}
+We may prove theorems ``by induction on $\LRTS \Gamma T A s$'', even if in reality this will be proofs by induction on $A \in s$ and cases on $\LRTS \Gamma T A s$.
+We write \fbox{$\RG \Gamma T A$} if $\LRTS \Gamma T A s$ for some sort $s$.
 
@@
   \item If\/ $\LRSize \Gamma a \alpha$ then $\LRSize{\Gamma'}{a\xi}\alpha$.
-  \item If\/ $\LRT {\Gamma} T A \ell$ then $\LRT{\Gamma'}{T\xi}A \ell$.
+  \item If\/ $\LRTS {\Gamma} T A s$ then $\LRTS{\Gamma'}{T\xi}A s$.
   \item If\/ $\LR \Gamma t T f A$ then $\LR{\Gamma'}{t\xi}{T\xi} f A$.
@@
 \end{lemma}
-\begin{lemma}[Compatibility with definitional equality]
-  Let $\Gamma \der T = T'$.
-  \begin{enumerate}
-  \item If\/ $\LRT \Gamma T A \ell$ then $\LRT \Gamma {T'} A \ell$.
-  \item If\/ $\LR \Gamma t T f A$ and $\Gamma \der t = t' : T$ then $\LR \Gamma {t'}{T'} f A$.
-  \end{enumerate}
-\end{lemma}
+
+%% NO LONGER TRUE AT THIS POINT
+% \begin{lemma}[Compatibility with definitional equality]
+%   Let $\Gamma \der T = T'$.
+%   \begin{enumerate}
+%   \item If\/ $\LRT \Gamma T A \ell$ then $\LRT \Gamma {T'} A \ell$.
+%   \item If\/ $\LR \Gamma t T f A$ and $\Gamma \der t = t' : T$ then $\LR \Gamma {t'}{T'} f A$.
+%   \end{enumerate}
+% \end{lemma}
 
@@
 \bla
-Let $A \in s$.
+Let $\LRTS{\Gamma} T A s$. Then:
+%Let $A \in s$.
 \begin{enumerate}
@@
 
-\item \label{it:outty}
-   If\/ $\LRTS{\Gamma} T A s$ then
-   $\Gamma \der T \doteq \tRty\,A : s$
-
 \item \label{it:out}
@@
   $\Gamma \der t \doteq \tR\,\down A f : T$.
+
+\item \label{it:outty}
+   %If\/ $\LRTS{\Gamma} T A s$ then
+   $\Gamma \der T \doteq \tRty\,A : s$.
 
@@
 We prove the propositions for $A'$ and $T'$ (instead of $A$ and $T$)
-simultaneously by induction on $A' \in \SET[\ell]$.
+simultaneously by induction on $A' \in s$.
 
@@
 $\Gamma' \der t\xi\, u \doteq \tRne\, (n\,\down A a) : T(\xi,u)$.
-The goal follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$.
+The goal follows by induction hypothesis (\ref{it:into}) for
+% $B[a] \in s$.
+$\LRTS{\Gamma'}{T(\xi,u)}{B[a]} s$.
 % This follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$ if
@@
   \label{cor:fresh}
-  If $\der \cext \Gamma T$ and $A \in \SET[\ell]$
+  If $\der \cext \Gamma T$ and %$A \in \SET[\ell]$
+  $\RG \Gamma T A$
   then $\LR {\cext \Gamma T} {\ind 0} {(T \slift)} {(\up A \var\Gamma)} A$.
@@
 \item
-   If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} {A'} {\ell'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \leq T'$.
-\end{enumerate}
-\end{lemma}
+   If $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \leq T'$.
+   % If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} {A'} {\ell'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \leq T'$.
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+  The proof is analogous to the one  for algorithmic subtyping to come (Lemma~\ref{lem:lrasub}).
+\end{proof}
 
@@
 \label{lem:lrsump}
-   If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} {A'} {\ell'}$ and $A \leq A' \in \TYPE$ then
+   If $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and $A \leq A' \in \TYPE$ then
    $\LR \Gamma t T f A$ implies $\LR \Gamma t {T'} f {A'}$.
+   % If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} {A'} {\ell'}$ and $A \leq A' \in \TYPE$ then
+   % $\LR \Gamma t T f A$ implies $\LR \Gamma t {T'} f {A'}$.
 \end{lemma}
@@
 \bla
-   If\/ $\LRTS \Gamma T A s$ and $\LRTS \Gamma {T'} {A'} {s'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \asub T'$.
-\end{lemma}
-\begin{proof}
-By induction on $A_1' \in s_1$ and $A_2' \in s_2$ and cases on the inductive relations
-$\LRTS \Gamma {T_1'} {A_1'} {s_1}$ and $\LRTS \Gamma {T_2'} {A_2'} {s_2}$ and $A_1' \leq A_2' \in \TYPE$
+   If\/ $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \asub T'$.
+\end{lemma}
+\begin{proof}
+By induction on % $A_1' \in s_1$ and $A_2' \in s_2$ and cases on the inductive relations
+$\LRTS \Gamma {T_1'} {A_1'} {s_1}$ and $\LRTS \Gamma {T_2'} {A_2'} {s_2}$
+and cases on $A_1' \leq A_2' \in \TYPE$
 we prove $\Gamma \der T_1' \asub T_2'$.
@@
   \item If $\DD :: \Gamma \der t \jchk T$ and $\resurrect\Gamma \der T \leq T'$ then $\Gamma' \der t \jchk T'$.
-  \item If $\DD :: \Gamma \der t \jinf T$ then $\Gamma' \der t \jinf T'$ and $\resurrect\Gamma' \der T \leq T'$.
+  \item If $\DD :: \Gamma \der t \jinf T$ then $\Gamma' \der t \jinf T'$ and $\resurrect{\Gamma'} \der T \leq T'$.
   \end{enumerate}

### 2017-02-25 nbe/icfp17.tex
[ ICFP 17 ] case and fix

@@
 \fi
+
+\makeatletter
+\newenvironment{proof*}[1][\proofname]{\par
+  \normalfont \topsep6\p@\@plus6\p@\relax
+  \trivlist
+  \item[\@proofindent\hskip\labelsep
+        {\@proofnamefont #1\@addpunct{.}}]\ignorespaces
+}{%
+  \endtrivlist\@endpefalse
+}
+\makeatother
 
@@
 \\[2ex]
-  \ru{t\,\ann \alpha\,(\lambda x.\, x@ \fixtt)\,c \evalsto w
+  \ru{t\,\ann \alpha\,(\lambda x.\, x\, \fixtt)\,c \evalsto w
      }{c @ \fixtt \evalsto w
@@
 \\[2ex]
-  \ru{T' \evalsto \Nat \beta \qquad
+  \ru{%T' \evalsto \Nat \beta \qquad
     }{(\up {T'} n) @ \case[\ell] T {t_z} {t_s} \evalsto
-      \up{T\,(\up{\Nat \infty} n)}
+      \up{T\,(\up{T'} n)}\,
+      % \up{T\,(\up{\Nat \infty} n)}
          n\,\case[\ell]
-             T %% {(\down {\Set[\infty]} T)}
+             {(\down {\Nat \infty \to \Set[\ell]} T)}
              {(\down {T\,{\zero\infty}} t_z)}
@@
 \\[2ex]
-  \ru{T' \evalsto \Nat \beta \qquad
+  \ru{%T' \evalsto \Nat \beta \qquad
     }{(\up {T'} n) @ \fix[\ell] T t \evalsto
@@
 \\[2ex]
-  \ru{\RTy k T V \qquad
+  \ru{\R k D V \qquad
       \R k {d_z} {v_z} \qquad
       \R k {d_s} {v_s}
-    }{\RElim k {(\case[\ell] T {d_z} {d_s})} {\case[\ell] V {v_z} {v_s}}}
+    }{\RElim k {(\case[\ell] D {d_z} {d_s})} {\case[\ell] V {v_z} {v_s}}}
 \qquad
@@
 \end{lemma}
+
+Now we define some PERs and PER constructors on values.
+All these PERs $\A$ are closed under weak head equality, meaning if $a = b \in \A$ and $a'$ has the same weak head normal form as $a$, then $a' = b in \A$.
+(By symmetry, $\A$ is also closed under weak head equality on the second argument.)
 
@@
 \]
-All relations involved here are closed under weak head expansion.
+All relations involved here are closed under weak head equality.
 \begin{lemma}[Well-definedness]
@@
 \end{corollary}
+
+\subsection{Computation}
+\label{sec:comp}
+
+The eliminations for natural numbers are acurrately modeled.
+\begin{lemma}[Case]
+  \label{lem:case}
+If $a = a' \in \Nat (\alpha+1)$
+and $B = B' \in \Nat (\alpha+1) \to \Set[\ell]$
+and $f_z = f_z' \in B\,(\zero \beta)$
+and $f_s = f_s' \in \funT x {\Nat \alpha} B\,(\suc \gamma x)$
+then $a\,\case[\ell] B {f_z}{f_s} = a'\,\case[\ell] {B'} {f_z'}{f_s'} \in B\,a$.
+\end{lemma}
+\begin{proof}
+By induction on $a = a' \in \Nat(\alpha+1)$.
+\begin{caselist}
+
+\nextcase $a \evalsto \zero \beta$ and $a' \evalsto \zero{\beta'}$.
+Since our PERs are closed under weak head equality,
+and, for instance, $a\, \case[\ell] B {f_z}{f_s}$ has the same weak head normal form as $f_z$,
+it suffices to show $f_z = f_z' \in B\,(\zero \beta)$, which is one of our assumptions.
+
+\nextcase $a \evalsto \suc \beta b$ and $a' \evalsto \suc {\beta'}{b'}$ with $b = b' \in \Nat \alpha$.
+Again, it suffices to show
+$f_s\,b = f_s'\,b' \in B\,(\suc \beta b)$, which is an instance of our last assumption.
+
+\nextcase $a \evalsto \up T n$ and $a' \evalsto \up {T'} n'$ with $n = n' \in \NE$.
+Let $D = \down{\Nat \infty \to \Set[\ell]} B$
+and $d_z = \down{B\,(\zero \infty)} f_z$
+and $d_s = \down{\funS x {\Nat\,\infty} {B\,(\suc \infty x)}} f_s$
+and $e = \case[\ell] D {d_z}{d_s}$.
+% $e = \case[\ell]
+%   {(\down{\Nat \infty \to \Set[\ell]} B)}
+%   {(\down{B\,(\zero \infty)} f_z)}
+%   {(\down{\funS x {\Nat\,\infty} {B\,(\suc \infty x)} f_s)}$
+Let $D',d_z',d_s',e'$ be defined analogously from $B',f_z',f_s'$.
+It suffices to show $e = e' \in \ELIM$, since then we have $n\,e = n'\,e' \in \NE$ by the closure properties of $\NE$ (Lemma~\ref{lem:closne}),
+and $\up{B[a]} (n\,e) = \up{B'[a']} (n'\,e') \in B[a]$ by reflection (Theorem~\ref{thm:rere}).
+The remaining goal $\case[\ell] D {d_z}{d_s} = \case[\ell] {D'} {d_z'}{d_s'} \in \ELIM$ follows by the closure properties for eliminations (Lemma~\ref{lem:closelim}), since $D = D' \in \NF$ and $d_z = d_z' \in \NF$ and $d_s = d_s' \in \NF$ all hold by reification (Theorem~\ref{thm:rere}).
+\end{caselist}
+\end{proof}
+
+\begin{lemma}[$\NAT$ is cocontinuous]
+  \label{lem:limit}
+  $\NAT(\infty) = \bigcup_{\alpha < \infty} \NAT(\alpha)$.
+\end{lemma}
+\begin{proof}
+  By induction on $a = a' \in \NAT(\infty)$, we can easily show $a = a' \in \NAT(\alpha)$ for some $\alpha < \infty$.
+  For instance, $\alpha$ could be number of uses of the successor rule plus one.
+\end{proof}
+
+%A type constructor $B \in \FixK\,\ell$ is \emph{admissible for recursion}, \fbox{$B \in \ADM$}, if
+%$\bigcap_{
+Let us write \fbox{$B = B' \in \Adm\ell$} iff $B = B' \in \FixK\,\ell$ and
+for all $\beta \in \SIZE$ and $a \in \Nat \beta$ we have
+$B\,\beta\, a \leq B\,\infty\,a \in \TYPE$ and
+$B'\,\beta\, a \leq B'\,\infty\,a \in \TYPE$.
+\begin{lemma}[Fix]
+Let $g = a \, \fix[\ell] B f$ and $g' = a'\,\fix[\ell]{B'}{f'}$.
+If $a = a' \in \Nat \alpha$
+and $B = B' \in \Adm\ell$
+and $f = f' \in \FixT\,B$
+then $g = g' \in B\,\alpha\,a$.
+\end{lemma}
+\begin{proof*}
+By wellfounded induction on $\alpha$.
+\begin{caselist}
+
+\nextcase $\alpha < \infty$ and $a \evalsto \up{T}n$ and $a' \evalsto \up{T'}{n'}$ and $n = n' \in \NE$.
+In this case $g$ and $g'$ evaluate to neutral applications of $\tfix$.  The proof proceeds analogously to Lemma~\ref{lem:case}.
+
+\nextcase $\alpha < \infty$ and $a \evalsto c$ and $a' \evalsto c'$.  Then $\alpha = \beta + 1$ with $\beta<\infty$.
+The weak head normal form of $g$ equals the weak head normal form of
+$h := f\,\ann \gamma\,(\lambda x.\,x\,\fix[\ell] B f)\,c$ where $\gamma$ is the size annotation of $c$.
+It suffices to show $h = h' \in B\,\alpha\,a$ for $h'$ defined analogously from $B',f',c'$.
+This follows from the assumption on $f, f'$ if we manage to show
+$(\lambda x.\,x\,\fix[\ell] B f) = (\lambda x.\,x\,\fix[\ell] {B'} {f'}) \in
+\funT x {\Nat\beta} {B\,\beta\,x}$.
+To this end, assume $b = b' \in \Nat\beta$ and show
+$b\,\fix[\ell] B f = b'\,\fix[\ell] {B'} {f'} \in B\,\beta\,b$.
+However, this is an instance of the induction hypothesis thanks to $\beta < \alpha$.
+
+\nextcase $\alpha = \infty$.
+Note that $a = a' \in \Nat \alpha$ for some $\alpha < \infty$ by Lemma~\ref{lem:limit}.
+By induction hypothesis, $g = g' \in B\,\alpha\,a$.  Since
+$B\,\alpha\,a \leq B\,\infty\,a$ by assumption, the goal
+$g = g' \in B\,\infty\,a$ follows by subsumption.
+\end{caselist}
+\end{proof*}
 

### 2017-02-25 nbe/icfp17.tex
[ ICFP 17 ] proof* without \qed

@@
 \end{theorem}
-\begin{proof}
+\begin{proof*}
 By induction on $T \in \SET[\ell]$ and cases on $T \shape S_1$ and $T \shape S_2$.
@@
   $\R k {\down{S_i} t_i} {\lambda v_i}$ by definition of read back.
-
+\qed
 \end{enumerate}
@@
 \end{caselist}
-\end{proof}
+\end{proof*}
 
@@
 \end{lemma}
-\begin{proof}
+\begin{proof*}
 By induction on $a = a' \in \Nat(\alpha+1)$.
@@
 The remaining goal $\case[\ell] D {d_z}{d_s} = \case[\ell] {D'} {d_z'}{d_s'} \in \ELIM$ follows by the closure properties for eliminations (Lemma~\ref{lem:closelim}), since $D = D' \in \NF$ and $d_z = d_z' \in \NF$ and $d_s = d_s' \in \NF$ all hold by reification (Theorem~\ref{thm:rere}).
+\qed
 \end{caselist}
-\end{proof}
+\end{proof*}
 
@@
 $g = g' \in B\,\infty\,a$ follows by subsumption.
+\qed
 \end{caselist}
@@
 \end{theorem}
-\begin{proof}
+\begin{proof*}
 Simultaneously, by induction on the derivation.
@@
 thus, by induction hypothesis, $b \rho \in \SIZE$.
+\qed
 \end{caselist}
-\end{proof}
+\end{proof*}
 
@@
 \end{lemma}
-\begin{proof}
+\begin{proof*}
 By induction on $\der \Gamma$.
@@
 $(\rho_\Gamma, \var\Gamma) \in \eext\Gamma \Size)$.
+\qed
 \end{caselist}
-\end{proof}
+\end{proof*}
 
@@
 
-\begin{proof}
+\begin{proof*}
 We prove the propositions for $A'$ and $T'$ (instead of $A$ and $T$)
@@
 % assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
+\qed
 \end{enumerate}
 \end{caselist}
-\end{proof}
+\end{proof*}
 
@@
 \end{theorem}
-\begin{proof}
+\begin{proof*}
 Each by induction on the derivation.
@@
 \item $\LRSize {\resurrect \Delta} {b\tau} {b\rho}$: follows from induction hypothesis on $\resurrect \Gamma \der b : \Size$ via $\LRSub {\resurrect \Delta} \tau \tau {\resurrect\Gamma} \rho \rho$ (from Lemma~\ref{lem:wklrsub}).
-\end{enumerate}
-
+\qed
+\end{enumerate}
 \end{caselist}
-\end{proof}
+\end{proof*}
 
@@
 \end{lemma}
-\begin{proof}
+\begin{proof*}
   By induction on $\der \Gamma$.
@@
 % $\LR {\cext \Gamma T} {\ind0} T f {T\rhog}$.
-
-\nextcase
-
+\qed
 \end{caselist}
-\end{proof}
+\end{proof*}
 
@@
 \end{corollary}
-\begin{proof}
+\begin{proof*}
 \bla
@@
    we erase all size annotations $\ann\alpha$ to $\ann\infty$ during readback.
-  \end{enumerate}
-\end{proof}
+\qed
+\end{enumerate}
+\end{proof*}
 
@@
 \end{lemma}
-\begin{proof}
+\begin{proof*}
 By induction on % $A_1' \in s_1$ and $A_2' \in s_2$ and cases on the inductive relations
@@
 Thus by induction hypothesis on $B_i[a] \in s_i$ we obtain $\cext \Gamma {U_2} \der T_1 \leq T_2$, together $\Gamma \der T_1' \asub T_2'$.
-
+\qed
 \end{caselist}
-\end{proof}
+\end{proof*}
 

### 2017-02-25 nbe/icfp17.tex
[ ICFP 17 ] Introduction up to mentioning ICC.

@@
 
-\newpage
+% higher rank sized types
+% Church-style and curry style
+% The problem of size irrelevance
+% Our solution (application rule)
+% Shape-irrelevance middle ground between Pfenning, ICC,
+% Agda's function types
+
+Dependently-typed programming languages and proof assistants, such as Agda~\citeyearpar{agdawiki}, Coq~\cite{inria:coq86}, and Idris~\cite{brady:jfp13} require programs to be total, for two reasons.
+First, for consistency:  since propositions are just types and proofs of a propositions just programs which inhabit the corresponding type, some types need to be empty; otherwise, each proposition would be true.
+However, in a partial language with general recursion, each type is inhabited by the looping program $f = f$.
+Secondly, totality is needed for decidability of type checking.
+Since types can be the result of a computation, we need computation to terminate during type checking, even for \emph{open terms}, \ie, terms with free variables.
+
+% \cite{boveKraussSozeau:mscs16}
+Consequently, each of the aforementioned languages based on Type Theory comes with a termination checker, which needs to reject all non-terminating programs, and should accept sufficiently many terminating programs to allow the user to express her algorithms.
+In the current termination checkers, programs are required to terminate by structural descent \cite{gimenez:guardeddefinitions}; the structural order may be extended to a lexicographic \cite{abelAltenkirch:jfp02} or size-change termination criterion \cite{jones:sizeChange,wahlstedt:PhD}.
+This fundamental limitation, since Type Theory allows many functions to be expressed in a structurally recursive manner, if needed by the help of a well-founded relation \cite{nordstrom:genrec}, inductive domain predicates \cite{boveCapretta:mscs05}, or inductive descriptions of the function graph \cite{bove:entcs09}.
+However, the syntactic termination check is very sensitive to reformulations of a program and hostile to abstraction \cite{abel:fics12}.
+
+Sized types \cite{hughesParetoSabry:popl96} delegate the checking for structural descent to the type system by annotating data types with a size parameter.
+The type checker can then ensure that in recursive calls, the size goes down, certifying termination.
+In the simplest setting \cite{gimenez:typeBased,abel:lmcs07}, the size is just an upper bound on the tree height of the data structure; however, more sophisticated size annotations have also been considered \cite{xi:terminationHOSC,blanqui:rta04}.
+Most sized type systems are non-dependent \cite{amadio:guardcondition,abelPientka:jfp16,bartheGregoireRiba:lernet08,bartheGregoireRiba:csl08,blanquiRiba:lpar06,dalLagoGrellois:probabilisticSizedTypes2017}, yet the combination of sized and dependent types has been studied as well \cite{blanqui:csl05,bartheGregoirePastawski:lpar06,gregoireSacchini:lpar10,sacchini:lics13,sacchini:flops14}.
+However, to the best of our knowledge, no study combines \emph{higher-rank size polymorphism} with full-fledged dependent types (Xi \citeyearpar{xi:terminationHOSC} only has indexed types).
+
+Higher-rank size quantification takes termination checking to the next level; it is necessary for abstraction and generic programming.
+For instance, it allows us to write a generic tree traversal which applies a user-given preprocessor on subtrees before recursively descending into these trees.
+The condition is that preprocessing does not increase the size of the subtree; otherwise, termination could not be guaranteed.
+Concretely, assume a type $\tT\,i$ of trees of size $< i$ with a constructor $\tnode : \forall i.\, \List\,(\tT\,i) \to \tT\,(i+1)$.
+In the following definition of $\ttrav$, the preprocessing $\vpre : \forall i.\, \tT\,i \to \tT\,i$ can be safely applied to input tree $t$ because the type of $\vpre$ bounds the size of $\vpre\,t$ by the size of $t$.
+In case $\vpre\,t = \tnode\,\vts$, the trees in the list $\vts$ are still guaranteed to be of strictly smaller size than $t$, thus, the recursive call to $\ttrav$, communicated via the $\tmap$ function for lists, is safe.
+\[
+\begin{array}{l}
+  \ttrav : (\vpre : \forall i.\, \tT i \to \tT i) (\vpost : \tT \infty \to \tT \infty) \to \forall i.\, \tT\,i \to \tT\,\infty
+  \\
+  \ttrav \; \vpre \; \vpost \; t = \vpost\;(\tcase\;\vpre\,t\;\tof \, \{ \, \tnode\,\vts \to \tnode\,(\tmap \; (\ttrav \, \vpre \, \vpost) \; \vts) \})
+\end{array}
+\]
+This is the \emph{Curry}-style program as accepted from the user, but state-of-the-art type checkers elaborate the program from surface syntax into an internal \emph{Church}-style syntax with explicit type abstractions and type applications.\footnote{Agda, Coq, Idris, and Haskell \cite{sulzmannChakravartyPeythonJonesDonnelly:tldi07} all have Church-style internal languages.}  With implicit type and size applications elaborated, $\ttrav$ would look as follows:
+\[
+\begin{array}{l}
+  \ttrav \; \vpre \; \vpost \; i\; t = \vpost\;(\tcase\;\vpre\,i\,t\;\tof \, \{ \, \tnode\,j\,\vts \to \tnode\,\infty\,(\tmap \; (\tT\,j) \; (\tT\,\infty) \; (\ttrav \, \vpre \, \vpost \, j) \; \vts) \})
+\end{array}
+\]
+From a dependent-type perspective, explicit size applications in terms can be problematic when the type checker compares terms for equality (which is necessary as types can depend on values).
+Inferred sizes may not be unique, as we have subtyping $\tT\,i \leq \tT\,j$ for $i \leq j$: we can always weaken an upper bound.
+For instance, given $\vts : \List\,(\tT\,i)$, all of $\tnode\,i\,\vts$ and $\tnode\,(i+1)\,\vts$ and $\dots$ and $\tnode\,\infty\,\vts$ have type $\tT\,\infty$.
+However, semantically, all these trees are equal, thus, the syntactic equality check should ignore the size argument to $\tnode$.
+Similarly, in the application $\vpre\,i\,t$ the size argument $i$ should be ignored by the equality check.
+Yet $\vpre\,i : \tT\,i \to \tT\,i$ and $\vpre\,j : \tT\,j \to \tT\,j$ have different types for $i \not= j$, and moreover these function types are not in the subtyping relation due to mixed-variant occurrence of the size parameter.
+It seems that during equality checking we have to consider terms of different types, for a while.
+Once we apply $\vpre\,i$ and $\vpre\,j$ to the same tree $t : \tT\,k$, which determines $i = j = k$, we are back in safety.
+However, allowing types to differ during an equality check can be hazardous, especially when the equality-check is type directed.
+
+Consider the analogous situation for the polymorphic lambda calculus System~F, be it the predicative variant or not, extended by a unit type $\One$.
+For Church-style, we can give a type-directed $\beta\eta$-equality test which equates all term at the unit type.
+The most interesting rules would be the $\eta$-rules for unit and function type and the congruence rule for type application:
+\begin{gather*}
+\renewcommand{\jchk}{:}
+\renewcommand{\jinf}{:}
+  \ru{}{\Gamma \der t = t' \jchk \One}
+\qquad
+  \ru{\Gamma, x \of A \der t\,x = t'\,x \jchk B
+    }{\Gamma \der t = t' \jchk A \to B}
+\qquad
+  \ru{\Gamma \der t = t' \jinf \forall X. B
+    }{\Gamma \der t\,A = t'\,A \jinf B[A/X]}
+\end{gather*}
+The Curry-style version would replace the last conclusion by $\Gamma \der t = t' : B[A/X]$ where the type $A$ to instantiate $X$ has to be guessed.
+However, in Curry-style more terms are equated than in Church-style, as for instance the Church-style terms $t\,A\,(\lambda x \of A.\, x)$ and $t\,B\,(\lambda x \of B.\, x)$ map to the same Curry-style term $t\,(\lambda x.\,x)$.
+How would we adopt the algorithm for Church-style such that it equates all terms that are equal in Curry-style?
+The conclusion of the last rule could be changed to $\Gamma \der t\,A = t'\,A' : B[A/X]$, but then the second term $t'\,A'$ does not have the ascribed type $B[A/X]$, and $\eta$-laws applied to this term would be unsound.
+For instance, the algorithm would yield $t\,\One\,x = t\,(A \to A)\,y$ even for $x \not= y$.
+We could also consider a heterogeneous check $(\Gamma \der t : A) = (\Gamma' \der t' : A')$ where each term is paired with its own type and context, but this leaves us with the dilemma of explaining the meaning of this judgement when $A$ and $A'$ are incompatible.
+
+Does the literature offer a solution to this problem?
+In fact, a Church-style calculus with Curry-style equality has been studied before, it is ICC$^*$ \cite{barrasBernardo:fossacs08} based on Miquel's Implicit Calculus of Constructions \cite{miquel:tlca01}.
 

### 2017-02-26 nbe/icfp17.tex
[ ICFP 17 ] On ICC*.

@@
 In fact, a Church-style calculus with Curry-style equality has been studied before, it is ICC$^*$ \cite{barrasBernardo:fossacs08} based on Miquel's Implicit Calculus of Constructions \cite{miquel:tlca01}.
+In ICC$^*$, equality is checked by erasing all type abstractions and applications, and comparing the remaining untyped terms for $\beta\eta$-equality.
+While this works for $\eta$-laws that can be formulated on untyped terms, such as $\eta$-contraction of functions $\lambda x.\,t\,x \redeta t$ (when $x$ not free in $t$), it does not extend to type-directed $\eta$-laws such as extensionality for the unit type.
+
+It has already been observed that for a type-directed equality check the precise type is not necessary, a \emph{shape} or \emph{skeleton} is sufficient that informs the algorithm whether the terms under comparison are functions, inhabitants of the unit type, or something else.
+For the Logical Framework (LF), the simplest dependent lambda-calculus, the skeletons are simple types that can be obtained from the original dependent types by erasing all dependencies.
+\citet{harperpfenning:equivalenceLF} present such an equality check for LF which is directed by simple types, and their technique should scale to all type theories that admit dependency erasure such as the Calculus of Constructions \cite{geuvers:shortFlexibleSNCC}.
 

### 2017-02-26 prototyping/Sit/src/InternalToAbstract.hs
[ Sit ] moved to new directory Sit/src

@@ -1,0 +1,4 @@
+
+-- | Print internal syntax to abstract syntax.
+
+module InternalToAbstract where

### 2017-02-26 prototyping/Sit/src/Abstract.hs
[ Sit ] fixed parsing of (F x) -> A

@@ -7,6 +7,7 @@
 module Abstract where
 
 import Control.Lens
+import Control.Monad
 
 import Sit.Abs as A
 
@@ -70,3 +71,12 @@ fromIdU :: A.IdU -> String
 fromIdU = \case
   A.Id (A.Ident x) -> x
   A.Under -> "_"
+
+-- | Try to convert an expression to a list of A.IdU
+
+parseIdUs :: A.Exp -> Maybe [A.IdU]
+parseIdUs e = do
+  let (f, es) = appView e
+  forM (f : es) $ \case
+    A.Var x -> Just x
+    _ -> Nothing

### 2017-02-26 prototyping/Sit/src/TypeChecker.hs

@@ -155,7 +155,10 @@ inferType e = do
       (t, l2) <- inferType b
       return (Pi (defaultDom u) (NoAbs "_" t) , maxSize l1 l2)
 
-    A.Pi xs a b -> inferPisType (map (, defaultDom a) xs) $ inferType b
+    A.Pi e a b -> do
+      let failure = throwError $ "Expected list of identifiers, found " ++ printTree e
+      xs <- maybe failure return $ parseIdUs e
+      inferPisType (map (, defaultDom a) xs) $ inferType b
 
     A.Forall bs c -> inferPisType (fromBind =<< bs) $ inferType c
       where
@@ -316,12 +319,17 @@ inferExp e0 = case (e0, appView e0) of
 
   (A.Infty, _) -> return (Infty, VSize)
 
-  (A.Plus x k, _) -> do
-    (u, t) <- inferId x
-    subType t VSize
-    return (sPlus u k, t)
+  (A.Plus e k, _) -> do
+    u <- checkSize e
+    return (sPlus u k, VSize)
 
-  (A.Var x, _) -> inferId x
+  -- (A.Plus x k, _) -> do
+  --   (u, t) <- inferId x
+  --   subType t VSize
+  --   return (sPlus u k, t)
+
+  (A.Var A.Under, _) -> throwError "Illegal expression: _"
+  (A.Var (A.Id x), _) -> inferId x
 
   (e0@(A.App f e), _) -> do
     (tf, t) <- inferExp f

### 2017-02-26 prototyping/Sit/src/Abstract.hs
[ Sit ] Switched from lens to data-lens-light.

@@ -1,15 +1,17 @@
 {-# LANGUAGE LambdaCase #-}
 
+{-# OPTIONS_GHC -fno-warn-typed-holes #-}
 {-# OPTIONS_GHC -fwarn-incomplete-patterns #-}
 
 -- | Auxiliary functions for abstract syntax.
 
 module Abstract where
 
-import Control.Lens
 import Control.Monad
 
 import Sit.Abs as A
+
+import Lens
 
 -- data AppView
 --   = FixV [Exp]

### 2017-02-26 prototyping/Sit/src/Evaluation.hs

@@ -10,8 +10,8 @@
 module Evaluation where
 
 import Control.Applicative
-import Control.Lens hiding (Index, Level)
 import Control.Monad
+import Control.Monad.Identity
 import Control.Monad.Reader
 
 import Data.Maybe
@@ -22,6 +22,7 @@ import Debug.Trace
 import Internal
 import Substitute
 
+import Lens
 import Impossible
 #include "undefined.h"
 
@@ -76,7 +77,7 @@ data VClos = VClos
 
 type Env = [Val]
 
-makeLenses ''VNe
+makeLens ''VNe
 
 -- | Variable
 

### 2017-02-26 prototyping/Sit/src/Internal.hs

@@ -8,10 +8,10 @@
 
 module Internal where
 
-import Control.Lens hiding (Level, Index)
-
 import Data.Foldable (Foldable)
 import Data.Traversable (Traversable)
+
+import Lens
 
 -- | Definition names are strings.
 
@@ -118,7 +118,7 @@ data Relevance
   | Irrelevant
   deriving (Eq, Ord, Show)
 
-makeLenses ''Dom
+makeLens ''Dom
 
 -- * Smart constructor.
 

### 2017-02-26 prototyping/Sit/src/TypeChecker.hs

@@ -7,7 +7,6 @@
 module TypeChecker where
 
 import Control.Applicative
-import Control.Lens hiding (Level)
 import Control.Monad
 import Control.Monad.Except
 import Control.Monad.Identity
@@ -31,6 +30,7 @@ import Internal
 import Substitute
 import Evaluation
 
+import Lens
 import Impossible
 #include "undefined.h"
 
@@ -46,7 +46,7 @@ data TCEnv = TCEnv
   , _envEnv :: Env  -- ^ Default environment.
   }
 
-makeLenses ''TCEnv
+makeLens ''TCEnv
 
 -- | Global state
 
@@ -55,7 +55,7 @@ data TCSt = TCSt
   , _stDefs   :: Map Id Val
   }
 
-makeLenses ''TCSt
+makeLens ''TCSt
 
 -- | The type checking monad
 type Check = ReaderT TCEnv (StateT TCSt (Except TypeError))

### 2017-02-26 prototyping/Sit/src/Evaluation.hs
[ Sit ] type annotation needed for ghc-8.

@@ -351,8 +351,9 @@ readbackNat = \case
 
 readbackNe  :: MonadEval m => VNe -> ReaderT Int m Term
 readbackNe (VNe x es) = do
-  i <- readback x
-  foldl App (Var i) <$> readback es
+  i            <- readback x
+  es' :: Elims <- readback es
+  return $ foldl App (Var i) es'
 
 readbackSize  :: MonadEval m => Val -> ReaderT Int m Term
 readbackSize = \case

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Literate Agda code.

@@
 \usepackage[cal=boondoxo]{mathalfa}
+\usepackage{latex/agda}
 
@@
 For the Logical Framework (LF), the simplest dependent lambda-calculus, the skeletons are simple types that can be obtained from the original dependent types by erasing all dependencies.
-\citet{harperpfenning:equivalenceLF} present such an equality check for LF which is directed by simple types, and their technique should scale to all type theories that admit dependency erasure such as the Calculus of Constructions \cite{geuvers:shortFlexibleSNCC}.
+\citet{harperpfenning:equivalenceLF} present such an equality check for LF which is directed by simple types, and their technique should scale to all type theories that admit dependency erasure.
+\footnote{For instance, the types of the Calculus of Constructions erase to F$^\omega$-types \cite{geuvers:shortFlexibleSNCC}, and the latter could be used to direct the equality check.}
+In the presence of large eliminations, which we have in type theories that feature both universes and data types, dependency erasure fails, and it is not clear what the skeleton of a dependent type should be.
+
+\section{Informal presentation}
+\label{sec:informal}
+
+\input{latex/SubDiagError}
 
 \section{Syntax}
+\label{sec:syntax}
 

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Agda code which passes.
[ ICFP 17 ] Finished introduction.

--- 2017-02-26 nbe/icfp17.tex
@@
 \usepackage[cal=boondoxo]{mathalfa}
+
+\usepackage{latex/agda}
+
+% NO EFFECT:
+% \renewcommand{\AgdaOperator}    [1]
+%     {\AgdaNoSpaceMath{\AgdaFontStyle{\textcolor{AgdaOperator}{\mathbf{#1}}}}}
 
@@
 \author{Andreas Abel}
-\authornote{with author1 note}          %% \authornote is optional;
+%\authornote{with author1 note}          %% \authornote is optional;
                                         %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\orcid{0000-0003-0420-4492}             %% \orcid is optional
 \affiliation{
@@
 %% Author with two affiliations and emails.
-\author{First2 Last2}
-\authornote{with author2 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\author{Andreas Vezzosi}
 \affiliation{
-  \position{Position2a}
-  \department{Department2a}             %% \department is recommended
-  \institution{Institution2a}           %% \institution is required
-  \streetaddress{Street2a Address2a}
-  \city{City2a}
-  \state{State2a}
-  \postcode{Post-Code2a}
-  \country{Country2a}
+  \department{Department of Computer Science and Engineering}
+  \institution{Chalmers}
+  \streetaddress{Rännvägen 6b}
+  \city{Göteborg}
+%  \state{State1}
+  \postcode{41296}
+  \country{Sweden}
 }
-\email{first2.last2@inst2a.com}         %% \email is recommended
+\email{vezzosi@chalmers.se}          %% \email is recommended
+
+\author{Theo Winterhalter}
 \affiliation{
-  \position{Position2b}
-  \department{Department2b}             %% \department is recommended
-  \institution{Institution2b}           %% \institution is required
-  \streetaddress{Street3b Address2b}
-  \city{City2b}
-  \state{State2b}
-  \postcode{Post-Code2b}
-  \country{Country2b}
+  \institution{École Normale Supérieure de Cachan}
 }
-\email{first2.last2@inst2b.org}         %% \email is recommended
+\email{theo.winterhalter@gmail.com}
 
@@
 <concept>
-<concept_id>10011007.10011006.10011008</concept_id>
-<concept_desc>Software and its engineering~General programming languages</concept_desc>
+<concept_id>10003752.10003790.10011740</concept_id>
+<concept_desc>Theory of computation~Type theory</concept_desc>
 <concept_significance>500</concept_significance>
@@
 <concept>
-<concept_id>10003456.10003457.10003521.10003525</concept_id>
-<concept_desc>Social and professional topics~History of programming languages</concept_desc>
+<concept_id>10003752.10010124.10010125.10010130</concept_id>
+<concept_desc>Theory of computation~Type structures</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10010124.10010138.10010142</concept_id>
+<concept_desc>Theory of computation~Program verification</concept_desc>
 <concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10010124.10010131.10010134</concept_id>
+<concept_desc>Theory of computation~Operational semantics</concept_desc>
+<concept_significance>100</concept_significance>
 </concept>
@@
 
-\ccsdesc[500]{Software and its engineering~General programming languages}
-\ccsdesc[300]{Social and professional topics~History of programming languages}
+\ccsdesc[500]{Theory of computation~Type theory}
+\ccsdesc[500]{Theory of computation~Type structures}
+\ccsdesc[300]{Theory of computation~Program verification}
+\ccsdesc[100]{Theory of computation~Operational semantics}
 %% End of generated code
@@
 %% comma separated list
-\keywords{keyword1, keyword2, keyword3}  %% \keywords is optional
+\keywords{dependent types, normalization-by-evaluation, proof irrelevance, sized types, subtyping, universes}  %% \keywords is optional
+%\keywords{Dependent Types, Normalization-by-Evaluation, Proof Irrelevance, Sized Types, Subtyping, Universes}  %% \keywords is optional
 
@@
 Does the literature offer a solution to this problem?
-In fact, a Church-style calculus with Curry-style equality has been studied before, it is ICC$^*$ \cite{barrasBernardo:fossacs08} based on Miquel's Implicit Calculus of Constructions \cite{miquel:tlca01}.
+In fact, a Church-style calculus with Curry-style equality has been studied before, it is ICC$^*$ \cite{barrasBernardo:fossacs08,mishraLingerSheard:fossacs08} based on Miquel's Implicit Calculus of Constructions \cite{miquel:tlca01}.
 In ICC$^*$, equality is checked by erasing all type abstractions and applications, and comparing the remaining untyped terms for $\beta\eta$-equality.
 While this works for $\eta$-laws that can be formulated on untyped terms, such as $\eta$-contraction of functions $\lambda x.\,t\,x \redeta t$ (when $x$ not free in $t$), it does not extend to type-directed $\eta$-laws such as extensionality for the unit type.
-
+Further, ICC$^*$ is not a type theory formulated with a typed equality judgement, which makes it hard to define its models \cite{miquel:lics00}---we wish not to go there, but stay within the framework of Martin-L\"of Type Theory \cite{martinlof:predicative75}.
+
+How can we have our cake and eat it, too?
 It has already been observed that for a type-directed equality check the precise type is not necessary, a \emph{shape} or \emph{skeleton} is sufficient that informs the algorithm whether the terms under comparison are functions, inhabitants of the unit type, or something else.
 For the Logical Framework (LF), the simplest dependent lambda-calculus, the skeletons are simple types that can be obtained from the original dependent types by erasing all dependencies.
-\citet{harperpfenning:equivalenceLF} present such an equality check for LF which is directed by simple types, and their technique should scale to all type theories that admit dependency erasure such as the Calculus of Constructions \cite{geuvers:shortFlexibleSNCC}.
+\citet{harperpfenning:equivalenceLF} present such an equality check for LF which is directed by simple types, and their technique should scale to all type theories that admit dependency erasure.\footnote{For instance, the types of the Calculus of Constructions erase to F$^\omega$-types \cite{geuvers:shortFlexibleSNCC}, and the latter could be used to direct the equality check.
+\cite{lovasPfenning:lmcs10} consider also \emph{refinement types for logical frameworks} which can be erased to simple types.}
+In the presence of large eliminations, which we have in type theories that feature both universes and data types, dependency erasure fails, and it is not clear what the skeleton of a dependent type should be.
+For instance consider the type $\funT n \NN \underbrace{A \to \dots \to A}_n \to A$ of $n$-ary functions; its shape is dependent on the value of $n$, thus cannot be determined statically.
+In the solution we present in this article, we do not introduce a new syntax for shapes, just a relation $T \shape S$ stating that type $S$ qualifies as shape for type $T$.
+We use a normalization function inspired by normalization-by-evaluation (NbE) that computes the $\eta$-long normal form of the terms under comparison, which we then check for identity.
+NbE \cite{bergerSchwichtenberg:lics91,danvy:tdpe} has proven a robust method to decide equality in powerful type theories with non-trivial $\eta$-laws.
+It scales to universes and large eliminations \cite{abelCoquandDybjer:lics07}, topped with singleton types or proof irrelevance \cite{abelCoquandPagano:lmcs11}, and even impredicativity \cite{abel:flops10}.
+At its hard there are reflection $\up T$ and reification $\down T$ functions directed by type $T$ and orchestrating just-in-time $\eta$-expansion.
+Reflection $\up T x$ maps variables $x$ into the realm of values of type $T$ and let's us compute with open terms.
+Reification $\down T a$ takes a value $a$ of type $T$ and computes its long normal form.
+For instance, the normal form of a closed function $f : U \to T$ would be $\lambda x.\,\down{T} (f\,(\up U x))$, and for its dependently-typed variant $f : \funT x U T[x]$ it would be $\lambda x.\,\down{T[\up U x]} (f\,(\up U x))$.
+
+The central technical observation is that reflection and reification do not need the precise type $T$, they work the same for any shape $S$ of $T$.
+For example, when reflecting a variable $x$ into the polymorphic function type $\forall i.\, F\,i$ where $F\,i = \Nat\,i \to \Nat\,i$, we obtain
+$(\up{\forall i.\/ F\,i}x)\,i = \up{F\,i} (x\,i)$ for size $i$ and
+$(\up{\forall i.\/ F\,i}x)\,j = \up{F\,j} (x\,j)$ for size $j$.
+The new types $F\,i$ and $F\,j$ we reflect at are no longer equal (and they are not subtypes of each other), but they still have the same shape, $\Nat\_ \to \Nat\_$.
+This means they will still move in lock-step what $\eta$-expansion concerns, which is sufficient to prove NbE correct for judgmental equality.
+We call the enabling property of $F$ \emph{shape irrelevance}, meaning that for any pair $i$, $j$ of legal arguments, $F\,i$ and $F\,j$ have the same shape.
+Whenever we form a irrelevant function type $\forall x \of U.\, T[x]$, we require $T[x]$ to be shape-irrelevant in $x$.
+This is the middle ground between ICC$^*$, where no restriction is placed on $T$ but $\eta$ for unit types is (at least, currently) out of reach, and Pfenning's \citeyearpar{pfenning:intextirr} irrelevance modality, adapted to full dependent types by \citet{abelScherer:types10}, which requires $T$ to be irrelevant in $x$ and, thus, has type equality $T[i] = T[j]$.
+
+For the time being, we do not (and cannot) develop a general theory of shape irrelevance.
+We confine ourselves to size-irrelevant function types $\forall i.\, T[i]$.
+This relieves us from defining a special shape-irrelevance modality, since all size-indexed types $T[i]$ are shape irrelevant in $i$, simply because there is no case distinction on size, and sizes appear relevantly only under a sized type constructor such as $\tNat$.
+Our technique would not extend to the polymorphic types $\forall X.\,B[X]$ of System~F.
+Even though there is no case distinction on types, shape irrelevance of $B[X]$ fails in general, as $X$ could appear as a type on the top-level, e.g.
+in $B[X] = X$ or $B[X] = X \to X$.
+
+This article makes the following novel contributions:
+\begin{enumerate}
+
+\item We present a dependent type theory \`a la Martin-L\"of with dependent function types, cumulative universes, subtyping, a judgmental equality with $\eta$-laws, a sized type of natural numbers and two size quantifiers: an irrelevant one $(\forall)$ for binding of sizes in irrelevant positions, and a relevant one $(\Pi)$ for binding of sizes in shape-irrelevant positions (Section~\ref{sec:syntax}).
+Judgmental equality features a ``Curry-style'' rule for irrelevant size application which ignores the size arguments, and consequently, the corresponding typing rule will also ignore the size argument.
+\[
+\ru{\Gamma \der t = t' : \forall i.\, T
+  }{\Gamma \der t\,a = t\,a' : T[b/i]}
+\qquad
+\ru{\Gamma \der t : \forall i.\, T
+  }{\Gamma \der t\,a : T[b/i]}
+\qquad
+\qquad
+\ru{\Gamma \der t = t' : \Pi i.\, T
+  }{\Gamma \der t\,a = t\,a : T[a/i]}
+\qquad
+\ru{\Gamma \der t : \Pi i.\, T
+  }{\Gamma \der t\,a : T[a/i]}
+\]
+
+\item We adapt normalization-by-evaluation (NbE) to sized types and size quantification and show that it decides judgmental equality (sections~\ref{sec:sem} and \ref{sec:soundnbe}).
+The novel technical tool is a relation $T \shape S$ which relates a type $T$ to its possible shapes $S$.
+This approximation relation allows reflection and reification at size-polymorphic types $\forall i.\, T$.
+As usual for the meta-theory of Type Theory with large eliminations, the machinery is involved, but just require the usual two logical relations:  First, a PER model to define the semantics of types and prove the completeness of NbE (Section~\ref{sec:sem}).
+Secondly, a relation between syntax and semantics to prove soundness of NbE (Section~\ref{sec:soundnbe}).
+
+\item We present an bidirectional type checking algorithm \cite{coquand:type} which takes the irrelevant size argument as reliable hint for the type checker (sections~\ref{sec:asub} and \ref{sec:tycheck}).
+It is complete for normal forms which can be typed with the restricted rule for $\forall$-elimination:
+\[
+\ru{\Gamma \der t : \forall i.\, T
+  }{\Gamma \der t\,a : T[a/i]}
+\]
+The algorithm employs the usual lazy reduction for types (just-in-time weak-head evaluation) in type and sub-type checker
+\cite{huet:wsscs89}.  In this, it improves on \citet{fridlenderPagano:tlca13} which instruments full normalization (NbE) at every step.
+\end{enumerate}
+This article is accompanied by a prototypical type checker \texttt{Sit} which implements the type system and type checking algorithm as described in the remainder of the paper.
+But first, we will motivate our type system from a practical perspective: reasoning about programs involving sized types in Agda.
+
+\section{Size irrelevance in practice}
+\label{sec:informal}
+
+\input{latex/SubDiagError}
+
+\input{latex/SubDiag}
 
 \section{Syntax}
+\label{sec:syntax}
 
@@
 \section{Soundness  of NbE}
-\label{sec:compl}
+\label{sec:soundnbe}
 

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Some polishing around the Agda code.
[ ICFP 17 ] Acknowledgements (not printed).

@@
 
-\section{Size irrelevance in practice}
+\section{Size Irrelevance in Practice}
 \label{sec:informal}
 
+In this section, we show how the lack of size irrelevance prevents us from reasoning naturally about programs involving sized types in Type Theory.
+We focus on Agda, the only mature implementation of Type Theory with an experimental integration of sized types.
+
 \input{latex/SubDiagError}
@@
 
-\section{Syntax}
+\section{A Type System WIth Irrelevant Size Application}
 \label{sec:syntax}
@@
 
-\section{Semantics}
+\section{Semantics and Completeness of Normalization by Evaluation}
 \label{sec:sem}
@@
 
-\section{Soundness  of NbE}
+\section{Soundness  of Normalization by Evaluation}
 \label{sec:soundnbe}
@@
 
-\section{Algorithmic subtyping}
+\section{Algorithmic Subtyping}
 \label{sec:asub}
@@
   This material is based upon work supported by the
-  \grantsponsor{GS100000001}{National Science
-    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-  conclusions or recommendations expressed in this material are those
-  of the author and do not necessarily reflect the views of the
-  National Science Foundation.
+  Swedish Research Council (Vetenskapsr\aa{}det)
+  under Grant
+  No.~621-2014-4864 \emph{Termination Certificates for Dependently-Typed Programs and Proofs via Refinement Types}.
+  The first author is grateful for recent discussions with Thierry Coquand, Nils Anders Danielsson, and Sandro Stucki
+  which helped clarifying the thoughts leading to this work.
+  He also acknowledges past discussions with Christoph-Simon Senjak.
 \end{acks}
@@
 
-%% Appendix
-\appendix
-\section{Appendix}
-
-Text of appendix \ldots
+% %% Appendix
+% \appendix
+% \section{Appendix}
+
+% Text of appendix \ldots
 

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Conclusions.

@@
 
-\item We present a dependent type theory \`a la Martin-L\"of with dependent function types, cumulative universes, subtyping, a judgmental equality with $\eta$-laws, a sized type of natural numbers and two size quantifiers: an irrelevant one $(\forall)$ for binding of sizes in irrelevant positions, and a relevant one $(\Pi)$ for binding of sizes in shape-irrelevant positions (Section~\ref{sec:syntax}).
+\item We present the first integration of a dependent type theory with higher-rank size polymorphism.
+Concretely, we consider a type theory \`a la Martin-L\"of with dependent function types, cumulative universes, subtyping, a judgmental equality with $\eta$-laws, a sized type of natural numbers and two size quantifiers: an irrelevant one $(\forall)$ for binding of sizes in irrelevant positions, and a relevant one $(\Pi)$ for binding of sizes in shape-irrelevant positions (Section~\ref{sec:syntax}).
 Judgmental equality features a ``Curry-style'' rule for irrelevant size application which ignores the size arguments, and consequently, the corresponding typing rule will also ignore the size argument.
@@
 
-\section{Conclusions}
+\section{Discussion and Conclusions}
 \label{sec:concl}
+
+In this article, we have described the first successful integration of higher-rank size polymorphism into a core type theory with dependent function types, a sized type of natural numbers, a predicative hierarchy of universes, subtyping, and $\eta$-equality.
+
+It is now straightforward to add a unit type $\One$ with extensional equality $t = {*} : \One$ for all $t : \One$.
+We simply extend reification such that $\down\One a = {*}$.
+Further, $\One$ is a new type shape with rule $\One \shape \One$.
+
+Reification erases all irrelevant size arguments to $\infty$.
+We also considered keeping the sizes, and define equality $v \approx v'$ of normal forms modulo differences in erased arguments.
+However, it seems difficult to prove that equal normal forms are judgmentally equal, \ie, if $\Gamma \der v,v' : T$ and $v \approx v'$ then $\Gamma \der v = v' : T$.
+Proving this statement seems to be the key to an incremental equality check; but
+the usual bidirectional argument \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types.
+In the long run, we wish for an incremental type-directed equality check, which is also at the heart of Agda's type checker, as this check can generate constraints for meta variables involved in type reconstruction \cite{norell:PhD}.
+
+For now, we have exploited shape-irrelevance of sized types, but this directly extends to universe levels.
+If we consider all universes as a single shape $\Set[\ell_1] \shape \Set[\ell_2]$, we can quantify over levels irrelevantly, as $\Set$ is a shape-irrelevant type constructor.
+This is a stepping stone for integrating universe cumulativity with Agda's explicit universe-polymorphism.
+If levels are no longer unique (because of subsumption), they will get in the way of proofs, analogously to sizes.
+With an irrelevant quantifier we can ignore levels where they do not matter.
+We will still respect them where they matter, thus, we keep consistency.
+
+The reflections on level irrelevance lead us to the question: can a type theory $T$ with a stratified universe hierarchy be understood as a sort of refinement of the inconsistent System~$U$ (Type:Type)?
+Intuitively, when checking two terms of $T$ for equality, could we ignore the stratification in the type $A$ which directs the equality check (thus, consider $A$ coming from $U$)?
+Such a perspective would put stratification in one pot with size assignment:  Size annotations and levels are both just annotations for the termination checker, but do not bear semantic relevance.
+We could switch the universe checker temporarily off as we do with the termination checker (see also the work of \citet{stumpSjobergWeirich:par10} on \emph{termination casts}).
+
+Finally, we would like a general theory of shape-irrelevance that extends beyond size-indexed types.
+For instance, any data type constructor could be considered shape-irrelevant in all its parameters.
+However, the current reification algorithm which erases irrelevant size arguments to $\infty$ would need some patching.
+\citet{abelCoquandPagano:lmcs11} erase all proofs of a proposition $P$ to a canonical element $\star$ which inhabits any proposition $P$ that is already inhabited.
+A similar technique could work in our setting and would be necessary unless we can justify the incremental equality check (see above).
 

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Abstract.

@@
 \begin{abstract}
-Text of abstract \ldots.
+Sized types have been developed to make termination checking more perspicuous, more powerful, and more modular by integrating termination into type checking.
+In dependently-typed proof assistants where proofs by induction are just recursive functional programs, the termination checker is an integral component of the trusted core, as validity of proofs depend on termination.
+However, a rigorous integration of full-fledged sized types into dependent type theory is lacking so far.
+Such an integration is non-trivial, as explicit sizes in proof terms might get in the way of equality checking, making terms appear distinct that should have the same semantics.
+
+In this article, we integrate dependent types and sized types with higher-rank size polymorphism, which is essential for generic programming and abstraction.
+We introduce a size quantifier $\forall$ which lets us ignore sizes in terms for equality checking, alongside with a second quantifier $\Pi$ for abstracting over sizes that do affect the semantics of types and terms.
+Judgemental equality is decided by an adaption of normalization-by-evaluation for our new type theory, which features \emph{type shape}-directed reflection and reification.
+It follows that subtyping and type checking of normal forms are decidable as well, the latter by a bidirectional algorithm.
 \end{abstract}
@@
 %% comma separated list
-\keywords{dependent types, normalization-by-evaluation, proof irrelevance, sized types, subtyping, universes}  %% \keywords is optional
+\keywords{dependent types, eta-equality, normalization-by-evaluation, proof irrelevance, sized types, subtyping, universes.}  %% \keywords is optional
 %\keywords{Dependent Types, Normalization-by-Evaluation, Proof Irrelevance, Sized Types, Subtyping, Universes}  %% \keywords is optional

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Polishing the flow of the introduction.

@@
 %% above the permission/copyright statement.
-\thanks{with paper note}                %% \thanks is optional
+%\thanks{with paper note}                %% \thanks is optional
                                         %% can be repeated if necesary
@@
 We introduce a size quantifier $\forall$ which lets us ignore sizes in terms for equality checking, alongside with a second quantifier $\Pi$ for abstracting over sizes that do affect the semantics of types and terms.
-Judgemental equality is decided by an adaption of normalization-by-evaluation for our new type theory, which features \emph{type shape}-directed reflection and reification.
+Judgmental equality is decided by an adaption of normalization-by-evaluation for our new type theory, which features \emph{type shape}-directed reflection and reification.
 It follows that subtyping and type checking of normal forms are decidable as well, the latter by a bidirectional algorithm.
@@
 Dependently-typed programming languages and proof assistants, such as Agda~\citeyearpar{agdawiki}, Coq~\cite{inria:coq86}, and Idris~\cite{brady:jfp13} require programs to be total, for two reasons.
-First, for consistency:  since propositions are just types and proofs of a propositions just programs which inhabit the corresponding type, some types need to be empty; otherwise, each proposition would be true.
+First, for consistency:  since propositions are just types and proofs of a proposition just programs which inhabit the corresponding type, some types need to be empty; otherwise, each proposition would be true.
 However, in a partial language with general recursion, each type is inhabited by the looping program $f = f$.
@@
 Consequently, each of the aforementioned languages based on Type Theory comes with a termination checker, which needs to reject all non-terminating programs, and should accept sufficiently many terminating programs to allow the user to express her algorithms.
-In the current termination checkers, programs are required to terminate by structural descent \cite{gimenez:guardeddefinitions}; the structural order may be extended to a lexicographic \cite{abelAltenkirch:jfp02} or size-change termination criterion \cite{jones:sizeChange,wahlstedt:PhD}.
-This fundamental limitation, since Type Theory allows many functions to be expressed in a structurally recursive manner, if needed by the help of a well-founded relation \cite{nordstrom:genrec}, inductive domain predicates \cite{boveCapretta:mscs05}, or inductive descriptions of the function graph \cite{bove:entcs09}.
-However, the syntactic termination check is very sensitive to reformulations of a program and hostile to abstraction \cite{abel:fics12}.
+In current termination checkers, programs are required to terminate by structural descent \cite{gimenez:guardeddefinitions}; the structural order may be extended to a lexicographic \cite{abelAltenkirch:jfp02} or size-change termination criterion \cite{jones:sizeChange,wahlstedt:PhD}.
+This is a fundamental limitation, since Type Theory allows many functions to be expressed in a structurally recursive manner, if needed by the help of a well-founded relation \cite{nordstrom:genrec}, inductive domain predicates \cite{boveCapretta:mscs05}, or inductive descriptions of the function graph \cite{bove:entcs09}.
+However, the syntactic termination check is very sensitive to reformulations of the program and hostile to abstraction \cite{abel:fics12}.
 
@@
 Most sized type systems are non-dependent \cite{amadio:guardcondition,abelPientka:jfp16,bartheGregoireRiba:lernet08,bartheGregoireRiba:csl08,blanquiRiba:lpar06,dalLagoGrellois:probabilisticSizedTypes2017}, yet the combination of sized and dependent types has been studied as well \cite{blanqui:csl05,bartheGregoirePastawski:lpar06,gregoireSacchini:lpar10,sacchini:lics13,sacchini:flops14}.
-However, to the best of our knowledge, no study combines \emph{higher-rank size polymorphism} with full-fledged dependent types (Xi \citeyearpar{xi:terminationHOSC} only has indexed types).
+However, to the best of our knowledge, no study combines \emph{higher-rank size polymorphism} with full-fledged dependent types.\footnote{\citet{xi:terminationHOSC} has first-class size polymorphism, but only indexed types, no universes or large eliminations.}
 
@@
 The condition is that preprocessing does not increase the size of the subtree; otherwise, termination could not be guaranteed.
-Concretely, assume a type $\tT\,i$ of trees of size $< i$ with a constructor $\tnode : \forall i.\, \List\,(\tT\,i) \to \tT\,(i+1)$.
+Concretely, assume a type $\tT\,i$ of trees of size $< i$ with a constructor $\tnode : \forall i.\, \List\,(\tT\,i) \to \tT\,(i+1)$ which takes a finite list of subtrees to form a new tree.
 In the following definition of $\ttrav$, the preprocessing $\vpre : \forall i.\, \tT\,i \to \tT\,i$ can be safely applied to input tree $t$ because the type of $\vpre$ bounds the size of $\vpre\,t$ by the size of $t$.
@@
   \\
-  \ttrav \; \vpre \; \vpost \; t = \vpost\;(\tcase\;\vpre\,t\;\tof \, \{ \, \tnode\,\vts \to \tnode\,(\tmap \; (\ttrav \, \vpre \, \vpost) \; \vts) \})
+  \ttrav \; \vpre \; \vpost \; t = \vpost\;(\tcase\;\vpre\,t\;\tof \, \{ \, \tnode\,\vts \to \tnode\,(\tmap \; (\ttrav \, \vpre \, \vpost) \; \vts)\, \})
 \end{array}
 \]
-This is the \emph{Curry}-style program as accepted from the user, but state-of-the-art type checkers elaborate the program from surface syntax into an internal \emph{Church}-style syntax with explicit type abstractions and type applications.\footnote{Agda, Coq, Idris, and Haskell \cite{sulzmannChakravartyPeythonJonesDonnelly:tldi07} all have Church-style internal languages.}  With implicit type and size applications elaborated, $\ttrav$ would look as follows:
+The display shows the \emph{Curry}-style program as provided by the user, but state-of-the-art type checkers elaborate the program from surface syntax into an internal \emph{Church}-style syntax with explicit type abstractions and type applications.\footnote{Agda, Coq, Idris, and Haskell \cite{sulzmannChakravartyPeythonJonesDonnelly:tldi07} all have Church-style internal languages.}  With implicit type and size applications elaborated, $\ttrav$ would look as follows:
 \[
@@
 \]
-From a dependent-type perspective, explicit size applications in terms can be problematic when the type checker compares terms for equality (which is necessary as types can depend on values).
-Inferred sizes may not be unique, as we have subtyping $\tT\,i \leq \tT\,j$ for $i \leq j$: we can always weaken an upper bound.
-For instance, given $\vts : \List\,(\tT\,i)$, all of $\tnode\,i\,\vts$ and $\tnode\,(i+1)\,\vts$ and $\dots$ and $\tnode\,\infty\,\vts$ have type $\tT\,\infty$.
-However, semantically, all these trees are equal, thus, the syntactic equality check should ignore the size argument to $\tnode$.
-Similarly, in the application $\vpre\,i\,t$ the size argument $i$ should be ignored by the equality check.
-Yet $\vpre\,i : \tT\,i \to \tT\,i$ and $\vpre\,j : \tT\,j \to \tT\,j$ have different types for $i \not= j$, and moreover these function types are not in the subtyping relation due to mixed-variant occurrence of the size parameter.
-It seems that during equality checking we have to consider terms of different types, for a while.
-Once we apply $\vpre\,i$ and $\vpre\,j$ to the same tree $t : \tT\,k$, which determines $i = j = k$, we are back in safety.
-However, allowing types to differ during an equality check can be hazardous, especially when the equality-check is type directed.
+Church-style syntax is the basis for all program analyses and transformations to follow and show be considered as the \emph{true} syntax.
+However, from a dependent-type perspective, explicit size applications
+in terms can be problematic when the type checker compares terms for
+equality---which is necessary as types can depend on values.  Inferred
+sizes may not be unique, as we have subtyping $\tT\,i \leq \tT\,j$ for
+$i \leq j$: we can always weaken an upper bound.  For instance, given
+$\vts : \List\,(\tT\,i)$, any of the terms
+$\tnode\,i\,\vts$, $\tnode\,(i+1)\,\vts$, $\dots$, $\tnode\,\infty\,\vts$
+%$\tnode\,i\,\vts$ or $\tnode\,(i+1)\,\vts$ or $\dots$ or $\tnode\,\infty\,\vts$
+has type
+$\tT\,\infty$.  Yet semantically, all these trees are equal,
+thus, the syntactic equality check should ignore the size argument to
+$\tnode$.  Similarly, in the application $\vpre\,i\,t$ the size
+argument $i$ should be ignored by the equality check.  Yet
+$\vpre\,i : \tT\,i \to \tT\,i$ and $\vpre\,j : \tT\,j \to \tT\,j$ have
+different types for $i \not= j$, and moreover these function types are
+not in the subtyping relation due to the mixed-variant occurrence of the
+size parameter.  It seems that during equality checking we have to
+consider terms of different types, at least for a while.  Once we apply
+$\vpre\,i$ and $\vpre\,j$ to the same tree $t : \tT\,k$, which
+determines $i = j = k$, we are back to safety.  However, allowing
+types to differ during an equality check can be hazardous, especially
+when the equality-check is type directed.
 
 Consider the analogous situation for the polymorphic lambda calculus System~F, be it the predicative variant or not, extended by a unit type $\One$.
-For Church-style, we can give a type-directed $\beta\eta$-equality test which equates all term at the unit type.
-The most interesting rules would be the $\eta$-rules for unit and function type and the congruence rule for type application:
+For Church-style, we can give a type-directed $\beta\eta$-equality test which equates all terms at the unit type.
+The most interesting rules are the $\eta$-rules for unit and function type and the congruence rule for type application:
 \begin{gather*}
@@
 \end{gather*}
-The Curry-style version would replace the last conclusion by $\Gamma \der t = t' : B[A/X]$ where the type $A$ to instantiate $X$ has to be guessed.
+The Curry-style version replaces the last conclusion by $\Gamma \der t = t' : B[A/X]$ where the type $A$ to instantiate $X$ has to be guessed.
 However, in Curry-style more terms are equated than in Church-style, as for instance the Church-style terms $t\,A\,(\lambda x \of A.\, x)$ and $t\,B\,(\lambda x \of B.\, x)$ map to the same Curry-style term $t\,(\lambda x.\,x)$.
@@
 Does the literature offer a solution to this problem?
-In fact, a Church-style calculus with Curry-style equality has been studied before, it is ICC$^*$ \cite{barrasBernardo:fossacs08,mishraLingerSheard:fossacs08} based on Miquel's Implicit Calculus of Constructions \cite{miquel:tlca01}.
+In fact, a Church-style calculus with Curry-style equality has been studied before, it is ICC$^*$ \cite{barrasBernardo:fossacs08,mishraLingerSheard:fossacs08} based on Miquel's Implicit Calculus of Constructions \citeyearpar{miquel:tlca01}.
 In ICC$^*$, equality is checked by erasing all type abstractions and applications, and comparing the remaining untyped terms for $\beta\eta$-equality.
 While this works for $\eta$-laws that can be formulated on untyped terms, such as $\eta$-contraction of functions $\lambda x.\,t\,x \redeta t$ (when $x$ not free in $t$), it does not extend to type-directed $\eta$-laws such as extensionality for the unit type.
-Further, ICC$^*$ is not a type theory formulated with a typed equality judgement, which makes it hard to define its models \cite{miquel:lics00}---we wish not to go there, but stay within the framework of Martin-L\"of Type Theory \cite{martinlof:predicative75}.
-
-How can we have our cake and eat it, too?
-It has already been observed that for a type-directed equality check the precise type is not necessary, a \emph{shape} or \emph{skeleton} is sufficient that informs the algorithm whether the terms under comparison are functions, inhabitants of the unit type, or something else.
-For the Logical Framework (LF), the simplest dependent lambda-calculus, the skeletons are simple types that can be obtained from the original dependent types by erasing all dependencies.
-\citet{harperpfenning:equivalenceLF} present such an equality check for LF which is directed by simple types, and their technique should scale to all type theories that admit dependency erasure.\footnote{For instance, the types of the Calculus of Constructions erase to F$^\omega$-types \cite{geuvers:shortFlexibleSNCC}, and the latter could be used to direct the equality check.
-\cite{lovasPfenning:lmcs10} consider also \emph{refinement types for logical frameworks} which can be erased to simple types.}
-In the presence of large eliminations, which we have in type theories that feature both universes and data types, dependency erasure fails, and it is not clear what the skeleton of a dependent type should be.
+Further, ICC$^*$ is not a type theory formulated with a typed equality judgement, which makes it hard to define its models \cite{miquel:lics00}---we wish not to go there, but stay within the framework of Martin-L\"of Type Theory \citeyearpar{martinlof:predicative75}.
+
+%How can we have our cake and eat it, too?
+Now, if the types of compared Curry-style terms are not equal, can they be sufficiently related to give a proper meaning to the algorithmic equality judgement?
+It has already been observed that for a type-directed equality check the precise type is not necessary, a \emph{shape} or \emph{skeleton} is sufficient.
+The skeleton informs the algorithm whether the terms under comparison are functions, inhabitants of the unit type, or something else, to possibly apply the appropriate $\eta$-law.
+For the Logical Framework (LF), the simplest dependent lambda-calculus, the skeletons are simple types that can be obtained from the original dependent types by erasing the dependencies:  dependent function types map to non-dependent ones and indexed data types to simple data types.
+\citet{harperpfenning:equivalenceLF} present such an equality check for LF which is directed by simple types, and their technique should scale to other type theories that admit dependency erasure.\footnote{For instance, the types of the Calculus of Constructions erase to F$^\omega$-types \cite{geuvers:shortFlexibleSNCC}, and the latter could be used to direct the equality check.
+\citet{lovasPfenning:lmcs10} consider also \emph{refinement types for logical frameworks} which can be erased to simple types.}
+
+By \emph{large eliminations} \cite{werner:types92} we refer to types computed by case distinction over values; they occur
+in type theories that feature both universes and data types.
+In the presence of large eliminations, dependency erasure fails, and it is not clear what the skeleton of a dependent type should be.
 For instance consider the type $\funT n \NN \underbrace{A \to \dots \to A}_n \to A$ of $n$-ary functions; its shape is dependent on the value of $n$, thus cannot be determined statically.
@@
 It scales to universes and large eliminations \cite{abelCoquandDybjer:lics07}, topped with singleton types or proof irrelevance \cite{abelCoquandPagano:lmcs11}, and even impredicativity \cite{abel:flops10}.
-At its hard there are reflection $\up T$ and reification $\down T$ functions directed by type $T$ and orchestrating just-in-time $\eta$-expansion.
+At its heart there are reflection $\up T$ and reification $\down T$ functions directed by type $T$ and orchestrating just-in-time $\eta$-expansion.
 Reflection $\up T x$ maps variables $x$ into the realm of values of type $T$ and let's us compute with open terms.
@@
 
-\section{A Type System WIth Irrelevant Size Application}
+\section{A Type System With Irrelevant Size Application}
 \label{sec:syntax}
@@
 to be used as generic values (unknowns) and type annotations
-$\up T t$ and $\down T t$ for lazy realisations of the
+$\up T t$ and $\down T t$ for lazy realizations of the
 reflection and reification operations of NbE.
@@
 First, we close the term $t$ with an environment $\eta$ that maps the free de Bruijn indices of $t$ to reflected de Bruijn levels.
-Reflection of de Bruijn levels follows the blue path $\Level \to \Dne \to \D$: Levels embed via constructor $\var{}$ into semantic neutrals $\DNe$ which are labelled with their type $A \in \D$ to become and element $\up A \lev j \in \D$.
+Reflection of de Bruijn levels follows the blue path $\Level \to \Dne \to \D$: Levels embed via constructor $\var{}$ into semantic neutrals $\DNe$ which are labeled with their type $A \in \D$ to become and element $\up A \lev j \in \D$.
 
@@
 
-The PERs $\NE$ characterizes neutral reifyable terms.
+The PERs $\NE$ characterizes neutral reifiable terms.
 % \[
@@
 For instance, assume a term $F : \Nat 0 \to \Set[0]$ which diverges if applied to a successor term.
-Then $T := \funT x {\Nat 0} {F\,x}$ is a welldefined type; we have $T \in \SET[0]$.
+Then $T := \funT x {\Nat 0} {F\,x}$ is a well-defined type; we have $T \in \SET[0]$.
 Now consider $S := \funT x {\Nat \infty} {F\,x}$.
@@
 
-The eliminations for natural numbers are acurrately modeled.
+The eliminations for natural numbers are accurately modeled.
 \begin{lemma}[Case]
@@
 \begin{proof*}
-By wellfounded induction on $\alpha$.
+By well-founded induction on $\alpha$.
 \begin{caselist}
@@
 
-In this section, we show that NbE is sound for judgemental equality, i.e., that same normal form implies definitional equality.
+In this section, we show that NbE is sound for judgmental equality, i.e., that same normal form implies definitional equality.
 The proof follows \citet{abelCoquandDybjer:lics07,abel:habil}.
@@
   \begin{enumerate}
-  \item Welltypedness: $\Gamma \der \sigma : \Delta$
+  \item Well-typedness: $\Gamma \der \sigma : \Delta$
    and $\Gamma \der \sigma = \sigma \eeq \tau : \Delta$
@@
 
-\item For the identity environment $\LRS \Gamma \sid \Gamma \rhog$ (Lemm~\ref{lem:idenv})
+\item For the identity environment $\LRS \Gamma \sid \Gamma \rhog$ (Lemma~\ref{lem:idenv})
 the Fundamental Theorem for Typing gives $\LR \Gamma t T {t\rhog} {T\rhog}$.
@@
    Here we use that the computed normal forms are identical because
-   we erase all size annotations $\ann\alpha$ to $\ann\infty$ during readback.
+   we erase all size annotations $\ann\alpha$ to $\ann\infty$ during read back.
 \qed
@@
 \begin{proof}
-  Follows directly by the fundamental theorem (Thm~\ref{thm:fund}) and inversion on semantic subtyping.
+  Follows directly by the fundamental theorem (Thm.~\ref{thm:fund}) and inversion on semantic subtyping.
 \end{proof}
@@
 \begin{proof}
-  By inversion of typing (Lem~\ref{lem:invty}), type constructor discrimination \label{lem:discr}
-  and inversion of subtyping (Lem~\ref{lem:invsubty}).
+  By inversion of typing (Lemma~\ref{lem:invty}), type constructor discrimination \label{lem:discr}
+  and inversion of subtyping (Lemma~\ref{lem:invsubty}).
 
@@
 \begin{proof}
-  By induction on $\Gamma \der T \asub T'$, using soundness of NbE and subject reduction (Thm~\ref{thm:sr}).
+  By induction on $\Gamma \der T \asub T'$, using soundness of NbE and subject reduction (Thm.~\ref{thm:sr}).
 \end{proof}

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Proof-read the introduction.

@@
 For instance consider the type $\funT n \NN \underbrace{A \to \dots \to A}_n \to A$ of $n$-ary functions; its shape is dependent on the value of $n$, thus cannot be determined statically.
-In the solution we present in this article, we do not introduce a new syntax for shapes, just a relation $T \shape S$ stating that type $S$ qualifies as shape for type $T$.
-We use a normalization function inspired by normalization-by-evaluation (NbE) that computes the $\eta$-long normal form of the terms under comparison, which we then check for identity.
-NbE \cite{bergerSchwichtenberg:lics91,danvy:tdpe} has proven a robust method to decide equality in powerful type theories with non-trivial $\eta$-laws.
+Thus, the ``skeleton'' idea is also not directly applicable.
+
+Going beyond the classical incremental equality check, there is a technique that can deal with dynamic $\eta$-expansion.
+It is a type-directed normalization function inspired by normalization-by-evaluation (NbE) that computes $\eta$-long normal forms \cite{bergerSchwichtenberg:lics91,danvy:tdpe}.
+We can check the computed normal forms for identity and, thus, decide definitional equality.
+NbE has proven to be a robust method to decide equality in powerful type theories with non-trivial $\eta$-laws.
 It scales to universes and large eliminations \cite{abelCoquandDybjer:lics07}, topped with singleton types or proof irrelevance \cite{abelCoquandPagano:lmcs11}, and even impredicativity \cite{abel:flops10}.
@@
 
-The central technical observation is that reflection and reification do not need the precise type $T$, they work the same for any shape $S$ of $T$.
+% In the solution we present in this article, we do not introduce a new syntax for shapes, just a relation $T \shape S$ stating that type $S$ qualifies as shape for type $T$.
+We use a normalization function inspired by normalization-by-evaluation (NbE) that computes the $\eta$-long normal form of the terms under comparison, which we then check for identity.
+
+The central technical observation is that reflection and reification do not need the precise type $T$, they work the same for any \emph{shape} $S$ of $T$.
+We managed, while not introducing a new syntax for shapes, to define a relation $T \shape S$ on type values stating that type $S$ qualifies as shape for type $T$.
+Shapes unfold dynamically during reflection and reification.
 For example, when reflecting a variable $x$ into the polymorphic function type $\forall i.\, F\,i$ where $F\,i = \Nat\,i \to \Nat\,i$, we obtain
@@
 Whenever we form a irrelevant function type $\forall x \of U.\, T[x]$, we require $T[x]$ to be shape-irrelevant in $x$.
-This is the middle ground between ICC$^*$, where no restriction is placed on $T$ but $\eta$ for unit types is (at least, currently) out of reach, and Pfenning's \citeyearpar{pfenning:intextirr} irrelevance modality, adapted to full dependent types by \citet{abelScherer:types10}, which requires $T$ to be irrelevant in $x$ and, thus, has type equality $T[i] = T[j]$.
+This is the middle ground between ICC$^*$, where no restriction is placed on $T$ but $\eta$ for unit types is out of reach (at least for the moment), and Pfenning's \citeyearpar{pfenning:intextirr} irrelevance modality, adapted to full dependent types by \citet{abelScherer:types10}, which requires $T$ to be irrelevant in $x$ and, thus, has type equality $T[i] = T[j]$.
 
@@
 We confine ourselves to size-irrelevant function types $\forall i.\, T[i]$.
-This relieves us from defining a special shape-irrelevance modality, since all size-indexed types $T[i]$ are shape irrelevant in $i$, simply because there is no case distinction on size, and sizes appear relevantly only under a sized type constructor such as $\tNat$.
+This relieves us from defining a special shape-irrelevance modality, since all size-indexed types $T[i]$ \emph{are} shape irrelevant in $i$, simply because there is no case distinction on size, and sizes appear relevantly only under a sized type constructor such as $\tNat$.
 Our technique would not extend to the polymorphic types $\forall X.\,B[X]$ of System~F.
 Even though there is no case distinction on types, shape irrelevance of $B[X]$ fails in general, as $X$ could appear as a type on the top-level, e.g.
-in $B[X] = X$ or $B[X] = X \to X$.
-
-This article makes the following novel contributions:
+in %$B[X] = X$ or
+$B[X] = X \to X$, and then $B[\One]$ and $B[A \to A]$ would have distinct shapes.
+
+To summarize, this article makes the following novel contributions:
 \begin{enumerate}
@@
 This approximation relation allows reflection and reification at size-polymorphic types $\forall i.\, T$.
-As usual for the meta-theory of Type Theory with large eliminations, the machinery is involved, but just require the usual two logical relations:  First, a PER model to define the semantics of types and prove the completeness of NbE (Section~\ref{sec:sem}).
+As usual for the meta-theory of Type Theory with large eliminations, the machinery is involved, but we just require the usual two logical relations:  First, a PER model to define the semantics of types and prove the completeness of NbE (Section~\ref{sec:sem}).
 Secondly, a relation between syntax and semantics to prove soundness of NbE (Section~\ref{sec:soundnbe}).
@@
 \]
-The algorithm employs the usual lazy reduction for types (just-in-time weak-head evaluation) in type and sub-type checker
+The algorithm employs the usual lazy reduction for types, \ie, just-in-time weak-head evaluation, in type and sub-type checker
 \cite{huet:wsscs89}.  In this, it improves on \citet{fridlenderPagano:tlca13} which instruments full normalization (NbE) at every step.
 \end{enumerate}
-This article is accompanied by a prototypical type checker \texttt{Sit} which implements the type system and type checking algorithm as described in the remainder of the paper.
-But first, we will motivate our type system from a practical perspective: reasoning about programs involving sized types in Agda.
+This article is accompanied by a prototypical type checker \textsf{Sit} which implements the type system and type checking algorithm as described in the remainder of the paper.
+But before going into the technical details, we will motivate our type system from a practical perspective: reasoning about programs involving sized types in Agda.
 

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Proof read part 2 and conclusion.

@@
 In this section, we show how the lack of size irrelevance prevents us from reasoning naturally about programs involving sized types in Type Theory.
-We focus on Agda, the only mature implementation of Type Theory with an experimental integration of sized types.
+We focus on Agda, at the time of writing the only mature implementation of Type Theory with an experimental integration of sized types.
 
@@
 In this article, we have described the first successful integration of higher-rank size polymorphism into a core type theory with dependent function types, a sized type of natural numbers, a predicative hierarchy of universes, subtyping, and $\eta$-equality.
+This is an important stepping stone for the smooth integration of sized types into dependently-typed proof assistants.
+In the final paragraphs, we discuss some questions and insights that follow from our work and go beyond it.
 
@@
 Reification erases all irrelevant size arguments to $\infty$.
-We also considered keeping the sizes, and define equality $v \approx v'$ of normal forms modulo differences in erased arguments.
-However, it seems difficult to prove that equal normal forms are judgmentally equal, \ie, if $\Gamma \der v,v' : T$ and $v \approx v'$ then $\Gamma \der v = v' : T$.
+We have also considered to keep the sizes there, and define equality $v \approx v'$ of normal forms up to differences in erased arguments.
+However, it seems difficult to prove that equal normal forms are judgmentally equal, \ie, that $\Gamma \der v,v' : T$ and $v \approx v'$ imply $\Gamma \der v = v' : T$.
 Proving this statement seems to be the key to an incremental equality check; but
@@
 
-For now, we have exploited shape-irrelevance of sized types, but this directly extends to universe levels.
+For now, we have only exploited shape-irrelevance of sized types, but this directly extends to universe levels.
 If we consider all universes as a single shape $\Set[\ell_1] \shape \Set[\ell_2]$, we can quantify over levels irrelevantly, as $\Set$ is a shape-irrelevant type constructor.
@@
 
-The reflections on level irrelevance lead us to the question: can a type theory $T$ with a stratified universe hierarchy be understood as a sort of refinement of the inconsistent System~$U$ (Type:Type)?
+Our reflections on level irrelevance lead us to the question: can a type theory $T$ with a stratified universe hierarchy be understood as a sort of refinement of the inconsistent System~$U$ (Type:Type)?
 Intuitively, when checking two terms of $T$ for equality, could we ignore the stratification in the type $A$ which directs the equality check (thus, consider $A$ coming from $U$)?
 Such a perspective would put stratification in one pot with size assignment:  Size annotations and levels are both just annotations for the termination checker, but do not bear semantic relevance.
-We could switch the universe checker temporarily off as we do with the termination checker (see also the work of \citet{stumpSjobergWeirich:par10} on \emph{termination casts}).
+We could switch the universe checker temporarily off as we do with the termination checker---cf.\ the work of \citet{stumpSjobergWeirich:par10} on \emph{termination casts}.
 

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Use auto-icfp17.bib
[ ICFP 17 ] Intro: forgotten "not"

@@
 In current termination checkers, programs are required to terminate by structural descent \cite{gimenez:guardeddefinitions}; the structural order may be extended to a lexicographic \cite{abelAltenkirch:jfp02} or size-change termination criterion \cite{jones:sizeChange,wahlstedt:PhD}.
-This is a fundamental limitation, since Type Theory allows many functions to be expressed in a structurally recursive manner, if needed by the help of a well-founded relation \cite{nordstrom:genrec}, inductive domain predicates \cite{boveCapretta:mscs05}, or inductive descriptions of the function graph \cite{bove:entcs09}.
+This is not a fundamental limitation, since Type Theory allows many functions to be expressed in a structurally recursive manner, if needed by the help of a well-founded relation \cite{nordstrom:genrec}, inductive domain predicates \cite{boveCapretta:mscs05}, or inductive descriptions of the function graph \cite{bove:entcs09}.
 However, the syntactic termination check is very sensitive to reformulations of the program and hostile to abstraction \cite{abel:fics12}.
@@
 %% Bibliography
-\bibliography{medium}
+\bibliography{auto-icfp17}
 

### 2017-02-27 nbe/icfp17.tex
Minor typos

--- 2017-02-27 nbe/icfp17.tex
@@
 %% above the permission/copyright statement.
-%\thanks{with paper note}                %% \thanks is optional
+\thanks{with paper note}                %% \thanks is optional
                                         %% can be repeated if necesary
@@
 We introduce a size quantifier $\forall$ which lets us ignore sizes in terms for equality checking, alongside with a second quantifier $\Pi$ for abstracting over sizes that do affect the semantics of types and terms.
-Judgmental equality is decided by an adaption of normalization-by-evaluation for our new type theory, which features \emph{type shape}-directed reflection and reification.
+Judgemental equality is decided by an adaption of normalization-by-evaluation for our new type theory, which features \emph{type shape}-directed reflection and reification.
 It follows that subtyping and type checking of normal forms are decidable as well, the latter by a bidirectional algorithm.
@@
 % \cite{boveKraussSozeau:mscs16}
-Consequently, each of the aforementioned languages based on Type Theory comes with a termination checker, which needs to reject all non-terminating programs, and should accept sufficiently many terminating programs to allow the user to express her algorithms.
-In current termination checkers, programs are required to terminate by structural descent \cite{gimenez:guardeddefinitions}; the structural order may be extended to a lexicographic \cite{abelAltenkirch:jfp02} or size-change termination criterion \cite{jones:sizeChange,wahlstedt:PhD}.
-This is a fundamental limitation, since Type Theory allows many functions to be expressed in a structurally recursive manner, if needed by the help of a well-founded relation \cite{nordstrom:genrec}, inductive domain predicates \cite{boveCapretta:mscs05}, or inductive descriptions of the function graph \cite{bove:entcs09}.
-However, the syntactic termination check is very sensitive to reformulations of the program and hostile to abstraction \cite{abel:fics12}.
+Consequently, each of the aforementioned languages based on Type Theory come with a termination checker, which needs to reject all non-terminating programs, and should accept sufficiently many terminating programs to allow the user to express her algorithms.
+In the current termination checkers, programs are required to terminate by structural descent \cite{gimenez:guardeddefinitions}; the structural order may be extended to a lexicographic \cite{abelAltenkirch:jfp02} or size-change termination criterion \cite{jones:sizeChange,wahlstedt:PhD}.
+This fundamental limitation, since Type Theory allows many functions to be expressed in a structurally recursive manner, if needed by the help of a well-founded relation \cite{nordstrom:genrec}, inductive domain predicates \cite{boveCapretta:mscs05}, or inductive descriptions of the function graph \cite{bove:entcs09}.
+However, the syntactic termination check is very sensitive to reformulations of a program and hostile to abstraction \cite{abel:fics12}.
 
@@
 Most sized type systems are non-dependent \cite{amadio:guardcondition,abelPientka:jfp16,bartheGregoireRiba:lernet08,bartheGregoireRiba:csl08,blanquiRiba:lpar06,dalLagoGrellois:probabilisticSizedTypes2017}, yet the combination of sized and dependent types has been studied as well \cite{blanqui:csl05,bartheGregoirePastawski:lpar06,gregoireSacchini:lpar10,sacchini:lics13,sacchini:flops14}.
-However, to the best of our knowledge, no study combines \emph{higher-rank size polymorphism} with full-fledged dependent types.\footnote{\citet{xi:terminationHOSC} has first-class size polymorphism, but only indexed types, no universes or large eliminations.}
+However, to the best of our knowledge, no study combines \emph{higher-rank size polymorphism} with full-fledged dependent types (Xi \citeyearpar{xi:terminationHOSC} only has indexed types).
 
@@
 The condition is that preprocessing does not increase the size of the subtree; otherwise, termination could not be guaranteed.
-Concretely, assume a type $\tT\,i$ of trees of size $< i$ with a constructor $\tnode : \forall i.\, \List\,(\tT\,i) \to \tT\,(i+1)$ which takes a finite list of subtrees to form a new tree.
+Concretely, assume a type $\tT\,i$ of trees of size $< i$ with a constructor $\tnode : \forall i.\, \List\,(\tT\,i) \to \tT\,(i+1)$.
 In the following definition of $\ttrav$, the preprocessing $\vpre : \forall i.\, \tT\,i \to \tT\,i$ can be safely applied to input tree $t$ because the type of $\vpre$ bounds the size of $\vpre\,t$ by the size of $t$.
@@
   \\
-  \ttrav \; \vpre \; \vpost \; t = \vpost\;(\tcase\;\vpre\,t\;\tof \, \{ \, \tnode\,\vts \to \tnode\,(\tmap \; (\ttrav \, \vpre \, \vpost) \; \vts)\, \})
+  \ttrav \; \vpre \; \vpost \; t = \vpost\;(\tcase\;\vpre\,t\;\tof \, \{ \, \tnode\,\vts \to \tnode\,(\tmap \; (\ttrav \, \vpre \, \vpost) \; \vts) \})
 \end{array}
 \]
-The display shows the \emph{Curry}-style program as provided by the user, but state-of-the-art type checkers elaborate the program from surface syntax into an internal \emph{Church}-style syntax with explicit type abstractions and type applications.\footnote{Agda, Coq, Idris, and Haskell \cite{sulzmannChakravartyPeythonJonesDonnelly:tldi07} all have Church-style internal languages.}  With implicit type and size applications elaborated, $\ttrav$ would look as follows:
+This is the \emph{Curry}-style program as accepted from the user, but state-of-the-art type checkers elaborate the program from surface syntax into an internal \emph{Church}-style syntax with explicit type abstractions and type applications.\footnote{Agda, Coq, Idris, and Haskell \cite{sulzmannChakravartyPeythonJonesDonnelly:tldi07} all have Church-style internal languages.}  With implicit type and size applications elaborated, $\ttrav$ would look as follows:
 \[
@@
 \]
-Church-style syntax is the basis for all program analyses and transformations to follow and show be considered as the \emph{true} syntax.
-However, from a dependent-type perspective, explicit size applications
-in terms can be problematic when the type checker compares terms for
-equality---which is necessary as types can depend on values.  Inferred
-sizes may not be unique, as we have subtyping $\tT\,i \leq \tT\,j$ for
-$i \leq j$: we can always weaken an upper bound.  For instance, given
-$\vts : \List\,(\tT\,i)$, any of the terms
-$\tnode\,i\,\vts$, $\tnode\,(i+1)\,\vts$, $\dots$, $\tnode\,\infty\,\vts$
-%$\tnode\,i\,\vts$ or $\tnode\,(i+1)\,\vts$ or $\dots$ or $\tnode\,\infty\,\vts$
-has type
-$\tT\,\infty$.  Yet semantically, all these trees are equal,
-thus, the syntactic equality check should ignore the size argument to
-$\tnode$.  Similarly, in the application $\vpre\,i\,t$ the size
-argument $i$ should be ignored by the equality check.  Yet
-$\vpre\,i : \tT\,i \to \tT\,i$ and $\vpre\,j : \tT\,j \to \tT\,j$ have
-different types for $i \not= j$, and moreover these function types are
-not in the subtyping relation due to the mixed-variant occurrence of the
-size parameter.  It seems that during equality checking we have to
-consider terms of different types, at least for a while.  Once we apply
-$\vpre\,i$ and $\vpre\,j$ to the same tree $t : \tT\,k$, which
-determines $i = j = k$, we are back to safety.  However, allowing
-types to differ during an equality check can be hazardous, especially
-when the equality-check is type directed.
+From a dependent-type perspective, explicit size applications in terms can be problematic when the type checker compares terms for equality (which is necessary as types can depend on values).
+Inferred sizes may not be unique, as we have subtyping $\tT\,i \leq \tT\,j$ for $i \leq j$: we can always weaken an upper bound.
+For instance, given $\vts : \List\,(\tT\,i)$, all of $\tnode\,i\,\vts$ and $\tnode\,(i+1)\,\vts$ and $\dots$ and $\tnode\,\infty\,\vts$ have type $\tT\,\infty$.
+However, semantically, all these trees are equal, thus, the syntactic equality check should ignore the size argument to $\tnode$.
+Similarly, in the application $\vpre\,i\,t$ the size argument $i$ should be ignored by the equality check.
+Yet $\vpre\,i : \tT\,i \to \tT\,i$ and $\vpre\,j : \tT\,j \to \tT\,j$ have different types for $i \not= j$, and moreover these function types are not in the subtyping relation due to mixed-variant occurrence of the size parameter.
+It seems that during equality checking we have to consider terms of different types, for a while.
+Once we apply $\vpre\,i$ and $\vpre\,j$ to the same tree $t : \tT\,k$, which determines $i = j = k$, we are back in safety.
+However, allowing types to differ during an equality check can be hazardous, especially when the equality-check is type directed.
 
@@
 For Church-style, we can give a type-directed $\beta\eta$-equality test which equates all terms at the unit type.
-The most interesting rules are the $\eta$-rules for unit and function type and the congruence rule for type application:
+The most interesting rules would be the $\eta$-rules for unit and function type and the congruence rule for type application:
 \begin{gather*}
@@
 \end{gather*}
-The Curry-style version replaces the last conclusion by $\Gamma \der t = t' : B[A/X]$ where the type $A$ to instantiate $X$ has to be guessed.
+The Curry-style version would replace the last conclusion by $\Gamma \der t = t' : B[A/X]$ where the type $A$ to instantiate $X$ has to be guessed.
 However, in Curry-style more terms are equated than in Church-style, as for instance the Church-style terms $t\,A\,(\lambda x \of A.\, x)$ and $t\,B\,(\lambda x \of B.\, x)$ map to the same Curry-style term $t\,(\lambda x.\,x)$.
@@
 Does the literature offer a solution to this problem?
-In fact, a Church-style calculus with Curry-style equality has been studied before, it is ICC$^*$ \cite{barrasBernardo:fossacs08,mishraLingerSheard:fossacs08} based on Miquel's Implicit Calculus of Constructions \citeyearpar{miquel:tlca01}.
+In fact, a Church-style calculus with Curry-style equality has been studied before, it is ICC$^*$ \cite{barrasBernardo:fossacs08,mishraLingerSheard:fossacs08} based on Miquel's Implicit Calculus of Constructions \cite{miquel:tlca01}.
 In ICC$^*$, equality is checked by erasing all type abstractions and applications, and comparing the remaining untyped terms for $\beta\eta$-equality.
 While this works for $\eta$-laws that can be formulated on untyped terms, such as $\eta$-contraction of functions $\lambda x.\,t\,x \redeta t$ (when $x$ not free in $t$), it does not extend to type-directed $\eta$-laws such as extensionality for the unit type.
-Further, ICC$^*$ is not a type theory formulated with a typed equality judgement, which makes it hard to define its models \cite{miquel:lics00}---we wish not to go there, but stay within the framework of Martin-L\"of Type Theory \citeyearpar{martinlof:predicative75}.
-
-%How can we have our cake and eat it, too?
-Now, if the types of compared Curry-style terms are not equal, can they be sufficiently related to give a proper meaning to the algorithmic equality judgement?
-It has already been observed that for a type-directed equality check the precise type is not necessary, a \emph{shape} or \emph{skeleton} is sufficient.
-The skeleton informs the algorithm whether the terms under comparison are functions, inhabitants of the unit type, or something else, to possibly apply the appropriate $\eta$-law.
-For the Logical Framework (LF), the simplest dependent lambda-calculus, the skeletons are simple types that can be obtained from the original dependent types by erasing the dependencies:  dependent function types map to non-dependent ones and indexed data types to simple data types.
-\citet{harperpfenning:equivalenceLF} present such an equality check for LF which is directed by simple types, and their technique should scale to other type theories that admit dependency erasure.\footnote{For instance, the types of the Calculus of Constructions erase to F$^\omega$-types \cite{geuvers:shortFlexibleSNCC}, and the latter could be used to direct the equality check.
-\citet{lovasPfenning:lmcs10} consider also \emph{refinement types for logical frameworks} which can be erased to simple types.}
-
-By \emph{large eliminations} \cite{werner:types92} we refer to types computed by case distinction over values; they occur
-in type theories that feature both universes and data types.
-In the presence of large eliminations, dependency erasure fails, and it is not clear what the skeleton of a dependent type should be.
+Further, ICC$^*$ is not a type theory formulated with a typed equality judgement, which makes it hard to define its models \cite{miquel:lics00}---we wish not to go there, but stay within the framework of Martin-L\"of Type Theory \cite{martinlof:predicative75}.
+
+How can we have our cake and eat it, too?
+It has already been observed that for a type-directed equality check the precise type is not necessary, a \emph{shape} or \emph{skeleton} is sufficient that informs the algorithm whether the terms under comparison are functions, inhabitants of the unit type, or something else.
+For the Logical Framework (LF), the simplest dependent lambda-calculus, the skeletons are simple types that can be obtained from the original dependent types by erasing all dependencies.
+\citet{harperpfenning:equivalenceLF} present such an equality check for LF which is directed by simple types, and their technique should scale to all type theories that admit dependency erasure.\footnote{For instance, the types of the Calculus of Constructions erase to F$^\omega$-types \cite{geuvers:shortFlexibleSNCC}, and the latter could be used to direct the equality check.
+\cite{lovasPfenning:lmcs10} consider also \emph{refinement types for logical frameworks} which can be erased to simple types.}
+In the presence of large eliminations, which we have in type theories that feature both universes and data types, dependency erasure fails, and it is not clear what the skeleton of a dependent type should be.
 For instance consider the type $\funT n \NN \underbrace{A \to \dots \to A}_n \to A$ of $n$-ary functions; its shape is dependent on the value of $n$, thus cannot be determined statically.
-Thus, the ``skeleton'' idea is also not directly applicable.
-
-Going beyond the classical incremental equality check, there is a technique that can deal with dynamic $\eta$-expansion.
-It is a type-directed normalization function inspired by normalization-by-evaluation (NbE) that computes $\eta$-long normal forms \cite{bergerSchwichtenberg:lics91,danvy:tdpe}.
-We can check the computed normal forms for identity and, thus, decide definitional equality.
-NbE has proven to be a robust method to decide equality in powerful type theories with non-trivial $\eta$-laws.
+In the solution we present in this article, we do not introduce a new syntax for shapes, just a relation $T \shape S$ stating that type $S$ qualifies as shape for type $T$.
+We use a normalization function inspired by normalization-by-evaluation (NbE) that computes the $\eta$-long normal form of the terms under comparison, which we then check for identity.
+NbE \cite{bergerSchwichtenberg:lics91,danvy:tdpe} has proven a robust method to decide equality in powerful type theories with non-trivial $\eta$-laws.
 It scales to universes and large eliminations \cite{abelCoquandDybjer:lics07}, topped with singleton types or proof irrelevance \cite{abelCoquandPagano:lmcs11}, and even impredicativity \cite{abel:flops10}.
-At its heart there are reflection $\up T$ and reification $\down T$ functions directed by type $T$ and orchestrating just-in-time $\eta$-expansion.
-Reflection $\up T x$ maps variables $x$ into the realm of values of type $T$ and let's us compute with open terms.
+At its hard there are reflection $\up T$ and reification $\down T$ functions directed by type $T$ and orchestrating just-in-time $\eta$-expansion.
+Reflection $\up T x$ maps variables $x$ into the realm of values of type $T$ and lets us compute with open terms.
 Reification $\down T a$ takes a value $a$ of type $T$ and computes its long normal form.
@@
 
-% In the solution we present in this article, we do not introduce a new syntax for shapes, just a relation $T \shape S$ stating that type $S$ qualifies as shape for type $T$.
-We use a normalization function inspired by normalization-by-evaluation (NbE) that computes the $\eta$-long normal form of the terms under comparison, which we then check for identity.
-
-The central technical observation is that reflection and reification do not need the precise type $T$, they work the same for any \emph{shape} $S$ of $T$.
-We managed, while not introducing a new syntax for shapes, to define a relation $T \shape S$ on type values stating that type $S$ qualifies as shape for type $T$.
-Shapes unfold dynamically during reflection and reification.
+The central technical observation is that reflection and reification do not need the precise type $T$, they work the same for any shape $S$ of $T$.
 For example, when reflecting a variable $x$ into the polymorphic function type $\forall i.\, F\,i$ where $F\,i = \Nat\,i \to \Nat\,i$, we obtain
@@
 Whenever we form a irrelevant function type $\forall x \of U.\, T[x]$, we require $T[x]$ to be shape-irrelevant in $x$.
-This is the middle ground between ICC$^*$, where no restriction is placed on $T$ but $\eta$ for unit types is out of reach (at least for the moment), and Pfenning's \citeyearpar{pfenning:intextirr} irrelevance modality, adapted to full dependent types by \citet{abelScherer:types10}, which requires $T$ to be irrelevant in $x$ and, thus, has type equality $T[i] = T[j]$.
+This is the middle ground between ICC$^*$, where no restriction is placed on $T$ but $\eta$ for unit types is (at least, currently) out of reach, and Pfenning's \citeyearpar{pfenning:intextirr} irrelevance modality, adapted to full dependent types by \citet{abelScherer:types10}, which requires $T$ to be irrelevant in $x$ and, thus, has type equality $T[i] = T[j]$.
 
@@
 We confine ourselves to size-irrelevant function types $\forall i.\, T[i]$.
-This relieves us from defining a special shape-irrelevance modality, since all size-indexed types $T[i]$ \emph{are} shape irrelevant in $i$, simply because there is no case distinction on size, and sizes appear relevantly only under a sized type constructor such as $\tNat$.
+This relieves us from defining a special shape-irrelevance modality, since all size-indexed types $T[i]$ are shape irrelevant in $i$, simply because there is no case distinction on size, and sizes appear relevantly only under a sized type constructor such as $\tNat$.
 Our technique would not extend to the polymorphic types $\forall X.\,B[X]$ of System~F.
 Even though there is no case distinction on types, shape irrelevance of $B[X]$ fails in general, as $X$ could appear as a type on the top-level, e.g.
-in %$B[X] = X$ or
-$B[X] = X \to X$, and then $B[\One]$ and $B[A \to A]$ would have distinct shapes.
-
-To summarize, this article makes the following novel contributions:
+in $B[X] = X$ or $B[X] = X \to X$.
+
+This article makes the following novel contributions:
 \begin{enumerate}
@@
 This approximation relation allows reflection and reification at size-polymorphic types $\forall i.\, T$.
-As usual for the meta-theory of Type Theory with large eliminations, the machinery is involved, but we just require the usual two logical relations:  First, a PER model to define the semantics of types and prove the completeness of NbE (Section~\ref{sec:sem}).
+As usual for the meta-theory of Type Theory with large eliminations, the machinery is involved, but just require the usual two logical relations:  First, a PER model to define the semantics of types and prove the completeness of NbE (Section~\ref{sec:sem}).
 Secondly, a relation between syntax and semantics to prove soundness of NbE (Section~\ref{sec:soundnbe}).
@@
 \]
-The algorithm employs the usual lazy reduction for types, \ie, just-in-time weak-head evaluation, in type and sub-type checker
+The algorithm employs the usual lazy reduction for types (just-in-time weak-head evaluation) in type and sub-type checker
 \cite{huet:wsscs89}.  In this, it improves on \citet{fridlenderPagano:tlca13} which instruments full normalization (NbE) at every step.
 \end{enumerate}
-This article is accompanied by a prototypical type checker \textsf{Sit} which implements the type system and type checking algorithm as described in the remainder of the paper.
-But before going into the technical details, we will motivate our type system from a practical perspective: reasoning about programs involving sized types in Agda.
+This article is accompanied by a prototypical type checker \texttt{Sit} which implements the type system and type checking algorithm as described in the remainder of the paper.
+But first, we will motivate our type system from a practical perspective: reasoning about programs involving sized types in Agda.
 
@@
 
-\section{A Type System With Irrelevant Size Application}
+\section{A Type System WIth Irrelevant Size Application}
 \label{sec:syntax}
@@
 to be used as generic values (unknowns) and type annotations
-$\up T t$ and $\down T t$ for lazy realizations of the
+$\up T t$ and $\down T t$ for lazy realisations of the
 reflection and reification operations of NbE.
@@
 First, we close the term $t$ with an environment $\eta$ that maps the free de Bruijn indices of $t$ to reflected de Bruijn levels.
-Reflection of de Bruijn levels follows the blue path $\Level \to \Dne \to \D$: Levels embed via constructor $\var{}$ into semantic neutrals $\DNe$ which are labeled with their type $A \in \D$ to become and element $\up A \lev j \in \D$.
+Reflection of de Bruijn levels follows the blue path $\Level \to \Dne \to \D$: Levels embed via constructor $\var{}$ into semantic neutrals $\DNe$ which are labelled with their type $A \in \D$ to become and element $\up A \lev j \in \D$.
 
@@
 
-The PERs $\NE$ characterizes neutral reifiable terms.
+The PERs $\NE$ characterizes neutral reifyable terms.
 % \[
@@
 For instance, assume a term $F : \Nat 0 \to \Set[0]$ which diverges if applied to a successor term.
-Then $T := \funT x {\Nat 0} {F\,x}$ is a well-defined type; we have $T \in \SET[0]$.
+Then $T := \funT x {\Nat 0} {F\,x}$ is a welldefined type; we have $T \in \SET[0]$.
 Now consider $S := \funT x {\Nat \infty} {F\,x}$.
@@
 
-The eliminations for natural numbers are accurately modeled.
+The eliminations for natural numbers are acurrately modeled.
 \begin{lemma}[Case]
@@
 \begin{proof*}
-By well-founded induction on $\alpha$.
+By wellfounded induction on $\alpha$.
 \begin{caselist}
@@
 
-In this section, we show that NbE is sound for judgmental equality, i.e., that same normal form implies definitional equality.
+In this section, we show that NbE is sound for judgemental equality, i.e., that same normal form implies definitional equality.
 The proof follows \citet{abelCoquandDybjer:lics07,abel:habil}.
@@
   \begin{enumerate}
-  \item Well-typedness: $\Gamma \der \sigma : \Delta$
+  \item Welltypedness: $\Gamma \der \sigma : \Delta$
    and $\Gamma \der \sigma = \sigma \eeq \tau : \Delta$
@@
 
-\item For the identity environment $\LRS \Gamma \sid \Gamma \rhog$ (Lemma~\ref{lem:idenv})
+\item For the identity environment $\LRS \Gamma \sid \Gamma \rhog$ (Lemm~\ref{lem:idenv})
 the Fundamental Theorem for Typing gives $\LR \Gamma t T {t\rhog} {T\rhog}$.
@@
    Here we use that the computed normal forms are identical because
-   we erase all size annotations $\ann\alpha$ to $\ann\infty$ during read back.
+   we erase all size annotations $\ann\alpha$ to $\ann\infty$ during readback.
 \qed
@@
 \begin{proof}
-  Follows directly by the fundamental theorem (Thm.~\ref{thm:fund}) and inversion on semantic subtyping.
+  Follows directly by the fundamental theorem (Thm~\ref{thm:fund}) and inversion on semantic subtyping.
 \end{proof}
@@
 \begin{proof}
-  By inversion of typing (Lemma~\ref{lem:invty}), type constructor discrimination \label{lem:discr}
-  and inversion of subtyping (Lemma~\ref{lem:invsubty}).
+  By inversion of typing (Lem~\ref{lem:invty}), type constructor discrimination \label{lem:discr}
+  and inversion of subtyping (Lem~\ref{lem:invsubty}).
 
@@
 \begin{proof}
-  By induction on $\Gamma \der T \asub T'$, using soundness of NbE and subject reduction (Thm.~\ref{thm:sr}).
+  By induction on $\Gamma \der T \asub T'$, using soundness of NbE and subject reduction (Thm~\ref{thm:sr}).
 \end{proof}

### 2017-02-27 nbe/icfp17.tex

--- 2017-02-27 nbe/icfp17.tex
@@
 % \cite{boveKraussSozeau:mscs16}
-Consequently, each of the aforementioned languages based on Type Theory comes with a termination checker, which needs to reject all non-terminating programs, and should accept sufficiently many terminating programs to allow the user to express her algorithms.
+Consequently, each of the aforementioned languages based on Type Theory come with a termination checker, which needs to reject all non-terminating programs, and should accept sufficiently many terminating programs to allow the user to express her algorithms.
 In current termination checkers, programs are required to terminate by structural descent \cite{gimenez:guardeddefinitions}; the structural order may be extended to a lexicographic \cite{abelAltenkirch:jfp02} or size-change termination criterion \cite{jones:sizeChange,wahlstedt:PhD}.

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Typos found by Andrea, beginning of sec 3

--- 2017-02-27 nbe/icfp17.tex
@@
 \]
-Church-style syntax is the basis for all program analyses and transformations to follow and show be considered as the \emph{true} syntax.
+Church-style syntax is the basis for all program analyses and transformations to follow and should be considered as the \emph{true} syntax.
 However, from a dependent-type perspective, explicit size applications
@@
 However, in Curry-style more terms are equated than in Church-style, as for instance the Church-style terms $t\,A\,(\lambda x \of A.\, x)$ and $t\,B\,(\lambda x \of B.\, x)$ map to the same Curry-style term $t\,(\lambda x.\,x)$.
-How would we adopt the algorithm for Church-style such that it equates all terms that are equal in Curry-style?
+How would we adapt the algorithm for Church-style such that it equates all terms that are equal in Curry-style?
 The conclusion of the last rule could be changed to $\Gamma \der t\,A = t'\,A' : B[A/X]$, but then the second term $t'\,A'$ does not have the ascribed type $B[A/X]$, and $\eta$-laws applied to this term would be unsound.
@@
 The new types $F\,i$ and $F\,j$ we reflect at are no longer equal (and they are not subtypes of each other), but they still have the same shape, $\Nat\_ \to \Nat\_$.
-This means they will still move in lock-step what $\eta$-expansion concerns, which is sufficient to prove NbE correct for judgmental equality.
+This means they will still move in lock-step in respect to $\eta$-expansion, which is sufficient to prove NbE correct for judgmental equality.
 We call the enabling property of $F$ \emph{shape irrelevance}, meaning that for any pair $i$, $j$ of legal arguments, $F\,i$ and $F\,j$ have the same shape.
@@
 
-Syntax (de Bruijn terms).
+In this section, we give syntax and \emph{declarative} typing, equality, and subtyping judgements.
+The typing relation $\Gamma \der t : T$ will \emph{not} be decidable; instead, we present algorithmic typing $\Gamma \der t \jchk T$ in Section~\ref{sec:tycheck}.
+However, equality and subtyping will be decidable for well-formed input, see sections~\ref{sec:sem}--\ref{sec:asub}.
+
+We present our type theory as \emph{pure type system} \cite{barendregt:lambdacube} with extra structure.
+\emph{Sorts} $s$ are drawn from an infinite predicative hierarchy of universes $\Set[\ell]$ for $\ell \in \NN$.
+Universes provide us with polymorphism and the capability to define types by recursion on values.
+Whether we have just two universes $\Set[0]$ and $\Set[1]$ or infinitely many, does not matter for the technical difficulty of the meta theory.
+The present setup have the advantage that every sort has again a sort since $\Set[\ell] : \Set[\ell+1]$, thus, we do not have to introduce a separate judgement $\Gamma \der T$ for well-formedness of types, we can define it as $\exists s.\ \Gamma \der T : s$.
+
+\begin{figure}[htbp]
+  \centering
 \[
@@
 \Elim & \ni & e
-    & ::= & t \mid a \mid \ann a \mid \case[\ell] T {t_1} t_2 \mid \fix[\ell] T t & \mbox{eliminations} \\
+    & ::= & t \mid a \mid \ann a \mid \case[\ell] T {t_z} {t_s} \mid \fix[\ell] T t & \mbox{eliminations} \\
 \SizeExp & \ni & a,b & ::= & \infty \mid o \mid \ind i + o ~(o \in \NN) & \mbox{size expressions} \\
@@
 \]
+  \caption{Syntax.}
+  \label{fig:syntax}
+\end{figure}
+
+For the expression syntax (see Fig.~\ref{fig:syntax}), we use de~Bruijn \citeyearpar{deBruijn:nameless} indices $\ind i$ to represent variables.
+Binders are lambda abstraction $\lambda t$ and dependent function types $\epiT U T$, which bind the $0$th index in $t$ and $T$, resp.
+The notation $\epiT U T$ is an umbrella for three kinds of function types, where $\evar \in \{ \erased, \noterased \}$ is a relevance annotation borrowed from \citet{pfenning:intextirr}: $\erpiT[\noterased] U T$, the ordinary dependent function type, $\erpiT[\noterased] \Size T$ for relevant size quantification, and $\erpiT \Size T$ for irrelevant size quantification.
+We omit the ``$\noterased$''-markers from $\Gamma$ and $\Pi$ by default and write $\forallT T$ for $\erpiT \Size T$.
+%
+Examples for relevant size quantification $\piT \Size T$, in syntax with named variables, are $\Size \to \Set[0]$ or $\funT i \Size \Nat i \to \Set[0]$.
+An instance of irrelevant quantification $\forall T$ would be $\forall i.\, \Nat i \to \Nat i$.
+Herein, $\Nat i$ denotes the type of natural numbers below $i$.
+The expression $\Size$ is a possible instance of $U$ in $\epiT U T$, or a possible type of a variable in a typing context $\Gamma$, but not a first-class type, \ie, we cannot construct our own function on sizes.
+
+Canonical natural numbers $c$ are constructed by $\zero a$ and $\suc a t$.
+Size expression $a$ is either a constant $o \in \NN$, a variable $\ind i + o$ possibly with increment $o$, or the limit ordinal $\infty$ which stands for $\omega$.
+The size expression $s$ in the constructors is a suggestion for the type checker but bears no semantic significance.
+For example, in the declarative typing presented here, we can have $\der \zero 5 : \Nat 1$; in the algorithmic typing however, $\der \zero 5 \jchk \Nat 1$ will be an error.
 
 %We write $\sigma$ and $\tau$ for substitutions, which are lists of terms, and
+Substitutions $\sigma$ are simply lists of terms which provide one term as replacement for each free de Bruijn index in a term $t$.
 We write $t \sigma$ for the application of substitution $\sigma$ to term $t$ which is defined as usual.
@@
 
-We omit the ``$\noterased$''-markers from $\Gamma$ and $\Pi$ by default.  We write $\forallT T$ for $\erpiT \Size T$.
 For concrete function types, we may use a named dependent function type notation as syntactic sugar for the corresponding de Bruijn representation.

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Finished description of the syntax.

@@
 Canonical natural numbers $c$ are constructed by $\zero a$ and $\suc a t$.
-Size expression $a$ is either a constant $o \in \NN$, a variable $\ind i + o$ possibly with increment $o$, or the limit ordinal $\infty$ which stands for $\omega$.
-The size expression $s$ in the constructors is a suggestion for the type checker but bears no semantic significance.
-For example, in the declarative typing presented here, we can have $\der \zero 5 : \Nat 1$; in the algorithmic typing however, $\der \zero 5 \jchk \Nat 1$ will be an error.
+Size expression $a$ is either a constant $o \in \NN$, a variable
+$\ind i + o$ possibly with increment $o$, or the limit ordinal
+$\infty$ which stands for $\omega$.  The size argument $a$ in the
+constructors is a suggestion for the type checker but bears no
+semantic significance.  For example, in the declarative typing
+presented here, we can have $\der \zero 5 : \Nat 1$.
+In the algorithmic typing however, $\der \zero 5 \jchk \Nat 1$ will be an error.
+
+Besides with regular application $t\,u$, relevant size application $t\,a$, and irrelevant size application $t \, \ann a$,
+we can eliminate a term $t$ by a case distinction $e = \case[\ell] T {t_z}{t_s}$ or a recursive function application $e' = \fix[\ell'] {T'} {t'}$.
+In both cases, the term should denote a natural number.
+Application of case distinction $\zero a \, e$ will reduce to the zero-branch $t_z$, and $\suc a t \,e$ to the instantiation of the successor branch $t_s\,t$.
+The type annotation $T$ in $\tcase$ allows us to infer the type of the whole case statement $t\,e$ as $T\,t$.
+The function call $c\,e'$ for a canonical number reduces to $f'\,(\lambda x.\,x\,e')\,c$ where we allowed ourself the use of a named abstraction in the presentation to the reader.
+The unfolding of fixed-points is thus restricted to application to canonical numbers; this is the usual reduction strategy which converges for terminating functions \cite{gimenez:typeBased}.
 
 %We write $\sigma$ and $\tau$ for substitutions, which are lists of terms, and
-Substitutions $\sigma$ are simply lists of terms which provide one term as replacement for each free de Bruijn index in a term $t$.
-We write $t \sigma$ for the application of substitution $\sigma$ to term $t$ which is defined as usual.
-We write $\slift^k_m$ for the substitution $(\ind{k+m-1},\dots,\ind{k+1},\ind k)$ which applies to a term with $m$ free indices and increases all of them by $k$.
-We write $\slift_m$ for the lifting $\slift^1_m$ and $\sid_m$ for the identity substitution $ \slift^0_m$.
+For ordinary $\beta$-reduction we employ
+\emph{substitutions} $\sigma$.
+These are simply lists of terms that provide one term as replacement for each free de Bruijn index in a term $t$.
+We write \fbox{$t \sigma$} for the application of substitution $\sigma$ to term $t$ which is defined as usual.
+Let \emph{lifting} \fbox{$\slift^k_m$} be the substitution $(\ind{k+m-1},\dots,\ind{k+1},\ind k)$ which accepts a term with $m$ free indices and increases each of them by $k$.
+We write $\slift_m$ for the lifting $\slift^1_m$ and \fbox{$\sid_m$} for the identity substitution $ \slift^0_m$.
 In general, we refer to liftings by letter $\xi$.
-The substitution $[u]_m = (\sid_m, u)$ replaces free index $\ind 0$ by term $u$ and decrements all other free indices by $1$. We drop subscript $m$ when clear from the context.
-Substitution composition $\sigma\tau$ is the pointwise application of substitution $\tau$ to the list of terms $\sigma$.  In the proofs to follow, we freely use the following identities:
+The substitution \fbox{$[u]_m$}${} = (\sid_m, u)$ replaces free index $\ind 0$ by term $u$ and decrements the other $m$ free indices by $1$.
+We drop subscript $m$ from liftings when clear from the context.
+Substitution composition \fbox{$\sigma\tau$} is the pointwise application of substitution $\tau$ to the list of terms $\sigma$.  In the proofs to follow, we freely use the following identities:
 \begin{gather*}
@@
 
-For concrete function types, we may use a named dependent function type notation as syntactic sugar for the corresponding de Bruijn representation.
+For concrete function types and terms, we may use a named dependent function type notation as syntactic sugar for the corresponding de Bruijn representation.
 For instance, $\funT i \Size \Nat i \to \Set[\ell]$ is sugar for $\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.
-We abbreviate this type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for $\allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$.
+We abbreviate this type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for
+$\forall i.\ (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$.
+% $\allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$.
 Similar as for $\Pi$, we use named lambda abstraction as sugar for de Bruijn abstraction.
-Named abstract takes care of proper lifting of de Bruijn indices, for instance, $\lambda x.\, t x = \lambda.
+Named abstraction takes care of proper lifting of de Bruijn indices, for instance, $\lambda x.\, t x = \lambda.
 (t \slift)\,\ind 0$ if $t$ is outside the scope of $x$.
@@
 
-Resurrection\cite{pfenning:lics01} \fbox{$\resurrect\Gamma$} removes
-the ``$\erased$''-markers from all bindings in $\Gamma$ (\ie, replaces
-them by ``$\noterased$''-markers).  Note that, trivially, resurrection is idempotent: $\resurrecttwice\Gamma = \resurrect{\Gamma}$.
+In typing contexts $\Gamma$, we distinguish relevant $(\noterased)$ and irrelevant $(\erased)$ bindings.
+When type checking a variable, it needs to be bound in the context relevantly.
+However, when entering a irrelevant position, for instance when checking size $a$ in term $\suc a t$ we declare previously irrelevant variables as relevant.
+This operation on the context has been coined
+\emph{resurrection} by \citet{pfenning:lics01}; formally \fbox{$\resurrect\Gamma$} removes
+the ``$\erased$''-markers from all bindings in $\Gamma$, \ie, replaces
+them by ``$\noterased$''-markers.  Note that, trivially, resurrection is idempotent: $\resurrecttwice\Gamma = \resurrect{\Gamma}$.
 
 Size increment \fbox{$a + o'$} for $o' \in \NN$ extends addition by $\infty + o' = \infty$ and $(\ind i + o) + o' = \ind i + (o + o')$.
-Size comparison \fbox{$a \leq b$} holds if either $b = \infty$ or  $o \leq o'$ where either $a = o$ and $b = o'$ or $a \in \{ o, \ind i + o \}$ and $b = \ind i + o'$.
-
-Judgements.
+Sizes are partially ordered; size comparison \fbox{$a \leq b$} holds as expected if either $b = \infty$ or  $o \leq o'$ where either $a = o$ and $b = o'$ or $a \in \{ o, \ind i + o \}$ and $b = \ind i + o'$.
+
+Our type theory is formulated via the following judgements, which are mutually inductively defined.
 \[
@@
 \]
-Derived judgements.
-\[
-\begin{array}{l@{~~}l@{~~}l@{~}l@{\qquad}l}
+In addition, we will refer to the following derived judgements.
+\[
+\begin{array}{l@{\quad}l@{\quad}l@{~}l@{~~}l}
   \Gamma \der T & \defas & \Gamma \der T : s & \mforsome s
@@
   \Gamma \der a = b : \Size & \defas &
-    \multicolumn 2 {@{}l@{\qquad}} {\Gamma \der a : \Size \mand a = b}
-    & \mbox{sizes $a$ and $b$ are well-formed and identical} \\
+    \multicolumn 2 {@{}l@{\quad}} {\Gamma \der a : \Size \mand a = b}
+    & \mbox{sizes $a$, $b$ are well-formed and identical} \\
   \Gamma \der a \leq b : \Size & \defas &
-     \multicolumn 2 {@{}l@{\qquad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
+     \multicolumn 2 {@{}l@{\quad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
 %     \multicolumn 2 {l@{\qquad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
-    & \mbox{sizes $a$ and $b$ are well-formed and comparable} \\
+    & \mbox{sizes $a$, $b$ are well-formed and comparable} \\
 %    & \mbox{in context $\Gamma$, well-formed size $a$ is less or equal than well-formed size $b$} \\
   \Gamma \der T : \Adm \ell & \defas &
-     \multicolumn 2 {@{}l@{\qquad}}
+     \multicolumn 2 {@{}l@{\quad}}
        {\Gamma \der T : \FixK \mand \Gamma.i\of\Size.x\of\Nat i \der T\,i\,x \leq T\,\infty\,x}

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Fixed overfull hbox for derived judgements.

@@
 Sizes are partially ordered; size comparison \fbox{$a \leq b$} holds as expected if either $b = \infty$ or  $o \leq o'$ where either $a = o$ and $b = o'$ or $a \in \{ o, \ind i + o \}$ and $b = \ind i + o'$.
+Different size variables are incomparable.
 
@@
   \Gamma \der T & \defas & \Gamma \der T : s & \mforsome s
-    & \mbox{in context $\Gamma$, type $T$ is well-formed} \\
+    & \mbox{in $\Gamma$, type $T$ is well-formed} \\
   \Gamma \der T = T' & \defas & \Gamma \der T = T' : s & \mforsome s
-    & \mbox{in context $\Gamma$, types $T$ and $T'$ are equal} \\
+    & \mbox{in $\Gamma$, types $T$ and $T'$ are equal} \\
   \Gamma \der a = b : \Size & \defas &
     \multicolumn 2 {@{}l@{\quad}} {\Gamma \der a : \Size \mand a = b}
-    & \mbox{sizes $a$, $b$ are well-formed and identical} \\
+    & \mbox{$a$, $b$ are identical sizes} \\
   \Gamma \der a \leq b : \Size & \defas &
@@
 %     \multicolumn 2 {l@{\qquad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
-    & \mbox{sizes $a$, $b$ are well-formed and comparable} \\
+    & \mbox{$a$, $b$ are comparable sizes} \\
 %    & \mbox{in context $\Gamma$, well-formed size $a$ is less or equal than well-formed size $b$} \\
@@
     & \mbox{% in context $\Gamma$
-            type constructor $T$ is admissible for recursion} \\
+            % type constructor
+            $T$ is admissible for recursion} \\
 \end{array}

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Moved syntactic judgements to own file.

@@
 
-\usepackage[right]{showlabels}
-\renewcommand{\showlabelfont}{\small\ttfamily\color{gray}}
+\usepackage[right]{showlabels}\renewcommand{\showlabelfont}{\small\ttfamily\color{gray}}
 
@@
 
+\input{figjudge}
+
 In typing contexts $\Gamma$, we distinguish relevant $(\noterased)$ and irrelevant $(\erased)$ bindings.
@@
 
-Our type theory is formulated via the following judgements, which are mutually inductively defined.
-\[
-\begin{array}{l@{\qquad}l}
-  \der \Gamma & \mbox{context $\Gamma$ is well-formed} \\
-  \Gamma(i) = \eann T & \mbox{in context $\Gamma$, de Bruijn index $i$ has type $T$ and annotation $\evar$} \\
-  \Gamma \der a : \Size & \mbox{in context $\Gamma$, size expression $a$ is well-formed} \\
-  \Gamma \der t : T & \mbox{in context $\Gamma$, term $t$ has type $T$} \\
-  \Gamma \der t = t' : T & \mbox{in context $\Gamma$, terms $t$ and $t'$ are equal of type $T$} \\
-%  \Gamma \mid U \der e : T & \mbox{in context $\Gamma$, elimination $e$ goes from $U$ to $T$} \\
-%  \Gamma \mid U \der e = e' : T & \mbox{in context $\Gamma$, eliminations $e$ and $e'$ are equal} \\
-  \Gamma \der T \leq T' & \mbox{in context $\Gamma$, type $T$ is a subtype of $T'$} \\
-%  \Gamma \der T : \Adm \ell & \mbox{in context $\Gamma$ type constructor $T$ is admissible for recursion} \\
-  \suty \Gamma \sigma \tau \Delta & \mbox{$\sigma$/$\tau$ is a valid term/type-level substitution for $\Delta$} \\
-  \sueq \Gamma \sigma {\sigma'} \tau \Delta & \mbox{$\sigma$/$\sigma'$/$\tau$ are a equal term/term/type-level substitutions for $\Delta$} \\
-\end{array}
-\]
-In addition, we will refer to the following derived judgements.
-\[
-\begin{array}{l@{\quad}l@{\quad}l@{~}l@{~~}l}
-  \Gamma \der T & \defas & \Gamma \der T : s & \mforsome s
-    & \mbox{in $\Gamma$, type $T$ is well-formed} \\
-  \Gamma \der T = T' & \defas & \Gamma \der T = T' : s & \mforsome s
-    & \mbox{in $\Gamma$, types $T$ and $T'$ are equal} \\
-  \Gamma \der a = b : \Size & \defas &
-    \multicolumn 2 {@{}l@{\quad}} {\Gamma \der a : \Size \mand a = b}
-    & \mbox{$a$, $b$ are identical sizes} \\
-  \Gamma \der a \leq b : \Size & \defas &
-     \multicolumn 2 {@{}l@{\quad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
-%     \multicolumn 2 {l@{\qquad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
-    & \mbox{$a$, $b$ are comparable sizes} \\
-%    & \mbox{in context $\Gamma$, well-formed size $a$ is less or equal than well-formed size $b$} \\
-  \Gamma \der T : \Adm \ell & \defas &
-     \multicolumn 2 {@{}l@{\quad}}
-       {\Gamma \der T : \FixK \mand \Gamma.i\of\Size.x\of\Nat i \der T\,i\,x \leq T\,\infty\,x}
-    & \mbox{% in context $\Gamma$
-            % type constructor
-            $T$ is admissible for recursion} \\
-\end{array}
-\]
+Fig.~\ref{fig:judge} lists the inductive and derived judgements of our type theory.
+
+%Our type theory is formulated via the following judgements, which are mutually inductively defined.
+
+%In addition, we will refer to the following derived judgements.
+
 Inference rules.
@@
 
-Liftings: $\fbox{$\xi : \Gamma \leq \Delta$}
-  \defas \Gamma \der \xi : \Delta \mand \xi = \slift^k_m \mforsome k,m$
-%  \defas \xi = \slift^k_m \mwith \Gamma \der \xi : \Delta$
-(implies $\der \Gamma$ and $\der \Delta$).
+% Liftings: $\fbox{$\xi : \Gamma \leq \Delta$}
+%   \defas \Gamma \der \xi : \Delta \mand \xi = \slift^k_m \mforsome k,m$
+% %  \defas \xi = \slift^k_m \mwith \Gamma \der \xi : \Delta$
+% (implies $\der \Gamma$ and $\der \Delta$).
+
 % (implies $\der \Gamma$ and $\der \Delta$ and $\Gamma \der \xi : \Delta$).
@@
 \item
-  If\/ $\Gamma \der \sigma = \sigma' \eeq tau : \Delta$ and $\Delta \der t : T$ then
+  If\/ $\Gamma \der \sigma = \sigma' \eeq \tau : \Delta$ and $\Delta \der t : T$ then
   $\Gamma \der t\sigma : T\tau$ and $\Gamma \der t\sigma' : T\tau$.

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Moved equality rules to own file.
[ ICFP 17 ] Moved typing rules etc. to own file.
\tau instead of tau
[ ICFP 17 ] Space correction after if.

@@
 (t \slift)\,\ind 0$ if $t$ is outside the scope of $x$.
-We may also use names when we construct concrete contexts, for instance, if $T$ is well-formed in context $\Gamma$, we may write $T\,i\,x$ in context $\Gamma.i\of\Size.x\of\Nat i$ to mean $T\slift^2\,\ind 1\,\ind 0$ in context
+We may also use names when we construct concrete contexts, for instance, if\/ $T$ is well-formed in context $\Gamma$, we may write $T\,i\,x$ in context $\Gamma.i\of\Size.x\of\Nat i$ to mean $T\slift^2\,\ind 1\,\ind 0$ in context
 $\Gamma.\Size.\Nat \ind 0$.
@@
 
-Fig.~\ref{fig:judge} lists the inductive and derived judgements of our type theory.
+Fig.~\ref{fig:judge} lists the inductive and derived judgements of our type theory and
+figures~\ref{fig:ty} and \ref{fig:eq} the inference rules.
+
+\input{figty}
+\input{figeq}
 
@@
 
-Inference rules.
-
-\fbox{$\der \Gamma$} (implies $\der \resurrect \Gamma$).
-\[
-  \ru{}{\der \cempty}
-\qquad
-  \ru{\der \Gamma \qquad
-      \resurrect \Gamma \der T
-    }{\der \cext \Gamma T}
-\qquad
-  \ru{\der \Gamma
-    }{\der \cext \Gamma \Size}
-\qquad
-  \ru{\der \Gamma
-    }{\der \erext \Gamma \Size}
-\]
-
-\fbox{$\Gamma(i) = \eann T$} (implies $\resurrect\Gamma \der T$).
-\[
-  \ru{}{(\eext \Gamma T)(0) = \eann T \slift}
-\qquad
-  \ru{\Gamma(i) = \eann T
-    }{(\cext \Gamma \_)(i+1) = \eann T \slift}
-\]
-
-\fbox{$\Gamma \der a : \Size$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der a : \Size$).
-\[
-  \ru{\der \Gamma}{\Gamma \der \infty : \Size}
-\qquad
-  \rux{\der \Gamma}{\Gamma \der o : \Size}{o \in \NN}
-\qquad
-  \ru{\der \Gamma \qquad \Gamma(i) = \erann[\noterased]\Size
-    }{\Gamma \der \ind i + o : \Size
-    }
-\]
-
-\fbox{$\Gamma \der t : T$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der T$ [and $\resurrect \Gamma \der t : T$]. Note: no rule for $\Gamma \der \Size : s$.)
-\begin{gather*}
-  \ru{\der \Gamma
-    }{\Gamma \der \Set[\ell] : \Set[\ell']
-    }{\ell {<} \ell'}
-\qquad
-  \ru{\Gamma \der U : s \qquad
-      \cext \Gamma U \der T : s
-    }{\Gamma \der \piT U T : s}
-\qquad
-  \ru{\cext \Gamma \Size \der T : s
-    }{\Gamma \der \epiT \Size T : s}
-\qquad
-  \ru{%\der \Gamma \qquad
-      \Gamma \der a : \Size
-    }{\Gamma \der \Nat a : \Set[0]}
-\\[2ex]
-  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
-    }{\Gamma \der \ind i : T
-    }{T \not= \Size}
-\qquad
-  \ru{\eext \Gamma U \der t : T
-    }{\Gamma \der \lambda t : \epiT U T}
-\qquad
-  \ru{\Gamma \der t : \piT U T \qquad
-      \Gamma \der u : U
-    }{\Gamma \der t\,u : T[u]}
+% Inference rules.
+
+% \fbox{$\der \Gamma$} (implies $\der \resurrect \Gamma$).
+% \[
+%   \ru{}{\der \cempty}
 % \qquad
-%   \ru{\Gamma \der t : T \qquad
-%       \Gamma \mid T \der e : T'
-%     }{\Gamma \der t\,e : T'}
-\qquad
-  \ru{\Gamma \der t : \piT \Size T \qquad
-      \Gamma \der a : \Size
-    }{\Gamma \der t\,a : T[a]}
-\\[2ex] % \qquad
-  \ru{%\Gamma \der a : \Size \qquad
-      \resurrect\Gamma \der a, b : \Size
-    }{\Gamma \der \zero{a} : \Nat (b + 1)}
-\qquad
-  \ru{\resurrect\Gamma \der a : \Size \qquad
-      \Gamma \der t : \Nat b
-    }{\Gamma \der \suc a t : \Nat (b + 1)}
-\qquad
-  \ru{\Gamma \der t : \forallT T \qquad
-      %\Gamma \der a : \Size \qquad
-      \resurrect\Gamma \der a,b : \Size
-    }{\Gamma \der t \ann a : T[b]}
-\\[2ex]
-  \ru{\Gamma \der u : \Nat (a + 1) \qquad
-      \Gamma \der T : \Nat (a + 1) \to \Set[\ell] \qquad
-      \Gamma \der t_z : T \, (\zero{a}) \qquad
-      \Gamma \der t_s : (x : \Nat a) \to T \, (\suc{a} x)
-    }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
-\\[2ex]
-  \ru{\Gamma \der u : \Nat a \qquad
-      \Gamma \der T : \Adm\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
-      \Gamma \der t : \FixT\, T
-     }{\Gamma \der u\,\fix[\ell] T t : T \,a \, u}
-\qquad
-  \ru{\Gamma \der t : T \qquad
-      \resurrect\Gamma \der T \leq T'
-     }{\Gamma \der t : T'}
-\end{gather*}
-Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
+%   \ru{\der \Gamma \qquad
+%       \resurrect \Gamma \der T
+%     }{\der \cext \Gamma T}
+% \qquad
+%   \ru{\der \Gamma
+%     }{\der \cext \Gamma \Size}
+% \qquad
+%   \ru{\der \Gamma
+%     }{\der \erext \Gamma \Size}
+% \]
+
+% \fbox{$\Gamma(i) = \eann T$} (implies $\resurrect\Gamma \der T$ if\/ $\der \Gamma$).
+% \[
+%   \ru{}{(\eext \Gamma T)(0) = \eann T \slift}
+% \qquad
+%   \ru{\Gamma(i) = \eann T
+%     }{(\cext \Gamma \_)(i+1) = \eann T \slift}
+% \]
+
+%Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
 % \fbox{$\Gamma \mid U \der e : T$}
@@
 
-\fbox{$\Gamma \der t = t' : T$} (implies $\resurrect \Gamma \der T$ and $\Gamma \der t,t' : T$ [and $\resurrect\Gamma \der t = t' : T$].)
-
-Computation rules.
-\begin{gather*}
-  \ru{\cext \Gamma U \der t : T \qquad
-      \Gamma \der u : U
-    }{\Gamma \der (\lambda t)\,u = t[u] : T[u]}
-\qquad
-  \ru{\cext \Gamma \Size \der t : T \qquad
-      \Gamma \der a : \Size
-    }{\Gamma \der (\lambda t)\,a = t[a] : T[a]}
-\qquad
-  \ru{\erext \Gamma \Size \der t : T \qquad
-      % \Gamma \der a : \Size \qquad
-      \resurrect \Gamma \der a, b : \Size
-    }{\Gamma \der (\lambda t)\,\ann a = t[a] : T[b]}
-\\[2ex]
-  \ru{\resurrect\Gamma \der a, b : \Size \qquad
-      \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
-      % \Gamma \der a : \Size \qquad
-      \Gamma \der t_z : T\,\zero{b} \qquad
-      \Gamma \der t_s : (x : \Nat b) \to T\,(\suc{b}x)
-    }{\Gamma \der (\zero{a})\,\case[\ell] T {t_z} {t_s} = t_z : T\,\zero{b}}
-\\[2ex]
-  \ru{\resurrect\Gamma \der a : \Size \qquad
-      \Gamma \der t : \Nat b \qquad
-      \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \qquad
-      \Gamma \der t_z : T\,\zero{b} \qquad
-      \Gamma \der t_s : (x : \Nat b) \to T\,(\suc{b}x)
-    }{\Gamma \der (\suc{a} t)\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,(\suc{b} t)}
-\\[2ex]
-  \ru{\Gamma \der c : \Nat b \qquad
-      \resurrect\Gamma \der a : \Size \qquad
-      \Gamma \der T : \Adm\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
-      \Gamma \der t : \FixT\,T %% \allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x
-     }{\Gamma \der c\,\fix[\ell] T t = t\ann a (\lambda x. x\,\fix[\ell] T t)\, c : T \,b \, c
-     }%{c \in \{\zero a, \suc a t'\}}
-\end{gather*}
-Extensionality rules.
-\begin{gather*}
-  \ru{\Gamma \der t : \piT U T
-    }{\Gamma \der t = \lambda x. t\,x : \piT U T}
-\qquad
-  \ru{\Gamma \der t : \forallT T \qquad
-      \cext {\resurrect\Gamma} \Size \der a : \Size
-      %\erext \Gamma \Size \der a : \Size
-    }{\Gamma \der t = \lambda x. t \ann a : \forallT T}
-\end{gather*}
-Congruence rules.
-\begin{gather*}
-  \ru{\der \Gamma
-    }{\Gamma \der \Set[\ell] = \Set[\ell]: \Set[\ell']
-    }{\ell {<} \ell'}
-\qquad
-  \ru{\Gamma \der U = U' : s \qquad
-      \cext \Gamma U \der T = T' : s
-    }{\Gamma \der \piT U T = \piT{U'}{T'}: s}
-\qquad
-  \ru{\cext \Gamma \Size \der T = T' : s
-    }{\Gamma \der \epiT \Size T = \epiT \Size T' : s}
-\\[2ex]
-  \ru{%\der \Gamma \qquad
-      \Gamma \der a : \Size
-    }{\Gamma \der \Nat a = \Nat a : \Set[0]}
-\\[2ex]
-  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
-    }{\Gamma \der \ind i = \ind i : T
-    }{T \not= \Size}
-\qquad
-  \ru{\eext \Gamma U \der t = t' : T
-    }{\Gamma \der \lambda t = \lambda t' : \epiT U T}
-\qquad
-  \ru{\Gamma \der t = t' : \piT U T \qquad
-      \Gamma \der u = u' : U
-    }{\Gamma \der t\,u = t'\,u' : T[u]}
-% \qquad
-%   \ru{\Gamma \der t : T \qquad
-%       \Gamma \mid T \der e : T'
-%     }{\Gamma \der t\,e : T'}
-\\[2ex]
-  \ru{\Gamma \der t = t' : \piT \Size T \qquad
-      \Gamma \der a : \Size
-    }{\Gamma \der t\,a = t'\,a : T[a]}
-\qquad
-  \ru{\Gamma \der t = t' : \forallT T \qquad
-      % \Gamma \der a, a' : \Size \qquad
-      \resurrect\Gamma \der a,a',b : \Size
-    }{\Gamma \der t \ann a = t' \ann{a'} : T[b]}
-\\[2ex] % \qquad
-  \ru{%\Gamma \der a, a' : \Size \qquad
-      \resurrect\Gamma \der a, a', b : \Size
-    }{\Gamma \der \zero a = \zero {a'}: \Nat (b + 1)}
-\qquad
-  \ru{\resurrect \Gamma \der a, a' : \Size \qquad
-      \Gamma \der t = t' : \Nat b
-    }{\Gamma \der \suc a t = \suc {a'} t' : \Nat (b + 1)}
-\\[2ex]
-  \ru{\Gamma \der u = u' : \Nat (a + 1) \qquad
-      \Gamma \der T = T' : \Nat (a + 1) \to \Set[\ell] \\
-      \Gamma \der t_z = t_z' : T \, (\zero a) \qquad
-      \Gamma \der t_s = t_s' : (x : \Nat a) \to T \, (\suc a x)
-    }{\Gamma \der u\,\case[\ell] T {t_z} t_s = u'\,\case[\ell] {T'} {t_z'} t_s' : T \, u}
-\\[2ex]
-  \ru{\Gamma \der u = u' : \Nat a \qquad
-      \Gamma \der T = T' : \Adm\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
-      \Gamma \der t = t' : \FixT\,T \qquad %% \allT i \Size (\funT x {\Nat i} T \,i\, x) \to \funT x {\Nat (i + 1)} T \,{i+1}\, x
-     }{\Gamma \der u\,\fix[\ell] T t = u'\,\fix[\ell] {T'} t' : T \, a \, u}
-\qquad
-  \ru{\Gamma \der t = t' : T \qquad
-      \resurrect\Gamma \der T \leq T'
-     }{\Gamma \der t = t' : T'}
-\end{gather*}
-Equivalence rules.  (Reflexivity is admissible.)
-\begin{gather*}
-  \ru{\Gamma \der t = t' : T
-    }{\Gamma \der t' = t : T}
-\qquad
-  \ru{\Gamma \der t_1 = t_2 : T \qquad
-      \Gamma \der t_2 = t_3 : T
-    }{\Gamma \der t_1 = t_3 : T}
-\end{gather*}
-
-\fbox{$\Gamma \der T \leq T'$} (implies $\Gamma \der T,T'$)
-% (implies $\resurrect \Gamma \der T,T'$)
-\begin{gather*}
-  \ru{\der \Gamma \qquad \ell {\leq} \ell'
-    }{\Gamma \der \Set[\ell] \leq \Set[\ell']}
-\qquad
-  \ru{\Gamma \der a \leq b : \Size
-    }{\Gamma \der \Nat a \leq \Nat b}
-\qquad
-  \ru{\Gamma \der T = T'
-    }{\Gamma \der T \leq T'}
-\\[2ex]
-  \ru{\Gamma \der U' \leq U \qquad
-      \cext \Gamma {U'} \der T \leq T'
-    }{\Gamma \der \piT U T \leq \piT {U'} {T'}}
-\qquad
-  \ru{\cext \Gamma {\Size} \der T \leq T'
-    }{\Gamma \der \epiT \Size T \leq \epiT {\Size} {T'}}
-\qquad
-  \ru{\Gamma \der T_1 \leq T_2 \qquad
-      \Gamma \der T_2 \leq T_3
-    }{\Gamma \der T_1 \leq T_3}
-\qquad
-\end{gather*}
-
-Substitution typing. % and equality.
-\fbox{$\Gamma \der \tau : \Delta$}
-(implies $\der \Gamma$ and $\der \Delta$
- %% and $\Gamma \der \tau : \resurrect \Delta$  %NO
- [and $\resurrect\Gamma \der \tau : \Delta$]
- and $\resurrect\Gamma \der \tau : \resurrect \Delta$).
-
-\begin{gather*}
-  \ru{\der \Gamma \qquad
-    }{\Gamma \der \sempty : \cempty}
-\qquad
-  \ru{\Gamma \der \tau : \Delta \qquad
-      \resurrect \Delta \der T \qquad
-      \Gamma \der t : T \tau
-    }{\Gamma \der \sext \tau t : \cext \Delta T}
-\qquad
-  \ru{\Gamma \der \tau : \Delta \qquad
-      \Gamma \der a : \Size
-    }{\Gamma \der \sext \tau a : \cext \Delta \Size}
-\qquad
-  \ru{\Gamma \der \tau : \Delta \qquad
-      \resurrect\Gamma \der a : \Size
-    }{\Gamma \der \sext \tau a : \erext \Delta \Size}
-\end{gather*}
+% \fbox{$\Gamma \der t = t' : T$} (implies $\resurrect \Gamma \der T$ and $\Gamma \der t,t' : T$ [and $\resurrect\Gamma \der t = t' : T$].)
 
@@
 % \fbox{$\suty \Gamma \sigma \tau \Delta$} and
-
-Substitution equality
-\fbox{$\sueq \Gamma \sigma {\sigma'} \tau \Delta$}
-(implies $\der \Gamma$ and $\der \Delta$ and $\Gamma \der \tau : \Delta$ and $\sueq \Gamma {\sigma'} {\sigma} \tau \Delta$).
-
 % \begin{gather*}
@@
 
-\begin{gather*}
-  \ru{\der \Gamma \qquad
-    }{\sueq \Gamma \sempty \sempty \sempty \cempty}
-\qquad
-  \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
-      \resurrect \Delta \der T \qquad
-      \Gamma \der u = u' = t : T \tau
-    }{\sueq \Gamma {\sext \sigma u } {\sext {\sigma'} {u'}} {\sext \tau t} {\cext \Delta T}}
-\\[2ex]
-  \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
-      \Gamma \der a = a' = b : T \tau
-    }{\sueq \Gamma {\sext \sigma a } {\sext {\sigma'} {a'}} {\sext \tau b} {\cext \Delta \Size}}
-\qquad
-  \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
-      \resurrect\Gamma \der a, a', b : T \tau
-    }{\sueq \Gamma {\sext \sigma a } {\sext {\sigma'} {a'}} {\sext \tau b} {\erext \Delta \Size}}
-\end{gather*}
-
-\begin{lemma}[Reflexivity]
-  If\/ $\Gamma \der t : T$ then $\Gamma \der t = t : T$.
-\end{lemma}
+%% COULD BE A LEMMA INSTEAD OF A RULE
+% \begin{lemma}[Reflexivity]
+%   If\/ $\Gamma \der t : T$ then $\Gamma \der t = t : T$.
+% \end{lemma}
 
@@
   \begin{enumerate}
-  \item If $\der \cext \Gamma \Delta$ then $\der \Gamma$
+  \item If\/ $\der \cext \Gamma \Delta$ then $\der \Gamma$
   \item If\/ $\Gamma \der J$ then $\der \Gamma$.
@@
   % Consequently, if\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
-  % Further, if $\Gamma \der J : \resurrect\Delta$ then $\Gamma \der J : \Delta$.
+  % Further, if\/ $\Gamma \der J : \resurrect\Delta$ then $\Gamma \der J : \Delta$.
 \item If\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
@@
 \begin{enumerate}
-\item If $\der \cext \Gamma \Delta$ then $\cext \Gamma \Delta \der \slift_{|\Gamma|}^{|\Delta|} : \Gamma$.
-      If $\der \cext \Gamma T$ then $\cext \Gamma T \der \slift : \Gamma$.
-\item If $\der \Gamma$ then $\Gamma \der \sid : \Gamma$.
-\item If $\Gamma \der u : U$ then $\Gamma \der \sg u : \cext \Gamma U$.
+\item If\/ $\der \cext \Gamma \Delta$ then $\cext \Gamma \Delta \der \slift_{|\Gamma|}^{|\Delta|} : \Gamma$.
+      If\/ $\der \cext \Gamma T$ then $\cext \Gamma T \der \slift : \Gamma$.
+\item If\/ $\der \Gamma$ then $\Gamma \der \sid : \Gamma$.
+\item If\/ $\Gamma \der u : U$ then $\Gamma \der \sg u : \cext \Gamma U$.
 \end{enumerate}
@@
   \item
-  If $\Delta \der t : T$ then $\Gamma \der t\sigma = t\sigma' : T\tau$.
+  If\/ $\Delta \der t : T$ then $\Gamma \der t\sigma = t\sigma' : T\tau$.
   \item
-  If $\Delta \der t = t' : T$ then $\Gamma \der t\sigma = t'\sigma' : T\tau$.
+  If\/ $\Delta \der t = t' : T$ then $\Gamma \der t\sigma = t'\sigma' : T\tau$.
   \item
-  Corollary: If $\Delta \der T \leq T'$ then $\Gamma \der T\sigma \leq T\sigma'$.
+  Corollary: If\/ $\Delta \der T \leq T'$ then $\Gamma \der T\sigma \leq T\sigma'$.
   \end{enumerate}
@@
 \begin{enumerate}
-\item If $\Gamma \der \Nat a : T'$ then $\Gamma \der a : \Size$ and $\resurrect\Gamma \der \Set[0] \leq T'$.
-\item If $\Gamma \der \Set[\ell] : T'$ then $\resurrect\Gamma \der \Set[\ell+1] \leq T'$.
-\item If $\Gamma \der \piT U T : T'$ then $\Gamma \der U : s$ and $\cext \Gamma U \der T : s$ and $\resurrect\Gamma \der s \leq T'$ for some $s$.
-\item If $\Gamma \der \epiT \Size T : T'$ then $\cext\Gamma \Size \der T : s$ and $\resurrect\Gamma \der s \leq T'$.
-\item If $\Gamma \der \ind i : T'$ then $\Gamma(i) = \erann[\noterased] T$ and $\resurrect\Gamma \der T \leq T'$ for some $T$.
-\item If $\Gamma \der \lambda t : T'$ then either
+\item If\/ $\Gamma \der \Nat a : T'$ then $\Gamma \der a : \Size$ and $\resurrect\Gamma \der \Set[0] \leq T'$.
+\item If\/ $\Gamma \der \Set[\ell] : T'$ then $\resurrect\Gamma \der \Set[\ell+1] \leq T'$.
+\item If\/ $\Gamma \der \piT U T : T'$ then $\Gamma \der U : s$ and $\cext \Gamma U \der T : s$ and $\resurrect\Gamma \der s \leq T'$ for some $s$.
+\item If\/ $\Gamma \der \epiT \Size T : T'$ then $\cext\Gamma \Size \der T : s$ and $\resurrect\Gamma \der s \leq T'$.
+\item If\/ $\Gamma \der \ind i : T'$ then $\Gamma(i) = \erann[\noterased] T$ and $\resurrect\Gamma \der T \leq T'$ for some $T$.
+\item If\/ $\Gamma \der \lambda t : T'$ then either
   $\cext \Gamma U \der t : T$
@@
   for some $T$.
-\item If $\Gamma \der t\,u : T'$ then $\Gamma \der t : \piT U T$ and $\Gamma \der u : U$ and $\resurrect\Gamma \der T[u] \leq T'$ for some $U$, $T$.
-\item If $\Gamma \der t\,a : T'$ then $\Gamma \der t : \piT \Size T$ and $\Gamma \der a : \Size$ and $\resurrect\Gamma \der T[a] \leq T'$ for some $T$.
-\item If $\Gamma \der t\,\ann a : T'$ then $\Gamma \der t : \forallT T$ and $\resurrect\Gamma \der a : \Size$ and $\resurrect\Gamma \der T[a] \leq T'$ for some $T$.
-\item If $\Gamma \der \zero a : T'$ then $\resurrect \Gamma \der a,b : \Size$ and $\resurrect\Gamma \der \Nat (b + 1) \leq T'$ for some $b$.
-\item If $\Gamma \der \suc a t : T'$ then $\resurrect \Gamma \der a,b : \Size$ and $\Gamma \der t : \Nat b$ and $\resurrect\Gamma \der \Nat(b+1) \leq T'$.
-\item If $\Gamma \der u \case[\ell] T {t_z} {t_s} : T'$ then $\Gamma \der u : \Nat (a+1)$ and $\Gamma \der T : \Nat(a+1) \to \Set[\ell]$ and $\Gamma \der t_z : T\,(\zero a)$ and $\Gamma \der t_s : \funT x {\Nat a} {T\,(\suc a x)}$ and $\resurrect\Gamma \der T\,u \leq T'$ for some $a$.
-\item If $\Gamma \der u \fix[\ell] T t : T'$ then $\Gamma \der u : \Nat a$ and $\Gamma \der T : \Adm\,\ell$ and $\Gamma \der t : \FixT\,T$ and $\resurrect\Gamma \der T\,a\,u \leq T'$.
+\item If\/ $\Gamma \der t\,u : T'$ then $\Gamma \der t : \piT U T$ and $\Gamma \der u : U$ and $\resurrect\Gamma \der T[u] \leq T'$ for some $U$, $T$.
+\item If\/ $\Gamma \der t\,a : T'$ then $\Gamma \der t : \piT \Size T$ and $\Gamma \der a : \Size$ and $\resurrect\Gamma \der T[a] \leq T'$ for some $T$.
+\item If\/ $\Gamma \der t\,\ann a : T'$ then $\Gamma \der t : \forallT T$ and $\resurrect\Gamma \der a : \Size$ and $\resurrect\Gamma \der T[a] \leq T'$ for some $T$.
+\item If\/ $\Gamma \der \zero a : T'$ then $\resurrect \Gamma \der a,b : \Size$ and $\resurrect\Gamma \der \Nat (b + 1) \leq T'$ for some $b$.
+\item If\/ $\Gamma \der \suc a t : T'$ then $\resurrect \Gamma \der a,b : \Size$ and $\Gamma \der t : \Nat b$ and $\resurrect\Gamma \der \Nat(b+1) \leq T'$.
+\item If\/ $\Gamma \der u \case[\ell] T {t_z} {t_s} : T'$ then $\Gamma \der u : \Nat (a+1)$ and $\Gamma \der T : \Nat(a+1) \to \Set[\ell]$ and $\Gamma \der t_z : T\,(\zero a)$ and $\Gamma \der t_s : \funT x {\Nat a} {T\,(\suc a x)}$ and $\resurrect\Gamma \der T\,u \leq T'$ for some $a$.
+\item If\/ $\Gamma \der u \fix[\ell] T t : T'$ then $\Gamma \der u : \Nat a$ and $\Gamma \der T : \Adm\,\ell$ and $\Gamma \der t : \FixT\,T$ and $\resurrect\Gamma \der T\,a\,u \leq T'$.
 \end{enumerate}
@@
   \item $\lev k = \lev k \in \NE$.
-  \item If $n = n' \in \NE$ and $e = e' \in \ELIM$ then $n\,e = n'\,e' \in \NE$.
+  \item If\/ $n = n' \in \NE$ and $e = e' \in \ELIM$ then $n\,e = n'\,e' \in \NE$.
   \end{enumerate}
@@
   \begin{enumerate}
-  \item If $d = d' \in \NF$ then $d = d' \in \ELIM$.
-  \item If $\alpha \in \SIZE$ then $\alpha = \alpha \in \ELIM$.
+  \item If\/ $d = d' \in \NF$ then $d = d' \in \ELIM$.
+  \item If\/ $\alpha \in \SIZE$ then $\alpha = \alpha \in \ELIM$.
   \item $\ann \alpha = \ann{\alpha'} \in \ELIM$.
-  \item If $A = A' \in \TY$ and $d_z = d_z' \in \NF$ and $d_s = d_s' \in \NF$
+  \item If\/ $A = A' \in \TY$ and $d_z = d_z' \in \NF$ and $d_s = d_s' \in \NF$
         then $\case[\ell] A {d_z}{d_s} = \case[\ell]{A'}{d_z'}{d_s'} \in \ELIM$.
-  \item If $D = D' \in \NF$ and $d = d' \in \NF$ then $\fix[\ell] D d = \fix[\ell]{D'}{d'} \in \ELIM$.
+  \item If\/ $D = D' \in \NF$ and $d = d' \in \NF$ then $\fix[\ell] D d = \fix[\ell]{D'}{d'} \in \ELIM$.
   \end{enumerate}
@@
 % \begin{enumerate}
-% \item If $\alpha \leq \beta$ then $\NAT(\alpha) \subseteq \NAT(\beta)$.
-% \item If $\F(\alpha) \leq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
-% \item IF $\A' \subseteq \A$ and $\F(u) \subseteq \F'(u)$ for all $u \in \A'$, then $\PIAF \subseteq \PI\,\A'\,\F'$.
+% \item If\/ $\alpha \leq \beta$ then $\NAT(\alpha) \subseteq \NAT(\beta)$.
+% \item If\/ $\F(\alpha) \leq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
+% \item IF\/ $\A' \subseteq \A$ and $\F(u) \subseteq \F'(u)$ for all $u \in \A'$, then $\PIAF \subseteq \PI\,\A'\,\F'$.
 % \end{enumerate}
@@
   \item Symmetry: $T_2 = T_1 \in \SET[\ell]$.
-  \item Transitivity: If $T_2 = T_3 \in \SET[\ell]$ then $T_1 = T_3 \in \SET[\ell]$.
+  \item Transitivity: If\/ $T_2 = T_3 \in \SET[\ell]$ then $T_1 = T_3 \in \SET[\ell]$.
   \item Extension: $\EL[\ell](T_1) = \EL[\ell](T_2)$ and ``both'' are PERs.
@@
 \label{lem:indep}
-  If $\DD_1 :: T = T_1 \in \SET[\ell_1]$
+  If\/ $\DD_1 :: T = T_1 \in \SET[\ell_1]$
   and $\DD_2 :: T_2 = T \in \SET[\ell_2]$ then $\EL[\ell_1](T) = \EL[\ell_2](T)$.
@@
 \begin{enumerate}
-\item \label{it:natsub} If $\alpha \leq \beta$ then $\NAT(\alpha) \subseteq \NAT(\beta)$.
-\item \label{it:allsub} If $\F(\alpha) \subseteq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
-\item \label{it:pisub}  If $\A' \subseteq \A$ and $\F(u) \subseteq \F'(u)$ for all $u \in \A'$, then $\PIAF \subseteq \PI\,\A'\,\F'$.
-\item \label{it:usub}  If $\ell \leq \ell'$ then $\SET[\ell] \subseteq \SET[\ell']$.
+\item \label{it:natsub} If\/ $\alpha \leq \beta$ then $\NAT(\alpha) \subseteq \NAT(\beta)$.
+\item \label{it:allsub} If\/ $\F(\alpha) \subseteq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
+\item \label{it:pisub}  If\/ $\A' \subseteq \A$ and $\F(u) \subseteq \F'(u)$ for all $u \in \A'$, then $\PIAF \subseteq \PI\,\A'\,\F'$.
+\item \label{it:usub}  If\/ $\ell \leq \ell'$ then $\SET[\ell] \subseteq \SET[\ell']$.
 \end{enumerate}
@@
 %   \label{lem:usubsump}
-%   If $\ell \leq \ell'$ then $\SET[\ell] \subseteq \SET[\ell']$.
+%   If\/ $\ell \leq \ell'$ then $\SET[\ell] \subseteq \SET[\ell']$.
 % \end{lemma}
@@
 \begin{enumerate}
-\item If $n_1 = n_2 \in \NE$ then $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$.
-\item If $t_1 = t_2 \in T$ then $\down{S_1}{t_1} = \down{S_2}{t_2} \in \NF$.
+\item If\/ $n_1 = n_2 \in \NE$ then $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$.
+\item If\/ $t_1 = t_2 \in T$ then $\down{S_1}{t_1} = \down{S_2}{t_2} \in \NF$.
 \end{enumerate}
@@
   \begin{enumerate}
-  \item If $n = n' \in \NE$ then $\up T n = \up T n' \in T$.
-  \item If $t = t' \in T$ then $\down T t = \down T {t'} \in \NF$.
+  \item If\/ $n = n' \in \NE$ then $\up T n = \up T n' \in T$.
+  \item If\/ $t = t' \in T$ then $\down T t = \down T {t'} \in \NF$.
   \end{enumerate}
@@
   \label{lem:case}
-If $a = a' \in \Nat (\alpha+1)$
+If\/ $a = a' \in \Nat (\alpha+1)$
 and $B = B' \in \Nat (\alpha+1) \to \Set[\ell]$
@@
 Let $g = a \, \fix[\ell] B f$ and $g' = a'\,\fix[\ell]{B'}{f'}$.
-If $a = a' \in \Nat \alpha$
+If\/ $a = a' \in \Nat \alpha$
 and $B = B' \in \Adm\ell$
@@
   \label{lem:resenv}
-  If $\eta = \eta' \eeq \rho \in \Gamma$ then $\rho \in \resurrect\Gamma$.
+  If\/ $\eta = \eta' \eeq \rho \in \Gamma$ then $\rho \in \resurrect\Gamma$.
 \end{lemma}
@@
   \item
-  If $\der \Gamma$ then $\models \Gamma$.
+  If\/ $\der \Gamma$ then $\models \Gamma$.
   \item
-  If $\Gamma \der J$ then % $\models \Gamma$ and
+  If\/ $\Gamma \der J$ then % $\models \Gamma$ and
   $\Gamma \models J$
@@
 \begin{lemma}[Identity environment]
-  If $\der \Gamma$ then
+  If\/ $\der \Gamma$ then
   $\rho_\Gamma \in \Gamma$.
@@
 By definitions, these relations are closed under subsumption and
-weakening, e.g., if $\RD \Gamma t d T$ and $\Gamma \der T \leq T'$ then $\RD \Gamma t d T'$,
-and if $\xi : \Gamma \leq \Gamma'$ then $\RD{\Gamma'}{t\xi}d{T\xi}$.
+weakening, e.g., if\/ $\RD \Gamma t d T$ and $\Gamma \der T \leq T'$ then $\RD \Gamma t d T'$,
+and if\/ $\xi : \Gamma \leq \Gamma'$ then $\RD{\Gamma'}{t\xi}d{T\xi}$.
 
@@
 We may prove theorems ``by induction on $\LRTS \Gamma T A s$'', even if in reality this will be proofs by induction on $A \in s$ and cases on $\LRTS \Gamma T A s$.
-We write \fbox{$\RG \Gamma T A$} if $\LRTS \Gamma T A s$ for some sort $s$.
+We write \fbox{$\RG \Gamma T A$} if\/ $\LRTS \Gamma T A s$ for some sort $s$.
 
@@
 \item
-If $\LRSize \Gamma a \alpha$ and $\LRSize \Gamma {a'} \alpha$ then $a = a'$.
+If\/ $\LRSize \Gamma a \alpha$ and $\LRSize \Gamma {a'} \alpha$ then $a = a'$.
 \item
-If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} A {\ell'}$ then $\Gamma \der T = T'$.
+If\/ $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} A {\ell'}$ then $\Gamma \der T = T'$.
 \end{enumerate}
@@
   \label{cor:fresh}
-  If $\der \cext \Gamma T$ and %$A \in \SET[\ell]$
+  If\/ $\der \cext \Gamma T$ and %$A \in \SET[\ell]$
   $\RG \Gamma T A$
@@
 \item
-   If $\LRSize \Gamma a \alpha$ and $\LRSize \Gamma {b} \beta$ and $\alpha \leq \beta$ then $a \leq b$.
+   If\/ $\LRSize \Gamma a \alpha$ and $\LRSize \Gamma {b} \beta$ and $\alpha \leq \beta$ then $a \leq b$.
 \item
-   If $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \leq T'$.
-   % If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} {A'} {\ell'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \leq T'$.
+   If\/ $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \leq T'$.
+   % If\/ $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} {A'} {\ell'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \leq T'$.
 \end{enumerate}
@@
 \label{lem:lrsump}
-   If $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and $A \leq A' \in \TYPE$ then
+   If\/ $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and $A \leq A' \in \TYPE$ then
    $\LR \Gamma t T f A$ implies $\LR \Gamma t {T'} f {A'}$.
-   % If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} {A'} {\ell'}$ and $A \leq A' \in \TYPE$ then
+   % If\/ $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} {A'} {\ell'}$ and $A \leq A' \in \TYPE$ then
    % $\LR \Gamma t T f A$ implies $\LR \Gamma t {T'} f {A'}$.
@@
   \item Weakening:
-  If $\xi : \Gamma' \leq \Gamma$ then
+  If\/ $\xi : \Gamma' \leq \Gamma$ then
   $\LRSub{\Gamma'}{\sigma\xi}{\tau\xi}{\Delta}\eta\rho$.
@@
   \item \label{it:sizesub} Size substitution:
-  If $\resurrect\Delta \der a : \Size$ then $a\eta \in \SIZE$.
+  If\/ $\resurrect\Delta \der a : \Size$ then $a\eta \in \SIZE$.
   \end{enumerate}
@@
 \item
-If $\Gamma \der t : T$ then $\Gamma \valid t : T$.
+If\/ $\Gamma \der t : T$ then $\Gamma \valid t : T$.
 \item
-If $\Gamma \der \sigma : \Delta$ then
+If\/ $\Gamma \der \sigma : \Delta$ then
 $\Gamma \valid \sigma : \Delta$
@@
   \label{lem:idenv}
-  If $\der \Gamma$ then
+  If\/ $\der \Gamma$ then
   $\LRS \Gamma \sid \Gamma \rhog$.
@@
 \begin{corollary}[Type weak head normalization]
-  If $\Gamma \der T$ then $T \evalsto W$ for some $W$.
+  If\/ $\Gamma \der T$ then $T \evalsto W$ for some $W$.
 \end{corollary}
@@
 \item \label{it:setinj}
-  If $\Gamma \der \Set[\ell] = \Set[\ell'] : s$ then $\ell = \ell'$.
+  If\/ $\Gamma \der \Set[\ell] = \Set[\ell'] : s$ then $\ell = \ell'$.
 \item \label{it:natinj}
-  If $\Gamma \der \Nat a = \Nat b : s$ then $a = b$.
+  If\/ $\Gamma \der \Nat a = \Nat b : s$ then $a = b$.
 \item \label{it:piinj}
-  If $\Gamma \der \piT U T = \piT{U'}{T'} : s$ then
+  If\/ $\Gamma \der \piT U T = \piT{U'}{T'} : s$ then
   $\Gamma \der U = U' : s$ and
@@
 \item \label{it:allinj}
-  If $\Gamma \der \epiT \Size T = \piT{\Size}{T'} : s$ then
+  If\/ $\Gamma \der \epiT \Size T = \piT{\Size}{T'} : s$ then
   $\cext \Gamma \Size \der T = T' : s$.
@@
 \item \label{it:setinvsub}
-  If $\Gamma \der \Set[\ell] \leq \Set[\ell'] : s$ then $\ell \leq \ell'$.
+  If\/ $\Gamma \der \Set[\ell] \leq \Set[\ell'] : s$ then $\ell \leq \ell'$.
 \item \label{it:natinvsub}
-  If $\Gamma \der \Nat a \leq \Nat b : s$ then $a \leq b$.
+  If\/ $\Gamma \der \Nat a \leq \Nat b : s$ then $a \leq b$.
 \item \label{it:piinvsub}
-  If $\Gamma \der \piT U T \leq \piT{U'}{T'} : s$ then
+  If\/ $\Gamma \der \piT U T \leq \piT{U'}{T'} : s$ then
   $\Gamma \der U' \leq U : s$ and
@@
 \item \label{it:allinvsub}
-  If $\Gamma \der \epiT \Size T \leq \piT{\Size}{T'} : s$ then
+  If\/ $\Gamma \der \epiT \Size T \leq \piT{\Size}{T'} : s$ then
   $\cext \Gamma \Size \der T \leq T' : s$.
@@
   \begin{enumerate}
-  \item   If $\Gamma \der \lambda t : \piT U T$ then
+  \item   If\/ $\Gamma \der \lambda t : \piT U T$ then
   $\cext \Gamma U \der t : T$.
-  \item   If $\Gamma \der \lambda t : \epiT \Size T$ then
+  \item   If\/ $\Gamma \der \lambda t : \epiT \Size T$ then
   $\eext \Gamma \Size \der t : T$.
@@
   \begin{enumerate}
-  \item   If $\Gamma \der (\lambda t)\,u : T'$ then
+  \item   If\/ $\Gamma \der (\lambda t)\,u : T'$ then
   $\cext \Gamma U \der t : T$ and $\Gamma \der u : U$ and $\Gamma \der T[u] \leq T'$
   for some $U,T$.
-  \item   If $\Gamma \der (\lambda t)\,a : T'$ then
+  \item   If\/ $\Gamma \der (\lambda t)\,a : T'$ then
   $\cext \Gamma \Size \der t : T$ and $\Gamma \der a : \Size$ and $\Gamma \der T[a] \leq T'$
@@
   \item \label{it:sinvall}
-  If $\Gamma \der (\lambda t)\,\ann a : T'$ then
+  If\/ $\Gamma \der (\lambda t)\,\ann a : T'$ then
   $\erext \Gamma \Size \der t : T$ and $\resurrect \Gamma \der a : \Size$ and $\Gamma \der T[a] \leq T'$
@@
   \item
-  If $\DD :: \Gamma \der t : T$ and $\DD' :: t \evalsto w$ then $\Gamma \der t = w : T$.
+  If\/ $\DD :: \Gamma \der t : T$ and $\DD' :: t \evalsto w$ then $\Gamma \der t = w : T$.
   \item
-  If $\DD :: \Gamma \der w\, e : T$ and $\DD' :: w @ e \evalsto w'$ then $\Gamma \der w\,e = w : T$.
+  If\/ $\DD :: \Gamma \der w\, e : T$ and $\DD' :: w @ e \evalsto w'$ then $\Gamma \der w\,e = w : T$.
   \end{enumerate}
@@
   \label{lem:asubsound}
-  If $\Gamma \der T \asub T'$ then $\Gamma \der T \leq T'$.
+  If\/ $\Gamma \der T \asub T'$ then $\Gamma \der T \leq T'$.
 \end{lemma}
@@
 \label{cor:complasub}
-If $\Gamma \der T \leq T'$ then $\Gamma \der T \asub T'$.
+If\/ $\Gamma \der T \leq T'$ then $\Gamma \der T \asub T'$.
 \end{corollary}
@@
   \label{lem:termasub}
-  If $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ then the query $\Gamma \der T \asub T'$ terminates.
+  If\/ $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ then the query $\Gamma \der T \asub T'$ terminates.
 \end{lemma}
@@
 \begin{theorem}[Decidability of subtyping]
-  If $\Gamma \der T,T'$, then $\Gamma \der T \leq T'$ is decided by the query $\Gamma \der T \asub T'$.
+  If\/ $\Gamma \der T,T'$, then $\Gamma \der T \leq T'$ is decided by the query $\Gamma \der T \asub T'$.
 \end{theorem}
@@
 We will refer to the restricted typing judgement as \fbox{$\Gamma \ders t : T$}, and obviously,
-if $\Gamma \ders t : T$ then $\Gamma \der t : T$.
+if\/ $\Gamma \ders t : T$ then $\Gamma \der t : T$.
 
@@
   \begin{enumerate}
-  \item If $\resurrect\Gamma \der T$ and $\DD :: \Gamma \der t \jchk T$ then $\Gamma \ders t : T$.
-  \item If $\DD :: \Gamma \der t \jinf T$ then $\resurrect\Gamma \der T$ and $\Gamma \ders t : T$.
+  \item If\/ $\resurrect\Gamma \der T$ and $\DD :: \Gamma \der t \jchk T$ then $\Gamma \ders t : T$.
+  \item If\/ $\DD :: \Gamma \der t \jinf T$ then $\resurrect\Gamma \der T$ and $\Gamma \ders t : T$.
   \end{enumerate}
@@
 \begin{enumerate}
-\item If $T' \evalsto \Nat a'$ then $T \evalsto \Nat a$ and $\Gamma \der a \asub a' : \Size$.
-\item If $T' \evalsto \Set[\ell']$ then $T \evalsto \Set[\ell]$ and $\ell \asub \ell'$.
-\item If $T' \evalsto \piT{A'}{B'}$ then $T \evalsto \piT {A} {B}$ and
+\item If\/ $T' \evalsto \Nat a'$ then $T \evalsto \Nat a$ and $\Gamma \der a \asub a' : \Size$.
+\item If\/ $T' \evalsto \Set[\ell']$ then $T \evalsto \Set[\ell]$ and $\ell \asub \ell'$.
+\item If\/ $T' \evalsto \piT{A'}{B'}$ then $T \evalsto \piT {A} {B}$ and
       $\Gamma \der A' \asub A$ and $\cext \Gamma {A'} \der B \asub B'$.
-\item If $T' \evalsto \epiT \Size {B'}$ and $T \evalsto \epiT \Size B$ and $\cext \Gamma \Size \der B \asub B'$.
+\item If\/ $T' \evalsto \epiT \Size {B'}$ and $T \evalsto \epiT \Size B$ and $\cext \Gamma \Size \der B \asub B'$.
 \end{enumerate}
@@
 \begin{enumerate}
-\item If $T \evalsto \Nat a$ then $T' \evalsto \Nat a'$ and $\Gamma \der a \asub a' : \Size$.
-\item If $T \evalsto \Set[\ell]$ then $T' \evalsto \Set[\ell']$ and $\ell \asub \ell'$.
-\item If $T \evalsto \piT A B$ then $T' \evalsto \piT {A'} {B'}$ and
+\item If\/ $T \evalsto \Nat a$ then $T' \evalsto \Nat a'$ and $\Gamma \der a \asub a' : \Size$.
+\item If\/ $T \evalsto \Set[\ell]$ then $T' \evalsto \Set[\ell']$ and $\ell \asub \ell'$.
+\item If\/ $T \evalsto \piT A B$ then $T' \evalsto \piT {A'} {B'}$ and
       $\Gamma \der A' \asub A$ and $\cext \Gamma {A'} \der B \asub B'$.
-\item If $T \evalsto \epiT \Size B$ and $T' \evalsto \epiT \Size B'$ and $\cext \Gamma \Size \der B \asub B'$.
+\item If\/ $T \evalsto \epiT \Size B$ and $T' \evalsto \epiT \Size B'$ and $\cext \Gamma \Size \der B \asub B'$.
 \end{enumerate}
@@
   \begin{enumerate}
-  \item If $\DD :: \Gamma \der t \jchk T$ and $\resurrect\Gamma \der T \leq T'$ then $\Gamma' \der t \jchk T'$.
-  \item If $\DD :: \Gamma \der t \jinf T$ then $\Gamma' \der t \jinf T'$ and $\resurrect{\Gamma'} \der T \leq T'$.
+  \item If\/ $\DD :: \Gamma \der t \jchk T$ and $\resurrect\Gamma \der T \leq T'$ then $\Gamma' \der t \jchk T'$.
+  \item If\/ $\DD :: \Gamma \der t \jinf T$ then $\Gamma' \der t \jinf T'$ and $\resurrect{\Gamma'} \der T \leq T'$.
   \end{enumerate}
@@
 \begin{enumerate}
-\item If $\DD :: \Gamma \ders v : T$ then $\Gamma \der v \jchk T$.
-\item If $\DD :: \Gamma \ders m : T$ then $\Gamma \der m \jinf U$ and $\resurrect\Gamma \der U \leq T$.
+\item If\/ $\DD :: \Gamma \ders v : T$ then $\Gamma \der v \jchk T$.
+\item If\/ $\DD :: \Gamma \ders m : T$ then $\Gamma \der m \jinf U$ and $\resurrect\Gamma \der U \leq T$.
 \end{enumerate}
@@
   \item The query $\Gamma \der t \jinf \mbox{?}$ terminates.
-  \item If $\resurrect\Gamma \der T$ then the query $\Gamma \der t \jchk T$ terminates.
+  \item If\/ $\resurrect\Gamma \der T$ then the query $\Gamma \der t \jchk T$ terminates.
   \end{enumerate}

### 2017-02-27 nbe/icfp17.tex
More typos
Hope I’m right this time.

@@
 A type $T$ will be interpreted as a partial equivalence relation (PER) $\A$ on terms, \ie, a relation which is symmetric and transitive.
-The domain $\dom(\A)$ of the relation can be thought of the set of terms which define the extension of the type; on $\dom(\A) = \{a \mid \exists a'.\ (a,a') \in \A \}$ the relation $\A$ is in fact an equivalence relation.
+The domain $\dom(\A)$ of the relation can be thought of as the set of terms which define the extension of the type; on $\dom(\A) = \{a \mid \exists a'.\ (a,a') \in \A \}$ the relation $\A$ is in fact an equivalence relation.
 We write $a = a' \in \A$ for relatedness in $\A$ and $a \in \A$ if $a \in \dom(\A)$.
@@
 together with the extension $\EL[\ell] T$ (for $T = T' \in \Set[\ell]$)
-which is PER of values of type $A$.
+which is a PER of values of type $A$.
 The rules for \fbox{$T = T' \in \SET[\ell]$} are:
@@
 We can define subtyping of type values \fbox{$T \leq T' \in \TYPE$} by induction on $T \in \SET[\ell]$ and $T' \in \SET[\ell']$.
-Simultaneously, we need to prove correctness, namely that $T \leq T' in \TYPE$ implies $\EL(T) \subseteq \EL(T')$.
+Simultaneously, we need to prove correctness, namely that $T \leq T' \in \TYPE$ implies $\EL(T) \subseteq \EL(T')$.
 The correctness follows from Lemma~\ref{lem:subsump} and we do not spell it out here.
@@
 Now consider $S := \funT x {\Nat \infty} {F\,x}$.
-We have $T \shape S$, but not $S$ is not well-defined; $S \not\in \SET[0]$.
+We have $T \shape S$, but $S$ is not well-defined; $S \not\in \SET[0]$.
 % $\funT x {\Nat 0} {F\,x} \shape , but the latter one is not a well-defined type.

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] First run over section 3 done.

@@
 %\documentclass[acmlarge]{acmart}\settopmatter{}
+
+% For having two versions of a paper in one file.
+% Stuff that does not fit into the short version can be encosed in \LONGVERSION{...}
+\ifdefined\LONGVERSION
+  \relax
+\else
+% short version:
+\newcommand{\LONGVERSION}[1]{}
+\newcommand{\SHORTVERSION}[1]{#1}
+% % long version:
+% \newcommand{\LONGVERSION}[1]{#1}
+% \newcommand{\SHORTVERSION}[1]{}
+% \newcommand{\SHORTVERSION}[1]{BEGIN~SHORT\ #1 \ END~SHORT}
+\fi
+\newcommand{\LONGSHORT}[2]{\LONGVERSION{#1}\SHORTVERSION{#2}}
+\newcommand{\SHORTLONG}[2]{\SHORTVERSION{#1}\LONGVERSION{#2}}
+\newcommand{\EXTENDED}[1]{}
 
@@
 
+\input{figty}
+\input{figeq}
+
 Fig.~\ref{fig:judge} lists the inductive and derived judgements of our type theory and
 figures~\ref{fig:ty} and \ref{fig:eq} the inference rules.
-
-\input{figty}
-\input{figeq}
-
-%Our type theory is formulated via the following judgements, which are mutually inductively defined.
-
-%In addition, we will refer to the following derived judgements.
-
-% Inference rules.
-
-% \fbox{$\der \Gamma$} (implies $\der \resurrect \Gamma$).
-% \[
-%   \ru{}{\der \cempty}
-% \qquad
-%   \ru{\der \Gamma \qquad
-%       \resurrect \Gamma \der T
-%     }{\der \cext \Gamma T}
-% \qquad
-%   \ru{\der \Gamma
-%     }{\der \cext \Gamma \Size}
-% \qquad
-%   \ru{\der \Gamma
-%     }{\der \erext \Gamma \Size}
-% \]
-
-% \fbox{$\Gamma(i) = \eann T$} (implies $\resurrect\Gamma \der T$ if\/ $\der \Gamma$).
-% \[
-%   \ru{}{(\eext \Gamma T)(0) = \eann T \slift}
-% \qquad
-%   \ru{\Gamma(i) = \eann T
-%     }{(\cext \Gamma \_)(i+1) = \eann T \slift}
-% \]
-
-%Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
-% \fbox{$\Gamma \mid U \der e : T$}
-% \begin{gather*}
-%   \ru{\cext \Gamma U \der T \qquad \Gamma \der u : U
-%     }{\Gamma \mid \piT U T \der u : T[u]}
-% \qquad
-%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
-%     }{\Gamma \mid \piT \Size T \der a : T[a]}
-% \qquad
-%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
-%     }{\Gamma \mid \forallT T \der \ann a : T[a]}
-% \\
-%   \ru{
-%     }{\Gamma \mid \Nat a \der \fix T t : T \ann a n}
-% \end{gather*}
-
-% \fbox{$\Gamma \der t = t' : T$} (implies $\resurrect \Gamma \der T$ and $\Gamma \der t,t' : T$ [and $\resurrect\Gamma \der t = t' : T$].)
-
-% Liftings: $\fbox{$\xi : \Gamma \leq \Delta$}
-%   \defas \Gamma \der \xi : \Delta \mand \xi = \slift^k_m \mforsome k,m$
-% %  \defas \xi = \slift^k_m \mwith \Gamma \der \xi : \Delta$
-% (implies $\der \Gamma$ and $\der \Delta$).
-
-% (implies $\der \Gamma$ and $\der \Delta$ and $\Gamma \der \xi : \Delta$).
-% \begin{gather*}
-%   \ru{\der \Gamma
-%     }{\sempty : \Gamma \leq \cempty}
-% \qquad
-%   \ru{\xi : \Gamma \leq \Delta \qquad
-%       \resurrect\Gamma \der T
-%     }{\xi \slift : \cext \Gamma T \leq \Delta}
-% \qquad
-%   \ru{\xi : \Gamma \leq \Delta \qquad
-%       \resurrect \Delta \der U \qquad
-%       \resurrect\Gamma \der T \leq U \xi
-%     }{(\xi \slift, \ind 0) : \cext \Gamma T \leq \cext \Delta U}
-% \end{gather*}
-
-% Substitution typing. % and equality.
-% \fbox{$\Gamma \der \sigma : \Delta$}
-% (implies $\der \Gamma$ and $\der \Delta$
-%  and $\resurrect\Gamma \der \sigma : \Delta$
-%  and $\resurrect\Gamma \der \sigma : \resurrect \Delta$).
-% % and
-% % \fbox{$\Gamma \der \sigma = \sigma' : \Delta$}.
-
-% \begin{gather*}
-%   \ru{\der \Gamma \qquad
-%     }{\Gamma \der \sempty : \cempty}
-% \qquad
-%   \ru{\Gamma \der \sigma : \Delta \qquad
-%       \resurrect \Delta \der T \qquad
-%       \Gamma \der t : T \sigma
-%     }{\Gamma \der \sext \sigma t : \cext \Delta T}
-% \qquad
-%   \ru{\Gamma \der \sigma : \Delta \qquad
-%       \Gamma \der a : \Size
-%     }{\Gamma \der \sext \sigma a : \eext \Delta \Size}
-% \end{gather*}
-
-% \begin{gather*}
-%   \ru{\der \Gamma \qquad
-%     }{\Gamma \der \sempty = \sempty : \cempty}
-% \qquad
-%   \ru{\Gamma \der \sigma = \sigma': \Delta \qquad
-%       \resurrect \Delta \der T \qquad
-%       \Gamma \der t = t' : T \sigma
-%     }{\Gamma \der \sext \sigma t = \sext {\sigma'} t': \cext \Delta T}
-% \end{gather*}
-
-% Substitution typing and equality.
-% \fbox{$\suty \Gamma \sigma \tau \Delta$} and
-% \begin{gather*}
-%   \ru{\der \Gamma \qquad
-%     }{\suty \Gamma \sempty \sempty \cempty}
-% \qquad
-%   \ru{\suty \Gamma \sigma \tau \Delta \qquad
-%       \resurrect \Delta \der T \qquad
-%       \Gamma \der u = t : T \tau
-%     }{\suty \Gamma {\sext \sigma u} {\sext \tau t} \cext \Delta T}
-% \\[2ex]
-%   \ru{\suty \Gamma \sigma \tau \Delta \qquad
-%       \Gamma \der a = b : \Size
-%     }{\suty \Gamma {\sext \sigma a} {\sext \tau b} \cext \Delta \Size}
-% \qquad
-%   \ru{\suty \Gamma \sigma \tau \Delta \qquad
-%       \Gamma \der a, b : \Size
-%     }{\suty \Gamma {\sext \sigma a} {\sext \tau b} \erext \Delta \Size}
-% \end{gather*}
+In the typing judgement $\Gamma \der t : T$, the term $t$ is in scope of
+$\Gamma$, \ie, may not mention irrelevant variables in relevant
+positions.  However, the type $T$ is in scope of the resurrected
+context $\resurrect\Gamma$, hence, can mention all variables declared
+in $\Gamma$.  The other judgements are organized similarly.  To
+understand this distinction, consider judgement
+$i \div \Size \der \Nat i$.  This would mean that $i$ is irrelevant in
+$\Nat i$ and thus, $\Gamma \der \Nat a = \Nat {a'}$ for all sizes
+$\resurrect\Gamma \der a,a' : \Size$.  But this is exactly wrong!  However,
+judgement $i \div \Size \der \zero i : \Nat (i + 1)$ is fine, it
+implies $\Gamma \der \zero a = \zero {a'} : \Nat (b+1)$ for all
+$\resurrect\Gamma \der a,a',b : \Size$.
+
+Our substitution theorem
+needs to reflect the distinct scope of things left of the colon vs.\
+things right of the colon.  In the last example we have applied the
+substitution triple $\Gamma \der [a] = [a'] \eeq [b] : (i \div \Size)$
+to judgement $i \div \Size \der \zero i : \Nat (i + 1)$.
+The first two substitutions apply to the term side while the third
+substitution applies to the type side.  The fact that we replace an
+irrelevant variable $i$ allows $a,a',b$ to refer to irrelevant
+variables from $\Gamma$, thus, they are in scope of $\resurrect\Gamma$.
 
@@
 
-In the following, let $J$ %, $L$, and $R$
+In the following, we list syntactic properties of our judgements.  To
+this end, let $J$ %, $L$, and $R$
 match a part of a judgement.
@@
 
+All types in a context are considered in the resurrected context,
+which justifies the first statement of the following lemma.
+A resurrected context is more permissive, as it brings more variable
+into scope.  As such, it is comparable to an extended context or a
+context where types have been replaced by subtypes.  This intuition
+accounts for the remaining statments but (\ref{it:ressubst}).  The
+latter is a defining property of substitutions: only replacement for
+irrelevant sizes may refer to irrelevant size variables.
 \begin{lemma}[Resurrection]
@@
 \begin{enumerate}
-\item $\der \Gamma$ iff $\der \resurrect\Gamma$;  then $\resurrect \Gamma \der \sid : \Gamma$.
+\item \label{it:reswf} $\der \Gamma$ iff $\der \resurrect\Gamma$;  then $\resurrect \Gamma \der \sid : \Gamma$.
   % Consequently, if\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
@@
   then $\Gamma \der J : \Delta$.
-\item If\/ $\Gamma \der \sigma : \Delta$ % then $\Gamma \der \sigma : \resurrect\Delta$.
+\item \label{it:ressubst} If\/ $\Gamma \der \sigma : \Delta$ % then $\Gamma \der \sigma : \resurrect\Delta$.
   then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
@@
 
+The relation $\sueq \Gamma \sigma {\sigma'} \tau \Delta$ is a partial equivalence
+relation (PER) on term-side substitutions $\sigma,\sigma'$. Note that
+we cannot resurrect this judgement to
+$\sueq{\resurrect\Gamma}{\sigma}{\sigma'}\tau \Delta$ in general!
+The type-side substitution $\tau$ can be converted to one that is
+equal in and for resurrected contexts.
 \begin{lemma}[Substitution equality]
@@
 \end{lemma}
+
+\LONGVERSION{
 \begin{corollary}[Partial resurrection for substitution equality]
@@
 \end{corollary}
+}
 

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Fixed typo in abstract

@@
 We introduce a size quantifier $\forall$ which lets us ignore sizes in terms for equality checking, alongside with a second quantifier $\Pi$ for abstracting over sizes that do affect the semantics of types and terms.
-Judgmental equality is decided by an adaption of normalization-by-evaluation for our new type theory, which features \emph{type shape}-directed reflection and reification.
+Judgmental equality is decided by an adaptation of normalization-by-evaluation for our new type theory, which features \emph{type shape}-directed reflection and reification.
 It follows that subtyping and type checking of normal forms are decidable as well, the latter by a bidirectional algorithm.

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Condensing the most wasteful displays.
[ ICFP 17 ] Delegating some proofs to the long version.

@@
 \begin{gather*}
-  \ru{T \evalsto \Nat \alpha \qquad
-      S \evalsto \Nat \beta
-    }{T \shape S}
-\qquad
   \ru{T \evalsto N \qquad
@@
 \qquad
-   \ru{T \evalsto \Set[\ell] \qquad
-       S \evalsto \Set[\ell]
-    }{T \shape S}
-\\[2ex]
   \ru{T \evalsto \piT A B \qquad
@@
 \\[2ex]
+   \ru{T \evalsto \Set[\ell] \qquad
+       S \evalsto \Set[\ell]
+    }{T \shape S}
+\qquad
   \ru{T \evalsto \piT \Size B \qquad
@@
 \\[2ex]
+  \ru{T \evalsto \Nat \alpha \qquad
+      S \evalsto \Nat \beta
+    }{T \shape S}
+\qquad
   \ru{T \evalsto \forallT B \qquad
@@
 
-\nextcase $T \evalsto \piT A B$
+\nextcase $T \evalsto \piT A B$ with $A \in \SET[\ell]$ and
+      $B[u] = B[u'] \in \SET[\ell] \mforall u = u' \in A$
 \begin{gather*}
-  \ru{A \in \SET[\ell] \qquad
-      B[u] = B[u'] \in \SET[\ell] \mforall u = u' \in A
-    }{T \in \SET[\ell]}
-\\[2ex]
+%   \ru{A \in \SET[\ell] \qquad
+%       B[u] = B[u'] \in \SET[\ell] \mforall u = u' \in A
+%     }{T \in \SET[\ell]}
+% \\[2ex]
   \ru{S_1 \evalsto \piT{A_1}{B_1} \qquad
@@
 
-\nextcase $T \evalsto \forallT B$
+\nextcase $T \evalsto \forallT B$ with $B[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE$
 \begin{gather*}
-  \ru{B[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
-    }{T \in \SET[\ell]}
-\\[2ex]
+%   \ru{B[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
+%     }{T \in \SET[\ell]}
+% \\[2ex]
   \ru{S_1 \evalsto \forallT{B_1} \qquad
@@
     }{T \shape S_1}
-\\[2ex]
+%\\[2ex]
+\qquad
   \ru{S_2 \evalsto \forallT{B_2} \qquad
@@
 \end{theorem}
+\SHORTVERSION{
+\begin{proof}
+Simultaneously, by induction on the derivation.
+\end{proof}
+}
+\LONGVERSION{  % BEGIN LONGVERSION
 \begin{proof*}
@@
 \end{proof*}
+} % END LONGVERSION
 
@@
 $\LRTS \Gamma {T'} {A'} s$ iff
-\[
-\begin{array}{l}
+\(
+\begin{array}[t]{l}
 \Gamma \der T' \evalsto \piT U T : s \mforsome U, T \mand \LRTS \Gamma U A s \mand \\
@@
 \end{array}
-\]
+\) \\
 $\LR \Gamma t {T'} f {A'}$ iff
-\[
-\begin{array}{l}
+\(
+\begin{array}[t]{l}
 \resurrect\Gamma \der T' \evalsto \piT U T : s \mforsome U, T \mand \LRTS {\resurrect\Gamma} U A s \mand \\
@@
 \end{array}
-\]
+\)
 
@@
 $\LRTS \Gamma {T'} {A'} s$ iff
-\[
-\begin{array}{l}
+\(
+\begin{array}[t]{l}
 \Gamma \der T' \evalsto \piT \Size T : s \mforsome T\\
@@
 \end{array}
-\]
+\) \\
 $\LR \Gamma t {T'} f {A'}$ iff
-\[
-\begin{array}{l}
+\(
+\begin{array}[t]{l}
 \resurrect\Gamma \der T' \evalsto \piT \Size T : s \mforsome T\\
@@
 $\LRTS \Gamma {T'} {A'} s$ iff
-\[
-\begin{array}{l}
+\(
+\begin{array}[t]{l}
 \Gamma \der T' \evalsto \forallT T : s \mforsome T\\
@@
 \end{array}
-\]
+\) \\
 $\LR \Gamma t {T'} f {A'}$ iff
-\[
-\begin{array}{l}
+\(
+\begin{array}[t]{l}
 \resurrect\Gamma \der T' \evalsto \forallT T : s \mforsome T\\
@@
 \end{array}
-\]
+\)
 \end{caselist}
@@
     }{\LRSub \Gamma {\sempty} {\sempty} {\cempty} {\sempty} {\sempty}}
-\\[2ex]
-  \ru{\LRSub \Gamma \sigma \tau \Delta \eta \rho \qquad
-      \resurrect \Delta \der T \qquad
-      \Gamma \der u = t : T\tau \qquad
-      \LR \Gamma t {T \tau} f {T \rho} \qquad
-      f = g \in T\rho
-    }{\LRSub \Gamma {(\sigma,u)} {(\tau,t)} {\cext \Delta T} {(\eta,f)} {(\rho,q)}}
-\\[2ex]
+\qquad
   \ru{\LRSub \Gamma \sigma \tau \Delta \eta \rho \qquad
@@
     }{\LRSub \Gamma {(\sigma,a)} {(\tau,b)} {\erext \Delta \Size} {(\eta,\alpha)} {(\rho,\beta)}}
+\\[2ex]
+  \ru{\LRSub \Gamma \sigma \tau \Delta \eta \rho \qquad
+      \resurrect \Delta \der T \qquad
+      \Gamma \der u = t : T\tau \qquad
+      \LR \Gamma t {T \tau} f {T \rho} \qquad
+      f = g \in T\rho
+    }{\LRSub \Gamma {(\sigma,u)} {(\tau,t)} {\cext \Delta T} {(\eta,f)} {(\rho,q)}}
 \end{gather*}
@@
 \end{lemma}
+\LONGVERSION{
 \begin{proof}
@@
 \end{proof}
-
+} % END LONGVERSION
 % \begin{lemma}[Typing of constructed data]
@@
 Each by induction on the derivation.
-
+\SHORTLONG{\qed}{ % LONGVERSION
 \begin{caselist}
@@
 \end{caselist}
+} % END LONGVERSION
 \end{proof*}
@@
   By induction on $\der \Gamma$.
+\SHORTLONG{\qed}{ % LONGVERSION
 \begin{caselist}
@@
 \end{corollary}
+\LONGVERSION{
 \begin{proof}
@@
 \end{proof}
-
+} % END LONGVERSION
 For normalization of types, we introduce \fbox{$\Nbe \Gamma T$}
@@
 \end{corollary}
+\LONGVERSION{
 \begin{proof}
   By the fundamental theorem of typing, $\RG \Gamma T {T\rhog}$,
-  which implies $T \evalsto W$ by definition of the logical relation..
+  which implies $T \evalsto W$ by definition of the logical relation.
 \end{proof}
-
+} % END LONGVERSION
+\LONGVERSION{ % BEGIN LONGVERSION
 \begin{corollary}[Type constructor injectivity]
@@
 \end{proof}
-
+} % END LONGVERSION
 \begin{lemma}[Type constructor discrimination]
@@
 \begin{proof}
-Similar to the proof of Cor.~\ref{cor:piinj}, using Lemma~\ref{lem:lrsub}.
+\LONGVERSION{Similar to the proof of Cor.~\ref{cor:piinj}, using
+  Lemma~\ref{lem:lrsub}.} % END LONGVERSION
 We recapitulate the proof for (\ref{it:piinvsub}).
@@
 \qquad
-  \ru{T \evalsto \Nat a \qquad
-      T' \evalsto \Nat a'
-    }{\Gamma \der T \asub T'
-    }{a \leq a'}
-\\[2ex]
   \ru{T_1' \evalsto \piT{U_1}{T_1} \qquad
@@
 \\[2ex]
+  \ru{T \evalsto \Nat a \qquad
+      T' \evalsto \Nat a'
+    }{\Gamma \der T \asub T'
+    }{a \leq a'}
+\qquad
   \ru{T_1' \evalsto \epiT\Size{T_1} \qquad
@@
     }{\Gamma \der \epiT \Size T \jchk T'}
-\\[2ex]
-  \ru{T' \evalsto \piT U T \qquad
-      \cext \Gamma U \der t \jchk T
-    }{\Gamma \der \lambda t \jchk T'}
-\qquad
-  \ru{T' \evalsto \epiT \Size T \qquad
-      \eext \Gamma \Size \der t \jchk T
-    }{\Gamma \der \lambda t \jchk T'}
+%\\[2ex]
+%   \ru{T' \evalsto \piT U T \qquad
+%       \cext \Gamma U \der t \jchk T
+%     }{\Gamma \der \lambda t \jchk T'}
+% \qquad
+%   \ru{T' \evalsto \epiT \Size T \qquad
+%       \eext \Gamma \Size \der t \jchk T
+%     }{\Gamma \der \lambda t \jchk T'}
 \\[2ex]
@@
 \\[2ex]
+  \ru{T' \evalsto \epiT U T \qquad
+      \eext \Gamma U \der t \jchk T
+    }{\Gamma \der \lambda t \jchk T'}
+\qquad
   \ru{\Gamma \der t \jinf T \qquad \Gamma \der T \asub T'

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Moved Sec. on type checking to long version.

@@
 
-In this section, we show that type checking for normal forms is decidable, and succeeds for those which can be type via the restricted rule for size polymorphism elimination:
+\LONGVERSION{
+\input{sec-tycheck}
+} % END LONG VERSION
+\SHORTVERSION{
+It is now straightforward to define a bidirectional type-checking
+algorithm which decides typing for normal forms, and succeeds for
+those that rely only on the following restricted rule for irrelevant
+size application:
 \[
@@
 \]
-We will refer to the restricted typing judgement as \fbox{$\Gamma \ders t : T$}, and obviously,
-if\/ $\Gamma \ders t : T$ then $\Gamma \der t : T$.
-
-Bidirectional typing of normal forms.
-
-Checking \fbox{$\Gamma \der t \jchk T$}.
-\begin{gather*}
-  \ru{T' \evalsto s \qquad
-      \Gamma \der a : \Size
-    }{\Gamma \der \Nat a \jchk T'}
-\qquad
-  \ru{T' \evalsto \Set[\ell'] \qquad \ell < \ell'
-    }{\Gamma \der \Set[\ell] \jchk T'}
-\\[2ex]
-  \ru{T' \evalsto s \qquad
-      \Gamma \der U \jchk s \qquad
-      \cext \Gamma U \der T \jchk s
-    }{\Gamma \der \piT U T \jchk T'}
-\qquad
-  \ru{T' \evalsto s \qquad
-      \cext \Gamma \Size \der T \jchk s
-    }{\Gamma \der \epiT \Size T \jchk T'}
-%\\[2ex]
-%   \ru{T' \evalsto \piT U T \qquad
-%       \cext \Gamma U \der t \jchk T
-%     }{\Gamma \der \lambda t \jchk T'}
-% \qquad
-%   \ru{T' \evalsto \epiT \Size T \qquad
-%       \eext \Gamma \Size \der t \jchk T
-%     }{\Gamma \der \lambda t \jchk T'}
-\\[2ex]
-  \ru{T' \evalsto \Nat b \qquad
-      \resurrect \Gamma \der a+1 \leq b : \Size
-    }{\Gamma \der \zero a \jchk T'}
-\qquad
-  \ru{T' \evalsto \Nat b \qquad
-      \resurrect \Gamma \der a+1 \leq b : \Size \qquad
-      \Gamma \der t \jchk \Nat a
-    }{\Gamma \der \suc a t \jchk T'}
-\\[2ex]
-  \ru{T' \evalsto \epiT U T \qquad
-      \eext \Gamma U \der t \jchk T
-    }{\Gamma \der \lambda t \jchk T'}
-\qquad
-  \ru{\Gamma \der t \jinf T \qquad \Gamma \der T \asub T'
-    }{\Gamma \der t \jchk T'}
-\end{gather*}
-
-Inference \fbox{$\Gamma \der t \jinf T$}.
-\begin{gather*}
-  \ru{\Gamma(i) = \erann[\noterased] T
-    }{\Gamma \der \ind i \jinf T}
-\qquad
-  \ru{\Gamma \der t \jinf T' \qquad
-      T' \evalsto \piT U T \qquad
-      \Gamma \der u \jchk U
-    }{\Gamma \der t\, u \jinf T[u]}
-\\[2ex]
-  \ru{\Gamma \der t \jinf T' \qquad
-      T' \evalsto \piT \Size T \qquad
-      \Gamma \der a : \Size
-    }{\Gamma \der t\,a \jinf T[a]}
-\qquad
-  \ru{\Gamma \der t \jinf T' \qquad
-      T' \evalsto \erpiT \Size T \qquad
-      \resurrect\Gamma \der a : \Size
-    }{\Gamma \der t\,\ann a \jinf T[a]}
-\\[2ex]
-  \ru{\Gamma \der u \jinf \Nat (a+1) \qquad
-      \Gamma \der T \jchk \Nat (a+1) \to \Set[\ell] \qquad
-      \Gamma \der t_z \jchk T\,(\zero a) \qquad
-      \Gamma \der t_s \jchk \funT x {\Nat a} {T\,(\suc a x)}
-    }{\Gamma \der u\, \case[\ell] T {t_z} {t_s} \jinf T\,u}
-\\[2ex]
-  \ru{\Gamma \der u \jinf \Nat a \qquad
-      \Gamma \der T \jchk \FixK\,\ell \qquad
-      \Gamma \der t \jchk \FixT\,T
-    }{\Gamma \der u\, \fix[\ell] T t \jinf T\,a\,u}
-\end{gather*}
-
-We could go beyond normal forms, by adding inference rules for the $\Nat$-constructors:
-\begin{gather*}
-  \ru{\resurrect \Gamma \der a : \Size
-    }{\Gamma \der \zero a \jinf \Nat(a+1)}
-\qquad
-  \ru{\resurrect \Gamma \der a : \Size \qquad
-      \Gamma \der t \jchk \Nat a
-    }{\Gamma \der \suc a t \jinf \Nat(a+1)}
-\end{gather*}
-
-\begin{theorem}[Soundness of type checking]
-  Let $\der \Gamma$.
-  \begin{enumerate}
-  \item If\/ $\resurrect\Gamma \der T$ and $\DD :: \Gamma \der t \jchk T$ then $\Gamma \ders t : T$.
-  \item If\/ $\DD :: \Gamma \der t \jinf T$ then $\resurrect\Gamma \der T$ and $\Gamma \ders t : T$.
-  \end{enumerate}
-\end{theorem}
-\begin{proof}
-  Simultaneously by induction on $\DD$, using subject reduction (Thm.~\ref{thm:sr})
-  and soundness of algorithmic subtyping (Lemma~\ref{lem:asubsound}).
-\end{proof}
-
-\begin{lemma}[Weak head reduction of subtypes]
-\label{lem:whdsub}
-Let $\DD :: \Gamma \der T \asub T'$.
-\begin{enumerate}
-\item If\/ $T' \evalsto \Nat a'$ then $T \evalsto \Nat a$ and $\Gamma \der a \asub a' : \Size$.
-\item If\/ $T' \evalsto \Set[\ell']$ then $T \evalsto \Set[\ell]$ and $\ell \asub \ell'$.
-\item If\/ $T' \evalsto \piT{A'}{B'}$ then $T \evalsto \piT {A} {B}$ and
-      $\Gamma \der A' \asub A$ and $\cext \Gamma {A'} \der B \asub B'$.
-\item If\/ $T' \evalsto \epiT \Size {B'}$ and $T \evalsto \epiT \Size B$ and $\cext \Gamma \Size \der B \asub B'$.
-\end{enumerate}
-\end{lemma}
-\begin{proof}
-  By cases on $\DD$, since weak head evaluation is deterministic.
-\end{proof}
-\begin{lemma}[Weak head reduction of supertypes]
-\label{lem:whdsuper}
-Let $\DD :: \Gamma \der T \asub T'$.
-\begin{enumerate}
-\item If\/ $T \evalsto \Nat a$ then $T' \evalsto \Nat a'$ and $\Gamma \der a \asub a' : \Size$.
-\item If\/ $T \evalsto \Set[\ell]$ then $T' \evalsto \Set[\ell']$ and $\ell \asub \ell'$.
-\item If\/ $T \evalsto \piT A B$ then $T' \evalsto \piT {A'} {B'}$ and
-      $\Gamma \der A' \asub A$ and $\cext \Gamma {A'} \der B \asub B'$.
-\item If\/ $T \evalsto \epiT \Size B$ and $T' \evalsto \epiT \Size B'$ and $\cext \Gamma \Size \der B \asub B'$.
-\end{enumerate}
-\end{lemma}
-\begin{proof}
-  By cases on $\DD$, since weak head evaluation is deterministic.
-\end{proof}
-\begin{lemma}[Subsumption for type checking]
-  \label{lem:subtc}
-  Let $\sid : \Gamma' \leq \Gamma$.
-  \begin{enumerate}
-  \item If\/ $\DD :: \Gamma \der t \jchk T$ and $\resurrect\Gamma \der T \leq T'$ then $\Gamma' \der t \jchk T'$.
-  \item If\/ $\DD :: \Gamma \der t \jinf T$ then $\Gamma' \der t \jinf T'$ and $\resurrect{\Gamma'} \der T \leq T'$.
-  \end{enumerate}
-\end{lemma}
-\begin{proof}
-  Simultaneously by induction on $\DD$, using % weak head normalization of types,
-  lemmata \ref{lem:whdsub} and \ref{lem:whdsuper},
-  and soundness and completeness of algorithmic subtyping.
-\end{proof}
-\begin{theorem}[Completeness of type checking for normal terms]
-\bla
-\begin{enumerate}
-\item If\/ $\DD :: \Gamma \ders v : T$ then $\Gamma \der v \jchk T$.
-\item If\/ $\DD :: \Gamma \ders m : T$ then $\Gamma \der m \jinf U$ and $\resurrect\Gamma \der U \leq T$.
-\end{enumerate}
-\end{theorem}
-\begin{proof}
-Simultaneously by induction on $\DD$, using (strong) inversion and Lemma~\ref{lem:subtc}.
-\end{proof}
-
-\begin{lemma}[Termination of type checking]
-  Let $\der \Gamma$.
-  \begin{enumerate}
-  \item The query $\Gamma \der t \jinf \mbox{?}$ terminates.
-  \item If\/ $\resurrect\Gamma \der T$ then the query $\Gamma \der t \jchk T$ terminates.
-  \end{enumerate}
-\end{lemma}
-\begin{proof}
-  By induction on $t$, using type weak head normalization and soundness of type checking, to maintain well-formedness of types.
-  And, of course, decidability of subtyping.
-\end{proof}
-
-\begin{theorem}[Decidability of type checking for normal terms]
-  Let $\der \Gamma$ and $\resurrect \Gamma \der T$.
-  Then $\Gamma \ders v : T$ is decided by $\Gamma \der v \jchk T$.
-\end{theorem}
+For lack of space, we delegate definition and soundness proof to the
+long version of this article.
+}
 

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Moved section algorithmic subtyping to long version.

@@
 
-Algorithmic subtyping \fbox{$\Gamma \der T \asub T'$}.
-\begin{gather*}
-  \ru{T \evalsto n \qquad
-      T' \evalsto n' \qquad
-      \Nbe \Gamma n \equiv \Nbe \Gamma n'
-    }{\Gamma \der T \asub T'}
-\\[2ex]
-  \rux{T \evalsto \Set[\ell] \qquad
-      T' \evalsto \Set[\ell']
-    }{\Gamma \der T \asub T'
-    }{\ell \leq \ell'}
-\qquad
-  \ru{T_1' \evalsto \piT{U_1}{T_1} \qquad
-      T_2' \evalsto \piT{U_2}{T_2} \qquad
-      \Gamma \der U_2 \asub U_1 \qquad
-      \cext \Gamma {U_2} \der T_1 \asub T_2
-    }{\Gamma \der T_1' \asub T_2'}
-\\[2ex]
-  \ru{T \evalsto \Nat a \qquad
-      T' \evalsto \Nat a'
-    }{\Gamma \der T \asub T'
-    }{a \leq a'}
-\qquad
-  \ru{T_1' \evalsto \epiT\Size{T_1} \qquad
-      T_2' \evalsto \epiT\Size{T_2} \qquad
-      \cext \Gamma {\Size} \der T_1 \asub T_2
-    }{\Gamma \der T_1' \asub T_2'}
-\end{gather*}
-
-\begin{lemma}[Soundness of algorithmic subtyping]
-  \label{lem:asubsound}
-  If\/ $\Gamma \der T \asub T'$ then $\Gamma \der T \leq T'$.
-\end{lemma}
-\begin{proof}
-  By induction on $\Gamma \der T \asub T'$, using soundness of NbE and subject reduction (Thm.~\ref{thm:sr}).
-\end{proof}
-
-\begin{lemma}[Semantic subtyping implies algorithmic subtyping]
-\label{lem:lrasub}
-\bla
-   If\/ $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \asub T'$.
-\end{lemma}
-\begin{proof*}
-By induction on % $A_1' \in s_1$ and $A_2' \in s_2$ and cases on the inductive relations
-$\LRTS \Gamma {T_1'} {A_1'} {s_1}$ and $\LRTS \Gamma {T_2'} {A_2'} {s_2}$
-and cases on $A_1' \leq A_2' \in \TYPE$
-we prove $\Gamma \der T_1' \asub T_2'$.
-
-\begin{caselist}
-
-\nextcase $A_i' \evalsto N_i$ and $T_i' \evalsto n_i$ and $N_1 = N_2 \in \NE$ and $\RTY \Gamma {n_i} {N_i} {s_i}$.
-By Cor.~\ref{cor:oneone} $\Gamma \der n_1 = n_2$, thus, by completeness of NbE, $\Nbe\Gamma n_1 \equiv \Nbe\Gamma n_2$, which entails $\Gamma \der T_1' \asub T_2'$.
-
-\nextcase $A_i' \evalsto \piT {A_i} {B_i}$ and  $T_i' \evalsto \piT{U_i}{T_i}$.
-From $\LRTS \Gamma {U_i} {A_i} {s_i}$ and $A_2 \leq A_1 \in \TYPE$ we get $\Gamma \der U_2 \asub U_1$
-by induction hypothesis on $A_i \in s_i$.
-Let $a := {\up{A_2} \var\Gamma}$.  With $\LR {\cext \Gamma {U_2}} {\ind 0} {(U_2\slift)} a {A_2}$ we get
-$\LRTS {\cext \Gamma {U_2}} {T_i} {B_i[a]} s$ and $B_1[a] \leq B_2[a] \in \TYPE$.
-Thus by induction hypothesis on $B_i[a] \in s_i$ we obtain $\cext \Gamma {U_2} \der T_1 \leq T_2$, together $\Gamma \der T_1' \asub T_2'$.
-\qed
-\end{caselist}
-\end{proof*}
-
-\begin{corollary}[Completeness of algorithmic subtyping]
-\label{cor:complasub}
-If\/ $\Gamma \der T \leq T'$ then $\Gamma \der T \asub T'$.
-\end{corollary}
-\begin{proof}
-  By the fundamental theorems $\RG \Gamma T {T\rhog}$ and $\RG \Gamma {T'} {T'\rhog}$ and $T\rhog \leq T'\rhog \in \TYPE$.
-  By Lemma~\ref{lem:lrasub}, $\Gamma \der T \asub T'$.
-\end{proof}
-
-Since the algorithmic subtyping relation is equivalent to the declarative one, we can freely swap one relation for the other.
-
-\begin{lemma}[Termination of algorithmic subtyping]
-  \label{lem:termasub}
-  If\/ $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ then the query $\Gamma \der T \asub T'$ terminates.
-\end{lemma}
-\begin{proof}
-  By induction on $A \in s$ and $A' \in s'$ and cases on $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$.
-\end{proof}
-
-\begin{theorem}[Decidability of subtyping]
-  If\/ $\Gamma \der T,T'$, then $\Gamma \der T \leq T'$ is decided by the query $\Gamma \der T \asub T'$.
-\end{theorem}
-\begin{proof}
-  By the fundamental theorem of typing, $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$,
-  thus, the query  $\Gamma \der T \asub T'$ terminates by Lemma~\ref{lem:termasub}.
-  If successfully, then $\Gamma \der T \leq T'$ by soundness of algorithmic equality.
-  If unsuccessfully, then $\Gamma \der T \leq T'$ is impossible by completeness of algorithmic equality.
-\end{proof}
+\LONGVERSION{
+\input{sec-tycheck}
+} % END LONG VERSION
+\SHORTVERSION{
+Subtyping is decidable, for instance by full normalization
+and comparison of normal forms
+\cite{fridlenderPagano:tlca13}.
+Alternatively, we can define the usual incremental subtyping algorithm
+based on weak head evaluation, which checks equality of neutral types
+by NbE.  For details, see the long version of this article.
+} % END SHORTVERSION
 
@@
 long version of this article.
-}
+} % END SHORTVERSION
 

### 2017-02-27 nbe/sec-asub.tex
[ ICFP 17 ] Moved section algorithmic subtyping to long version.

% Section on algorithmic subtyping delegated to long version

Algorithmic subtyping \fbox{$\Gamma \der T \asub T'$}.
\begin{gather*}
  \ru{T \evalsto n \qquad
      T' \evalsto n' \qquad
      \Nbe \Gamma n \equiv \Nbe \Gamma n'
    }{\Gamma \der T \asub T'}
\\[2ex]
  \rux{T \evalsto \Set[\ell] \qquad
      T' \evalsto \Set[\ell']
    }{\Gamma \der T \asub T'
    }{\ell \leq \ell'}
\qquad
  \ru{T_1' \evalsto \piT{U_1}{T_1} \qquad
      T_2' \evalsto \piT{U_2}{T_2} \qquad
      \Gamma \der U_2 \asub U_1 \qquad
      \cext \Gamma {U_2} \der T_1 \asub T_2
    }{\Gamma \der T_1' \asub T_2'}
\\[2ex]
  \ru{T \evalsto \Nat a \qquad
      T' \evalsto \Nat a'
    }{\Gamma \der T \asub T'
    }{a \leq a'}
\qquad
  \ru{T_1' \evalsto \epiT\Size{T_1} \qquad
      T_2' \evalsto \epiT\Size{T_2} \qquad
      \cext \Gamma {\Size} \der T_1 \asub T_2
    }{\Gamma \der T_1' \asub T_2'}
\end{gather*}

\begin{lemma}[Soundness of algorithmic subtyping]
  \label{lem:asubsound}
  If\/ $\Gamma \der T \asub T'$ then $\Gamma \der T \leq T'$.
\end{lemma}
\begin{proof}
  By induction on $\Gamma \der T \asub T'$, using soundness of NbE and subject reduction (Thm.~\ref{thm:sr}).
\end{proof}

\begin{lemma}[Semantic subtyping implies algorithmic subtyping]
\label{lem:lrasub}
\bla
   If\/ $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \asub T'$.
\end{lemma}
\begin{proof*}
By induction on % $A_1' \in s_1$ and $A_2' \in s_2$ and cases on the inductive relations
$\LRTS \Gamma {T_1'} {A_1'} {s_1}$ and $\LRTS \Gamma {T_2'} {A_2'} {s_2}$
and cases on $A_1' \leq A_2' \in \TYPE$
we prove $\Gamma \der T_1' \asub T_2'$.

\begin{caselist}

\nextcase $A_i' \evalsto N_i$ and $T_i' \evalsto n_i$ and $N_1 = N_2 \in \NE$ and $\RTY \Gamma {n_i} {N_i} {s_i}$.
By Cor.~\ref{cor:oneone} $\Gamma \der n_1 = n_2$, thus, by completeness of NbE, $\Nbe\Gamma n_1 \equiv \Nbe\Gamma n_2$, which entails $\Gamma \der T_1' \asub T_2'$.

\nextcase $A_i' \evalsto \piT {A_i} {B_i}$ and  $T_i' \evalsto \piT{U_i}{T_i}$.
From $\LRTS \Gamma {U_i} {A_i} {s_i}$ and $A_2 \leq A_1 \in \TYPE$ we get $\Gamma \der U_2 \asub U_1$
by induction hypothesis on $A_i \in s_i$.
Let $a := {\up{A_2} \var\Gamma}$.  With $\LR {\cext \Gamma {U_2}} {\ind 0} {(U_2\slift)} a {A_2}$ we get
$\LRTS {\cext \Gamma {U_2}} {T_i} {B_i[a]} s$ and $B_1[a] \leq B_2[a] \in \TYPE$.  Thus by induction hypothesis on $B_i[a] \in s_i$ we obtain $\cext \Gamma {U_2} \der T_1 \leq T_2$, together $\Gamma \der T_1' \asub T_2'$.
\qed
\end{caselist}
\end{proof*}

\begin{corollary}[Completeness of algorithmic subtyping]
\label{cor:complasub}
If\/ $\Gamma \der T \leq T'$ then $\Gamma \der T \asub T'$.
\end{corollary}
\begin{proof}
  By the fundamental theorems $\RG \Gamma T {T\rhog}$ and $\RG \Gamma {T'} {T'\rhog}$ and $T\rhog \leq T'\rhog \in \TYPE$.
  By Lemma~\ref{lem:lrasub}, $\Gamma \der T \asub T'$.
\end{proof}

Since the algorithmic subtyping relation is equivalent to the declarative one, we can freely swap one relation for the other.

\begin{lemma}[Termination of algorithmic subtyping]
  \label{lem:termasub}
  If\/ $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ then the query $\Gamma \der T \asub T'$ terminates.
\end{lemma}
\begin{proof}
  By induction on $A \in s$ and $A' \in s'$ and cases on $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$.
\end{proof}

\begin{theorem}[Decidability of subtyping]
  If\/ $\Gamma \der T,T'$, then $\Gamma \der T \leq T'$ is decided by the query $\Gamma \der T \asub T'$.
\end{theorem}
\begin{proof}
  By the fundamental theorem of typing, $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$,
  thus, the query  $\Gamma \der T \asub T'$ terminates by Lemma~\ref{lem:termasub}.
  If successfully, then $\Gamma \der T \leq T'$ by soundness of algorithmic equality.
  If unsuccessfully, then $\Gamma \der T \leq T'$ is impossible by completeness of algorithmic equality.
\end{proof}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "icfp17-long"
%%% End:

### 2017-02-28 nbe/icfp17.tex
[ ICFP 17 ] Moved proof of subject reduction to long version.

@@
 For the Logical Framework (LF), the simplest dependent lambda-calculus, the skeletons are simple types that can be obtained from the original dependent types by erasing the dependencies:  dependent function types map to non-dependent ones and indexed data types to simple data types.
-\citet{harperpfenning:equivalenceLF} present such an equality check for LF which is directed by simple types, and their technique should scale to other type theories that admit dependency erasure.\footnote{For instance, the types of the Calculus of Constructions erase to F$^\omega$-types \cite{geuvers:shortFlexibleSNCC}, and the latter could be used to direct the equality check.
+\citet{harperPfenning:equivalenceLF} present such an equality check for LF which is directed by simple types, and their technique should scale to other type theories that admit dependency erasure.\footnote{For instance, the types of the Calculus of Constructions erase to F$^\omega$-types \cite{geuvers:shortFlexibleSNCC}, and the latter could be used to direct the equality check.
 \citet{lovasPfenning:lmcs10} consider also \emph{refinement types for logical frameworks} which can be erased to simple types.}
@@
 We write \fbox{$\RG \Gamma T A$} if\/ $\LRTS \Gamma T A s$ for some sort $s$.
-
+\SHORTVERSION{The logical relations are closed under weakening.}
+\LONGVERSION{ % BEGIN LONG
 \begin{lemma}[Weakening]
@@
 \end{lemma}
+} % END LONG
 
@@
 simultaneously by induction on $A' \in s$.
-
+\SHORTLONG{\qed}{ % LONG VERSION
 \begin{caselist}
@@
 \end{caselist}
+} % END LONGVERSION
 \end{proof*}
 
+\LONGVERSION{
 \begin{corollary}[One-to-one]
@@
 \end{corollary}
-
+} % END LONGVERSION
 \begin{corollary}[Fresh variable]
@@
 \end{proof}
+}
 
@@
 
+\LONGVERSION{ % BEGIN LONGVERSION
 \begin{lemma}[Properties of the logical relation for substitution]
@@
 \end{lemma}
+} % END LONGVERSION
 \LONGVERSION{
@@
 \end{lemma}
+\LONGVERSION{ % BEGIN LONGVERSION
 \begin{proof*}
   By induction on $\der \Gamma$.
-\SHORTLONG{\qed}{ % LONGVERSION
 \begin{caselist}
@@
 \end{caselist}
+\end{proof*}
 } % END LONGVERSION
-\end{proof*}
 
@@
 } % END LONGVERSION
-For normalization of types, we introduce \fbox{$\Nbe \Gamma T$}
-via the relation $\Nbe \Gamma T \evalsto V \defas \RTy \Gamma {T\rhog} V$.
-Analogously, we show that the test $\Nbe \Gamma T \equiv \Nbe \Gamma T'$
-decides type equality $\Gamma \der T = T'$.
-
-\begin{corollary}[Type weak head normalization]
-  If\/ $\Gamma \der T$ then $T \evalsto W$ for some $W$.
-\end{corollary}
+
 \LONGVERSION{
-\begin{proof}
-  By the fundamental theorem of typing, $\RG \Gamma T {T\rhog}$,
-  which implies $T \evalsto W$ by definition of the logical relation.
-\end{proof}
-} % END LONGVERSION
-\LONGVERSION{ % BEGIN LONGVERSION
-\begin{corollary}[Type constructor injectivity]
-\label{cor:piinj}
-\bla
-\begin{enumerate}
-\item \label{it:setinj}
-  If\/ $\Gamma \der \Set[\ell] = \Set[\ell'] : s$ then $\ell = \ell'$.
-\item \label{it:natinj}
-  If\/ $\Gamma \der \Nat a = \Nat b : s$ then $a = b$.
-\item \label{it:piinj}
-  If\/ $\Gamma \der \piT U T = \piT{U'}{T'} : s$ then
-  $\Gamma \der U = U' : s$ and
-  $\cext \Gamma U \der T = T' : s$.
-\item \label{it:allinj}
-  If\/ $\Gamma \der \epiT \Size T = \piT{\Size}{T'} : s$ then
-  $\cext \Gamma \Size \der T = T' : s$.
-\end{enumerate}
-\end{corollary}
-\begin{proof}
-\bla
-Statement (\ref{it:setinj}) follows by inversion on  $\Set[\ell] = \Set[\ell'] \in s$, which is a direct consequence of the fundamental theorem (Thm.~\ref{thm:fund}).
-
-For (\ref{it:natinj}), observe that $\Nat (a\rhog) = \Nat (b\rhog) \in s$
-by the fundamental theorem (Thm.~\ref{thm:fund}), which by definition implies $a\rhog = b\rhog$. % \in \SIZE$.
-By the fundamental theorem of typing (Thm.~\ref{thm:fundty}), $\LRTS \Gamma {\Nat a} {\Nat (a\rhog)} s$
-which by definition implies $\LRSize \Gamma a {a\rhog}$.  Analogously, we get $\LRSize \Gamma b {b\rhog}$.
-By Cor.~\ref{cor:oneone}, $a = b$.
-
-Last, we prove statement (\ref{it:piinj}), the last statement follows analogously.
-By the fundamental theorem,
-$(\piT U T)\rhog = (\piT{U'}{T'})\rhog \in s$ which by definition means
-$U\rhog = U'\rhog \in s$ and $T(\rhog,u) = T'(\rhog,u') \in s$ for all $u = u' \in s$.
-By reification (Thm.~\ref{thm:rere}),
-$\RTy\Gamma {U\rhog} V$ and $\RTy\Gamma {U'\rhog} V$ for some normal form
-$V$.
-Since by inversion we have $\Gamma \der U,U' : s$ we get $\Gamma \der U = U' : s$ by soundness of NbE.
-Now, choosing $u := u' := \up {U\rhog} \var\Gamma \in U$, we obtain
-$T\rhogu = T' \rhogu \in s$, which analogously gives us $\cext\Gamma U \der T = T' : s$.
-\end{proof}
-} % END LONGVERSION
-\begin{lemma}[Type constructor discrimination]
-  \label{lem:discr}
-  Different type constructors are not related by subtyping.  For instance,
-  $\Gamma \der \Nat a \leq \piT U T$ and
-  $\Gamma \der \erpiT \Size T \leq \piT \Size T'$ are both impossible.
-%  There are $4\times4 = 16$ statements of that sort.
-  % The following judgments are \emph{not} derivable.
-  % \begin{enumerate}
-  % \item $\Gamma \der \Nat a \leq \Set[\ell]$ and $\Gamma \der \
-  % \end{enumerate}
-\end{lemma}
-\begin{proof}
-  Follows directly by the fundamental theorem (Thm.~\ref{thm:fund}) and inversion on semantic subtyping.
-\end{proof}
-% \section{Algorithmic Subtyping}
-% \label{sec:algosub}
-
-\begin{lemma}[Inversion of subtyping \cite{fridlenderPagano:tlca13}]
-\label{lem:invsubty}
-\bla
-\begin{enumerate}
-\item \label{it:setinvsub}
-  If\/ $\Gamma \der \Set[\ell] \leq \Set[\ell'] : s$ then $\ell \leq \ell'$.
-\item \label{it:natinvsub}
-  If\/ $\Gamma \der \Nat a \leq \Nat b : s$ then $a \leq b$.
-\item \label{it:piinvsub}
-  If\/ $\Gamma \der \piT U T \leq \piT{U'}{T'} : s$ then
-  $\Gamma \der U' \leq U : s$ and
-  $\cext \Gamma {U'} \der T \leq T' : s$.
-\item \label{it:allinvsub}
-  If\/ $\Gamma \der \epiT \Size T \leq \piT{\Size}{T'} : s$ then
-  $\cext \Gamma \Size \der T \leq T' : s$.
-\end{enumerate}
-\end{lemma}
-\begin{proof}
-\LONGVERSION{Similar to the proof of Cor.~\ref{cor:piinj}, using
-  Lemma~\ref{lem:lrsub}.} % END LONGVERSION
-We recapitulate the proof for (\ref{it:piinvsub}).
-By the fundamental theorems, we get
-$\RG \Gamma {\piT U T} {(\piT U T)\rhog}$ and
-$\RG \Gamma {\piT {U'} {T'}} {(\piT {U'} {T'})\rhog}$ and
-${(\piT U T)\rhog} \leq {(\piT {U'} {T'})\rhog} \in \TYPE$.
-By inversion, first
-$\RG \Gamma U {U\rhog}$ and
-$\RG \Gamma {U'} {U'\rhog}$ and
-$U'\rhog \leq U\rhog \in \TYPE$ which imply $\Gamma \der U' \leq U$ by Lemma~\ref{lem:lrsub}.
-Recall that
-$\rhogup = (\rhog, u)$ where
-$u = \up{U'\rhog} \var\Gamma \in U'\rhog$ and
-$\LR {\cext \Gamma U'} {\ind0} {U'\slift} {u} {U'\rhog}$ by Cor.~\ref{cor:fresh}.
-From
-$\LRTS {\cext \Gamma U'} T {T(\rhogup)} s$ and
-$\LRTS {\cext \Gamma U'} {T'} {T'(\rhogup)} s$ and
-$T(\rhogup) \leq T'(\rhogup) \in s$, again by Lemma~\ref{lem:lrsub},
-$\cext \Gamma {U'} \der T \leq T' : s$.
-\end{proof}
-
-\begin{lemma}[Strong inversion for abstraction]
-  \label{lem:invabs}
-  \bla
-  \begin{enumerate}
-  \item   If\/ $\Gamma \der \lambda t : \piT U T$ then
-  $\cext \Gamma U \der t : T$.
-  \item   If\/ $\Gamma \der \lambda t : \epiT \Size T$ then
-  $\eext \Gamma \Size \der t : T$.
-  \end{enumerate}
-\end{lemma}
-\begin{proof}
-  By inversion of typing (Lemma~\ref{lem:invty}), type constructor discrimination \label{lem:discr}
-  and inversion of subtyping (Lemma~\ref{lem:invsubty}).
-
-  For instance, inversion on $\Gamma \der \lambda t : \piT U T$ gives us $\cext \Gamma {U'} \der t : {T'}$ with $\Gamma \der \piT{U'}{T'} \leq \piT U T$ (the other case, $\epiT \Size T$ is excluded by discrimination).
-By inversion of subtyping, $\Gamma \der U \leq U'$ and $\cext \Gamma {U} \der T' \leq T$.
-Since $\sid : \cext \Gamma U \leq \cext \Gamma {U'}$, we have $\cext \Gamma U \der t : T'$ by weakening, and our goal follows by subsumption.
-\end{proof}
-
-\begin{lemma}[Strong inversion of redexes]
-  \label{lem:stronginv}
-  \bla
-  \begin{enumerate}
-  \item   If\/ $\Gamma \der (\lambda t)\,u : T'$ then
-  $\cext \Gamma U \der t : T$ and $\Gamma \der u : U$ and $\Gamma \der T[u] \leq T'$
-  for some $U,T$.
-  \item   If\/ $\Gamma \der (\lambda t)\,a : T'$ then
-  $\cext \Gamma \Size \der t : T$ and $\Gamma \der a : \Size$ and $\Gamma \der T[a] \leq T'$
-  for some $T$.
-  \item \label{it:sinvall}
-  If\/ $\Gamma \der (\lambda t)\,\ann a : T'$ then
-  $\erext \Gamma \Size \der t : T$ and $\resurrect \Gamma \der a : \Size$ and $\Gamma \der T[a] \leq T'$
-  for some $T$.
-  \end{enumerate}
-\end{lemma}
-\begin{proof}
-  For (\ref{it:sinvall}), from $\Gamma \der (\lambda t) \ann a : T'$ we get $\Gamma \der \lambda t : \forall T$ and $\resurrect \Gamma \der a,b : \Size$ and $\Gamma \der T[b] \leq T'$.
-Strong inversion for abstraction gives us $\erext \Gamma \Size \der t : T$.
-Since $\Gamma \der [a] = [a] \eeq [b] : \erext \Gamma \Size$,  the substitution lemma yields
-$\Gamma \der t[a] : T[b]$, thus $\Gamma \der t[a] : T'$ by subsumption.
-\end{proof}
-
-\begin{theorem}[Subject reduction]
-  \label{thm:sr}
-  \bla
-  \begin{enumerate}
-  \item
-  If\/ $\DD :: \Gamma \der t : T$ and $\DD' :: t \evalsto w$ then $\Gamma \der t = w : T$.
-  \item
-  If\/ $\DD :: \Gamma \der w\, e : T$ and $\DD' :: w @ e \evalsto w'$ then $\Gamma \der w\,e = w : T$.
-  \end{enumerate}
-\end{theorem}
-\begin{proof}
-  Simultaneously by induction on $\DD'$
-  % By induction on $t \evalsto w$,
-  using inversion (Lemma~\ref{lem:invty}) and strong inversion (Lemma~\ref{lem:stronginv})
-  on the typing derivation $\DD$. % $\Gamma \der t : T$.
-\end{proof}
+\input{sec-sr}
+} % END LONG VERSION
+\SHORTVERSION{
+From correctness of NbE and the logical relations we can further prove
+injectivity of type constructors, inversion of subtyping, and subject reduction.
+The proofs follow roughly \citet{fridlenderPagano:tlca13}, for
+details, see the long version of this article.
+} % END SHORTVERSION
 
@@
 \LONGVERSION{
-\input{sec-tycheck}
+\input{sec-asub}
 } % END LONG VERSION

### 2017-02-28 nbe/sec-sr.tex
[ ICFP 17 ] Moved proof of subject reduction to long version.

% Proof of subject reduction delegated to long version

For normalization of types, we introduce \fbox{$\Nbe \Gamma T$}
via the relation $\Nbe \Gamma T \evalsto V \defas \RTy \Gamma {T\rhog} V$.
Analogously, we show that the test $\Nbe \Gamma T \equiv \Nbe \Gamma T'$
decides type equality $\Gamma \der T = T'$.

\begin{corollary}[Type weak head normalization]
  If\/ $\Gamma \der T$ then $T \evalsto W$ for some $W$.
\end{corollary}
\LONGVERSION{
\begin{proof}
  By the fundamental theorem of typing, $\RG \Gamma T {T\rhog}$,
  which implies $T \evalsto W$ by definition of the logical relation.
\end{proof}
} % END LONGVERSION
\LONGVERSION{ % BEGIN LONGVERSION
\begin{corollary}[Type constructor injectivity]
\label{cor:piinj}
\bla
\begin{enumerate}
\item \label{it:setinj}
  If\/ $\Gamma \der \Set[\ell] = \Set[\ell'] : s$ then $\ell = \ell'$.
\item \label{it:natinj}
  If\/ $\Gamma \der \Nat a = \Nat b : s$ then $a = b$.
\item \label{it:piinj}
  If\/ $\Gamma \der \piT U T = \piT{U'}{T'} : s$ then
  $\Gamma \der U = U' : s$ and
  $\cext \Gamma U \der T = T' : s$.
\item \label{it:allinj}
  If\/ $\Gamma \der \epiT \Size T = \piT{\Size}{T'} : s$ then
  $\cext \Gamma \Size \der T = T' : s$.
\end{enumerate}
\end{corollary}
\begin{proof}
\bla
Statement (\ref{it:setinj}) follows by inversion on  $\Set[\ell] = \Set[\ell'] \in s$, which is a direct consequence of the fundamental theorem (Thm.~\ref{thm:fund}).

For (\ref{it:natinj}), observe that $\Nat (a\rhog) = \Nat (b\rhog) \in s$
by the fundamental theorem (Thm.~\ref{thm:fund}), which by definition implies $a\rhog = b\rhog$. % \in \SIZE$.
By the fundamental theorem of typing (Thm.~\ref{thm:fundty}), $\LRTS \Gamma {\Nat a} {\Nat (a\rhog)} s$
which by definition implies $\LRSize \Gamma a {a\rhog}$.  Analogously, we get $\LRSize \Gamma b {b\rhog}$.
By Cor.~\ref{cor:oneone}, $a = b$.

Last, we prove statement (\ref{it:piinj}), the last statement follows analogously.
By the fundamental theorem,
$(\piT U T)\rhog = (\piT{U'}{T'})\rhog \in s$ which by definition means
$U\rhog = U'\rhog \in s$ and $T(\rhog,u) = T'(\rhog,u') \in s$ for all $u = u' \in s$.
By reification (Thm.~\ref{thm:rere}),
$\RTy\Gamma {U\rhog} V$ and $\RTy\Gamma {U'\rhog} V$ for some normal form
$V$.
Since by inversion we have $\Gamma \der U,U' : s$ we get $\Gamma \der U = U' : s$ by soundness of NbE.
Now, choosing $u := u' := \up {U\rhog} \var\Gamma \in U$, we obtain
$T\rhogu = T' \rhogu \in s$, which analogously gives us $\cext\Gamma U \der T = T' : s$.
\end{proof}
} % END LONGVERSION
\begin{lemma}[Type constructor discrimination]
  \label{lem:discr}
  Different type constructors are not related by subtyping.  For instance,
  $\Gamma \der \Nat a \leq \piT U T$ and
  $\Gamma \der \erpiT \Size T \leq \piT \Size T'$ are both impossible.
%  There are $4\times4 = 16$ statements of that sort.
  % The following judgments are \emph{not} derivable.
  % \begin{enumerate}
  % \item $\Gamma \der \Nat a \leq \Set[\ell]$ and $\Gamma \der \
  % \end{enumerate}
\end{lemma}
\begin{proof}
  Follows directly by the fundamental theorem (Thm.~\ref{thm:fund}) and inversion on semantic subtyping.
\end{proof}
% \section{Algorithmic Subtyping}
% \label{sec:algosub}

\begin{lemma}[Inversion of subtyping \cite{fridlenderPagano:tlca13}]
\label{lem:invsubty}
\bla
\begin{enumerate}
\item \label{it:setinvsub}
  If\/ $\Gamma \der \Set[\ell] \leq \Set[\ell'] : s$ then $\ell \leq \ell'$.
\item \label{it:natinvsub}
  If\/ $\Gamma \der \Nat a \leq \Nat b : s$ then $a \leq b$.
\item \label{it:piinvsub}
  If\/ $\Gamma \der \piT U T \leq \piT{U'}{T'} : s$ then
  $\Gamma \der U' \leq U : s$ and
  $\cext \Gamma {U'} \der T \leq T' : s$.
\item \label{it:allinvsub}
  If\/ $\Gamma \der \epiT \Size T \leq \piT{\Size}{T'} : s$ then
  $\cext \Gamma \Size \der T \leq T' : s$.
\end{enumerate}
\end{lemma}
\begin{proof}
\LONGVERSION{Similar to the proof of Cor.~\ref{cor:piinj}, using
  Lemma~\ref{lem:lrsub}.} % END LONGVERSION
We recapitulate the proof for (\ref{it:piinvsub}).
By the fundamental theorems, we get
$\RG \Gamma {\piT U T} {(\piT U T)\rhog}$ and
$\RG \Gamma {\piT {U'} {T'}} {(\piT {U'} {T'})\rhog}$ and
${(\piT U T)\rhog} \leq {(\piT {U'} {T'})\rhog} \in \TYPE$.
By inversion, first
$\RG \Gamma U {U\rhog}$ and
$\RG \Gamma {U'} {U'\rhog}$ and
$U'\rhog \leq U\rhog \in \TYPE$ which imply $\Gamma \der U' \leq U$ by Lemma~\ref{lem:lrsub}.
Recall that
$\rhogup = (\rhog, u)$ where
$u = \up{U'\rhog} \var\Gamma \in U'\rhog$ and
$\LR {\cext \Gamma U'} {\ind0} {U'\slift} {u} {U'\rhog}$ by Cor.~\ref{cor:fresh}.
From
$\LRTS {\cext \Gamma U'} T {T(\rhogup)} s$ and
$\LRTS {\cext \Gamma U'} {T'} {T'(\rhogup)} s$ and
$T(\rhogup) \leq T'(\rhogup) \in s$, again by Lemma~\ref{lem:lrsub},
$\cext \Gamma {U'} \der T \leq T' : s$.
\end{proof}

\begin{lemma}[Strong inversion for abstraction]
  \label{lem:invabs}
  \bla
  \begin{enumerate}
  \item   If\/ $\Gamma \der \lambda t : \piT U T$ then
  $\cext \Gamma U \der t : T$.
  \item   If\/ $\Gamma \der \lambda t : \epiT \Size T$ then
  $\eext \Gamma \Size \der t : T$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  By inversion of typing (Lemma~\ref{lem:invty}), type constructor discrimination (Lemma~\ref{lem:discr})
  and inversion of subtyping (Lemma~\ref{lem:invsubty}).

  For instance, inversion on $\Gamma \der \lambda t : \piT U T$ gives us $\cext \Gamma {U'} \der t : {T'}$ with $\Gamma \der \piT{U'}{T'} \leq \piT U T$ (the other case, $\epiT \Size T$ is excluded by discrimination).  By inversion of subtyping, $\Gamma \der U \leq U'$ and $\cext \Gamma {U} \der T' \leq T$.  Since $\sid : \cext \Gamma U \leq \cext \Gamma {U'}$, we have $\cext \Gamma U \der t : T'$ by weakening, and our goal follows by subsumption.
\end{proof}

\begin{lemma}[Strong inversion of redexes]
  \label{lem:stronginv}
  \bla
  \begin{enumerate}
  \item   If\/ $\Gamma \der (\lambda t)\,u : T'$ then
  $\cext \Gamma U \der t : T$ and $\Gamma \der u : U$ and $\Gamma \der T[u] \leq T'$
  for some $U,T$.
  \item   If\/ $\Gamma \der (\lambda t)\,a : T'$ then
  $\cext \Gamma \Size \der t : T$ and $\Gamma \der a : \Size$ and $\Gamma \der T[a] \leq T'$
  for some $T$.
  \item \label{it:sinvall}
  If\/ $\Gamma \der (\lambda t)\,\ann a : T'$ then
  $\erext \Gamma \Size \der t : T$ and $\resurrect \Gamma \der a : \Size$ and $\Gamma \der T[a] \leq T'$
  for some $T$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  For (\ref{it:sinvall}), from $\Gamma \der (\lambda t) \ann a : T'$ we get $\Gamma \der \lambda t : \forall T$ and $\resurrect \Gamma \der a,b : \Size$ and $\Gamma \der T[b] \leq T'$.  Strong inversion for abstraction gives us $\erext \Gamma \Size \der t : T$.  Since $\Gamma \der [a] = [a] \eeq [b] : \erext \Gamma \Size$,  the substitution lemma yields
$\Gamma \der t[a] : T[b]$, thus $\Gamma \der t[a] : T'$ by subsumption.
\end{proof}

\begin{theorem}[Subject reduction]
  \label{thm:sr}
  \bla
  \begin{enumerate}
  \item
  If\/ $\DD :: \Gamma \der t : T$ and $\DD' :: t \evalsto w$ then $\Gamma \der t = w : T$.
  \item
  If\/ $\DD :: \Gamma \der w\, e : T$ and $\DD' :: w @ e \evalsto w'$ then $\Gamma \der w\,e = w : T$.
  \end{enumerate}
\end{theorem}
\begin{proof}
  Simultaneously by induction on $\DD'$
  % By induction on $t \evalsto w$,
  using inversion (Lemma~\ref{lem:invty}) and strong inversion (Lemma~\ref{lem:stronginv})
  on the typing derivation $\DD$. % $\Gamma \der t : T$.
\end{proof}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "icfp17-long"
%%% End:

### 2017-02-28 nbe/icfp17.tex
fixed some typos in section 3

@@
 The type annotation $T$ in $\tcase$ allows us to infer the type of the whole case statement $t\,e$ as $T\,t$.
-The function call $c\,e'$ for a canonical number reduces to $f'\,(\lambda x.\,x\,e')\,c$ where we allowed ourself the use of a named abstraction in the presentation to the reader.
+The function call $c\,e'$ for a canonical number reduces to $f'\,(\lambda x.\,x\,e')\,c$ where we allowed ourselves the use of a named abstraction in the presentation to the reader.
 The unfolding of fixed-points is thus restricted to application to canonical numbers; this is the usual reduction strategy which converges for terminating functions \cite{gimenez:typeBased}.
@@
 When type checking a variable, it needs to be bound in the context relevantly.
-However, when entering a irrelevant position, for instance when checking size $a$ in term $\suc a t$ we declare previously irrelevant variables as relevant.
+However, when entering an irrelevant position, for instance when checking size $a$ in term $\suc a t$ we declare previously irrelevant variables as relevant.
 This operation on the context has been coined
@@
 \item If\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
-\item If\/ $\Gamma \der J : \resurrect \Delta$ % and $\der \Delta$
-  then $\Gamma \der J : \Delta$.
+\item If\/ $\Gamma \der \sigma : \resurrect \Delta$ % and $\der \Delta$
+  then $\Gamma \der \sigma : \Delta$.
 \item \label{it:ressubst} If\/ $\Gamma \der \sigma : \Delta$ % then $\Gamma \der \sigma : \resurrect\Delta$.

### 2017-02-28 nbe/icfp17.tex
[ ICFP 17 ] Description of values.

@@
    & ::= & \zero a \mid \suc a t & \mbox{constructed data} \\
-\Ne & \ni & n
+\NeExp & \ni & n
     & ::= & \ind i \mid n \, e & \mbox{neutral expressions} \\
@@
 \item If\/ $\Gamma \der \suc a t : T'$ then $\resurrect \Gamma \der a,b : \Size$ and $\Gamma \der t : \Nat b$ and $\resurrect\Gamma \der \Nat(b+1) \leq T'$.
-\item If\/ $\Gamma \der u \case[\ell] T {t_z} {t_s} : T'$ then $\Gamma \der u : \Nat (a+1)$ and $\Gamma \der T : \Nat(a+1) \to \Set[\ell]$ and $\Gamma \der t_z : T\,(\zero a)$ and $\Gamma \der t_s : \funT x {\Nat a} {T\,(\suc a x)}$ and $\resurrect\Gamma \der T\,u \leq T'$ for some $a$.
-\item If\/ $\Gamma \der u \fix[\ell] T t : T'$ then $\Gamma \der u : \Nat a$ and $\Gamma \der T : \Adm\,\ell$ and $\Gamma \der t : \FixT\,T$ and $\resurrect\Gamma \der T\,a\,u \leq T'$.
+\item If\/ $\Gamma \der u\,\case[\ell] T {t_z} {t_s} : T'$ then $\Gamma \der u : \Nat (a+1)$ and $\Gamma \der T : \Nat(a+1) \to \Set[\ell]$ and $\Gamma \der t_z : T\,(\zero a)$ and $\Gamma \der t_s : \funT x {\Nat a} {T\,(\suc a x)}$ and $\resurrect\Gamma \der T\,u \leq T'$ for some $a$.
+\item If\/ $\Gamma \der u\,\fix[\ell] T t : T'$ then $\Gamma \der u : \Nat a$ and $\Gamma \der T : \Adm\,\ell$ and $\Gamma \der t : \FixT\,T$ and $\resurrect\Gamma \der T\,a\,u \leq T'$.
 \end{enumerate}
@@
 
-We extend the syntax by de Bruijn levels $\lev k$
-to be used as generic values (unknowns) and type annotations
-$\up T t$ and $\down T t$ for lazy realizations of the
+In this section we present an operational semantics of our language,
+define the NbE algorithm, construct a PER model, and demonstrate that
+NbE is complete for definitional equality, \ie, if
+$\Gamma \der t = t' : T$, then $t$ and $t'$ have the same normal form.
+\[
+\begin{array}{lllll@{\qquad}l}
+  \Ne & \ni &
+  m & ::= & \ind i \mid m\,v \mid m\,a \mid m\,\ann a \mid
+            m\,\case[\ell] V{v_z}{v_s} \mid m\,\fix[\ell] V v
+    & \mbox{normal neutral term} \\
+  \Nf & \ni &
+  v & ::= & m \mid \lambda v \mid \zero a \mid \suc a v \mid
+            \Set[\ell] \mid \Nat a \mid \piT {V_u}{V_t} \mid
+            \epiT\Size V
+    & \mbox{normal form} \\
+\end{array}
+\]
+For the operational semantics,
+instead of defining a separate language of values,
+we extend the syntax of expressions by de Bruijn levels $\lev k$
+to be used as generic values (unknowns), and type annotations
+$\up A n$ and $\down A t$ for lazy realizations of the
 reflection and reification operations of NbE.
+\emph{Terms} are expressions that do not contain these new expression
+forms. \emph{Values} \fbox{$f,g,A,B,F \in \D$} are expressions with no free de Bruijn
+indices, where each neutral $n$ is under a reflection marker $\up A n$.
+The types $A$ that direct reflection $\up A n$ and reification
+$\down A f$ also live in the value world.
 \[
@@
 %  \Exp  & \ni & t & ::= & \dots \mid \lev k \mid \up T t \mid \down T t \\
-  \Ne   & \ni & n & ::= & \dots \mid \lev k \\
-  \tUp  & \ni & N & ::= & \up T n \\
-  \Whnf & \ni & w & ::= & \dots \mid N \\
-  \tDown & \ni & d & ::= & \down T t \\
-  \Exp  & \ni & t & ::= & \dots \mid d \\
+  \tUp  & \ni & N & ::= & \up A n
+     & \mbox{reflection of neutral term $n$ into the value world $\D$, at
+       type $A$} \\
+  \tDown & \ni & d & ::= & \down A f
+     & \mbox{reification of value $f$ at type $A$} \\
+  \NeExp & \ni & n & ::= & \dots \mid \lev k
+    & \mbox{de Bruijn level $k$} \\
+  \Whnf & \ni & w & ::= & \dots \mid N
+    & \mbox{reflected neutral are weak head normal} \\
+  \Exp  & \ni & t & ::= & \dots \mid d
+    & \mbox{reified values may occur in expressions} \\
 %  \Whnf & \ni & w,d,f,g,A,B,F,G & ::= & \dots \mid  \up T n \mid \down T t \\
+\end{array}
+\]
+\emph{De~Bruijn levels} are the mirror images of de Bruijn indices.  While de Bruijn
+indices index the context from the right, \ie, $\ind 0$ refers the
+last type that entered the context, de Bruijn levels index it from the
+left, \ie, $\lev 0$ refers to the first type in the context.  This
+way, de Bruijn levels are stable under context extensions, and
+suitable to represent unknowns.
+
+\emph{Size values} \fbox{$\alpha, \beta \in \SIZE$} are size
+expressions that use de Bruijn levels instead of de Bruijn
+indices. Comparison for size values $\alpha \leq \beta$ is the
+analogue of the one for size terms $a \leq b$.  In the following, we
+will reuse letter $a$ for a value if it cannot be confused with a size
+term.
+\[
+\begin{array}{lllll@{\qquad}l}
+  \DNe & \ni &
+  n & ::= & \ind i \mid n\,d \mid n\,\alpha \mid n\,\ann \alpha \mid
+            n\,\case[\ell] D{d_z}{d_s} \mid n\,\fix[\ell] D d
+    & \mbox{unreflected neutral value} \\
+  \DNf & \ni &
+  d & ::= & \down A f
+    & \mbox{reified value} \\
 \end{array}
@@
 First, we close the term $t$ with an environment $\eta$ that maps the free de Bruijn indices of $t$ to reflected de Bruijn levels.
-Reflection of de Bruijn levels follows the blue path $\Level \to \Dne \to \D$: Levels embed via constructor $\var{}$ into semantic neutrals $\DNe$ which are labeled with their type $A \in \D$ to become and element $\up A \lev j \in \D$.
+Reflection of de Bruijn levels follows the blue path $\Level \to \DNe \to \D$: Levels embed via constructor $\var{}$ into semantic neutrals $\DNe$ which are labeled with their type $A \in \D$ to become and element $\up A \lev j \in \D$.
 
 \item
-Then, we label value $t \eta \in \D$ with its type $A$ to obtain $\down A t \eta \in \Dnf$.
+Then, we label value $t \eta \in \D$ with its type $A$ to obtain $\down A t \eta \in \DNf$.
 \item

### 2017-02-28 nbe/icfp17.tex
[ ICFP 17 ] Comment on read back.

--- 2017-02-28 nbe/icfp17.tex
@@
 \begin{array}{lllrl@{\qquad}l}
-%  \Exp  & \ni & t & ::= & \dots \mid \lev k \mid \up T t \mid \down T t \\
+  \NeExp & \ni & n & ::= & \dots \mid \lev k
+    & \mbox{de Bruijn level $k$} \\
   \tUp  & \ni & N & ::= & \up A n
@@
        type $A$} \\
-  \tDown & \ni & d & ::= & \down A f
-     & \mbox{reification of value $f$ at type $A$} \\
-  \NeExp & \ni & n & ::= & \dots \mid \lev k
-    & \mbox{de Bruijn level $k$} \\
   \Whnf & \ni & w & ::= & \dots \mid N
     & \mbox{reflected neutral are weak head normal} \\
-  \Exp  & \ni & t & ::= & \dots \mid d
-    & \mbox{reified values may occur in expressions} \\
-%  \Whnf & \ni & w,d,f,g,A,B,F,G & ::= & \dots \mid  \up T n \mid \down T t \\
+  \Exp  & \ni & t & ::= & \dots \mid \down A f
+     & \mbox{reification of value $f$ at type $A$} \\
 \end{array}
 \]
+% \[
+% \begin{array}{lllrl@{\qquad}l}
+% %  \Exp  & \ni & t & ::= & \dots \mid \lev k \mid \up T t \mid \down T t \\
+%   \tUp  & \ni & N & ::= & \up A n
+%      & \mbox{reflection of neutral term $n$ into the value world $\D$, at
+%        type $A$} \\
+%   \tDown & \ni & d & ::= & \down A f
+%      & \mbox{reification of value $f$ at type $A$} \\
+%   \NeExp & \ni & n & ::= & \dots \mid \lev k
+%     & \mbox{de Bruijn level $k$} \\
+%   \Whnf & \ni & w & ::= & \dots \mid N
+%     & \mbox{reflected neutral are weak head normal} \\
+%   \Exp  & \ni & t & ::= & \dots \mid d
+%     & \mbox{reified values may occur in expressions} \\
+% %  \Whnf & \ni & w,d,f,g,A,B,F,G & ::= & \dots \mid  \up T n \mid \down T t \\
+% \end{array}
+% \]
 \emph{De~Bruijn levels} are the mirror images of de Bruijn indices.  While de Bruijn
@@
 term.
+
+Finally, we identify two expression classes for NbE. Neutrals
+$n \in \DNe$ are the ones that will appear in values under the
+reflection marker $\up{A}$.  Reified values $d \in \DNf$ are values
+under a reification marker $\down{A}$.
 \[
@@
 
-Figure~\ref{fig:typedchart}, adapted from \citet{abel:habil} shows the syntactic categories and main operations involved in NbE in what we call \emph{locally nameless style}.
-The red path $\Exp \to \D \to \Dnf \to \Nf$ decomposes $\beta\eta$-normalization into three steps.
+\newcommand{\redto}{\mathrel{\color{red}{\to}}}
+\newcommand{\blueto}{\mathrel{\color{blue}{\to}}}
+
+Figure~\ref{fig:typedchart}, adapted from \citet{abel:habil} summarized the syntactic categories and main operations involved in NbE in what is called \emph{locally nameless style}.
+The \textcolor{red}{red path} $\Exp \redto \D \redto \Dnf \redto \Nf$ decomposes $\beta\eta$-normalization into three steps.
 \begin{enumerate}
 \item
-First, we close the term $t$ with an environment $\eta$ that maps the free de Bruijn indices of $t$ to reflected de Bruijn levels.
-Reflection of de Bruijn levels follows the blue path $\Level \to \DNe \to \D$: Levels embed via constructor $\var{}$ into semantic neutrals $\DNe$ which are labeled with their type $A \in \D$ to become and element $\up A \lev j \in \D$.
+First, we close the term $t$ with an environment \textcolor{blue}{$\eta$} that maps the
+free de Bruijn indices of $t$ to reflected de Bruijn levels.
+Reflection of de Bruijn levels follows the \textcolor{blue}{blue path}
+$\Level \blueto \DNe \blueto \D$: Levels embed via constructor $\var{}$
+into semantic neutrals $\DNe$ which are labeled with their type $A \in \D$ to become and element $\up A \lev j \in \D$.
 
@@
 
-\subsection{Operational semantics}
-
-Judgements.
-\[
-\begin{array}{l@{\qquad}l}
-  t \evalsto w & \mbox{term $t$ evaluates to weak head normal form $w$} \\
-  w @ e \evalsto w' & \mbox{eliminating $w$ with $e$ evaluates to $w'$} \\
-%  w @ \vec e \evalsto w' & \mbox{applying $w$ to eliminations $\vec e$ evaluates to $w'$} \\
-\end{array}
-\]
+\subsection{Weak head reduction}
+
+We define the operational semantics of our language by the weak head
+evaluation relation \fbox{$t \evalsto w$} which is defined on
+expressions, thus works on values as on terms.  It is defined mutually
+with auxiliary relation \fbox{$w @ e \evalsto w'$} stating that weak
+head normal form $w$ is eliminated by $e$ into weak head normal form $w'$.
+% \[
+% \begin{array}{l@{\qquad}l}
+%   t \evalsto w & \mbox{term $t$ evaluates to weak head normal form $w$} \\
+%   w @ e \evalsto w' & \mbox{eliminating $w$ with $e$ evaluates to $w'$} \\
+% %  w @ \vec e \evalsto w' & \mbox{applying $w$ to eliminations $\vec e$ evaluates to $w'$} \\
+% \end{array}
+% \]
 
 \fbox{$t \evalsto w$}
+\vspace{-4ex}
 \begin{gather*}
@@
     }{(\lambda t) @ \ann \alpha \evalsto w}
-\\[2ex]
+\qquad
   \ru{t_z \evalsto w
     }{(\zero \alpha) @ \casett \evalsto w}
-\qquad
+\\[2ex]
   \ru{t_s\, t \evalsto w
     }{(\suc \alpha t) @ \casett \evalsto w}
-\\[2ex]
+\qquad
   \ru{t\,\ann \alpha\,(\lambda x.\, x\, \fixtt)\,c \evalsto w
@@
 \end{gather*}
-The rules added for NbE deal with elimination of delayed reflection:
+%The rules added for NbE deal with elimination of delayed reflection:
+For NbE, we add evaluation rules that deal with elimination of delayed reflection:
 \begin{gather*}
@@
 
-Judgements.
-\[
-\begin{array}{l@{\qquad}l}
-   \R k d  v     & \mbox{} \\
-   \RTy k T  V  & \mbox{} \\
-   \RNat k t  v & \mbox{} \\
-   \RNe k t  m  & \mbox{} \\
-\end{array}
-\]
-\fbox{$\R k d v$}
+The \emph{read back} phase of NbE \cite{gregoireLeroy:icfp02}
+transforms a reified value $d$ into a normal form $v$.  It is
+specified via an inductively defined relation $\R k d v$ and several auxiliary relations.
+The number $k$, will be instantiated by the length of the context
+$\Gamma$ later.  It allows us to transform a de Bruijn level $l$ into a de
+Bruijn index $i$, via the law $i + l + 1 = k$.  At this point, we do
+not ensure that the $k$ is large enough to accommodate the de Bruijn
+levels in $d$.  Levels $l \geq k$ which are to big will simply mapped
+to de Bruijn index $0$.  The correct $k$ is later ensured by our
+logical relation (Section~\ref{sec:soundnbe}).
+% Judgements.
+% \[
+% \begin{array}{l@{\qquad}l}
+%    \R k d  v     & \mbox{} \\
+%    \RTy k T  V  & \mbox{} \\
+%    \RNat k t  v & \mbox{} \\
+%    \RNe k t  m  & \mbox{} \\
+% \end{array}
+% \]
+Even though read back operates on values in practice, formally it is
+simply defined on expressions.
+
+\noindent
+\fbox{$\R k d v$}\ \ Read back reified value $d$.
 \begin{gather*}
@@
 
-\fbox{$\RUp k t m$}\ \  Read back neutrals under annotation, which is ignored.
+\noindent
+\fbox{$\RUp k t m$}\ \  Read back neutrals under annotation.  (The annotation is ignored.)
 \begin{gather*}
@@
 
-\fbox{$\RNat k t v $}
+\noindent
+\fbox{$\RNat k t v $}\ \ Read back natural number value.
 \begin{gather*}
@@
 
-\fbox{$\RSize k {\alpha} a$}
+\noindent
+\fbox{$\RSize k {\alpha} a$}\ \ Read back size value $\alpha$.
 \begin{gather*}
@@
 
-\fbox{$\RNe k n m$} and \fbox{$\RElim k e e^v$}
+\noindent
+\fbox{$\RNe k n m$} and \fbox{$\RElim k e e^v$}\ \ Read back unreflected neutral.
 \begin{gather*}
@@
 \end{gather*}
-\fbox{$\RTy k T V$}
+
+\noindent
+\fbox{$\RTy k T V$} \ \ Read back type value.
 \begin{gather*}

### 2017-02-28 nbe/icfp17.tex
[ ICFP 17 ] Finished pass through the techncial sections.

@@
 \begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
-  \fbox{$n = n' \in \NE$}
+  n = n' \in \NE
     & \defas &
     \RNe k n m &\mand& \RNe k {n'} {m} &\mforall k \\
-  \fbox{$d = d' \in \NF$}
+  d = d' \in \NF
     & \defas &
     \R k d v &\mand& \R k {d'} {v} &\mforall k \\
-  \fbox{$e = e' \in \ELIM$}
+  e = e' \in \ELIM
     & \defas &
     \RElim k e e_v &\mand& \R k {e'} {e_v} &\mforall k \\
-  \fbox{$A = A' \in \TY$}
+  A = A' \in \TY
     & \defas &
     \RTy k A V &\mand& \RTy k {A'} {V} &\mforall k \\
+  % \fbox{$n = n' \in \NE$}
+  %   & \defas &
+  %   \RNe k n m &\mand& \RNe k {n'} {m} &\mforall k \\
+  % \fbox{$d = d' \in \NF$}
+  %   & \defas &
+  %   \R k d v &\mand& \R k {d'} {v} &\mforall k \\
+  % \fbox{$e = e' \in \ELIM$}
+  %   & \defas &
+  %   \RElim k e e_v &\mand& \R k {e'} {e_v} &\mforall k \\
+  % \fbox{$A = A' \in \TY$}
+  %   & \defas &
+  %   \RTy k A V &\mand& \RTy k {A'} {V} &\mforall k \\
 \end{array}
@@
 Now we define some PERs and PER constructors on values.
-All these PERs $\A$ are closed under weak head equality, meaning if $a = b \in \A$ and $a'$ has the same weak head normal form as $a$, then $a' = b in \A$.
+All these PERs $\A$ are closed under weak head equality, meaning if $a = b \in \A$ and $a'$ has the same weak head normal form as $a$, then $a' = b \in \A$.
 (By symmetry, $\A$ is also closed under weak head equality on the second argument.)
 
-Neutrals in the value world $\D$.
+%Neutrals in the value world $\D$.
+PER $\NEE$ interprets all neutral types.
 \[
@@
 \]
-\fbox{$\NAT(\alpha)$}
+
+\fbox{$\NAT(\alpha)$} interprets $\Nat \alpha$ and is defined inductively by the following rules.
 \begin{gather*}
@@
     }{t = t' \in \NAT(\beta)}
-\qquad
+\quad
   \ru{t \evalsto \zero \alpha \qquad
@@
     }{t = t' \in \NAT(\beta+1)}
-\qquad
+\quad
   \ru{t \evalsto \suc \alpha u \qquad
@@
 
-\fbox{$\SIZE$} is a discrete PER of size values:
+\fbox{$\SIZE$} interprets $\Size$ and is a discrete PER of size values:
+\vspace{-1ex}
 \begin{gather*}
@@
 
-Let $\A$ be a PER and $\F$ a family of PERs over $\A$ such that
+Let $\A$ be a PER (including $\A = \SIZE$) and $\F$ a family of PERs over $\A$ such that
 $\F(u) = \F(u')$ whenever $u = u' \in \A$.  We define
@@
 \]
-This works also for $\A = \SIZE$.
+%This works also for $\A = \SIZE$.
+
 For a family $\F$ over $\SIZE$ we also have the irrelevant function space
@@
 \subsection{PER model}
-\label{sec:permode}
+\label{sec:permodel}
+
+Semantic types and their interpretation as PERs are now defined via a
+family of inductive-recursive definitions \cite{dybjer:jsl00}, one for
+each universe level $\ell$.  The construction follows \citet{abelCoquandDybjer:lics07}.
 
@@
 \end{proof}
-Sine $\EL[\ell](T)$ does not depend $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$,  we may simply write $t = t' \in \EL(T)$ or even $t = t' \in T$.
+}
+Since $\EL[\ell](T)$ does not depend $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$,  we may simply write $t = t' \in \EL(T)$ or even $t = t' \in T$.
 
@@
 \end{proof}
-We can define subtyping of type values \fbox{$T \leq T' \in \TYPE$} by induction on $T \in \SET[\ell]$ and $T' \in \SET[\ell']$.
+}
+We define subtyping of type values \fbox{$T \leq T' \in \TYPE$} by induction on $T \in \SET[\ell]$ and $T' \in \SET[\ell']$.
 Simultaneously, we need to prove correctness, namely that $T \leq T' \in \TYPE$ implies $\EL(T) \subseteq \EL(T')$.
@@
 To perform the $\eta$-expansion, not the precise type is needed, just the approximate shape, in particular, whether it is a function type (do expand) or a base type (do not expand).
-For the logical framework, the shape of a dependent type is just its underlying simple type \cite{harperPfenning:equivalenceLF}.
-However, in the presence of universes and large eliminations, there is no underlying simple type.
-Of course, we can take a type as its own shape, but we want at least that $\Nat \alpha$ and $\Nat \beta$ have the same shape even for different $\alpha, \beta$.
-Thus, we define a relation $T \shape S$ to express that $S$ is a possible shape of type $T$.
+For the logical framework, the shape of a dependent type is just its
+underlying simple type \cite{harperPfenning:equivalenceLF}. However,
+in the presence of universes and large eliminations, there is no
+underlying simple type.  Of course, we can take a type as its own
+shape, but we want at least that $\Nat \alpha$ and $\Nat \beta$ have
+the same shape even for different $\alpha, \beta$.
+Also all neutral types can be summarized under a single shape.
+We make our intuition precise by defining a relation $T \shape S$
+between type values, to express that $S$ is a possible shape of type $T$.
 
@@
 
-The eliminations for natural numbers are accurately modeled.
+In this section we show that the eliminations for natural numbers are accurately modeled.
 \begin{lemma}[Case]
@@
 
+In this section we show that the declarative judgements are sound, in
+particular, well-formed syntactic types map to semantic types, and
+definitionally equal terms map to related values in the PER model.
+The proof is standard.  First, we define inductively a
 PER of substitutions
@@
 
-Semantic judgements:
+Then, we define semantic counterparts of our declarative judgements:
 \[
@@
 
-We write $\var\Gamma$ for $\var{|\Gamma|}$.
-We define the identity environment $\rho_\Gamma$ by induction on $\Gamma$.
-We set $\rho_{\cempty} = \sempty$ and
-\[
-\begin{array}{lll}
-  \rho_{\cext \Gamma T} & = & (\rho_\Gamma, \up{T\rho_\Gamma} \var\Gamma) \\
-  \rho_{\eext \Gamma \Size} & = & (\rho_\Gamma, \var\Gamma) \\
-\end{array}
-\]
+From the fundamental theorem, we harvest completeness of NbE in this
+section, \ie, we show that definitionally equal terms have the same
+normal form.
+We may write simply $\Gamma$ for the length $|\Gamma|$ where there is no
+danger of confusion, for instance in de Bruijn level $\var\Gamma$ or
+in read back $\tR_{\Gamma}$.
+%We write $\var\Gamma$ for $\var{|\Gamma|}$.
+We define the \emph{identity environment} \fbox{$\rho_\Gamma$} by induction on $\Gamma$,
+setting $\rho_{\cempty} = \sempty$ and
+$\rho_{\eext \Gamma \Size} = (\rho_\Gamma, \var\Gamma)$ and
+$\rho_{\cext \Gamma T} = (\rho_\Gamma, \up{T\rho_\Gamma} \var\Gamma)$.
+% \[
+% \begin{array}{lll}
+%   \rho_{\cext \Gamma T} & = & (\rho_\Gamma, \up{T\rho_\Gamma} \var\Gamma) \\
+%   \rho_{\eext \Gamma \Size} & = & (\rho_\Gamma, \var\Gamma) \\
+% \end{array}
+% \]
 \begin{lemma}[Identity environment]
@@
 \nextcase
+\vspace{-3ex}
 \[
@@
 
-We can now define the normalization relation
-$\fbox{$\nbe \Gamma T t \evalsto v$} \iff \R {\Gamma} {\down{T \rho_\Gamma} (t \rho_\Gamma)} v$.
+We now define the normalization relation
+$\fbox{$\nbe \Gamma T t \evalsto v$} \defas \R {\Gamma} {\down{T \rho_\Gamma} (t \rho_\Gamma)} v$.
 Whenever $\nbe \Gamma T t \evalsto v$, we may write $\nbe \Gamma T t$ for $v$.
@@
 
-In this section, we show that NbE is sound for judgmental equality, i.e., that same normal form implies definitional equality.
-The proof follows \citet{abelCoquandDybjer:lics07,abel:habil}.
-However, the logical relation defined in the following will also yield a weak head normalization theorem.
+In this section, we show that NbE is sound for judgmental equality,
+i.e., that same normal form implies definitional equality.  The proof
+follows \citet{abelCoquandDybjer:lics07} and
+\citet{fridlenderPagano:tlca13} and defines a Kripke logical relation
+$\LR \Gamma t T f A$ between a well-typed term $\Gamma \der t : T$ and
+a value $f \in A$.
+However, in contrast to the cited works,
+the logical relation defined in the following will also yield a weak
+head normalization theorem.
+
+First, let us define some auxiliary judgements that relate a
+well-formed syntactic object to a value, via read back.
+They will constitute the logical relation for base types,
+yet, are not general enough for function types.
 \begin{alignat*}{8}
@@
 \end{theorem}
-
+\SHORTVERSION{
+\begin{proof}
+Simultaneously by induction on $\LRTS \Gamma T A s$.
+\end{proof}
+} % END SHORT VERSION
+\LONGVERSION{
 \begin{proof*}
@@
 simultaneously by induction on $A' \in s$.
-\SHORTLONG{\qed}{ % LONG VERSION
 \begin{caselist}

### 2017-02-28 nbe/icfp17.tex
[ ICFP 17 ] Spell checker run.
[ ICFP 17 ] Documentation for the Adm judgement
(required cutting some inversion properties for space reasons)

@@
 
-\usepackage[right]{showlabels}\renewcommand{\showlabelfont}{\small\ttfamily\color{gray}}
+%\usepackage[right]{showlabels}\renewcommand{\showlabelfont}{\small\ttfamily\color{gray}}
 
@@
 \]
+\vspace{-4ex}
   \caption{Syntax.}
@@
 figures~\ref{fig:ty} and \ref{fig:eq} the inference rules.
+Judgement $\Gamma \der T : \Adm \ell$ characterizes the valid type
+annotations $T$ in recursion $\fix[\ell] T t$.  The type constructor $T$
+has to be monotone in the size argument; this is a technical condition
+for type-based termination \cite{gimenez:typeBased}.  We will make use
+of it in Section~\ref{sec:comp}.
+
 In the typing judgement $\Gamma \der t : T$, the term $t$ is in scope of
@@
 context where types have been replaced by subtypes.  This intuition
-accounts for the remaining statments but (\ref{it:ressubst}).  The
+accounts for the remaining statements but (\ref{it:ressubst}).  The
 latter is a defining property of substitutions: only replacement for
@@
 \begin{enumerate}
+\LONGVERSION{
 \item If\/ $\Gamma \der \Nat a : T'$ then $\Gamma \der a : \Size$ and $\resurrect\Gamma \der \Set[0] \leq T'$.
@@
 \item If\/ $\Gamma \der \piT U T : T'$ then $\Gamma \der U : s$ and $\cext \Gamma U \der T : s$ and $\resurrect\Gamma \der s \leq T'$ for some $s$.
-\item If\/ $\Gamma \der \epiT \Size T : T'$ then $\cext\Gamma \Size \der T : s$ and $\resurrect\Gamma \der s \leq T'$.
+\item If\/ $\Gamma \der \epiT \Size T : T'$ then
+  $\cext\Gamma \Size \der T : s$ and $\resurrect\Gamma \der s \leq T'$.
+}
 \item If\/ $\Gamma \der \ind i : T'$ then $\Gamma(i) = \erann[\noterased] T$ and $\resurrect\Gamma \der T \leq T'$ for some $T$.
@@
 \item If\/ $\Gamma \der t\,a : T'$ then $\Gamma \der t : \piT \Size T$ and $\Gamma \der a : \Size$ and $\resurrect\Gamma \der T[a] \leq T'$ for some $T$.
-\item If\/ $\Gamma \der t\,\ann a : T'$ then $\Gamma \der t : \forallT T$ and $\resurrect\Gamma \der a : \Size$ and $\resurrect\Gamma \der T[a] \leq T'$ for some $T$.
+\item If\/ $\Gamma \der t\,\ann a : T'$ then
+   $\Gamma \der t : \forallT T$ and
+   $\resurrect\Gamma \der a,b : \Size$ and
+   $\resurrect\Gamma \der T[b] \leq T'$ for some $T$, $b$.
+\LONGVERSION{
 \item If\/ $\Gamma \der \zero a : T'$ then $\resurrect \Gamma \der a,b : \Size$ and $\resurrect\Gamma \der \Nat (b + 1) \leq T'$ for some $b$.
@@
 \item If\/ $\Gamma \der u\,\case[\ell] T {t_z} {t_s} : T'$ then $\Gamma \der u : \Nat (a+1)$ and $\Gamma \der T : \Nat(a+1) \to \Set[\ell]$ and $\Gamma \der t_z : T\,(\zero a)$ and $\Gamma \der t_s : \funT x {\Nat a} {T\,(\suc a x)}$ and $\resurrect\Gamma \der T\,u \leq T'$ for some $a$.
-\item If\/ $\Gamma \der u\,\fix[\ell] T t : T'$ then $\Gamma \der u : \Nat a$ and $\Gamma \der T : \Adm\,\ell$ and $\Gamma \der t : \FixT\,T$ and $\resurrect\Gamma \der T\,a\,u \leq T'$.
+\item If\/ $\Gamma \der u\,\fix[\ell] T t : T'$ then
+  $\Gamma \der u : \Nat a$ and
+  $\Gamma \der T : \Adm\,\ell$ and
+  $\Gamma \der t : \FixT\,T$ and
+  $\resurrect\Gamma \der T\,a\,u \leq T'$.
+} % END LONGVERSION
+\SHORTVERSION{
+\item ... Analogous properties for the remaining term and type constructors.
+}
 \end{enumerate}

### 2017-02-28 nbe/icfp17.tex
some "the"
[ ICFP 17 ] Switched to "all" bibliography (no abbrevs)
[ ICFP 17 ] Updated notation on reflection evaluation

@@
 
-In this section, we give syntax and \emph{declarative} typing, equality, and subtyping judgements.
+In this section, we give the syntax and the \emph{declarative} typing, equality, and subtyping judgements.
 The typing relation $\Gamma \der t : T$ will \emph{not} be decidable; instead, we present algorithmic typing $\Gamma \der t \jchk T$ in Section~\ref{sec:tycheck}.
@@
   \Whnf & \ni & w & ::= & \dots \mid N
-    & \mbox{reflected neutral are weak head normal} \\
+    & \mbox{reflected neutral is weak head normal} \\
   \Exp  & \ni & t & ::= & \dots \mid \down A f
@@
 expressions that use de Bruijn levels instead of de Bruijn
-indices. Comparison for size values $\alpha \leq \beta$ is the
-analogue of the one for size terms $a \leq b$.  In the following, we
-will reuse letter $a$ for a value if it cannot be confused with a size
+indices. Comparison of size values $\alpha \leq \beta$ is
+analoguous to comparison of size terms $a \leq b$.  In the following, we
+will reuse letter $a$ for a value if it cannot be confused for a size
 term.
@@
 evaluation relation \fbox{$t \evalsto w$} which is defined on
-expressions, thus works on values as on terms.  It is defined mutually
+expressions, thus works on values as well as on terms.  It is defined mutually
 with auxiliary relation \fbox{$w @ e \evalsto w'$} stating that weak
@@
 \begin{gather*}
-  \ru{T' \evalsto \piT U T
-    }{(\up {T'} n) @ u \evalsto \up{T[u]} (n \, \down U u) }
-\qquad
-  \ru{T' \evalsto \piT \Size T
-    }{(\up {T'} n) @ \alpha \evalsto \up{T[\alpha]} (n \, \alpha)}
-\qquad
-  \ru{T' \evalsto \forallT T
-    }{(\up {T'} n) @ \ann \alpha \evalsto \up{T[\alpha]} (n \ann \alpha)}
-%   \ru{T \evalsto \piT A B \qquad
-%       \up{B[u]} (n \, \down A u) \evalsto w
-%     }{(\up T n) @ u \evalsto w}
-% \qquad
-%   \ru{T \evalsto \piT \Size B \qquad
-%       \up{B[u]} (n \, \alpha) \evalsto w
-%     }{(\up T n) @ \alpha \evalsto w}
-% \qquad
-%   \ru{T \evalsto \forallT B \qquad
-%       \up{B[\alpha]} (n \ann \alpha) \evalsto w
-%     }{(\up T n) @ \ann \alpha \evalsto w}
+  \ru{A' \evalsto \piT A B
+    }{(\up {A'} n) @ u \evalsto \up{B[u]} (n \, \down A u) }
+\qquad
+  \ru{A \evalsto \piT \Size B
+    }{(\up {A} n) @ \alpha \evalsto \up{B[\alpha]} (n \, \alpha)}
+\qquad
+  \ru{A \evalsto \forallT B
+    }{(\up {A} n) @ \ann \alpha \evalsto \up{B[\alpha]} (n \ann \alpha)}
 \\[2ex]
-  \ru{%T' \evalsto \Nat \beta \qquad
-    }{(\up {T'} n) @ \case[\ell] T {t_z} {t_s} \evalsto
-      \up{T\,(\up{T'} n)}\,
-      % \up{T\,(\up{\Nat \infty} n)}
+  \ru{%A \evalsto \Nat \beta \qquad
+    }{(\up {A} n) @ \case[\ell] B {f_z} {f_s} \evalsto
+      \up{B\,(\up{A} n)}\,
+      % \up{B\,(\up{\Nat \infty} n)}
          n\,\case[\ell]
-             {(\down {\Nat \infty \to \Set[\ell]} T)}
-             {(\down {T\,{\zero\infty}} t_z)}
-             {(\down {\funS x {\Nat \infty} T\,(\suc \infty x)} t_s)}
+             {(\down {\Nat \infty \to \Set[\ell]} B)}
+             {(\down {B\,{\zero\infty}} f_z)}
+             {(\down {\funS x {\Nat \infty} B\,(\suc \infty x)} f_s)}
     }
 \\[2ex]
-  \ru{%T' \evalsto \Nat \beta \qquad
-    }{(\up {T'} n) @ \fix[\ell] T t \evalsto
-      n\,\fix[\ell] {(\down {\FixK} T)} {(\down {\FixT\, T} t)}
+  \ru{%A \evalsto \Nat \beta \qquad
+    }{(\up {A} n) @ \fix[\ell] B f \evalsto
+      n\,\fix[\ell] {(\down {\FixK} B)} {(\down {\FixT\, B} f)}
     }
@@
 
+\clearpage
+
 %% Bibliography

### 2017-02-28 nbe/icfp17.tex
[ ICFP 17 ] Another read of section 4.

@@
 not ensure that the $k$ is large enough to accommodate the de Bruijn
-levels in $d$.  Levels $l \geq k$ which are to big will simply mapped
+levels in $d$.  Levels $l \geq k$ which are to big will simply be mapped
 to de Bruijn index $0$.  The correct $k$ is later ensured by our
@@
 Even though read back operates on values in practice, formally it is
-simply defined on expressions.
+defined on expressions.
 
@@
 A type $T$ will be interpreted as a partial equivalence relation (PER) $\A$ on terms, \ie, a relation which is symmetric and transitive.
-The domain $\dom(\A)$ of the relation can be thought of as the set of terms which define the extension of the type; on $\dom(\A) = \{a \mid \exists a'.\ (a,a') \in \A \}$ the relation $\A$ is in fact an equivalence relation.
+The domain $\dom(\A)$ of the relation can be thought of as the set of terms which denotes the extension of the type; on $\dom(\A) = \{a \mid \exists a'.\ (a,a') \in \A \}$ the relation $\A$ is in fact an equivalence relation.
 We write $a = a' \in \A$ for relatedness in $\A$ and $a \in \A$ if $a \in \dom(\A)$.
@@
   n = n' \in \NE
+  .
 % \fbox{$\NEE$} \defas \{ (\up T n, \up{T'}{n'})
@@
 Also all neutral types can be summarized under a single shape.
+
 We make our intuition precise by defining a relation $T \shape S$
-between type values, to express that $S$ is a possible shape of type $T$.
+between type values, to express that $S$ is a possible shape of type
+$T$.  The asymmetry of this relation stems from the case for function
+types.  At function types
+$\piT A B \shape \piT R S$, we take $S$ to be a family over domain
+$A$, not $R$!  We cannot take $R$ since we have to compare families $B$ and $S$
+at a common domain, and $A$ and $R$ are not equal.
 
@@
 Given $\piT U T \shape \piT R S$ and $\piT R S = \piT {R'}{S'} \in \SET[\ell]$ we would need to show that $T[u] \shape S'[u']$ for all $u = u' \in U$, but we only have $S[u] = S'[u'] \in \SET[\ell]$ for all $u = u' \in R$, thus the induction does not go through.
-The fact that $U \shape R$ does not give us a handle on their inhabitants, we would need $U \leq R \in \TYPE$.
+The fact that $U \shape R$ does not give us a handle on their inhabitants, we would need a stronger relation such as $U \leq R \in \TYPE$.
 It is possible to construct an actual counterexample, using
@@
 By induction hypothesis, $g = g' \in B\,\alpha\,a$.  Since
-$B\,\alpha\,a \leq B\,\infty\,a$ by assumption, the goal
+$B\,\alpha\,a \leq B\,\infty\,a$ by admissibility of $B$, the goal
 $g = g' \in B\,\infty\,a$ follows by subsumption.
@@
 normal form.
-We may write simply $\Gamma$ for the length $|\Gamma|$ where there is no
+We may write simply $\Gamma$ for its length $|\Gamma|$ when there is no
 danger of confusion, for instance in de Bruijn level $\var\Gamma$ or

### 2017-02-28 nbe/icfp17.tex
[ ICFP 17 ] Quick reading of Sec 5

@@
 In this section, we show that NbE is sound for judgmental equality,
-i.e., that same normal form implies definitional equality.  The proof
+i.e., that \emph{same normal form} implies \emph{definitional equality}.  The proof
 follows \citet{abelCoquandDybjer:lics07} and
@@
 a value $f \in A$.
+\LONGVERSION{
 However, in contrast to the cited works,
@@
 head normalization theorem.
+} % END LONG VERSION
 

### 2017-04-20 nbe/rebuttal-icfp17.txt
[ ICFP 17 ] rebuttal

We thank the reviewers for the time spent on careful reading and the
valuable feedback they provide to us.

We will address all points raised by the reviewers when preparing the
final version.  In the following, we comment on the most important
concerns raised by the reviewers.

On sections 6 and 7: They do not offer any new technical ideas over
the state of the art and thus have been temporarily confined to the
long version.  They will appear in the final version where the page
limit is relaxed.

We shall also move some technical stuff (typing rules, proofs) to the
appendix.

> Reviewer A

>  * It is not clear why a relevant use of Size is needed. Couldn't Size
>    be irrelevant and any other type be relevant?

Nat : Size -> Set is an example of a relevant (more precisely,
shape-irrelevant) function type over Size.  If Nat i would be equal to
Nat j for any i,j, there was no point in having sizes in the first place.

sizeSuc : Size -> Size would be an example of relevant, non
shape-irrelevant function type (but we dropped these for simplicity).

> p2l20: what is post, and why do we care?

Postprocessing, that need not preserve the size.  Will explain better.

> p4l34: what is b?

A size expression (unrelated to a).

> p5l23: I couldn't understand what is an inaccessible pattern.

A pattern forced by other pattern matches (due to dependencies).  It
is not matched against at runtime (would be pointless).

> p6l7: what is the notation {! !}?

Inside {! !} there are candidate terms, which need not type-check.

> p7l24: why de Bruijn indices? what's the point of using them in the
> presentation of the theory?

We are studying properties of a dependently-typed core language, de
Bruijn indices are the most popular representation of variables (Coq,
Agda).  We wish to do proofs about a theory close to an implementation.
Where de Bruijn indices are hard to read, e.g. in concrete terms, we
fall back to named variables for the sake of presentation.

> Reviewer B

> however I have strong concerns about the relaxation of the
> size checking being able to be employed too often in such a
> circumstance as to make the size types redundant in circumstances in
> which they would help catch errors.

We are in essence describing Church-style size quantification with a
Curry-style semantics, i.e., we ignore sizes where they act as type
arguments, e.g., in the constructors of Nat.  We do not ignore sizes
where they act as indices in types, e.g., in Nat itself.  Curry-style
is sound in the same way as Church-style is, thus, we do not miss any
type errors.  We use the irrelevant size arguments only in the type
*checker*.  Here, they must be correct.  However, in the semantics,
i.e., the typing and equality rules, they are ignored as just being
hints for the type checker.

> Reviewer C

> - ... Definition of the PER 'NE'... I don't understand why there is no constraint
>   on the types $T$ and $T'$?

NE contains all terms that are stuck.  There is no need to ask for
well-typedness.  It is common in normalization proofs that the
semantic types are over-approximations.  They may contain lots of junk,
as long as normalization is not violated.
In contrast, the logical relation of Section 5 restricts to well-typed
terms.

> Page 4, lines 34-36: should the 't's on the right hand side of the
> conclusions of the first and third judgement here be primed?

Yes. Thanks for spotting this typo.

> Page 7, line 5/6: "pure type system". I thought that pure type systems
> necessarily had untyped equality, but maybe not.

You are right.  We mean "Pure Type System with Judgmental Equality"
[Herbelin, Siles, LICS 2010].

> Pages 10 and 11. Why are some of the rules in boxes?

These mostly involve irrelevant size application.  We will explain.

> Page 16, line 16: Is the denotation of $\forall \mathcal{F}$ the
> intersection?

In some sense, if you ignore the irrelevant size application.

> Page 16, line 43: I assume that $\mathcal{D}$ refers to the derivation
> of the equality, but this isn't explicitly mentioned.

Yes.  We will explain.

> Page 18, line 41: "... by lemma 4.1, and by induction
> hypothesis". Also using Lemma 4.7? because we need $B_1[u_1]
> \sqsubseteq B_1[u_1]$?

Actually, no. We need $B[u_1] \sqsubseteq B_1[u_1]$ which we get by last
premise of the rule.

> Also, on line 44, there is a use of Lemma 4.1(1) when
> applying the induction hypothesis?

Yes, correct.

> Page 19, lines 39-40: Maybe point out that this definition is
> mirroring the derived judgement for 'Adm' on page 8, line 37.

Shall do.

> Page 24, lines 17-23: what is an incremental equality checker?

One that does not do full normalization in one go, but just weak-head
normalizes before recursively comparing the subterms.  Shall explain.

### 2017-04-20 nbe/icfp17.tex
[ ICFP 17 ] removed a comma
[ ICFP 17 ] removed post traversal in intro
[ ICFP 17 ] Spacing before post

@@
 Sized types \cite{hughesParetoSabry:popl96} delegate the checking for structural descent to the type system by annotating data types with a size parameter.
-The type checker can then ensure that in recursive calls, the size goes down, certifying termination.
+The type checker can then ensure that in recursive calls the size goes down, certifying termination.
 In the simplest setting \cite{gimenez:typeBased,abel:lmcs07}, the size is just an upper bound on the tree height of the data structure; however, more sophisticated size annotations have also been considered \cite{xi:terminationHOSC,blanqui:rta04}.
@@
 
-Higher-rank size quantification takes termination checking to the next level; it is necessary for abstraction and generic programming.
-For instance, it allows us to write a generic tree traversal which applies a user-given preprocessor on subtrees before recursively descending into these trees.
+Higher-rank size quantification takes termination checking to the next
+level; it is necessary for abstraction and generic programming.  For
+instance, it allows us to write a generic tree traversal which applies
+a user-given preprocessor on subtrees before recursively descending
+into these trees, and a postprocessor after surfacing from the
+descent.
 The condition is that preprocessing does not increase the size of the subtree; otherwise, termination could not be guaranteed.
@@
 \begin{array}{l}
-  \ttrav : (\vpre : \forall i.\, \tT i \to \tT i) (\vpost : \tT \infty \to \tT \infty) \to \forall i.\, \tT\,i \to \tT\,\infty
+  \ttrav : (\vpre : \forall i.\, \tT i \to \tT i)\; (\vpost : \tT \infty \to \tT \infty) \to \forall i.\, \tT\,i \to \tT\,\infty
   \\

### 2017-05-06 types17/types17.tex
[ types17 ] initial commit

% !TEX encoding = UTF-8 Unicode
\documentclass[a4paper]{easychair}

\usepackage[T1]{fontenc} %
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}

\usepackage{amsmath, amssymb, amsfonts, stmaryrd}
\usepackage{pifont}
\PrerenderUnicode{é} % For the author names in the heading

% Add some colors
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{hyperref}
\hypersetup{
 linktocpage,
 colorlinks,
 citecolor=BlueViolet,
 filecolor=red,
 linkcolor=Blue,
 urlcolor=BrickRed
}

\usepackage{latex/agda}

\usepackage{graphicx}
\usepackage{placeins}

% Meta comment
\newcommand\meta[1]{\textcolor{blue}{\emph{#1}}}

% Include the macro file
\input{macros}

\title{Normalization by Evaluation for Sized Dependent Types}
\titlerunning{NbE for Sized Types}

\author{
  Andreas Abel\inst{1}
  \and
  Andrea Vezzosi\inst{1}
  \and
  Th\'{e}o Winterhalter\inst{2}
}
\authorrunning{Abel and Winterhalter}

\institute{
  Department of Computer Science and Eng., Gothenburg
  University, Sweden \\
  \texttt{\{abela,vezzosi\}@chalmers.se}
  \and
  \' Ecole Normale Supérieure de Cachan, France \\
  \texttt{theo.winterhalter@ens-cachan.fr}
}

\begin{document}

\maketitle

\begin{abstract}
Sized types have been developed to make termination checking more perspicuous, more powerful, and more modular by integrating termination into type checking.  In dependently-typed proof assistants where proofs by induction are just recursive functional programs, the termination checker is an integral component of the trusted core, as validity of proofs depend on termination.  However, a rigorous integration of full-fledged sized types into dependent type theory is lacking so far.  Such an integration is non-trivial, as explicit sizes in proof terms might get in the way of equality checking, making terms appear distinct that should have the same semantics.
%
In this work, we integrate dependent types and sized types with
higher-rank size polymorphism, which is essential for generic
programming and abstraction.  We introduce a size quantifier $\forall$
which lets us ignore sizes in terms for equality checking, alongside
with a second quantifier $\Pi$ for abstracting over sizes that do
affect the semantics of types and terms.  Judgmental equality is
decided by an adaptation of normalization-by-evaluation. % for our new type theory, which features \emph{type shape}-directed reflection and reification.
\end{abstract}

\medskip
%\section{The Need for an Irrelevant Size Quantifier}

Agda \cite{agdawiki} features first class size polymorphism
\cite{abel:fics12} in contrast to theoretical accounts of sized
dependent types
\cite{bartheGregoirePastawski:lpar06,blanqui:rta04,sacchini:lics13}
who typically just have prenex (ML-style) size quantification.
Consequently, Agda's internal language contains size expressions in
terms wherever a size quantifier is instantiated.  However, these
size expressions, which are not unique due to subtyping, can get in
the way of reasoning about sizeful programs.
Consider the type of sized natural numbers.

\input{latex/SubDiagError}

%\section{Relevant and Irrelevant Size Quantification}

The problem is solved by distinguishing relevant ($\Pi$) from irrelevant
($\forall$) size quantification.  The relevant quantifier is the usual
dependent function space over sizes.  In particular, the congruence
rule for size application requires matching size arguments:
\[
  \ru{\Gamma \der t = t' : \Pi i.\, T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a = t'\,a : T[a/i]}
\]
Typically, the relevant quantifier is used in types of types, for
instance, in its non-dependent form, in $\Nat : \aSize \to \Set$.
In contrast, the irrelevant size quantifier is used in types of
programs and ignores sizes in size applications.  The rules for
application, while Church-style, \textit{de facto} implement Curry-style
quantification:
\[
  \ru{\Gamma \der t = t' : \forall i.\, T \qquad
      \resurrect\Gamma \der a,a',b : \Size
    }{\Gamma \der t\,a = t'\,a' : T[b/i]}
\qquad
  \ru{\Gamma \der t : \forall i.\, T \qquad
      \resurrect\Gamma \der a,b : \Size
    }{\Gamma \der t\,a : T[b/i]}
\]
Further, the size arguments are scoped in the \emph{resurrected}
\cite{pfenning:lics01} context
$\resurrect \Gamma$ and, thus, are allowed to mention irrelevant size
variables.  Those are introduced by irrelevant size abstraction and
marked by the $\div$-symbol in the context.
In contrast, the quantified size variable may occur relevantly in the
\emph{type}.
\[
  \ru{\Gamma, i \erof \Size \der t : T
    }{\Gamma \der \lambda i.\, t : \forall i.\, T}
\qquad
  \ru{\Gamma, i \of \Size \der T : \Set
    }{\Gamma \der \forall i.\, T : \Set}
\]

The lack of type unicity arising from the size application rule has
prevented us from adopting the usual incremental algorithm for
equality checking
\cite{harperPfenning:equivalenceLF,abelScherer:types10}.
However, we have succeeded to employ normalization by evaluation
%\cite{bergerSchwichtenberg:lics91,abelCoquandDybjer:lics07}
\cite{abelCoquandDybjer:lics07}
for deciding judgmental equality \cite{abelVezzosiWinterhalter:icfp17}.

\bibliographystyle{plain}
\bibliography{auto-types17} % ../article/biblio,

\end{document}

### 2017-05-08 types17/types17.tex
[ TYPES 17 ] Proof reading.

@@
 \begin{abstract}
-Sized types have been developed to make termination checking more perspicuous, more powerful, and more modular by integrating termination into type checking.
-In dependently-typed proof assistants where proofs by induction are just recursive functional programs, the termination checker is an integral component of the trusted core, as validity of proofs depend on termination.
-However, a rigorous integration of full-fledged sized types into dependent type theory is lacking so far.
-Such an integration is non-trivial, as explicit sizes in proof terms might get in the way of equality checking, making terms appear distinct that should have the same semantics.
+
+Sized types have been developed to make termination checking more
+perspicuous, more powerful, and more modular by integrating
+termination into type checking.  In dependently-typed proof
+assistants where proofs by induction are just recursive functional
+programs, the termination checker is an integral component of the
+trusted core, as validity of proofs depends on termination.
+However, a rigorous integration of full-fledged sized types into
+dependent type theory is lacking so far.  Such an integration is
+non-trivial, as explicit sizes in proof terms might get in the way
+of equality checking, making terms appear distinct that should have
+the same semantics.
 %
@@
 affect the semantics of types and terms.  Judgmental equality is
-decided by an adaptation of normalization-by-evaluation. % for our new type theory, which features \emph{type shape}-directed reflection and reification.
+decided by an adaptation of normalization-by-evaluation.
+% for our new type theory, which features \emph{type shape}-directed reflection and reification.
 \end{abstract}
@@
 
-Agda \cite{agdawiki} features first class size polymorphism
+Agda \cite{agdawiki} features first-class size polymorphism
 \cite{abel:fics12} in contrast to theoretical accounts of sized

### 2017-06-08 nbe/icfp17.tex
[ ICFP 17 ] Fixed Into/Out-of lemma (needs shape).

@@
 weakening, e.g., if\/ $\RD \Gamma t d T$ and $\Gamma \der T \leq T'$ then $\RD \Gamma t d T'$,
-and if\/ $\xi : \Gamma \leq \Gamma'$ then $\RD{\Gamma'}{t\xi}d{T\xi}$.
+and if\/ $\xi : \Gamma' \leq \Gamma$ then $\RD{\Gamma'}{t\xi}d{T\xi}$.
 
@@
 \bla
-Let $\LRTS{\Gamma} T A s$. Then:
+Let $\LRTS{\Gamma} T A s$ and $A \shape S$. Then:
 %Let $A \in s$.
@@
    If\/ $\Gamma \der t \doteq \tRne\,n : T$ then
-   $\LR \Gamma t T {\up A n} A$.
+   $\LR \Gamma t T {\up S n} A$.
 
@@
   If\/ $\LR \Gamma t T f A$ then
-  $\Gamma \der t \doteq \tR\,\down A f : T$.
+  $\Gamma \der t \doteq \tR\,\down S f : T$.
 
@@
 \begin{proof*}
-We prove the propositions for $A'$ and $T'$ (instead of $A$ and $T$)
+We prove the propositions for $A'$, $S'$, and $T'$ (instead of $A$, $S$, and $T$)
 simultaneously by induction on $A' \in s$.
@@
 
-\nextcase $A' \evalsto \piT A B$ and $\resurrect\Gamma \der T' = \piT U T$.
+\nextcase $A' \evalsto \piT A B$ and $S' \evalsto \piT R S$
+and $\resurrect\Gamma \der T' = \piT U T$.
+
 \begin{enumerate}
@@
 The premise is, after type conversion, $\Gamma \der t \doteq \tRne\, n : \piT U T$.
-To demonstrate $\LR \Gamma t {T'} {\up {A'} n} {A'}$ we assume
+To demonstrate $\LR \Gamma t {T'} {\up {S'} n} {A'}$ we assume
 $\xi : \Gamma' \leq \Gamma$ and $\LR{\Gamma'}{u}{U\xi} a A$ and show
-$\LR{\Gamma'} {(t\xi\,u)} {T(\xi,u)} {\up{B[a]}(n\,\down A a)} {B[a]}$.
-By induction hypothesis (\ref{it:out}) we have $\RD {\Gamma'} u {\down A a} {U\xi}$, and
+$\LR{\Gamma'} {(t\xi\,u)} {T(\xi,u)} {\up{S[a]}(n\,\down A a)} {B[a]}$.
+By induction hypothesis (\ref{it:out}) we have $\RD {\Gamma'} u {\down R a} {U\xi}$, and
 together with the weakened premise $\Gamma' \der t\xi \doteq \tRne\, n : (\piT U T) \xi$ we get
-$\Gamma' \der t\xi\, u \doteq \tRne\, (n\,\down A a) : T(\xi,u)$.
+$\Gamma' \der t\xi\, u \doteq \tRne\, (n\,\down R a) : T(\xi,u)$.
 The goal follows by induction hypothesis (\ref{it:into}) for
@@
 \end{enumerate}
+
+\nextcase $A' \evalsto \forallT A$ and $S' \evalsto \forallT S$ and
+$\resurrect\Gamma \der T' = \forallT T$.
+
+\begin{enumerate}
+\item We assume $\xi : \Gamma' \leq \Gamma$ and
+$\resurrect{\Gamma'} \der a, b : \Size$ and $\alpha,\beta \in \SIZE$ with
+$\LRSize {\resurrect{\Gamma'}} b \beta$ and show
+$\LR {\Gamma'} {t\xi\ann a} {T(\xi,b)} {(\up {S'} n) \ann \alpha} {A[\beta]}$.
+By the evaluation rules for reflection, it is sufficient to show
+% It is sufficient to show
+$\LR {\Gamma'} {t\xi\ann a} {T(\xi,b)} {\up {S[\alpha]} (n \ann \alpha)} {A[\beta]}$.
+This follows by induction hypothesis (\ref{it:into}) if
+$\LRTS {\Gamma'} {T(\xi,b)} {A[\beta]} s$ and $A[\beta] \shape S[\alpha]$ and
+$\RNE {\Gamma'} {t\xi\ann a} {n \ann \alpha} {T(\xi,b)}$.
+The first two of these subgoals are immediate.  For the third, assume
+$\xi' : \Gamma'' \leq \Gamma'$.
+We have to show that there exists a neutral normal form $m'$ such that
+$\RNe {\Gamma''} {n \ann \alpha} m'$ and
+$\Gamma'' \der (t\xi \ann a)\xi' = m' : (T(\xi,b))\xi'$.
+Sizes can always be read back, thus, let $\RSize {\Gamma''} \alpha {a_0}$.
+By assmption $\RNE \Gamma t n {T'}$, there is an $m$ with
+$\RNe {\Gamma''} n m$ and
+$\Gamma'' \der t\xi\xi' = m : T'\xi\xi'$.
+The latter equation implies
+$\Gamma'' \der t\xi\xi' \ann {a\xi'} = m \ann {a_0} : T(\xi\xi',b\xi')$
+by irrelevant size application to
+$\resurrect{\Gamma''} \der a\xi', a_0, b \xi' : \Size$,
+thus we are done by choosing $m' : = m \ann{a_0}$.
+
+\end{enumerate}
+
 \end{caselist}
@@
 \end{document}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "icfp17-long"
+%%% End:

### 2017-06-08 nbe/icfp17.tex
[ ICFP 17 ] Factored out proof part into proof of Lemma 5.1(3)

@@
 \item If\/ $\RNE \Gamma t n \piT \Size T$ and $\RSIZE \Gamma a \alpha$ then $\RNE \Gamma {t\,a} {n\,\alpha} {T[a]}$.
-\item If\/ $\RNE \Gamma t n \forallT T$ and $\RSIZE {\resurrect\Gamma} a \alpha$ then $\RNE \Gamma {t\,\ann a} {n\,\ann \alpha} {T[a]}$.
+\item \label{it:closneall}
+If\/ $\RNE \Gamma t n \forallT T$ and $\resurrect\Gamma \der a,b : \Size$ and $\alpha \in \SIZE$ then $\RNE \Gamma {t\,\ann a} {n\,\ann \alpha} {T[b]}$.
+%% TOO WEAK:
+% \item If\/ $\RNE \Gamma t n \forallT T$ and $\RSIZE {\resurrect\Gamma} a \alpha$ then $\RNE \Gamma {t\,\ann a} {n\,\ann \alpha} {T[a]}$.
 \end{enumerate}
 \end{lemma}
+\LONGVERSION{
+\begin{proof*}
+For (\ref{it:closneall}), assume
+$\xi : \Gamma' \leq \Gamma$.
+We have to show that there exists a neutral normal form $m'$ such that
+$\RNe {\Gamma'} {n\, \ann \alpha} m'$ and
+$\Gamma' \der (t\, \ann a)\xi = m' : T[b]\xi$.
+Sizes can always be read back, thus, let $\RSize {\Gamma'} \alpha {a_0}$.
+By assumption $\RNE \Gamma t n {\forallT T}$, there is an $m$ with
+$\RNe {\Gamma'} n m$ and
+$\Gamma' \der t\xi = m : (\forallT T)\xi$.
+The latter equation implies
+$\Gamma' \der t\xi \ann {a\xi} = m \ann {a_0} : T(\xi,b\xi)$
+by irrelevant size application to
+$\resurrect{\Gamma'} \der a\xi, a_0, b \xi : \Size$,
+thus we are done with $m' := m \ann{a_0}$.
+\qed
+\end{proof*}
+} % END LONGVERSION
 
@@
 % assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
-\qed
 \end{enumerate}
@@
 $\RNE {\Gamma'} {t\xi\ann a} {n \ann \alpha} {T(\xi,b)}$.
-The first two of these subgoals are immediate.  For the third, assume
-$\xi' : \Gamma'' \leq \Gamma'$.
-We have to show that there exists a neutral normal form $m'$ such that
-$\RNe {\Gamma''} {n \ann \alpha} m'$ and
-$\Gamma'' \der (t\xi \ann a)\xi' = m' : (T(\xi,b))\xi'$.
-Sizes can always be read back, thus, let $\RSize {\Gamma''} \alpha {a_0}$.
-By assmption $\RNE \Gamma t n {T'}$, there is an $m$ with
-$\RNe {\Gamma''} n m$ and
-$\Gamma'' \der t\xi\xi' = m : T'\xi\xi'$.
-The latter equation implies
-$\Gamma'' \der t\xi\xi' \ann {a\xi'} = m \ann {a_0} : T(\xi\xi',b\xi')$
-by irrelevant size application to
-$\resurrect{\Gamma''} \der a\xi', a_0, b \xi' : \Size$,
-thus we are done by choosing $m' : = m \ann{a_0}$.
-
-\end{enumerate}
-
+The first two of these subgoals are immediate.
+The third follows by Lemma~\ref{lem:rclosne} (\ref{it:closneall}) from the
+weakened assumption
+ $\RNE {\Gamma'} {t\xi} n {T'\xi}$.
+% For the third, assume
+% $\xi' : \Gamma'' \leq \Gamma'$.
+% We have to show that there exists a neutral normal form $m'$ such that
+% $\RNe {\Gamma''} {n \ann \alpha} m'$ and
+% $\Gamma'' \der (t\xi \ann a)\xi' = m' : (T(\xi,b))\xi'$.
+% Sizes can always be read back, thus, let $\RSize {\Gamma''} \alpha {a_0}$.
+% By assmption $\RNE \Gamma t n {T'}$, there is an $m$ with
+% $\RNe {\Gamma''} n m$ and
+% $\Gamma'' \der t\xi\xi' = m : T'\xi\xi'$.
+% The latter equation implies
+% $\Gamma'' \der t\xi\xi' \ann {a\xi'} = m \ann {a_0} : T(\xi\xi',b\xi')$
+% by irrelevant size application to
+% $\resurrect{\Gamma''} \der a\xi', a_0, b \xi' : \Size$,
+% thus we are done by choosing $m' : = m \ann{a_0}$.
+%
+\qed
+\end{enumerate}
+%
 \end{caselist}

### 2017-06-08 nbe/icfp17.tex
[ ICFP 17 ] Factored out lemma "fresh variable readback"

@@
 and if\/ $\xi : \Gamma' \leq \Gamma$ then $\RD{\Gamma'}{t\xi}d{T\xi}$.
-
+\begin{lemma}[Fresh variable readback]
+  \label{lem:freshread}
+  If\/ $\Gamma \der U$ then $\RNE{\cext \Gamma U} {\ind 0} {\var \Gamma} {U \slift}$.
+\end{lemma}
+\begin{proof}
+  Assume $\xi : \Gamma' \leq \cext \Gamma U$.
+  Note that $\RNe{\Gamma'} {\var\Gamma} {\ind i}$ where
+  $i = |\Gamma'| - |\cext \Gamma U|$ is the length of the context extension.
+  Since $\ind 0 \xi = \ind i$, we conclude
+  $\Gamma' \der \ind 0 \xi = \ind i : U \slift \xi$
+  by weakening of the judgement $\cext \Gamma U \der \ind 0 = \ind 0 : U \slift$.
+\end{proof}
 \begin{lemma}[Closure properties for neutrals]
@@
 $\Gamma' \der (t\, \ann a)\xi = m' : T[b]\xi$.
-Sizes can always be read back, thus, let $\RSize {\Gamma'} \alpha {a_0}$.
+Sizes can always be read back, thus, let $\RSize {\Gamma'} \alpha {a_0}$,
+which guarantees $\resurrect{\Gamma'} \der a_0 : \Size$.
 By assumption $\RNE \Gamma t n {\forallT T}$, there is an $m$ with
@@
 \begin{enumerate}
-\item
+\item % into
 The premise is, after type conversion, $\Gamma \der t \doteq \tRne\, n : \piT U T$.
@@
 \begin{enumerate}
-\item We assume $\xi : \Gamma' \leq \Gamma$ and
+\item % into
+We assume $\xi : \Gamma' \leq \Gamma$ and
 $\resurrect{\Gamma'} \der a, b : \Size$ and $\alpha,\beta \in \SIZE$ with
@@
 } % END LONGVERSION
-
-\LONGVERSION{
-\begin{corollary}[One-to-one]
-\label{cor:oneone}
-\bla
-\begin{enumerate}
-\item
-If\/ $\LRSize \Gamma a \alpha$ and $\LRSize \Gamma {a'} \alpha$ then $a = a'$.
-\item
-If\/ $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} A {\ell'}$ then $\Gamma \der T = T'$.
-\end{enumerate}
-\end{corollary}
-} % END LONGVERSION
-\begin{corollary}[Fresh variable]
+\begin{corollary}[Fresh variable injection]
   \label{cor:fresh}
@@
 \begin{proof}
-  Assume $\xi : \Gamma' \leq \cext \Gamma T$.
-  Note that for the length
-  $i = |\Gamma'| - |\cext \Gamma T|$ of the context extension
-  we have $\ind i = \ind 0 \xi$, thus
-  $\Gamma' \der \ind 0 \xi = \ind i : T \slift \xi$.
-  Since $\RNe{\Gamma'} {\var\Gamma} {\ind i}$,
-  the goal follows now from Thm.~\ref{thm:into} part (\ref{it:into}).
+  Since $\RNE {\cext \Gamma T} {\ind 0} {\var \Gamma} {T \slift}$
+  by Lemma~\ref{lem:freshread},
+  the goal follows from Thm.~\ref{thm:into} part (\ref{it:into}).
+  % Assume $\xi : \Gamma' \leq \cext \Gamma T$.
+  % Note that for the length
+  % $i = |\Gamma'| - |\cext \Gamma T|$ of the context extension
+  % we have $\ind i = \ind 0 \xi$, thus
+  % $\Gamma' \der \ind 0 \xi = \ind i : T \slift \xi$.
+  % Since $\RNe{\Gamma'} {\var\Gamma} {\ind i}$,
+  % the goal follows now from Thm.~\ref{thm:into} part (\ref{it:into}).
 \end{proof}
+
+\LONGVERSION{
+\begin{corollary}[One-to-one]
+\label{cor:oneone}
+\bla
+\begin{enumerate}
+\item
+If\/ $\LRSize \Gamma a \alpha$ and $\LRSize \Gamma {a'} \alpha$ then $a = a'$.
+\item
+If\/ $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} A {\ell'}$ then $\Gamma \der T = T'$.
+\end{enumerate}
+\end{corollary}
+} % END LONGVERSION
 

### 2017-06-09 nbe/icfp17.tex
[ ICFP 17 ] "out of LR" proofs

@@
 % assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
+
+\item % out of
+The premise is $\LR \Gamma t {T'} f {A'}$,
+which means that for all $\xi : \Gamma' \leq \Gamma$ and
+$\LR {\Gamma'} u {U\xi} a A$ we have
+$\LR {\Gamma'} {t\xi\,u} {T(\xi,u)} {f\,a} {B[a]}$.
+
+To show $\Gamma \der t \doteq \tR\,\down {S'} f : {T'}$
+we assume $\xi : \Gamma' \leq \Gamma$ and produce a normal form $v$
+such that $\R {\Gamma'} {\down{S'} f} {\lambda v}$ and
+$\Gamma' \der t\xi = \lambda v : T'\xi$.
+Induction hypothesis (\ref{it:into}) on
+$\RNE {\cext{\Gamma'}{U\xi}} {\ind 0} {\var{\Gamma'}} {U\xi\slift}$
+gives us $\LR {\cext{\Gamma'}{U\xi}} {\ind 0} {U\xi\slift} a A$
+with $a := \up R {\var{\Gamma'}}$.
+Thus, we can instantiate the assumption $\LR \Gamma t {T'} f {A'}$
+to obtain
+$\LR {\cext{\Gamma'}{U\xi}} {(t\xi\slift)\,\ind 0} {T(\xi\slift,\ind 0)} {f\,a} {B[a]}$.
+Now induction hypothesis (\ref{it:out}) yields a normal form $v$ such that
+$\R {\cext{\Gamma'}{U\xi}} {\down {S[a]} (f\,a)} v$
+and
+$\cext{\Gamma'}{U\xi} \der (t\xi\slift)\,\ind 0 = v : T(\xi\slift,\ind 0)$.
+Since $a$ is the semantic version of the last bound variable,
+$\R {\Gamma'} {\down{S'} f} {\lambda v}$ follows by definition of reification.
+For the final goal, we $\lambda$-abstract the definitional equality to
+$\Gamma' \der \lambda.\, (t\xi\slift)\,\ind 0 = \lambda v : \piT{(U\xi)} {(T(\xi\slift,\ind 0))}$
+and with $\eta$-equality and the substitution laws we get
+$\Gamma' \der t\xi = \lambda v : (\piT U T)\xi$.
 \end{enumerate}
@@
 \begin{enumerate}
+
 \item % into
@@
 The third follows by Lemma~\ref{lem:rclosne} (\ref{it:closneall}) from the
-weakened assumption
- $\RNE {\Gamma'} {t\xi} n {T'\xi}$.
+weakened assumption $\RNE {\Gamma'} {t\xi} n {T'\xi}$.
 % For the third, assume
@@
 %
+
+\item % out of
+To show $\RD \Gamma t {\down{S'} f} {T'}$ we assume $\xi : \Gamma' \leq \Gamma$ and
+produce a normal form $v$ such that $\R{\Gamma'} {\down{S'}f} {\lambda v}$ and
+$\Gamma' \der t\xi = \lambda v : T'\xi$.
+It is sufficient to show
+$\R{\erext{\Gamma'}\Size} {\down{S[\var{\Gamma'}]}(f \ann{\var{\Gamma'}})} v$
+and
+$\erext{\Gamma'}\Size \der (t\xi\slift)\,\ind 0 = v : T(\xi\slift,\ind 0)$.
+These goals, in turn, follow by induction hypothesis (\ref{it:out}) on
+$\LR{\erext{\Gamma'}\Size}{(t\xi\slift)\,\ind 0}{T(\xi\slift,\ind 0)}{f \ann{\var{\Gamma'}}}{A[\var{\Gamma'}]}$ which is an instance of our assumption
+$\LR \Gamma t {T'} f {A'}$.
 \qed

### 2017-06-09 nbe/icfp17.tex
[ ICFP 17 ] Started polishing final version.

@@
 %% For double-blind review submission
-\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
+\documentclass[acmlarge%,review,anonymous
+]{acmart}\settopmatter{printfolios=true}
 %% For single-blind review submission
@@
 %\copyrightyear{2017}           %% If different from \acmYear
+
+% \acmBadgeR[http://icfp17.sigplan.org/track/icfp-2017-Artifacts]{artifact_evaluated-functional.png}
+\acmBadgeR[http://icfp17.sigplan.org/track/icfp-2017-Artifacts]{artifact_evaluated-reusable.png}
+%\acmBadgeL[http://icfp17.sigplan.org/track/icfp-2017-Artifacts]{artifact_available.png}
+\acmBadgeL[https://hackage.haskell.org/package/Sit]{artifact_available.png}
 
@@
 
-Dependently-typed programming languages and proof assistants, such as Agda~\citeyearpar{agdawiki}, Coq~\cite{inria:coq86}, and Idris~\cite{brady:jfp13} require programs to be total, for two reasons.
+Dependently-typed programming languages and proof assistants, such as Agda~\citeyearpar{agdawiki} and Coq~\cite{inria:coq86}, require programs to be total, for two reasons.
 First, for consistency:  since propositions are just types and proofs of a proposition just programs which inhabit the corresponding type, some types need to be empty; otherwise, each proposition would be true.
@@
 % \cite{boveKraussSozeau:mscs16}
-Consequently, each of the aforementioned languages based on Type Theory comes with a termination checker, which needs to reject all non-terminating programs, and should accept sufficiently many terminating programs to allow the user to express her algorithms.
+Consequently, the aforementioned languages based on Type Theory come with a termination checker, which needs to reject all non-terminating programs, and should accept sufficiently many terminating programs to allow the user to express her algorithms.
 In current termination checkers, programs are required to terminate by structural descent \cite{gimenez:guardeddefinitions}; the structural order may be extended to a lexicographic \cite{abelAltenkirch:jfp02} or size-change termination criterion \cite{jones:sizeChange,wahlstedt:PhD}.
@@
 \]
-The display shows the \emph{Curry}-style program as provided by the user, but state-of-the-art type checkers elaborate the program from surface syntax into an internal \emph{Church}-style syntax with explicit type abstractions and type applications.\footnote{Agda, Coq, Idris, and Haskell \cite{sulzmannChakravartyPeythonJonesDonnelly:tldi07} all have Church-style internal languages.}  With implicit type and size applications elaborated, $\ttrav$ would look as follows:
+The display shows the \emph{Curry}-style program as provided by the user, but state-of-the-art type checkers elaborate the program from surface syntax into an internal \emph{Church}-style syntax with explicit type abstractions and type applications.\footnote{Agda, Coq, Idris~\cite{brady:jfp13}, and Haskell~\cite{sulzmannChakravartyPeythonJonesDonnelly:tldi07} all have Church-style internal languages.}  With implicit type and size applications elaborated, $\ttrav$ would look as follows:
 \[
@@
 determines $i = j = k$, we are back to safety.  However, allowing
-types to differ during an equality check can be hazardous, especially
+types to differ during an equality check needs special consideration, especially
 when the equality-check is type directed.
@@
 
-Going beyond the classical incremental equality check, there is a technique that can deal with dynamic $\eta$-expansion.
+Going beyond the standard syntax-directed equality check, there is a technique that can deal with dynamic $\eta$-expansion.
 It is a type-directed normalization function inspired by normalization-by-evaluation (NbE) that computes $\eta$-long normal forms \cite{bergerSchwichtenberg:lics91,danvy:tdpe}.
@@
 
-We present our type theory as \emph{pure type system} \cite{barendregt:lambdacube} with extra structure.
-\emph{Sorts} $s$ are drawn from an infinite predicative hierarchy of universes $\Set[\ell]$ for $\ell \in \NN$.
+We present our type theory as (domain-free) \emph{pure type system} \cite{barendregt:lambdacube} with extra structure.
+The \emph{sorts} $s$ are drawn from an infinite predicative hierarchy of universes $\Set[\ell]$ for $\ell \in \NN$.
 Universes provide us with polymorphism and the capability to define types by recursion on values.
@@
 
-For the expression syntax (see Fig.~\ref{fig:syntax}), we use de~Bruijn \citeyearpar{deBruijn:nameless} indices $\ind i$ to represent variables.
-Binders are lambda abstraction $\lambda t$ and dependent function types $\epiT U T$, which bind the $0$th index in $t$ and $T$, resp.
-The notation $\epiT U T$ is an umbrella for three kinds of function types, where $\evar \in \{ \erased, \noterased \}$ is a relevance annotation borrowed from \citet{pfenning:intextirr}: $\erpiT[\noterased] U T$, the ordinary dependent function type, $\erpiT[\noterased] \Size T$ for relevant size quantification, and $\erpiT \Size T$ for irrelevant size quantification.
-We omit the ``$\noterased$''-markers from $\Gamma$ and $\Pi$ by default and write $\forallT T$ for $\erpiT \Size T$.
+For the expression syntax (see Fig.~\ref{fig:syntax}), we use
+de~Bruijn \citeyearpar{deBruijn:nameless} indices $\ind i$ to
+represent variables.  The index $i \in \NN$ points to the $i$th
+enclosing binder of variable $\ind i$.  Binders are lambda abstraction
+$\lambda t$ and dependent function types $\epiT U T$, which bind the
+$0$th index in $t$ and $T$, resp.  For instance, the term
+$\lambda x.\,x\,(\lambda z.\,z)\, (\lambda y.\,x\,y)$
+with named variables $x,y,z$ has de Bruijn representation
+$\lambda.\,\ind 0\,(\lambda.\, \ind 0)\,(\lambda.\,\ind 1\,\ind0)$.
+
+The notation $\epiT U T$ is an
+umbrella for three kinds of function types, where
+$\evar \in \{ \erased, \noterased \}$ is a relevance annotation
+borrowed from \citet{pfenning:intextirr}. $\erpiT[\noterased] U T$ is
+the ordinary dependent function type, $\erpiT[\noterased] \Size T$ is
+relevant size quantification, and $\erpiT \Size T$ is irrelevant size
+quantification.  We omit the ``$\noterased$''-markers
+from $\Pi$ by default (and also in contexts $\Gamma$)
+and write $\forallT T$ for $\erpiT \Size T$.
 %
@@
 
-Besides with regular application $t\,u$, relevant size application $t\,a$, and irrelevant size application $t \, \ann a$,
-we can eliminate a term $t$ by a case distinction $e = \case[\ell] T {t_z}{t_s}$ or a recursive function application $e' = \fix[\ell'] {T'} {t'}$.
-In both cases, the term should denote a natural number.
-Application of case distinction $\zero a \, e$ will reduce to the zero-branch $t_z$, and $\suc a t \,e$ to the instantiation of the successor branch $t_s\,t$.
-The type annotation $T$ in $\tcase$ allows us to infer the type of the whole case statement $t\,e$ as $T\,t$.
-The function call $c\,e'$ for a canonical number reduces to $f'\,(\lambda x.\,x\,e')\,c$ where we allowed ourselves the use of a named abstraction in the presentation to the reader.
-The unfolding of fixed-points is thus restricted to application to canonical numbers; this is the usual reduction strategy which converges for terminating functions \cite{gimenez:typeBased}.
+Besides with regular application $t\,u$, relevant size application
+$t\,a$, and irrelevant size application $t \, \ann a$, we can
+eliminate a term $t$ by a case distinction
+$e = \case[\ell] T {t_z}{t_s}$ or a recursive function application
+$e' = \fix[\ell'] {T'} {t'}$.  In both cases, the term should denote a
+natural number.  Application of case distinction $\zero a \, e$ will
+reduce to the zero-branch $t_z$, and $(\suc a t) \,e$ to the
+instantiation of the successor branch $t_s\,t$.  The type annotation
+$T$ in $\tcase$ allows us to infer the type of the whole case
+statement $t\,e$ as $T\,t$.  The function call $c\,e'$ for a canonical
+number reduces to $f'\,(\lambda x.\,x\,e')\,c$ where we allowed
+ourselves the use of a named abstraction in the presentation to the
+reader.  The unfolding of fixed-points is thus restricted to
+application to canonical numbers; this is the usual reduction strategy
+which converges for terminating functions \cite{gimenez:typeBased}.
 
@@
 
-\clearpage
+% \clearpage
 

### 2017-06-09 nbe/icfp17.tex
[ ICFP 17 ] Some polishing in Section 3.

@@
 Judgmental equality features a ``Curry-style'' rule for irrelevant size application which ignores the size arguments, and consequently, the corresponding typing rule will also ignore the size argument.
+(In the following rules, $a$, $a'$, and $b$ stand for arbitrary size expressions.)
 \[
@@
 %
-Examples for relevant size quantification $\piT \Size T$, in syntax with named variables, are $\Size \to \Set[0]$ or $\funT i \Size \Nat i \to \Set[0]$.
-An instance of irrelevant quantification $\forall T$ would be $\forall i.\, \Nat i \to \Nat i$.
-Herein, $\Nat i$ denotes the type of natural numbers below $i$.
-The expression $\Size$ is a possible instance of $U$ in $\epiT U T$, or a possible type of a variable in a typing context $\Gamma$, but not a first-class type, \ie, we cannot construct our own function on sizes.
+Examples for relevant size quantification $\piT \Size T$ are
+$\piT \Size {\Set[0]}$ and
+$\piT \Size {\piT {(\Nat{\ind 0})} {\Set[0]}}$.
+In a syntax with named variables and non-dependent function type
+they could be written as $\Size \to \Set[0]$ and
+$\funT z \Size \Nat z \to \Set[0]$, resp.
+%
+An instance of irrelevant
+quantification $\forall T$ would be
+$\forall.\, \piT {(\Nat{\ind 0})} {(\Nat{\ind 1})}$
+which is
+$\forall z.\, \Nat z \to \Nat z$
+in a named syntax.
+Herein, $\Nat z$ denotes the type of natural numbers below $z$. The
+expression $\Size$ is a possible instance of $U$ in $\epiT U T$, or a
+possible type of a variable in a typing context $\Gamma$, but not a
+first-class type, \ie, we cannot construct our own functions on sizes.
 
 Canonical natural numbers $c$ are constructed by $\zero a$ and $\suc a t$.
-Size expression $a$ is either a constant $o \in \NN$, a variable
+A size expression $a$ is either a constant $o \in \NN$, a variable
 $\ind i + o$ possibly with increment $o$, or the limit ordinal
 $\infty$ which stands for $\omega$.  The size argument $a$ in the
-constructors is a suggestion for the type checker but bears no
+constructors $\tzero$ and $\tsuc$
+is a suggestion for the type checker but bears no
 semantic significance.  For example, in the declarative typing
@@
 In the algorithmic typing however, $\der \zero 5 \jchk \Nat 1$ will be an error.
-
-Besides with regular application $t\,u$, relevant size application
-$t\,a$, and irrelevant size application $t \, \ann a$, we can
-eliminate a term $t$ by a case distinction
-$e = \case[\ell] T {t_z}{t_s}$ or a recursive function application
-$e' = \fix[\ell'] {T'} {t'}$.  In both cases, the term should denote a
-natural number.  Application of case distinction $\zero a \, e$ will
-reduce to the zero-branch $t_z$, and $(\suc a t) \,e$ to the
-instantiation of the successor branch $t_s\,t$.  The type annotation
+Note, however, that $\der \zero a : \Nat 0$ is impossible for any $a$,
+as $\tzero$ is not strictly below $0$
+(when both term and size are interpreted as natural numbers).
+
+Regular application $t\,u$, relevant size application
+$t\,a$, and irrelevant size application $t \, \ann a$ eliminate
+functions $t$ and are subsumed under the form $t\,e$ with $e ::= u \mid a \mid \ann a$.
+We have two further eliminations, which make sense when $t$ stands for a natural number.
+These are case distinction
+$e = \case[\ell] T {t_z}{t_s}$ and recursive function application
+$e' = \fix[\ell'] {T'} {t'}$.
+Application of case distinction $\zero a \, e$ will
+reduce to the zero-branch $t_z$, and application $(\suc a t) \,e$ to the
+instantiation $t_s\,t$ of the successor branch.  The type annotation
 $T$ in $\tcase$ allows us to infer the type of the whole case
 statement $t\,e$ as $T\,t$.  The function call $c\,e'$ for a canonical
-number reduces to $f'\,(\lambda x.\,x\,e')\,c$ where we allowed
+number $c$ and elimination $e' = \fix[\ell'] {T'} {t'}$
+reduces to $t'\,(\lambda x.\,x\,e')\,c$ where we allowed
 ourselves the use of a named abstraction in the presentation to the
@@
 %We write $\sigma$ and $\tau$ for substitutions, which are lists of terms, and
-For ordinary $\beta$-reduction we employ
-\emph{substitutions} $\sigma$.
-These are simply lists of terms that provide one term as replacement for each free de Bruijn index in a term $t$.
-We write \fbox{$t \sigma$} for the application of substitution $\sigma$ to term $t$ which is defined as usual.
-Let \emph{lifting} \fbox{$\slift^k_m$} be the substitution $(\ind{k+m-1},\dots,\ind{k+1},\ind k)$ which accepts a term with $m$ free indices and increases each of them by $k$.
-We write $\slift_m$ for the lifting $\slift^1_m$ and \fbox{$\sid_m$} for the identity substitution $ \slift^0_m$.
-In general, we refer to liftings by letter $\xi$.
-The substitution \fbox{$[u]_m$}${} = (\sid_m, u)$ replaces free index $\ind 0$ by term $u$ and decrements the other $m$ free indices by $1$.
-We drop subscript $m$ from liftings when clear from the context.
-Substitution composition \fbox{$\sigma\tau$} is the pointwise application of substitution $\tau$ to the list of terms $\sigma$.  In the proofs to follow, we freely use the following identities:
+For ordinary $\beta$-reduction we employ \emph{substitutions}
+$\sigma$.  These are simply lists of terms that provide one term as
+replacement for each free de Bruijn index in a term $t$. We write
+\fbox{$t \sigma$} for the application of substitution $\sigma$ to term
+$t$ which is defined as usual.  Let \emph{lifting} \fbox{$\slift^k_m$}
+be the substitution $(\ind{k+m-1},\dots,\ind{k+1},\ind k)$ which
+accepts a term with $m$ free indices and increases each of them by
+$k$.  We write $\slift_m$ for the lifting $\slift^1_m$ and
+\fbox{$\sid_m$} for the identity substitution $ \slift^0_m$. In
+general, we refer to liftings by letter $\xi$.
+%
+The substitution \fbox{$[u]_m$}${} = (\sid_m, u)$ replaces free index
+$\ind 0$ by term $u$ and decrements the other $m$ free indices by
+$1$. We drop subscript $m$ from liftings and substitutions when clear from the context.
+%
+Substitution composition \fbox{$\sigma\tau$} is the pointwise
+application of substitution $\tau$ to the list of terms $\sigma$.  In
+the proofs to follow, we freely use the following identities:
 \begin{gather*}
-  t\sid \equiv t
-\quad
+  t\,\sid \equiv t
+\qquad
   (t\sigma)\tau \equiv t(\sigma\tau)
-\qquad
-  \sigma \sid \equiv \sigma
-\quad
-  \sid \tau \equiv \tau
-\quad
+\qquad\quad
+  \sigma\, \sid \equiv \sigma
+\qquad
+  \sid\, \tau \equiv \tau
+\qquad
   (\rho\sigma)\tau \equiv \rho(\sigma\tau)
@@
   \ind 0 (\sigma,t) \equiv t
-\quad
+\qquad
   \slift(\sigma,t) \equiv \sigma
-\qquad
+\qquad\quad
   [t]\sigma \equiv (\sigma, t\sigma)
-\quad
+\qquad
   \slift[t] \equiv \sid
@@
 
-For concrete function types and terms, we may use a named dependent function type notation as syntactic sugar for the corresponding de Bruijn representation.
-For instance, $\funT i \Size \Nat i \to \Set[\ell]$ is sugar for $\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.
-We abbreviate this type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for
-$\forall i.\ (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$.
-% $\allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$.
-Similar as for $\Pi$, we use named lambda abstraction as sugar for de Bruijn abstraction.
-Named abstraction takes care of proper lifting of de Bruijn indices, for instance, $\lambda x.\, t x = \lambda.
-(t \slift)\,\ind 0$ if $t$ is outside the scope of $x$.
-We may also use names when we construct concrete contexts, for instance, if\/ $T$ is well-formed in context $\Gamma$, we may write $T\,i\,x$ in context $\Gamma.i\of\Size.x\of\Nat i$ to mean $T\slift^2\,\ind 1\,\ind 0$ in context
-$\Gamma.\Size.\Nat \ind 0$.
+As already done in some examples, we may use a named dependent
+function type notation as syntactic sugar for the corresponding de
+Bruijn representation.  For instance,
+$\funT z \Size \Nat z \to \Set[\ell]$ is sugar for
+$\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.  We abbreviate this
+type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for
+$\forall z.\ (\sfunT x {\Nat z} T\, z\, x) \to \sfunT x {\Nat (z + 1)} T \,(z+1)\, x$.
+%
+Similar as for $\Pi$, we use named lambda abstraction as sugar for de
+Bruijn abstraction.  Named abstraction takes care of proper lifting of
+de Bruijn indices, for instance,
+$\lambda x.\, t x = \lambda. (t \slift)\,\ind 0$ if $t$ is outside the
+scope of $x$.
+%
+We may also use names when we construct concrete contexts, for
+instance, if\/ $T$ is well-formed in context $\Gamma$, we may write
+$T\,z\,x$ in context $\Gamma.z\of\Size.x\of\Nat z$ to mean
+$T\slift^2\,\ind 1\,\ind 0$ in context $\Gamma.\Size.\Nat \ind 0$.
 

### 2017-06-09 nbe/icfp17.tex
[ ICFP 17 ] Lines in figures and size variable z

@@
+\nonstopmode
 %% For double-blind review submission
@@
 \centering
+\hrulefill
 \[
@@
 \]
-\vspace{-4ex}
+\hrulefill
+\vspace{-2ex}
   \caption{Syntax.}
@@
 understand this distinction, consider judgement
-$i \div \Size \der \Nat i$.  This would mean that $i$ is irrelevant in
-$\Nat i$ and thus, $\Gamma \der \Nat a = \Nat {a'}$ for all sizes
+$z \div \Size \der \Nat z$.  This would mean that $z$ is irrelevant in
+$\Nat z$ and thus, $\Gamma \der \Nat a = \Nat {a'}$ for all sizes
 $\resurrect\Gamma \der a,a' : \Size$.  But this is exactly wrong!  However,
-judgement $i \div \Size \der \zero i : \Nat (i + 1)$ is fine, it
+judgement $z \div \Size \der \zero z : \Nat (z + 1)$ is fine, it
 implies $\Gamma \der \zero a = \zero {a'} : \Nat (b+1)$ for all
@@
 things right of the colon.  In the last example we have applied the
-substitution triple $\Gamma \der [a] = [a'] \eeq [b] : (i \div \Size)$
-to judgement $i \div \Size \der \zero i : \Nat (i + 1)$.
+substitution triple $\Gamma \der [a] = [a'] \eeq [b] : (z \div \Size)$
+to judgement $z \div \Size \der \zero z : \Nat (z + 1)$.
 The first two substitutions apply to the term side while the third
 substitution applies to the type side.  The fact that we replace an
-irrelevant variable $i$ allows $a,a',b$ to refer to irrelevant
+irrelevant variable $z$ allows $a,a',b$ to refer to irrelevant
 variables from $\Gamma$, thus, they are in scope of $\resurrect\Gamma$.
@@
 \begin{figure}
+\hrulefill
 $$
@@
 $$
+\hrulefill
 \caption{Type-assignment NbE in locally nameless

### 2017-06-10 nbe/icfp17.tex
[ ICFP 17 ] Added annotation erasure operation.

@@
 figures~\ref{fig:ty} and \ref{fig:eq} the inference rules.
+We have \fbox{boxed} the rules dealing with irrelevant size application.
 Judgement $\Gamma \der T : \Adm \ell$ characterizes the valid type
@@
 
+Typing requires from annotations $\ann a$ in a term only that they are well-scoped size expressions, \ie, just mention relevant size variables.
+Let \fbox{$t^\infty$} denote the \emph{erasure} of term $t$, meaning that we replace all annotations $\ann a$ in $t$ by $\ann \infty$.
+Let \fbox{$t \approx u$} relate terms that only differ in their annotations, \ie, $t \approx u \defiff t^\infty = u^\infty$.
+Erasure does not change the term modulo judgmental equality:
+\begin{lemma}[Erasure and similarity]
+\label{lem:erase}
+\bla
+\begin{enumerate}
+\item If\/ $\Gamma \der t : T$ then $\Gamma \der t = t^\infty : T$.
+\item If\/ $\Gamma \der t, u : T$ and $t \approx u$ then $\Gamma \der t = u : T$.
+\end{enumerate}
+\end{lemma}
+
 %% COULD BE A LEMMA INSTEAD OF A RULE
@@
 
+\noindent
 In the following, we list syntactic properties of our judgements.  To
@@
 \begin{enumerate}
-\item \label{it:reswf} $\der \Gamma$ iff $\der \resurrect\Gamma$;  then $\resurrect \Gamma \der \sid : \Gamma$.
+\item \label{it:reswf} $\der \Gamma$ iff $\der \resurrect\Gamma$.
+  Then $\resurrect \Gamma \der \sid : \Gamma$, which can be written
+  $\sid : \resurrect\Gamma \leq \Gamma$.
   % Consequently, if\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
@@
 relation (PER) on term-side substitutions $\sigma,\sigma'$. Note that
-we cannot resurrect this judgement to
-$\sueq{\resurrect\Gamma}{\sigma}{\sigma'}\tau \Delta$ in general!
-The type-side substitution $\tau$ can be converted to one that is
-equal in and for resurrected contexts.
+usually we cannot resurrect this judgement to
+$\sueq{\resurrect\Gamma}{\sigma}{\sigma'}\tau {\resurrect\Delta}$.
+For instance,
+$\sueq{z_1 {\div} \Size.\,z_2 {\div} \Size} {\sg{z_1}} {\sg{z_2}} {\sg{\infty}} {z {\div} \Size}$
+holds but
+$\sueq{z_1 {:} \Size.\,z_2 {:} \Size} {\sg{z_1}} {\sg{z_2}} {\sg{\infty}} {z {:} \Size}$
+clearly not.
+% The type-side substitution $\tau$ can be converted to one that is
+% equal in and for resurrected contexts.
 \begin{lemma}[Substitution equality]

### 2017-06-10 nbe/icfp17.tex
[ ICFP 17 ] No longer erase to infty during readback.

@@
 \qquad
-  \ru{
-    }{\RElim k {\ann \alpha} {\ann \infty}}
-  % \ru{\RSize k \alpha b
-  %   }{\RElim k {\ann \alpha} {\ann b}}
+  % \ru{
+  %   }{\RElim k {\ann \alpha} {\ann \infty}}
+  \ru{\RSize k \alpha b
+    }{\RElim k {\ann \alpha} {\ann b}}
 \\[2ex]
@@
 
-The PERs $\NE$ characterizes neutral reifiable terms.
+%The PERs $\NE$ characterizes neutral reifiable terms.
 % \[
@@
 \[
-\begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
-  n = n' \in \NE
-    & \defas &
-    \RNe k n m &\mand& \RNe k {n'} {m} &\mforall k \\
-  d = d' \in \NF
-    & \defas &
-    \R k d v &\mand& \R k {d'} {v} &\mforall k \\
-  e = e' \in \ELIM
-    & \defas &
-    \RElim k e e_v &\mand& \R k {e'} {e_v} &\mforall k \\
-  A = A' \in \TY
-    & \defas &
-    \RTy k A V &\mand& \RTy k {A'} {V} &\mforall k \\
-  % \fbox{$n = n' \in \NE$}
+% \begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
+%   n = n' \in \NE
 %     & \defas &
 %     \RNe k n m &\mand& \RNe k {n'} {m} &\mforall k \\
-  % \fbox{$d = d' \in \NF$}
+%   d = d' \in \NF
 %     & \defas &
 %     \R k d v &\mand& \R k {d'} {v} &\mforall k \\
-  % \fbox{$e = e' \in \ELIM$}
+%   e = e' \in \ELIM
 %     & \defas &
 %     \RElim k e e_v &\mand& \R k {e'} {e_v} &\mforall k \\
-  % \fbox{$A = A' \in \TY$}
+%   A = A' \in \TY
 %     & \defas &
 %     \RTy k A V &\mand& \RTy k {A'} {V} &\mforall k \\
+% \end{array}
+%\newcommand{\myfbox}[1]{\begin{array}[c]{@{}l@{}}\framebox(60,12)[l]{~\,#1}\end{array}}
+\newcommand{\myfbox}[1]{\framebox(60,12)[l]{~\,#1}}
+\begin{array}{l@{\quad}l@{\quad}l@{~}l@{\quad}l@{~}l@{~\;}r@{~}l@{~}l@{~}l}
+  \raisebox{-0.5ex}{\myfbox{$n = n' \in \NE$}}
+    & \defas &
+    \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m &\nfeq& m' &\mforall k
+\\[1ex]
+  \raisebox{-0.8ex}{\myfbox{$d = d' \in \NF$}}
+    & \defas &
+    \R k d v &\mand& \R k {d'} {v'} &\mand& v &\nfeq& v' &\mforall k
+\\[1ex]
+  \raisebox{-0.6ex}{\myfbox{$e = e' \in \ELIM$}}
+    & \defas &
+    \RElim k e e_v &\mand& \R k {e'} {e_v'} &\mand& e_v &\nfeq& e_v' &\mforall k
+\\[1ex]
+  \raisebox{-0.8ex}{\myfbox{$A = A' \in \TY$}}
+    & \defas &
+    \RTy k A V &\mand& \RTy k {A'} {V'} &\mand& V &\nfeq& V' &\mforall k \\
 \end{array}
-% \begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
-%   \fbox{$n = n' \in \NE$}
-%     & \defas &
-%     \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m \nfeq m' &\mforall k \\
-%   \fbox{$d = d' \in \NF$}
-%     & \defas &
-%     \R k d v &\mand& \R k {d'} {v'} &\mand& v \nfeq v' &\mforall k \\
-%   \fbox{$e = e' \in \ELIM$}
-%     & \defas &
-%     \RElim k e e_v &\mand& \R k {e'} {e_v'} &\mand& e_v \nfeq e_v' &\mforall k \\
-%   \fbox{$A = A' \in \TY$}
-%     & \defas &
-%     \RTy k A V &\mand& \RTy k {A'} {V'} &\mand& V \nfeq V' &\mforall k \\
-% \end{array}
   % \ru{\RNe k t m \mand \RNe k {t'} {m'} \mand m \nfeq m' \mforall k
@@
 \item We assume $k \in \NN$ and show $\R k {\down{S_i}{t_i}} \lambda v_i$ for some normal forms
-$v_1,v_2$.  Let $u_i = \up{A_i} \lev k$.  Note that $u_1 = u_2 \in A$ by induction hypothesis.
+$v_1 \nfeq v_2$.  Let $u_i = \up{A_i} \lev k$.  Note that $u_1 = u_2 \in A$ by induction hypothesis.
 It is sufficient to show
@@
   $t_1\,\ann{\lev k} = t_2\,\ann{\lev k} \in B[\lev k]$.  Thus, by induction hypothesis,
-  $\R{k+1}{\down{B_i[\lev k]}(t_i\,\ann{\lev k})}{v_i}$, and finally
+  $\R{k+1}{\down{B_i[\lev k]}(t_i\,\ann{\lev k})}{v_i}$ with $v_1 \nfeq v_2$, and finally
   $\R k {\down{S_i} t_i} {\lambda v_i}$ by definition of read back.
@@
 \begin{theorem}[Completeness of NbE]
-  If\/ $\Gamma \der t = t' : T$ then there is a normal form $v$ such that
-  $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v$.
-  % If\/ $\Gamma \der t = t' : T$ then there are normal forms $v \nfeq v'$ such that
-  % $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v'$.
+  % If\/ $\Gamma \der t = t' : T$ then there is a normal form $v$ such that
+  % $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v$.
+  If\/ $\Gamma \der t = t' : T$ then there are normal forms $v \nfeq v'$ such that
+  $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v'$.
 \end{theorem}
@@
   \item
-  If\/ $\Gamma \der t,t' : T$ and $\nbe \Gamma T t \equiv \nbe \Gamma T {t'}$ then
+  If\/ $\Gamma \der t,t' : T$ and $\nbe \Gamma T t \nfeq \nbe \Gamma T {t'}$ then
   $\Gamma \der t = t' : T$.
@@
 
-\item By transitivity of judgmental equality from (\ref{it:sound}).
-   Here we use that the computed normal forms are identical because
-   we erase all size annotations $\ann\alpha$ to $\ann\infty$ during read back.
+\item By transitivity of judgmental equality from (\ref{it:sound}), using
+  Lemma~\ref{lem:erase}:
+  $\Gamma \der t = \nbe \Gamma T t = \nbe \Gamma T {t'} = t' : T$.
+   % Here we use that the computed normal forms are identical because
+   % we erase all size annotations $\ann\alpha$ to $\ann\infty$ during read back.
 \qed
@@
 \begin{corollary}[Decidability of judgemental equality]
-  If\/ $\Gamma \der t,t' : T$ then the test whether $\nbe\Gamma T t \equiv \nbe\Gamma T{t'}$
+  If\/ $\Gamma \der t,t' : T$ then the test whether $\nbe\Gamma T t \nfeq \nbe\Gamma T{t'}$
   terminates and
@@
 
-Reification erases all irrelevant size arguments to $\infty$.
+% Reification erases all irrelevant size arguments to $\infty$.
 We have also considered to keep the sizes there, and define equality $v \approx v'$ of normal forms up to differences in erased arguments.
@@
 Proving this statement seems to be the key to an incremental equality check; but
-the usual bidirectional argument \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types.
+% the usual bidirectional argument \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types.
 In the long run, we wish for an incremental type-directed equality check, which is also at the heart of Agda's type checker, as this check can generate constraints for meta variables involved in type reconstruction \cite{norell:PhD}.
+However, the usual bidirectional construction \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types.
 

### 2017-06-10 nbe/sec-asub.tex
[ ICFP 17 ] No longer erase to infty during readback.

@@
       T' \evalsto n' \qquad
-      \Nbe \Gamma n \equiv \Nbe \Gamma n'
+      \Nbe \Gamma n \nfeq \Nbe \Gamma n'
     }{\Gamma \der T \asub T'}
@@
 \nextcase $A_i' \evalsto N_i$ and $T_i' \evalsto n_i$ and $N_1 = N_2 \in \NE$ and $\RTY \Gamma {n_i} {N_i} {s_i}$.
-By Cor.~\ref{cor:oneone} $\Gamma \der n_1 = n_2$, thus, by completeness of NbE, $\Nbe\Gamma n_1 \equiv \Nbe\Gamma n_2$, which entails $\Gamma \der T_1' \asub T_2'$.
+By Cor.~\ref{cor:oneone} $\Gamma \der n_1 = n_2$, thus, by completeness of NbE, $\Nbe\Gamma n_1 \nfeq \Nbe\Gamma n_2$, which entails $\Gamma \der T_1' \asub T_2'$.
 

### 2017-06-10 nbe/sec-sr.tex
[ ICFP 17 ] No longer erase to infty during readback.

@@
 via the relation $\Nbe \Gamma T \evalsto V \defas \RTy \Gamma {T\rhog} V$.
-Analogously, we show that the test $\Nbe \Gamma T \equiv \Nbe \Gamma T'$
+Analogously, we show that the test $\Nbe \Gamma T \nfeq \Nbe \Gamma T'$
 decides type equality $\Gamma \der T = T'$.
@@
 By reification (Thm.~\ref{thm:rere}),
-$\RTy\Gamma {U\rhog} V$ and $\RTy\Gamma {U'\rhog} V$ for some normal form
-$V$.
+$\RTy\Gamma {U\rhog} V$ and $\RTy\Gamma {U'\rhog} V'$ for some normal forms
+$V \nfeq V'$.
 Since by inversion we have $\Gamma \der U,U' : s$ we get $\Gamma \der U = U' : s$ by soundness of NbE.

### 2017-06-10 nbe/icfp17.tex
[ ICFP 17 ] Some polishing in Section 4.

@@
 NbE is complete for definitional equality, \ie, if
-$\Gamma \der t = t' : T$, then $t$ and $t'$ have the same normal form.
+$\Gamma \der t = t' : T$, then $t$ and $t'$ have the same normal form
+up to annotations.
 \[
@@
 $\Level \blueto \DNe \blueto \D$: Levels embed via constructor $\var{}$
-into semantic neutrals $\DNe$ which are labeled with their type $A \in \D$ to become and element $\up A \lev j \in \D$.
+into semantic neutrals $\DNe$ which are labeled with their type $A \in \D$
+to become an element $\up A \lev j \in \D$.
 
@@
 expressions, thus works on values as well as on terms.  It is defined mutually
-with auxiliary relation \fbox{$w @ e \evalsto w'$} stating that weak
+with auxiliary relation \fbox{$w \at e \evalsto w'$} stating that weak
 head normal form $w$ is eliminated by $e$ into weak head normal form $w'$.
@@
 %   t \evalsto w & \mbox{term $t$ evaluates to weak head normal form $w$} \\
-%   w @ e \evalsto w' & \mbox{eliminating $w$ with $e$ evaluates to $w'$} \\
-% %  w @ \vec e \evalsto w' & \mbox{applying $w$ to eliminations $\vec e$ evaluates to $w'$} \\
+%   w \at e \evalsto w' & \mbox{eliminating $w$ with $e$ evaluates to $w'$} \\
+% %  w \at \vec e \evalsto w' & \mbox{applying $w$ to eliminations $\vec e$ evaluates to $w'$} \\
 % \end{array}
@@
 
+\noindent
 \fbox{$t \evalsto w$}
-\vspace{-4ex}
+\ and \
+\fbox{$w \at e \evalsto w'$}
+\vspace{-6ex}
 \begin{gather*}
@@
 \qquad
-  \ru{t \evalsto w \qquad w @ e \evalsto w'}{t\,e \evalsto w'}
-\end{gather*}
-
-% \fbox{$w @ \vec e \evalsto w'$}
+  \ru{t \evalsto w \qquad w \at e \evalsto w'}{t\,e \evalsto w'}
+% \end{gather*}
+% \fbox{$w \at e \evalsto w'$}
 % \begin{gather*}
-%   \ru{}{w @ () \evalsto w}
-% \qquad
-%   \ru{w @ e \evalsto w' \qquad
-%       w' @ \vec e \evalsto w''
-%     }{w @ (e, \vec e) \evalsto w''}
-% \end{gather*}
-
-\fbox{$w @ e \evalsto w'$}
-\begin{gather*}
+\\[1.5ex]
   \ru{t[u] \evalsto w
-    }{(\lambda t) @ u \evalsto w}
+    }{(\lambda t) \at u \evalsto w}
 \qquad
   \ru{t[\alpha] \evalsto w
-    }{(\lambda t) @ \alpha \evalsto w}
+    }{(\lambda t) \at \alpha \evalsto w}
 \qquad
   \ru{t[\alpha] \evalsto w
-    }{(\lambda t) @ \ann \alpha \evalsto w}
+    }{(\lambda t) \at \ann \alpha \evalsto w}
 \qquad
   \ru{t_z \evalsto w
-    }{(\zero \alpha) @ \casett \evalsto w}
-\\[2ex]
+    }{(\zero \alpha) \at \casett \evalsto w}
+\\[1.5ex]
   \ru{t_s\, t \evalsto w
-    }{(\suc \alpha t) @ \casett \evalsto w}
-\qquad
-  \ru{t\,\ann \alpha\,(\lambda x.\, x\, \fixtt)\,c \evalsto w
-     }{c @ \fixtt \evalsto w
+    }{(\suc \alpha t) \at \casett \evalsto w}
+\qquad
+  \rux{t\,\ann \alpha\,(\lambda x.\, x\, \fixtt)\,c \evalsto w
+     }{c \at \fixtt \evalsto w
      }{c\in \{ \zero \alpha, \suc \alpha u \}}
@@
   \ru{A' \evalsto \piT A B
-    }{(\up {A'} n) @ u \evalsto \up{B[u]} (n \, \down A u) }
+    }{(\up {A'} n) \at u \evalsto \up{B[u]} (n \, \down A u) }
 \qquad
   \ru{A \evalsto \piT \Size B
-    }{(\up {A} n) @ \alpha \evalsto \up{B[\alpha]} (n \, \alpha)}
+    }{(\up {A} n) \at \alpha \evalsto \up{B[\alpha]} (n \, \alpha)}
 \qquad
   \ru{A \evalsto \forallT B
-    }{(\up {A} n) @ \ann \alpha \evalsto \up{B[\alpha]} (n \ann \alpha)}
-\\[2ex]
+    }{(\up {A} n) \at \ann \alpha \evalsto \up{B[\alpha]} (n \ann \alpha)}
+\\[1.2ex]
   \ru{%A \evalsto \Nat \beta \qquad
-    }{(\up {A} n) @ \case[\ell] B {f_z} {f_s} \evalsto
+    }{(\up {A} n) \at \case[\ell] B {f_z} {f_s} \evalsto
       \up{B\,(\up{A} n)}\,
@@
     }
-\\[2ex]
+\\[1ex]
   \ru{%A \evalsto \Nat \beta \qquad
-    }{(\up {A} n) @ \fix[\ell] B f \evalsto
+    }{(\up {A} n) \at \fix[\ell] B f \evalsto
       n\,\fix[\ell] {(\down {\FixK} B)} {(\down {\FixT\, B} f)}
@@
 \fbox{$\RSize k {\alpha} a$}\ \ Read back size value $\alpha$.
+\vspace{-1.5ex}
 \begin{gather*}
@@
 
-A type $T$ will be interpreted as a partial equivalence relation (PER) $\A$ on terms, \ie, a relation which is symmetric and transitive.
-The domain $\dom(\A)$ of the relation can be thought of as the set of terms which denotes the extension of the type; on $\dom(\A) = \{a \mid \exists a'.\ (a,a') \in \A \}$ the relation $\A$ is in fact an equivalence relation.
-We write $a = a' \in \A$ for relatedness in $\A$ and $a \in \A$ if $a \in \dom(\A)$.
-
-%The PERs $\NE$ characterizes neutral reifiable terms.
-% \[
-% \begin{array}{lll}%{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
-%   \fbox{$\up T n = \up {T'} {n'} \in \NE$}
-%     & \defas &
-%     T \evalsto N \mand T' \evalsto N' \mand
-%     \mforall k \mboth
-%     \RNe k n m \mand \RNe k {n'} {m'} \mand m \nfeq m'
-%     .\\
-% \end{array}
-% \]
-The PERs $\NE$ and $\NF$ characterize (neutral) normalizing terms.
+A type $T$ will be interpreted as a partial equivalence relation (PER)
+$\A$ on terms, \ie, a relation which is symmetric and transitive. The
+domain $\dom(\A)$ of the relation can be thought of as the set of
+terms which denotes the extension of the type; on
+$\dom(\A) = \{a \mid \exists a'.\ (a,a') \in \A \}$ the relation $\A$
+is in fact an equivalence relation.  We write $a = a' \in \A$ for
+relatedness in $\A$ and $a \in \A$ if $a \in \dom(\A)$.
+
+The PERs $\NE$ and $\NF$ characterize (neutral) normalizing values.
+For instance, two values $n$ and $n'$ are related in $\NE$ if at any $k \in \NN$
+they can be read back to neutral normal forms $m$ and $m'$
+which are identical up to annotations.
 \[
@@
 %     & \defas &
-%     \RElim k e e_v &\mand& \R k {e'} {e_v} &\mforall k \\
+%     \RElim k e e_v &\mand& \RElim k {e'} {e_v} &\mforall k \\
 %   A = A' \in \TY
@@
 %\newcommand{\myfbox}[1]{\begin{array}[c]{@{}l@{}}\framebox(60,12)[l]{~\,#1}\end{array}}
-\newcommand{\myfbox}[1]{\framebox(60,12)[l]{~\,#1}}
-\begin{array}{l@{\quad}l@{\quad}l@{~}l@{\quad}l@{~}l@{~\;}r@{~}l@{~}l@{~}l}
+\newcommand{\myfbox}[1]{\framebox(60,13)[l]{~\,#1}}
+\begin{array}{l@{\quad}l@{\quad}l@{~}l@{\,}l@{\quad}l@{\,}l@{~}l@{~\;}r@{~}l@{~}l@{~}l}
   \raisebox{-0.5ex}{\myfbox{$n = n' \in \NE$}}
     & \defas &
-    \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m &\nfeq& m' &\mforall k
+    \tRne _k & n \evalsto m &\mand& \tRne _k & {n'} \evalsto {m'} &\mand& m &\nfeq& m' &\mforall k
 \\[1ex]
@@
     & \defas &
-    \R k d v &\mand& \R k {d'} {v'} &\mand& v &\nfeq& v' &\mforall k
+    \tR _k & d \evalsto v &\mand& \tR _k & {d'} \evalsto {v'} &\mand& v &\nfeq& v' &\mforall k
 \\[1ex]
@@
     & \defas &
-    \RElim k e e_v &\mand& \R k {e'} {e_v'} &\mand& e_v &\nfeq& e_v' &\mforall k
+    \tRelim _k & e \evalsto e_v &\mand& \tRelim _k & {e'} \evalsto {e_v'} &\mand& e_v &\nfeq& e_v' &\mforall k
 \\[1ex]
@@
     & \defas &
-    \RTy k A V &\mand& \RTy k {A'} {V'} &\mand& V &\nfeq& V' &\mforall k \\
+    \tRty _k & A \evalsto V &\mand& \tRty _k & {A'} \evalsto {V'} &\mand& V &\nfeq& V' &\mforall k \\
 \end{array}
-  % \ru{\RNe k t m \mand \RNe k {t'} {m'} \mand m \nfeq m' \mforall k
-  %   }{t = t' \in \NE}
-  % \ru{t \evalsto \up N n \qquad t' \evalsto \up{N'}{n'} \qquad n \nfeq n'
-  %   }{t = t' \in \NE}
+% \begin{array}{l@{\quad}l@{\quad}l@{~}l@{\quad}l@{~}l@{~\;}r@{~}l@{~}l@{~}l}
+%   \raisebox{-0.5ex}{\myfbox{$n = n' \in \NE$}}
+%     & \defas &
+%     \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m &\nfeq& m' &\mforall k
+% \\[1ex]
+%   \raisebox{-0.8ex}{\myfbox{$d = d' \in \NF$}}
+%     & \defas &
+%     \R k d v &\mand& \R k {d'} {v'} &\mand& v &\nfeq& v' &\mforall k
+% \\[1ex]
+%   \raisebox{-0.6ex}{\myfbox{$e = e' \in \ELIM$}}
+%     & \defas &
+%     \RElim k e e_v &\mand& \RElim k {e'} {e_v'} &\mand& e_v &\nfeq& e_v' &\mforall k
+% \\[1ex]
+%   \raisebox{-0.8ex}{\myfbox{$A = A' \in \TY$}}
+%     & \defas &
+%     \RTy k A V &\mand& \RTy k {A'} {V'} &\mand& V &\nfeq& V' &\mforall k \\
+% \end{array}
 \]
-%All PERs $\A$ we consider are sandwiched $\NE \subseteq \A \subseteq \NF$ between $\NE$ and $\NF$.
+Once we have established useful closure properties of these PERs, they
+abstract most of the reasoning about the read-back relation from our
+proofs.  This idea is due to Coquand \cite{abelCoquandPagano:tlca09}.
 
@@
 together with the extension $\EL[\ell] T$ (for $T = T' \in \Set[\ell]$)
-which is a PER of values of type $A$.
+which is a PER of values of type $T$.
 The rules for \fbox{$T = T' \in \SET[\ell]$} are:
@@
 }
-Since $\EL[\ell](T)$ does not depend $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$,  we may simply write $t = t' \in \EL(T)$ or even $t = t' \in T$.
+Since $\EL[\ell](T)$ does not depend on $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$,  we may simply write $t = t' \in \EL(T)$ or even $t = t' \in T$.
 
@@
 
-Reflection and reification perform $\eta$-expansion such that we arrive at an $\eta$-long $\beta$-normal form.
+Reflection and reification perform $\eta$-expansion so that we arrive at an $\eta$-long $\beta$-normal form.
 To perform the $\eta$-expansion, not the precise type is needed, just the approximate shape, in particular, whether it is a function type (do expand) or a base type (do not expand).
@@
 
-However, templates are not closed under subtyping in either direction because of contravariance.
+However, templates are not closed under subtyping in either direction because subtyping is contravariant for function type domains but the shape relation is covariant.
 
@@
 %$\bigcap_{
+\noindent
 Let us write \fbox{$B = B' \in \Adm\ell$} iff $B = B' \in \FixK\,\ell$ and
@@
 %% Bibliography
-\bibliography{auto-icfp17}
+\bibliography{auto-icfp17-long}
 

### 2017-06-10 nbe/icfp17.tex
[ ICFP 17 ] DOIs for all references (where available)

@@
 In the simplest setting \cite{gimenez:typeBased,abel:lmcs07}, the size is just an upper bound on the tree height of the data structure; however, more sophisticated size annotations have also been considered \cite{xi:terminationHOSC,blanqui:rta04}.
-Most sized type systems are non-dependent \cite{amadio:guardcondition,abelPientka:jfp16,bartheGregoireRiba:lernet08,bartheGregoireRiba:csl08,blanquiRiba:lpar06,dalLagoGrellois:probabilisticSizedTypes2017}, yet the combination of sized and dependent types has been studied as well \cite{blanqui:csl05,bartheGregoirePastawski:lpar06,gregoireSacchini:lpar10,sacchini:lics13,sacchini:flops14}.
+Most sized type systems are non-dependent \cite{amadio:guardcondition,abelPientka:jfp16,bartheGregoireRiba:lernet08,bartheGregoireRiba:csl08,blanquiRiba:lpar06,dalLagoGrellois:esop17}, yet the combination of sized and dependent types has been studied as well \cite{blanqui:csl05,bartheGregoirePastawski:lpar06,gregoireSacchini:lpar10,sacchini:lics13,sacchini:flops14}.
 However, to the best of our knowledge, no study combines \emph{higher-rank size polymorphism} with full-fledged dependent types.\footnote{\citet{xi:terminationHOSC} has first-class size polymorphism, but only indexed types, no universes or large eliminations.}
@@
 Whenever we form a irrelevant function type $\forall x \of U.\, T[x]$, we require $T[x]$ to be shape-irrelevant in $x$.
-This is the middle ground between ICC$^*$, where no restriction is placed on $T$ but $\eta$ for unit types is out of reach (at least for the moment), and Pfenning's \citeyearpar{pfenning:intextirr} irrelevance modality, adapted to full dependent types by \citet{abelScherer:types10}, which requires $T$ to be irrelevant in $x$ and, thus, has type equality $T[i] = T[j]$.
+This is the middle ground between ICC$^*$, where no restriction is placed on $T$ but $\eta$ for unit types is out of reach (at least for the moment), and Pfenning's \citeyearpar{pfenning:lics01} irrelevance modality, adapted to full dependent types by \citet{abelScherer:types10}, which requires $T$ to be irrelevant in $x$ and, thus, has type equality $T[i] = T[j]$.
 
@@
 $\evar \in \{ \erased, \noterased \}$ is a relevance annotation
-borrowed from \citet{pfenning:intextirr}. $\erpiT[\noterased] U T$ is
+borrowed from \citet{pfenning:lics01}. $\erpiT[\noterased] U T$ is
 the ordinary dependent function type, $\erpiT[\noterased] \Size T$ is

### 2017-06-10 nbe/icfp17.tex
[ ICFP 17 ] Missing resurrects and missing principal types.

@@
 \]
+Our substitution theorem distinguishes term- from type-side substitutions.
 
@@
 \noindent
+We should remark here that we have \emph{neither type unicity nor principal types}
+due to the irrelevant size application rule.
 In the following, we list syntactic properties of our judgements.  To
@@
 \noindent
-Let us write \fbox{$B = B' \in \Adm\ell$} iff $B = B' \in \FixK\,\ell$ and
+As the semantic counterpart of judgement $\Gamma \der T : \Adm \ell$,
+let us write \fbox{$B = B' \in \Adm\ell$} iff $B = B' \in \FixK\,\ell$ and
 for all $\beta \in \SIZE$ and $a \in \Nat \beta$ we have
@@
 definitionally equal terms map to related values in the PER model.
-The proof is standard.  First, we define inductively a
+The proof runs the usual course.  First, we define inductively a
 PER of substitutions
-\fbox{$\eta = \eta' \eeq \rho \in \Gamma$}.
+\fbox{$\eta = \eta' \eeq \rho \in \Gamma$}\,.
 \begin{gather*}
@@
 
+\noindent
 Then, we define semantic counterparts of our declarative judgements:
@@
   \models \eext \Gamma \Size & \defas & \models \Gamma \\
+  \models \cext \Gamma s     & \defas & \models \Gamma \\
   \models \cext \Gamma T     & \defas & \models \Gamma \mand \resurrect\Gamma \models T
@@
 \\[2ex]
+  \Gamma \models T : \Adm\,\ell & \defas & \Gamma \models T : \FixK\,\ell \mand
+    T\eta = T'\eta' \in \Adm\,\ell
+    % \EL(T\eta) \subseteq \EL(T'\eta')
+    \mforall \eta = \eta' \eeq \rho \in \Gamma
+\\
   \Gamma \models T \leq T' & \defas & \Gamma \models T \mand \Gamma \models T' \mand
@@
   If\/ $\Gamma \der J$ then % $\models \Gamma$ and
-  $\Gamma \models J$
+  $\Gamma \models J$.
   \end{enumerate}
@@
 By induction hypothesis $\rho_\Gamma \in \Gamma$.  Since $\var\Gamma \in \SIZE$ we conclude
-$(\rho_\Gamma, \var\Gamma) \in \eext\Gamma \Size)$.
+$(\rho_\Gamma, \var\Gamma) \in \eext\Gamma \Size$.
 \qed
@@
   $\Gamma \der T' \evalsto n : s$ for some neutral $n$ and
-  $\Gamma \der T' \doteq \tRty\,A' : s$
+  $\Gamma \der T' \doteq \tRty\,A' : s$.
   %$\forall \xi : \Gamma' \leq \Gamma.\ \ \exists M.\ \ \RTy{\Gamma'}{A'}M \mand \Gamma' \der T'\xi = M : \Set[\ell]$.
@@
   $\LR{\Gamma}{t}{T'}{f}{A'}$ iff
-  $\Gamma \der t \doteq \tR \, \down {A'} f : T'$
+  $\Gamma \der t \doteq \tR \, \down {A'} f : T'$.
   % $\forall \xi : \Gamma' \leq \Gamma.\ \
@@
   $\LRSize {\resurrect\Gamma} a \alpha$ and
-  $\Gamma \der t \doteq \tR\,\down{A'} f : \Nat a$
+  $\Gamma \der t \doteq \tR\,\down{A'} f : \Nat a$.
   % \\
@@
   \begin{enumerate}
-  \item If\/ $\LRSize \Gamma a \alpha$ then $\LRSize{\Gamma'}{a\xi}\alpha$.
+  %% SEE ABOVE
+  % \item If\/ $\LRSize \Gamma a \alpha$ then $\LRSize{\Gamma'}{a\xi}\alpha$.
   \item If\/ $\LRTS {\Gamma} T A s$ then $\LRTS{\Gamma'}{T\xi}A s$.
@@
 If\/ $\Gamma \der \sigma : \Delta$ then
-$\Gamma \valid \sigma : \Delta$
+$\Gamma \valid \sigma : \Delta$.
 \end{enumerate}
@@
 By induction hypothesis we have $\LR \Delta {t \sigma} {(\forall T)\tau} {t \eta} {(\forall T) \rho}$.
-It suffices to show
+It suffices to show the following:
 \begin{enumerate}
@@
 In the long run, we wish for an incremental type-directed equality check, which is also at the heart of Agda's type checker, as this check can generate constraints for meta variables involved in type reconstruction \cite{norell:PhD}.
-However, the usual bidirectional construction \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types.
+However, the usual bidirectional construction \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types, and even principal types.
 
@@
 
-Finally, we would like a general theory of shape-irrelevance that extends beyond size-indexed types.
-For instance, any data type constructor could be considered shape-irrelevant in all its parameters.
-However, the current reification algorithm which erases irrelevant size arguments to $\infty$ would need some patching.
-\citet{abelCoquandPagano:lmcs11} erase all proofs of a proposition $P$ to a canonical element $\star$ which inhabits any proposition $P$ that is already inhabited.
-A similar technique could work in our setting and would be necessary unless we can justify the incremental equality check (see above).
+Finally, we would like a general theory of shape-irrelevance that
+extends beyond size-indexed types.  For instance, any data type
+constructor could be considered shape-irrelevant in all its
+indices, with the consequence that index arguments in the data constructors
+could be declared irrelevant.
+However, our notion of judgmental equality does not support
+irrelevant arguments of dependent type.
+It works for the non-dependent type $\Size$, but we also
+relied on having a closed inhabitant $\infty$ in $\Size$.
+More research is needed to tell a more general story of shape-irrelevance.
+
+% Finally, we would like a general theory of shape-irrelevance that
+% extends beyond size-indexed types.  For instance, any data type
+% constructor could be considered shape-irrelevant in all its
+% parameters.
+% However, the current reification algorithm which erases
+% irrelevant size arguments to $\infty$ would need some patching.
+% \citet{abelCoquandPagano:lmcs11} erase all proofs of a proposition $P$
+% to a canonical element $\star$ which inhabits any proposition $P$ that
+% is already inhabited.  A similar technique could work in our setting
+% and would be necessary unless we can justify the incremental equality
+% check (see above).
 
@@
   He also acknowledges past discussions with Christoph-Simon Senjak.
+  The incentive to write this article came during the EU Cost Action CA15123 EUTYPES meeting in Ljubljana in January 2017; thanks to Andrej Bauer for organizing it.
 \end{acks}

### 2017-06-10 nbe/icfp17.tex
[ ICFP 17 ] Addressed comments of reviewer C.

@@
 
-Going beyond the standard syntax-directed equality check, there is a technique that can deal with dynamic $\eta$-expansion.
-It is a type-directed normalization function inspired by normalization-by-evaluation (NbE) that computes $\eta$-long normal forms \cite{bergerSchwichtenberg:lics91,danvy:tdpe}.
-We can check the computed normal forms for identity and, thus, decide definitional equality.
-NbE has proven to be a robust method to decide equality in powerful type theories with non-trivial $\eta$-laws.
-It scales to universes and large eliminations \cite{abelCoquandDybjer:lics07}, topped with singleton types or proof irrelevance \cite{abelCoquandPagano:lmcs11}, and even impredicativity \cite{abel:flops10}.
-At its heart there are reflection $\up T$ and reification $\down T$ functions directed by type $T$ and orchestrating just-in-time $\eta$-expansion.
-Reflection $\up T x$ maps variables $x$ into the realm of values of type $T$ and let's us compute with open terms.
-Reification $\down T a$ takes a value $a$ of type $T$ and computes its long normal form.
-For instance, the normal form of a closed function $f : U \to T$ would be $\lambda x.\,\down{T} (f\,(\up U x))$, and for its dependently-typed variant $f : \funT x U T[x]$ it would be $\lambda x.\,\down{T[\up U x]} (f\,(\up U x))$.
+Going beyond the standard syntax-directed equality check, there is a
+technique that can deal with dynamic $\eta$-expansion.  It is a
+type-directed normalization function inspired by
+normalization-by-evaluation (NbE) that computes $\eta$-long normal
+forms \cite{bergerSchwichtenberg:lics91,danvy:tdpe}.  We can check the
+computed normal forms for identity and, thus, decide definitional
+equality.  NbE has proven to be a robust method to decide equality in
+powerful type theories with non-trivial $\eta$-laws.  It scales to
+universes and large eliminations \cite{abelCoquandDybjer:lics07},
+topped with singleton types or proof irrelevance
+\cite{abelCoquandPagano:lmcs11}, and even impredicativity
+\cite{abel:flops10}.  At its heart there are reflection $\up T$ and
+reification $\down T$ functions directed by type $T$ and orchestrating
+just-in-time $\eta$-expansion.  Reflection $\up T x$ maps variables
+$x$ into the realm of values of type $T$ and lets us compute with
+open terms.  Reification $\down T a$ takes a value $a$ of type $T$ and
+computes its long normal form.  For instance, the normal form of a
+closed function $f : U \to T$ would be
+$\lambda x.\,\down{T} (f\,(\up U x))$, and for its dependently-typed
+variant $f : \funT x U T[x]$ it would be
+$\lambda x.\,\down{T[\up U x]} (f\,(\up U x))$.
 
@@
 \ru{\Gamma \der t = t' : \forall i.\, T
-  }{\Gamma \der t\,a = t\,a' : T[b/i]}
+  }{\Gamma \der t\,a = t'\,a' : T[b/i]}
 \qquad
@@
 \ru{\Gamma \der t = t' : \Pi i.\, T
-  }{\Gamma \der t\,a = t\,a : T[a/i]}
+  }{\Gamma \der t\,a = t'\,a : T[a/i]}
 \qquad
@@
 %
-Similar as for $\Pi$, we use named lambda abstraction as sugar for de
+Similarly to for $\Pi$, we use named lambda abstraction as sugar for de
 Bruijn abstraction.  Named abstraction takes care of proper lifting of
@@
 of it in Section~\ref{sec:comp}.
+We write \fbox{$\DD :: J$} to express that $\DD$ is a derivation of judgement $J$.
 
@@
 indices. Comparison of size values $\alpha \leq \beta$ is
-analoguous to comparison of size terms $a \leq b$.  In the following, we
+analogous to comparison of size terms $a \leq b$.  In the following, we
 will reuse letter $a$ for a value if it cannot be confused for a size
@@
 \end{gather*}
-\begin{lemma}[Subtyping is preorder]
+\begin{lemma}[Subtyping is a preorder]
 \label{lem:preord}
@@
 Reflection and reification perform $\eta$-expansion so that we arrive at an $\eta$-long $\beta$-normal form.
-To perform the $\eta$-expansion, not the precise type is needed, just the approximate shape, in particular, whether it is a function type (do expand) or a base type (do not expand).
+To perform the $\eta$-expansion, the precise type is not needed, just the approximate shape, in particular, whether it is a function type (do expand) or a base type (do not expand).
 For the logical framework, the shape of a dependent type is just its
@@
 \begin{enumerate}
-\item If\/ $n_1 = n_2 \in \NE$ then $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$.
-\item If\/ $t_1 = t_2 \in T$ then $\down{S_1}{t_1} = \down{S_2}{t_2} \in \NF$.
+\item \label{it:up}
+If\/ $n_1 = n_2 \in \NE$ then $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$.
+\item \label{it:down}
+If\/ $t_1 = t_2 \in T$ then $\down{S_1}{t_1} = \down{S_2}{t_2} \in \NF$.
 \end{enumerate}
@@
   Let $d_i = \down{A_i}{u_i}$.
-  By induction hypothesis, $d_1 = d_2 \in \NF$.
+  By induction hypothesis (\ref{it:down}) with shapes $A \shape A_1$ and $A \shape A_2$
+  we get $d_1 = d_2 \in \NF$.
   Thus, $n_1\,d_1 = n_2\,d_2 \in \NE$ by Lemma~\ref{lem:closne}, and
-  by induction hypothesis,  $\up{B_1[u_1]}(n_1\,d_1) = \up{B_2[u_2]}(n_2\,d_2) \in B[u_1]$.
+  by induction hypothesis (\ref{it:up}) with shapes
+  $B[u_1] \shape B_1[u_1]$ and $B[u_1] \shape B_2[u_2]$ we obtain
+  $\up{B_1[u_1]}(n_1\,d_1) = \up{B_2[u_2]}(n_2\,d_2) \in B[u_1]$.
   With $(\up{S_i}{n_i})\,u_i \evalsto \up{B_i[u_i]}(n_i\,d_i)$ we are done by definition of $\EL(T)$.
@@
 \item We assume $k \in \NN$ and show $\R k {\down{S_i}{t_i}} \lambda v_i$ for some normal forms
-$v_1 \nfeq v_2$.  Let $u_i = \up{A_i} \lev k$.  Note that $u_1 = u_2 \in A$ by induction hypothesis.
+$v_1 \nfeq v_2$.  Let $u_i = \up{A_i} \lev k$.  Note that $u_1 = u_2 \in A$ by induction hypothesis, since $\var k = \var k \in \NE$ by Lemma~\ref{lem:closne}.
 It is sufficient to show
@@
 They will constitute the logical relation for base types,
-yet, are not general enough for function types.
+but need to be strengthened for function types.
 \begin{alignat*}{8}
@@
 % \]
-By definitions, these relations are closed under subsumption and
+By definition, these relations are closed under subsumption and
 weakening, e.g., if\/ $\RD \Gamma t d T$ and $\Gamma \der T \leq T'$ then $\RD \Gamma t d T'$,
@@
 This is an important stepping stone for the smooth integration of sized types into dependently-typed proof assistants.
-In the final paragraphs, we discuss some questions and insights that follow from our work and go beyond it.
+In these final paragraphs, we discuss some questions and insights that follow from our work and go beyond it.
 
@@
 % the usual bidirectional argument \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types.
-In the long run, we wish for an incremental type-directed equality check, which is also at the heart of Agda's type checker, as this check can generate constraints for meta variables involved in type reconstruction \cite{norell:PhD}.
-However, the usual bidirectional construction \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types, and even principal types.
+
+In the long run, we wish for a type-directed equality check that does
+not normalization in one go, but interleaves weak head normalization
+with structural comparison.
+Such a equality test is at the heart of Agda's type
+checker and it generates constraints for meta variables
+involved in type reconstruction \cite{norell:PhD}.  However, the usual
+bidirectional construction \cite{abelScherer:types10} does not seem to
+go through as we lack uniqueness of types (and even principal types).
 

### 2017-06-11 nbe/icfp17.tex
[ ICFP 17 ] Switched to acm small and fixed some displays.

@@
 %% For double-blind review submission
-\documentclass[acmlarge%,review,anonymous
+\documentclass[acmsmall%acmlarge%,review,anonymous
 ]{acmart}\settopmatter{printfolios=true}
@@
 %\usepackage[right]{showlabels}\renewcommand{\showlabelfont}{\small\ttfamily\color{gray}}
+
+% controlling flushleft/center for math displays
+% http://www.golatex.de/wechsel-zwischen-leqno-und-reqno-fleqn-uvm-t2488.html
+\makeatletter
+%\def\leqn{\tagsleft@true}
+%\def\reqn{\tagsleft@false}
+\def\fleq{\@fleqntrue\let\mathindent\@mathmargin \@mathmargin=\leftmargini}
+\def\cneq{\@fleqnfalse}
+\g@addto@macro{\endsubequations}{\addtocounter{equation}{-1}}
+\makeatother
 
@@
 \[
-\begin{array}{lllrl@{\qquad}l}
+\begin{array}{l@{~}l@{~\,}l@{~\,}r@{~\,}l@{\quad}l}
 \Sort & \ni & s
@@
 \[
-\begin{array}{lllll@{\qquad}l}
+\begin{array}{l@{~\,}l@{~\,}l@{~\,}l@{~\,}l@{\qquad}l}
   \Ne & \ni &
@@
             m\,\case[\ell] V{v_z}{v_s} \mid m\,\fix[\ell] V v
-    & \mbox{normal neutral term} \\
+    & \mbox{neutral n.f.} \\
+    % & \mbox{normal neutral term} \\
   \Nf & \ni &
@@
   \tUp  & \ni & N & ::= & \up A n
-     & \mbox{reflection of neutral term $n$ into the value world $\D$, at
-       type $A$} \\
+     & \mbox{reflection of neutral term $n$ as value of type $A$} \\
+     % & \mbox{reflection of neutral term $n$ into the value world $\D$, at type $A$} \\
   \Whnf & \ni & w & ::= & \dots \mid N
@@
     }{t = t' \in \NAT(\beta)}
-\quad
-  \ru{t \evalsto \zero \alpha \qquad
+\qquad
+  \ru{t \evalsto \zero \alpha \\
       t' \evalsto \zero{\alpha'}
     }{t = t' \in \NAT(\beta+1)}
-\quad
+\qquad
   \ru{t \evalsto \suc \alpha u \qquad
-      t' \evalsto \suc {\alpha'} {u'} \qquad
+      t' \evalsto \suc {\alpha'} {u'} \\
       u = u' \in \NAT(\beta)
@@
 
+\fleq
 Let $\A$ be a PER (including $\A = \SIZE$) and $\F$ a family of PERs over $\A$ such that
@@
 \[
-  \fbox{$\FORALL\,\F$} \defas
+  \fbox{$\FORALL\,\F$} \quad \, \defas
   \{(t,t') \mid t \ann \alpha = t' \ann{\alpha'} \in \F(\beta) \mforall \alpha,\alpha',\beta \in \SIZE \}
@@
 \]
+\cneq
 
@@
 \\[3ex]
-  \rul{T \evalsto \piT A B \qquad
-      T' \evalsto \piT{A'}{B'} \\
-      A = A' \in \SET[\ell] \qquad
+  \rul{T \evalsto \piT A B \quad \
+      T' \evalsto \piT{A'}{B'} \quad \
+      A = A' \in \SET[\ell] \\
       B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
@@
 \\[3ex]
-  \rul{T \evalsto \piT \Size B \quad
-      T' \evalsto \piT{\Size}{B'} \quad
+  \rul{T \evalsto \piT \Size B \qquad
+      T' \evalsto \piT{\Size}{B'} \\
       B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
@@
 \\[3ex]
-  \rul{T \evalsto \forallT B \quad
-      T' \evalsto \forallT{B'} \quad
+  \rul{T \evalsto \forallT B \qquad
+      T' \evalsto \forallT{B'} \\
       B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
@@
 \qquad
-  \ru{T \evalsto \Nat \alpha \qquad
-      T' \evalsto \Nat \alpha' \qquad
+  \ru{T \evalsto \Nat \alpha \quad
+      T' \evalsto \Nat \alpha' \quad
       \alpha \leq \alpha'
@@
 \qquad
-  \ru{T \evalsto \Set[\ell_0] \qquad
-      T' \evalsto \Set[\ell_0'] \qquad
+  \ru{T \evalsto \Set[\ell_0] \quad
+      T' \evalsto \Set[\ell_0'] \quad
       \ell_0 \leq \ell_0'
@@
 \begin{gather*}
-  \ru{T \evalsto N \qquad
+  \ru{T \evalsto N \quad \
       S \evalsto N'
@@
 \qquad
-  \ru{T \evalsto \piT A B \qquad
-      S \evalsto \piT{A'}{B'} \qquad
-      A \shape A' \qquad
+  \ru{T \evalsto \piT A B \quad \
+      S \evalsto \piT{A'}{B'} \quad \
+      A \shape A' \quad \
       B[u] \shape B'[u'] \mforall u = u' \in A
@@
     }{T \shape S_1}
-\\[2ex]
+\\[1.5ex]
   \ru{S_2 \evalsto \piT{A_2}{B_2} \qquad
@@
     }{T \shape S_1}
-%\\[2ex]
-\qquad
+\\[1.5ex]
   \ru{S_2 \evalsto \forallT{B_2} \qquad
@@
 
-\subsection{Computation}
+\subsection{Computation with natural numbers}
 \label{sec:comp}

### 2017-06-11 nbe/icfp17.tex
[ ICFP 17 ] Reformatted LR
[ ICFP 17 ] Put rules for case and recursion into separate figure.
[ ICFP 17 ] Removed some proofs from short version
[ ICFP 17 ] Figurized two large displays.

@@
 \usepackage[cal=boondoxo]{mathalfa}
+\usepackage{calc} % \widthof
 
@@
 \end{enumerate}
-This article is accompanied by a prototypical type checker \textsf{Sit} which implements the type system and type checking algorithm as described in the remainder of the paper.
+This article is accompanied by a prototypical type checker \href{https://hackage.haskell.org/package/Sit}{\texttt{Sit}} which implements the type system and type checking algorithm as described in the remainder of the paper.
 But before going into the technical details, we will motivate our type system from a practical perspective: reasoning about programs involving sized types in Agda.
@@
 \input{figeq}
+\input{fignat}
 
@@
 We have \fbox{boxed} the rules dealing with irrelevant size application.
+Fig.~\ref{fig:nat} add the typing and equality rules for case distinction
+and recursion on natural numbers.
 Judgement $\Gamma \der T : \Adm \ell$ characterizes the valid type
@@
 
+\fleq
 %Neutrals in the value world $\D$.
@@
 
-\fleq
 Let $\A$ be a PER (including $\A = \SIZE$) and $\F$ a family of PERs over $\A$ such that
@@
 which is a PER of values of type $T$.
-The rules for \fbox{$T = T' \in \SET[\ell]$} are:
+The rules for \fbox{$T = T' \in \SET[\ell]$} are listed in Fig.~\ref{fig:per}.
+\begin{figure}
+\hrulefill
 \[
@@
 \]
+\hrulefill
+\caption{Semantic types and their interpretation.\label{fig:per}}
+\end{figure}
 All relations involved here are closed under weak head equality.
@@
 By induction on $T \in \SET[\ell]$ and cases on $T \shape S_1$ and $T \shape S_2$.
+\SHORTLONG{\qed}{
 \begin{caselist}
@@
 \end{caselist}
+} % END SHORTLONG
 \end{proof*}
@@
 By induction on $a = a' \in \Nat(\alpha+1)$.
+\SHORTLONG{\qed}{
 \begin{caselist}
@@
 By well-founded induction on $\alpha$.
+\SHORTLONG{\qed}{
 \begin{caselist}
@@
 \noindent
-Then, we define semantic counterparts of our declarative judgements:
+Then, in Fig.~\ref{fig:sem}, we define semantic counterparts of our declarative judgements by recursion on the length of the context.
+\begin{figure}
+\hrulefill
 \[
@@
 \]
+\hrulefill
+\caption{Semantic judgements.\label{fig:sem}}
+\end{figure}
 
@@
 \end{lemma}
+\LONGVERSION{
 \begin{proof*}
@@
 \begin{caselist}
+\fleq
+\newlength{\colw}
+\setlength{\colw}{\widthof{$\LR{\Gamma}{U}{T'}{B}{A'}$}}
 
@@
 \nextcase $A' \evalsto N$ neutral.
-
-  \noindent
-  $\LRTS{\Gamma}{T'}{A'} s$ iff
-  $\Gamma \der T' \evalsto n : s$ for some neutral $n$ and
-  $\Gamma \der T' \doteq \tRty\,A' : s$.
-  %$\forall \xi : \Gamma' \leq \Gamma.\ \ \exists M.\ \ \RTy{\Gamma'}{A'}M \mand \Gamma' \der T'\xi = M : \Set[\ell]$.
-
-  \noindent
-  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff
-  $\Gamma \der t \doteq \tR \, \down {A'} f : T'$.
-  % $\forall \xi : \Gamma' \leq \Gamma.\ \
-  %  \exists m. \ \
-  %  \R {\Gamma'} {\down {A'} f} m \mand
-  %  \Gamma' \der t\xi = m : T'\xi$.
-
-\vspace{2ex}
+\[
+\begin{array}{p{\colw}ll}
+  $\LRTS{\Gamma}{T'}{A'} s$ & \defiff &
+  \mbox{$\Gamma \der T' \evalsto n : s$ for some neutral $n$ and
+  $\Gamma \der T' \doteq \tRty\,A' : s$.}
+\\
+  $\LR{\Gamma}{t}{T'}{f}{A'}$ & \defiff &
+  \Gamma \der t \doteq \tR \, \down {A'} f : T' .
+\end{array}
+\]
+
+%   \noindent
+%   $\LRTS{\Gamma}{T'}{A'} s$ iff
+%   $\Gamma \der T' \evalsto n : s$ for some neutral $n$ and
+%   $\Gamma \der T' \doteq \tRty\,A' : s$.
+%   %$\forall \xi : \Gamma' \leq \Gamma.\ \ \exists M.\ \ \RTy{\Gamma'}{A'}M \mand \Gamma' \der T'\xi = M : \Set[\ell]$.
+
+%   \noindent
+%   $\LR{\Gamma}{t}{T'}{f}{A'}$ iff
+%   $\Gamma \der t \doteq \tR \, \down {A'} f : T'$.
+%   % $\forall \xi : \Gamma' \leq \Gamma.\ \
+%   %  \exists m. \ \
+%   %  \R {\Gamma'} {\down {A'} f} m \mand
+%   %  \Gamma' \der t\xi = m : T'\xi$.
+
+\vspace{0ex}
 \nextcase $A' \evalsto \Nat \alpha$.
-
-  \noindent
-  $\LRTS{\Gamma}{T'}{A'} s$ iff $\Gamma \der T' \evalsto \Nat a : s$ for some $a$ and $\LRSize \Gamma a \alpha$.
-
-  \noindent
-  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff $\resurrect\Gamma \der T' \evalsto \Nat a$ for some $a$ and
-  $\LRSize {\resurrect\Gamma} a \alpha$ and
-  $\Gamma \der t \doteq \tR\,\down{A'} f : \Nat a$.
-  % \\
-  % $\forall \xi : \Gamma' \leq \Gamma.\ \
-  %  \exists v. \ \
-  %  \R {\Gamma'} {\down {A'} f} v \mand
-  %  \Gamma' \der t\xi = v : \Nat\,a\xi$.
-
-\vspace{2ex}
+\[
+\begin{array}{p{\colw}ll}
+  $\LRTS{\Gamma}{T'}{A'} s$ & \defiff &
+    \mbox{$\Gamma \der T' \evalsto \Nat a : s$ for some $a$ and $\LRSize \Gamma a \alpha$.}
+\\
+  $\LR{\Gamma}{t}{T'}{f}{A'}$ & \defiff &
+    \mbox{$\resurrect\Gamma \der T' \evalsto \Nat a : s$ for some $a$ and
+  $\LRSize {\resurrect\Gamma} a \alpha$}
+\\
+\multicolumn 3 {@{\qquad}l} {
+   \mand \Gamma \der t \doteq \tR\,\down{A'} f : \Nat a.
+  }
+\end{array}
+\]
+%   \noindent
+%   $\LRTS{\Gamma}{T'}{A'} s$ iff $\Gamma \der T' \evalsto \Nat a : s$ for some $a$ and $\LRSize \Gamma a \alpha$.
+
+%   \noindent
+%   $\LR{\Gamma}{t}{T'}{f}{A'}$ iff $\resurrect\Gamma \der T' \evalsto \Nat a$ for some $a$ and
+%   $\LRSize {\resurrect\Gamma} a \alpha$ and
+%   $\Gamma \der t \doteq \tR\,\down{A'} f : \Nat a$.
+%   % \\
+%   % $\forall \xi : \Gamma' \leq \Gamma.\ \
+%   %  \exists v. \ \
+%   %  \R {\Gamma'} {\down {A'} f} v \mand
+%   %  \Gamma' \der t\xi = v : \Nat\,a\xi$.
+
+\vspace{0ex}
 \nextcase $A' \evalsto \Set[\ell']$.
-
-  \noindent
-  $\LRTS {\Gamma} {T'} {A'} s$ iff $\Gamma \der T' \evalsto \Set[\ell'] : s$.
-
-  \noindent
-  $\LR \Gamma U {T'} B {A'}$ iff
-  $\resurrect\Gamma \der T' \evalsto \Set[\ell'] : s$ and $\LRT \Gamma U B {\ell'}$.
-
-\vspace{2ex}
+\[
+\begin{array}{p{\colw}ll}
+  $\LRTS {\Gamma} {T'} {A'} s$ & \defiff & \Gamma \der T' \evalsto \Set[\ell'] : s.
+\\
+  $\LR \Gamma U {T'} B {A'}$  & \defiff &
+  \resurrect\Gamma \der T' \evalsto \Set[\ell'] : s \mand \LRT \Gamma U B {\ell'}.
+\end{array}
+\]
+  % \noindent
+  % $\LRTS {\Gamma} {T'} {A'} s$ iff $\Gamma \der T' \evalsto \Set[\ell'] : s$.
+
+  % \noindent
+  % $\LR \Gamma U {T'} B {A'}$ iff
+  % $\resurrect\Gamma \der T' \evalsto \Set[\ell'] : s$ and $\LRT \Gamma U B {\ell'}$.
+
+\vspace{0ex}
 \nextcase $A' \evalsto \piT A B$.
-
-\noindent
-$\LRTS \Gamma {T'} {A'} s$ iff
-\(
-\begin{array}[t]{l}
-\Gamma \der T' \evalsto \piT U T : s \mforsome U, T \mand \LRTS \Gamma U A s \mand \\
-\forall \xi : \Gamma' \leq \Gamma.\ \
+\[
+\begin{array}{p{\colw}ll}
+$\LRTS \Gamma {T'} {A'} s$ & \defiff &
+  \Gamma \der T' \evalsto \piT U T : s \mforsome U, T \mand \LRTS \Gamma U A s \\
+\multicolumn 3 {@{\qquad}l} {
+  \mand \forall \xi : \Gamma' \leq \Gamma.\ \
   \LR {\Gamma'} u {U\xi} a A \implies \LRTS {\Gamma'} {T(\xi,u)} {B[a]} s
-. \\
-\end{array}
-\) \\
-$\LR \Gamma t {T'} f {A'}$ iff
-\(
-\begin{array}[t]{l}
-\resurrect\Gamma \der T' \evalsto \piT U T : s \mforsome U, T \mand \LRTS {\resurrect\Gamma} U A s \mand \\
-\forall \xi : \Gamma' \leq \Gamma.\ \
+  .
+}
+\\[1ex]
+$\LR \Gamma t {T'} f {A'}$ & \defiff &
+  \resurrect\Gamma \der T' \evalsto \piT U T : s \mforsome U, T \mand \LRTS {\resurrect\Gamma} U A s \\
+\multicolumn 3 {@{\qquad}l} {
+  \mand \forall \xi : \Gamma' \leq \Gamma.\ \
   \LR {\Gamma'} u {U\xi} a A \implies \LR {\Gamma'} {t\xi\,u} {T(\xi,u)} {f\,a} {B[a]}
   .
+}
 \end{array}
-\)
-
-\vspace{2ex}
+\]
+% \noindent
+% $\LRTS \Gamma {T'} {A'} s$ iff
+% \(
+% \begin{array}[t]{l}
+% \Gamma \der T' \evalsto \piT U T : s \mforsome U, T \mand \LRTS \Gamma U A s \mand \\
+% \forall \xi : \Gamma' \leq \Gamma.\ \
+% \LR {\Gamma'} u {U\xi} a A \implies \LRTS {\Gamma'} {T(\xi,u)} {B[a]} s
+% . \\
+% \end{array}
+% \) \\
+% $\LR \Gamma t {T'} f {A'}$ iff
+% \(
+% \begin{array}[t]{l}
+% \resurrect\Gamma \der T' \evalsto \piT U T : s \mforsome U, T \mand \LRTS {\resurrect\Gamma} U A s \mand \\
+% \forall \xi : \Gamma' \leq \Gamma.\ \
+% \LR {\Gamma'} u {U\xi} a A \implies \LR {\Gamma'} {t\xi\,u} {T(\xi,u)} {f\,a} {B[a]}
+% .
+% \end{array}
+% \)
+
+\vspace{0ex}
 \nextcase $A' \evalsto \piT \Size B$.
-
-\noindent
-$\LRTS \Gamma {T'} {A'} s$ iff
-\(
-\begin{array}[t]{l}
-\Gamma \der T' \evalsto \piT \Size T : s \mforsome T\\
-\forall \xi : \Gamma' \leq \Gamma.\ \
+\[
+\begin{array}{p{\colw}ll}
+$\LRTS \Gamma {T'} {A'} s$ & \defiff &
+  \Gamma \der T' \evalsto \piT \Size T : s \mforsome T
+\\
+\multicolumn 3 {@{\qquad}l} {
+  \mand \forall \xi : \Gamma' \leq \Gamma.\ \
   \LRSize {\Gamma'} a \alpha
@@
   .
-\end{array}
-\) \\
-$\LR \Gamma t {T'} f {A'}$ iff
-\(
-\begin{array}[t]{l}
-\resurrect\Gamma \der T' \evalsto \piT \Size T : s \mforsome T\\
+}
+\\[1ex]
+$\LR \Gamma t {T'} f {A'}$ & \defiff &
+\resurrect\Gamma \der T' \evalsto \piT \Size T : s \mforsome T
+\\
+\multicolumn 3 {@{\qquad}l} {
+  \mand
   \forall \xi : \Gamma' \leq \Gamma.\ \
@@
   .
+}
 \end{array}
-\)
-
-\vspace{2ex}
+\]
+% \noindent
+% $\LRTS \Gamma {T'} {A'} s$ iff
+% \(
+% \begin{array}[t]{l}
+% \Gamma \der T' \evalsto \piT \Size T : s \mforsome T\\
+% \forall \xi : \Gamma' \leq \Gamma.\ \
+% \LRSize {\Gamma'} a \alpha
+% \implies \LRTS {\Gamma'} {T(\xi,a)} {B[\alpha]} s
+% .
+% \end{array}
+% \) \\
+% $\LR \Gamma t {T'} f {A'}$ iff
+% \(
+% \begin{array}[t]{l}
+% \resurrect\Gamma \der T' \evalsto \piT \Size T : s \mforsome T\\
+% \forall \xi : \Gamma' \leq \Gamma.\ \
+% \LRSize {\Gamma'} a \alpha
+% \implies \LR {\Gamma'} {t\xi\,a} {T(\xi,a)} {f\,\alpha} {B[\alpha]}
+% .
+% \end{array}
+% \)
+
+\vspace{0ex}
 \nextcase $A' \evalsto \forallT B$.
-
-\noindent
-$\LRTS \Gamma {T'} {A'} s$ iff
-\(
-\begin{array}[t]{l}
-\Gamma \der T' \evalsto \forallT T : s \mforsome T\\
+\[
+\begin{array}{p{\colw}ll}
+$\LRTS \Gamma {T'} {A'} s$ & \defiff &
+  \Gamma \der T' \evalsto \forallT T : s \mforsome T
+\\
+\multicolumn 3 {@{\qquad}l} {
+  \mand
   \forall \xi : \Gamma' \leq \Gamma,\
@@
   .
-\end{array}
-\) \\
-$\LR \Gamma t {T'} f {A'}$ iff
-\(
-\begin{array}[t]{l}
-\resurrect\Gamma \der T' \evalsto \forallT T : s \mforsome T\\
+}
+\\[1ex]
+$\LR \Gamma t {T'} f {A'}$ & \defiff &
+  \resurrect\Gamma \der T' \evalsto \forallT T : s \mforsome T
+\\
+\multicolumn 3 {@{\qquad}l} {
+  \mand
   \forall \xi : \Gamma' \leq \Gamma,\
@@
   \resurrect{\Gamma'} \der a,b : \Size,\
-\alpha,\beta \in \SIZE. \ \\ \qquad
+  \alpha,\beta \in \SIZE. \
+}
+\\
+\multicolumn 3 {@{\qquad\qquad}l} {
   \LRSize {\resurrect{\Gamma'}} b \beta
@@
   .
+}
 \end{array}
-\)
+\]
+% \noindent
+% $\LRTS \Gamma {T'} {A'} s$ iff
+% \(
+% \begin{array}[t]{l}
+% \Gamma \der T' \evalsto \forallT T : s \mforsome T\\
+% \forall \xi : \Gamma' \leq \Gamma,\
+% {\Gamma'} \der b : \Size,\
+% \beta \in \SIZE. \ \
+% \LRSize {\Gamma'} b \beta
+% \implies \LRTS {\Gamma'} {T(\xi,b)} {B[\beta]} s
+% .
+% \end{array}
+% \) \\
+% $\LR \Gamma t {T'} f {A'}$ iff
+% \(
+% \begin{array}[t]{l}
+% \resurrect\Gamma \der T' \evalsto \forallT T : s \mforsome T\\
+% \forall \xi : \Gamma' \leq \Gamma,\
+% %\Gamma' \der a : \Size,\
+% \resurrect{\Gamma'} \der a,b : \Size,\
+% \alpha,\beta \in \SIZE. \ \\ \qquad
+% \LRSize {\resurrect{\Gamma'}} b \beta
+% \implies \LR {\Gamma'} {t\xi\,\ann a} {T(\xi,b)} {f\,\ann\alpha} {B[\beta]}
+% .
+% \end{array}
+% \)
 \end{caselist}
@@
 %%% mode: latex
-%%% TeX-master: "icfp17-long"
+%%% TeX-master: "icfp17"
 %%% End:

### 2017-06-11 nbe/icfp17.tex
[ ICFP 17 ] Shortened text where it lead to ugly line break.

@@
 and if\/ $\xi : \Gamma' \leq \Gamma$ then $\RD{\Gamma'}{t\xi}d{T\xi}$.
+\LONGVERSION{
 \begin{lemma}[Fresh variable readback]
@@
 \end{proof}
+} % END LONGVERSION
 \begin{lemma}[Closure properties for neutrals]
@@
 \end{proof*}
-} % END LONGVERSION
-\begin{corollary}[Fresh variable injection]
+%\begin{corollary}[Fresh variable injection]
+\begin{corollary}[Fresh variable]
   \label{cor:fresh}
@@
 \end{proof}
+} % END LONGVERSION
 
@@
 
-\begin{lemma}[Semantic subtyping implies judgmental subtyping \cite{fridlenderPagano:tlca13}]
+%\begin{lemma}[Semantic subtyping implies judgmental subtyping \cite{fridlenderPagano:tlca13}]
+\begin{lemma}[Semantic implies judgmental subtyping \cite{fridlenderPagano:tlca13}]
 \label{lem:lrsub}
@@
   \label{lem:wklrsub}
+  \bla \\
   Let $\LRSub \Gamma \sigma \tau \Delta \eta \rho$. Then:
@@
   By induction on $\der \Gamma$.
-\begin{caselist}
-
-\nextcase
+  Consider case:
+% \begin{caselist}
+% \nextcase
 \[
@@
 \qed
-\end{caselist}
+%\end{caselist}
 \end{proof*}
@@
 
-\item By transitivity of judgmental equality from (\ref{it:sound}), using
-  Lemma~\ref{lem:erase}:
+\item %By transitivity of judgmental equality from
+  From (\ref{it:sound}), using Lemma~\ref{lem:erase}:
   $\Gamma \der t = \nbe \Gamma T t = \nbe \Gamma T {t'} = t' : T$.
@@
 \begin{proof}
-  Follows directly from soundness (which includes termination) and completeness of NbE.
+  Follows directly from soundness (including termination) and completeness of NbE.
 \end{proof}

### 2017-06-11 nbe/sec-asub.tex
[ ICFP 17 ] Switched to acm small and fixed some displays.
[ ICFP 17 ] Shortened text where it lead to ugly line break.

@@
     }{\Gamma \der T \asub T'}
-\\[2ex]
+\qquad
   \rux{T \evalsto \Set[\ell] \qquad
@@
     }{\ell \leq \ell'}
+\\[2ex]
+  \rux{T \evalsto \Nat a \qquad
+      T' \evalsto \Nat a'
+    }{\Gamma \der T \asub T'
+    }{a \leq a'}
 \qquad
+  \ru{T_1' \evalsto \epiT\Size{T_1} \quad \
+      T_2' \evalsto \epiT\Size{T_2} \quad \
+      \cext \Gamma {\Size} \der T_1 \asub T_2
+    }{\Gamma \der T_1' \asub T_2'}
+\\[2ex]
   \ru{T_1' \evalsto \piT{U_1}{T_1} \qquad
@@
       \cext \Gamma {U_2} \der T_1 \asub T_2
-    }{\Gamma \der T_1' \asub T_2'}
-\\[2ex]
-  \ru{T \evalsto \Nat a \qquad
-      T' \evalsto \Nat a'
-    }{\Gamma \der T \asub T'
-    }{a \leq a'}
-\qquad
-  \ru{T_1' \evalsto \epiT\Size{T_1} \qquad
-      T_2' \evalsto \epiT\Size{T_2} \qquad
-      \cext \Gamma {\Size} \der T_1 \asub T_2
     }{\Gamma \der T_1' \asub T_2'}
@@
 \begin{proof}
-  By induction on $\Gamma \der T \asub T'$, using soundness of NbE and subject reduction (Thm.~\ref{thm:sr}).
+  By induction on $\Gamma \der T \asub T'$, soundness of NbE, and subject reduction (Thm.~\ref{thm:sr}).
 \end{proof}
@@
   If successfully, then $\Gamma \der T \leq T'$ by soundness of algorithmic equality.
-  If unsuccessfully, then $\Gamma \der T \leq T'$ is impossible by completeness of algorithmic equality.
+  % If unsuccessfully, then
+  Otherwise  $\Gamma \der T \leq T'$ is impossible by completeness of algorithmic equality.
 \end{proof}

### 2017-06-11 nbe/icfp17.tex
[ ICFP 17 ] Put shape definition into figure.
[ ICFP 17 ] Figurized the lr for substitutions

@@
 
-\fbox{$T \shape S$} for $T \in \SET[\ell]$.  We call $T$ the \emph{template} and $S$ one of its possible \emph{shapes}.
+Fig.~\ref{fig:shape} defines \fbox{$T \shape S$} for $T \in \SET[\ell]$.
+We call $T$ the \emph{template} and $S$ one of its possible \emph{shapes}.
+%% WORKAROUND SINCE \shape does not seem to work in \caption
+\newbox{\TshapeS}\savebox{\TshapeS}{\fbox{$T \shape S$}}
+\begin{figure}
+\hrulefill
 \begin{gather*}
@@
 \end{gather*}
-% The last rule is the crucial one
-
+\hrulefill
+\caption{Type shapes \usebox{\TshapeS}\,.}
+\label{fig:shape}
+\end{figure}
 Note that $T \in \SET[\ell]$ and $T \shape S$ do not imply $S \in \SET[\ell]$.
@@
 
-\begin{lemma}[Equal types make equally good templates]
+\begin{lemma}[Templates are up to equality]
   If\/ $T = T' \in \SET[\ell]$ and $T' \shape S$ then $T \shape S$.
@@
 
-Logical relation for substitutions.
-\fbox{$\LRSub \Gamma \sigma \tau \Delta \eta \rho$}.
+\begin{figure}[htbp]
+\hrulefill
 \begin{gather*}
@@
 \end{gather*}
+\hrulefill
+\caption{Logical relation for substitutions
+  \fbox{$\LRSub \Gamma \sigma \tau \Delta \eta \rho$}.}
+\label{fig:lrsub}
+\end{figure}
+
+Fig.~\ref{fig:lrsub} defines a logical relation for substitutions
+\fbox{$\LRSub \Gamma \sigma \tau \Delta \eta \rho$}.
 We write \fbox{$\LRS \Gamma \tau \Delta \rho$} for $\LRSub \Gamma \tau \tau \Delta \rho \rho$.
@@
 
-Validity judgements.
+The following judgements are used to state the fundamental theorem of typing.
 \[
@@
     % \resurrect\Gamma \valid T \mand
-    \valid \cext \Gamma T \mand
+    % \valid \cext \Gamma T \mand
     \LR {\Gamma'} {t \sigma} {T \tau} {t \eta} {T \rho}

### 2017-06-11 nbe/icfp17.tex
[ ICFP 17 ] Added sections 6 and 7 to short version.

@@
 \LONGVERSION{ % BEGIN LONGVERSION
-\begin{lemma}[Properties of the logical relation for substitution]
+\begin{lemma}[Properties of the logical relation for substitutions]
   \label{lem:wklrsub}
@@
 
-\LONGVERSION{
 \input{sec-asub}
-} % END LONG VERSION
-\SHORTVERSION{
-Subtyping is decidable, for instance by full normalization
-and comparison of normal forms
-\cite{fridlenderPagano:tlca13}.
-Alternatively, we can define the usual incremental subtyping algorithm
-based on weak head evaluation, which checks equality of neutral types
-by NbE.  For details, see the long version of this article.
-} % END SHORTVERSION
+% \LONGVERSION{
+% \input{sec-asub}
+% } % END LONG VERSION
+% \SHORTVERSION{
+% Subtyping is decidable, for instance by full normalization
+% and comparison of normal forms
+% \cite{fridlenderPagano:tlca13}.
+% Alternatively, we can define the usual incremental subtyping algorithm
+% based on weak head evaluation, which checks equality of neutral types
+% by NbE.  For details, see the long version of this article.
+% } % END SHORTVERSION
 
@@
 
-\LONGVERSION{
 \input{sec-tycheck}
-} % END LONG VERSION
-\SHORTVERSION{
-It is now straightforward to define a bidirectional type-checking
-algorithm which decides typing for normal forms, and succeeds for
-those that rely only on the following restricted rule for irrelevant
-size application:
-\[
-  \ru{\Gamma \ders t : \forall T \qquad
-      \resurrect\Gamma \der a : \Size
-    }{\Gamma \ders t\, \ann a : T[a]}
-\]
-For lack of space, we delegate definition and soundness proof to the
-long version of this article.
-} % END SHORTVERSION
+% \LONGVERSION{
+% \input{sec-tycheck}
+% } % END LONG VERSION
+% \SHORTVERSION{
+% It is now straightforward to define a bidirectional type-checking
+% algorithm which decides typing for normal forms, and succeeds for
+% those that rely only on the following restricted rule for irrelevant
+% size application:
+% \[
+%   \ru{\Gamma \ders t : \forall T \qquad
+%       \resurrect\Gamma \der a : \Size
+%     }{\Gamma \ders t\, \ann a : T[a]}
+% \]
+% For lack of space, we delegate definition and soundness proof to the
+% long version of this article.
+% } % END SHORTVERSION
 

### 2017-06-11 nbe/sec-asub.tex
[ ICFP 17 ] Figurized algorithmic subtyping.
[ ICFP 17 ] Added sections 6 and 7 to short version.

@@
 
-Algorithmic subtyping \fbox{$\Gamma \der T \asub T'$}.
+Fig~\ref{fig:asub} defines an incremental subtyping algorithm
+\fbox{$\Gamma \der T \asub T'$}.
+Neutral types are subtypes iff they are equal, which is checked using NbE.
+\begin{figure}[htbp]
+\hrulefill
 \begin{gather*}
@@
 \end{gather*}
+\hrulefill
+\caption{Algorithmic subtyping \fbox{$\Gamma \der T \asub T'$}.}
+\label{fig:asub}
+\end{figure}
 
@@
 \begin{proof}
-  By induction on $\Gamma \der T \asub T'$, soundness of NbE, and subject reduction (Thm.~\ref{thm:sr}).
+  By induction on $\Gamma \der T \asub T'$, soundness of NbE, and
+  subject reduction\LONGVERSION{~(Thm.~\ref{thm:sr})}.
 \end{proof}
@@
 \label{lem:lrasub}
-\bla
+\bla \\
    If\/ $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \asub T'$.
 \end{lemma}
+\SHORTLONG{
+\begin{proof}
+  By induction on $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and
+  cases on $A \leq A' \in \TYPE$.
+\end{proof}
+}{ % LONG VERSION
 \begin{proof*}
@@
 \end{proof*}
+} % END LONG VERSION
 

### 2017-06-12 nbe/icfp17.tex
Typos

@@
 We have \fbox{boxed} the rules dealing with irrelevant size application.
-Fig.~\ref{fig:nat} add the typing and equality rules for case distinction
+Fig.~\ref{fig:nat} adds the typing and equality rules for case distinction
 and recursion on natural numbers.
@@
   By induction on $a = a' \in \NAT(\infty)$, we can easily show $a = a' \in \NAT(\alpha)$ for some $\alpha < \infty$.
-  For instance, $\alpha$ could be number of uses of the successor rule plus one.
+  For instance, $\alpha$ could be the number of uses of the successor rule plus one.
 \end{proof}
@@
 In the long run, we wish for a type-directed equality check that does
-not normalization in one go, but interleaves weak head normalization
+not do normalization in one go, but interleaves weak head normalization
 with structural comparison.
-Such a equality test is at the heart of Agda's type
+Such an equality test is at the heart of Agda's type
 checker and it generates constraints for meta variables

### 2017-06-14 nbe/icfp17.tex
[ ICFP 17 ] Added alpha(') \in SIZE.
[ ICFP 17 ] Put reflection/reification proof back, shortened elsewhere

@@
   \item If\/ $\alpha \in \SIZE$ then $\alpha = \alpha \in \ELIM$.
-  \item $\ann \alpha = \ann{\alpha'} \in \ELIM$.
-  \item If\/ $A = A' \in \TY$ and $d_z = d_z' \in \NF$ and $d_s = d_s' \in \NF$
+  \item If\/ $\alpha,\alpha' \in \SIZE$ then $\ann \alpha = \ann{\alpha'} \in \ELIM$.
+  \item If\/ $A = A' \in \TY$ and $d_z {=} d_z' \in \NF$ and $d_s {=} d_s' \in \NF$
         then $\case[\ell] A {d_z}{d_s} = \case[\ell]{A'}{d_z'}{d_s'} \in \ELIM$.
@@
 By induction on $T \in \SET[\ell]$ and cases on $T \shape S_1$ and $T \shape S_2$.
-\SHORTLONG{\qed}{
+%\SHORTLONG{\qed}{
 \begin{caselist}
 
+\LONGVERSION{
 \nextcase $T \evalsto \piT A B$ with $A \in \SET[\ell]$ and
@@
 \end{enumerate}
+} % END LONG VERSION
 
@@
 \end{caselist}
-} % END SHORTLONG
+%} % END SHORTLONG
 \end{proof*}
@@
   If\/ $\Gamma \der t : T$ then $\Gamma \der t = \nbe \Gamma T t : T$.
-  \item
+  \item \label{it:soundeq}
   If\/ $\Gamma \der t,t' : T$ and $\nbe \Gamma T t \nfeq \nbe \Gamma T {t'}$ then
@@
 \begin{proof*}
-\bla
-\begin{enumerate}
-
-\item For the identity environment $\LRS \Gamma \sid \Gamma \rhog$ (Lemma~\ref{lem:idenv})
+(\ref{it:sound}) For the identity environment $\LRS \Gamma \sid \Gamma \rhog$ (Lemma~\ref{lem:idenv})
 the Fundamental Theorem for Typing gives $\LR \Gamma t T {t\rhog} {T\rhog}$.
@@
 $\Gamma \der t = v : T$ by Thm.~\ref{thm:into}.
-
-\item %By transitivity of judgmental equality from
+%
+Then (\ref{it:soundeq}):
   From (\ref{it:sound}), using Lemma~\ref{lem:erase}:
@@
 \qed
-\end{enumerate}
+% \bla
+% \begin{enumerate}
+
+% \item For the identity environment $\LRS \Gamma \sid \Gamma \rhog$ (Lemma~\ref{lem:idenv})
+% the Fundamental Theorem for Typing gives $\LR \Gamma t T {t\rhog} {T\rhog}$.
+% This implies $\R \Gamma {\down{(T\rhog)} (t\rhog)} v$ for some normal form $v$ and
+% $\Gamma \der t = v : T$ by Thm.~\ref{thm:into}.
+
+% \item %By transitivity of judgmental equality from
+%   From (\ref{it:sound}), using Lemma~\ref{lem:erase}:
+%   $\Gamma \der t = \nbe \Gamma T t = \nbe \Gamma T {t'} = t' : T$.
+%    % Here we use that the computed normal forms are identical because
+%    % we erase all size annotations $\ann\alpha$ to $\ann\infty$ during read back.
+% \qed
+% \end{enumerate}
 \end{proof*}
@@
 
-% \clearpage
+\SHORTVERSION{\clearpage}
 

### 2017-07-04 nbe/icfp17.tex
[ ICFP 2017 ] changes for camera-ready submission
[ ICFP 2017 ] submission: removed badges, fixed Theo affiliation

@@
 %% For double-blind review submission
-\documentclass[acmsmall%acmlarge%,review,anonymous
-]{acmart}\settopmatter{printfolios=true}
+%%\documentclass[acmsmall%acmlarge%,review,anonymous
+%%]{acmart}\settopmatter{printfolios=true}
 %% For single-blind review submission
@@
 %% For final camera-ready submission
-%\documentclass[acmlarge]{acmart}\settopmatter{}
+\documentclass[acmsmall,screen]{acmart}\settopmatter{}
 
@@
 
-\makeatletter\if@ACM@journal\makeatother
-%% Journal information (used by PACMPL format)
-%% Supplied to authors by publisher for camera-ready submission
+\setcopyright{rightsretained}
 \acmJournal{PACMPL}
+\acmYear{2017}
 \acmVolume{1}
-\acmNumber{1}
-\acmArticle{1}
-\acmYear{2017}
-\acmMonth{1}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\else\makeatother
-%% Conference information (used by SIGPLAN proceedings format)
-%% Supplied to authors by publisher for camera-ready submission
-\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
-\acmYear{2017}
-\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\fi
+\acmNumber{ICFP}
+\acmArticle{33}
+\acmMonth{9}
+\acmDOI{10.1145/3110277}
+\acmPrice{}
+
+%% \makeatletter\if@ACM@journal\makeatother
+%% %% Journal information (used by PACMPL format)
+%% %% Supplied to authors by publisher for camera-ready submission
+%% \acmJournal{PACMPL}
+%% \acmVolume{1}
+%% \acmNumber{1}
+%% \acmArticle{33}
+%% \acmYear{2017}
+%% \acmMonth{9}
+%% \acmDOI{10.1145/nnnnnnn.nnnnnnn}
+%% \startPage{1}
+%% \else\makeatother
+%% %% Conference information (used by SIGPLAN proceedings format)
+%% %% Supplied to authors by publisher for camera-ready submission
+%% \acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
+%% \acmYear{2017}
+%% \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
+%% \acmDOI{10.1145/nnnnnnn.nnnnnnn}
+%% \startPage{1}
+%% \fi
 
@@
 %% see authors.acm.org) by publisher for camera-ready submission
-\setcopyright{none}             %% For review submission
+%%\setcopyright{none}             %% For review submission
 %\setcopyright{acmcopyright}
@@
 % \acmBadgeR[http://icfp17.sigplan.org/track/icfp-2017-Artifacts]{artifact_evaluated-functional.png}
-\acmBadgeR[http://icfp17.sigplan.org/track/icfp-2017-Artifacts]{artifact_evaluated-reusable.png}
-%\acmBadgeL[http://icfp17.sigplan.org/track/icfp-2017-Artifacts]{artifact_available.png}
-\acmBadgeL[https://hackage.haskell.org/package/Sit]{artifact_available.png}
+% \acmBadgeR[http://icfp17.sigplan.org/track/icfp-2017-Artifacts]{artifact_evaluated-reusable.png}
+% \acmBadgeL[http://icfp17.sigplan.org/track/icfp-2017-Artifacts]{artifact_available.png}
+% \acmBadgeL[https://hackage.haskell.org/package/Sit]{artifact_available.png}
 
@@
   \department{Department of Computer Science and Engineering}
-  \institution{Chalmers}
+  \institution{Chalmers University of Technology}
   \streetaddress{Rännvägen 6b}
@@
 \affiliation{
-  \institution{École Normale Supérieure de Cachan}
+  \institution{École normale supérieure Paris-Saclay}
+  \country{France}
 }

### 2017-08-13 nbe/icfp17.tex
[ ICFP 2017 ] break overfull box
[ ICFP 2017 ] fixed most issues raised by publisher

@@
 %% Title information
-\title[NbE for Sized Types]{Normalization by Evaluation for Sized Dependent Types}
+\title{Normalization by Evaluation for Sized Dependent Types}
 %\titlenote{with title note}             %% \titlenote is optional;
@@
   \department{Department of Computer Science and Engineering}
-  \institution{Gothenburg University}
+  \institution{Gothenburg University, Sweden}
   \streetaddress{Rännvägen 6b}
@@
   \department{Department of Computer Science and Engineering}
-  \institution{Chalmers University of Technology}
+  \institution{Chalmers University of Technology, Sweden}
   \streetaddress{Rännvägen 6b}
@@
 \affiliation{
-  \institution{École normale supérieure Paris-Saclay}
+  \institution{École normale supérieure Paris-Saclay, France}
   \country{France}
@@
 \begin{array}{l}
-  \ttrav \; \vpre \; \vpost \; i\; t = \vpost\;(\tcase\;\vpre\,i\,t\;\tof \, \{ \, \tnode\,j\,\vts \to \tnode\,\infty\,(\tmap \; (\tT\,j) \; (\tT\,\infty) \; (\ttrav \, \vpre \, \vpost \, j) \; \vts) \})
+  \ttrav \; \vpre \; \vpost \; i\; t =\\
+  \phantom{\ttrav\,}\vpost\;(\tcase\;\vpre\,i\,t\;\tof \, \{ \, \tnode\,j\,\vts \to \tnode\,\infty\,(\tmap \; (\tT\,j) \; (\tT\,\infty) \; (\ttrav \, \vpre \, \vpost \, j) \; \vts) \})
 \end{array}
@@
 
-\section{A Type System With Irrelevant Size Application}
+\section{A Type System with Irrelevant Size Application}
 \label{sec:syntax}
@@
 
-\subsection{Weak head reduction}
+\subsection{Weak Head Reduction}
 
@@
 
-\subsection{Read back}
+\subsection{Read Back}
 \label{sec:read}
@@
 
-\subsection{Partial equivalence relations}
+\subsection{Partial Equivalence Relations}
 \label{sec:per}
@@
 
-\subsection{PER model}
+\subsection{PER Model}
 \label{sec:permodel}
@@
 
-\subsection{Type shapes}
+\subsection{Type Shapes}
 \label{sec:shape}
@@
 
-\subsection{Computation with natural numbers}
+\subsection{Computation with Natural Numbers}
 \label{sec:comp}
@@
 
-\SHORTVERSION{\clearpage}
+%%\SHORTVERSION{\clearpage}
 

### 2018-03-02 nbe/icfp17.tex
ICFP17: fixed proof of core lemma

@@
 \begin{enumerate}
-\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $\alpha_1,\alpha_2 \in \SIZE$.
+\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $\alpha_1,\alpha_2,\beta \in \SIZE$.
   Since $n_1 \ann{\alpha_1} = n_2 \ann{\alpha_2} \in \NE$ by Lemma~\ref{lem:closne}, we obtain
-  $\up{B_1[\alpha_1]}{(n_1 \ann{\alpha_1})} = \up{B_2[\alpha_2]}{(n_2 \ann{\alpha_2})} \in B[\alpha_1]$
+  $\up{B_1[\alpha_1]}{(n_1 \ann{\alpha_1})} = \up{B_2[\alpha_2]}{(n_2 \ann{\alpha_2})} \in B[\beta]$
   by induction hypothesis.
-  Thus, $(\up{S_1}{n_1}) \ann{\alpha_1} = (\up{S_2}{n_2}) \ann{\alpha_2} \in B[\alpha_1]$
+  Thus, $(\up{S_1}{n_1}) \ann{\alpha_1} = (\up{S_2}{n_2}) \ann{\alpha_2} \in B[\beta]$
   by weak head expansion, which entails the goal by definition of $\EL(T)$.
